### 2024-05-14 report/markdown/report.Rmd
Analysis report draft
Legends for Figures and Tables

---
title: "FGF- and FGFR-coding gene alterations in urothelial cancers"
subtitle: "Short analysis report"
author: "Piotr Tymoszuk for the Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: blca.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "CairoPNG", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Results

## Characteristic of investigated cohorts and frequency of genetic alterations

```{r fig-general-freq, fig.width = figur::convert(rep_figs$general_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$general_freq, to = 'in')$h, fig.cap = 'The most frequent somatic mutations and gene copy alterations in urothelial cancer.'}

rep_figs$general_freq$plot

```

__Figure \@ref(fig:fig-general-freq). The most frequent somatic mutations and gene copy alterations in urothelial cancer.__

_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% paste(collapse = ', ')`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots. Alterations affecting FGF- and FGFR-coding genes are highlighted with bold font. Numbers of analyzed cancer samples are indicated in the plot legend._

\newpage

```{r fig-fgfr-freq, fig.width = figur::convert(rep_figs$fgfr_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_freq, to = 'in')$h, fig.cap = 'Frequency of somatic mutations and copy number alterations of FGF- and FGFR-coding genes.'}

rep_figs$fgfr_freq$plot

```

__Figure \@ref(fig:fig-fgfr-freq). Frequency of somatic mutations and copy number alterations of FGF- and FGFR-coding genes.__

_Percentages of somatic mutations, deletions and amplifications of FGF- and FGFR-coding genes in the investigated cohort are presented in bar plots. Note that amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort. Numbers of analyzed cancer samples are indicated in the plot legend._

\newpage

```{r tab-cohorts, tab.cap = 'Characteristic of the investigated cohorts. Numeric features are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages and observation counts of the categories.'}

flextable::flextable(rep_tabs$cohorts) %>% 
  set_widths(c(3, rep(3.5, 3), 2.8, 2.8)) %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-genetics, tab.cap = "Frequency of the most common somatic mutations and copy number variants in the investigated cohorts. Frequencies of alterations present in at least 5% of cancer samples are displayed. Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$genetics[1, 1] %>% 
  mutate(`Alteration type` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes in the investigated cohorts. Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$fgf_genetics[1, 1] %>% 
  mutate(`Alteration type` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

## Classification of mutations in FGFR-coding genes, and their possible effects on protein structure and function

```{r fig-mutation-type, fig.width = figur::convert(rep_figs$mutation_type, to = 'in')$w, fig.height = figur::convert(rep_figs$mutation_type, to = 'in')$h, fig.cap = 'Frequency of somatic mutations in FGFR1/2/3/4 genes split by mutation type.'}

rep_figs$mutation_type$plot

```

__Figure \@ref(fig:fig-mutation-type). Frequency of somatic mutations in FGFR1/2/3/4 genes split by mutation type.__

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by their effects on the coding sequence. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. _

\newpage

```{r fig-variant-type, fig.width = figur::convert(rep_figs$variant_type, to = 'in')$w, fig.height = figur::convert(rep_figs$variant_type, to = 'in')$h, fig.cap = 'Frequency of somatic mutations in FGFR1/2/3/4 genes split by variant type.'}

rep_figs$variant_type$plot

```

__Figure \@ref(fig:fig-variant-type). Frequency of somatic mutations in FGFR1/2/3/4 genes split by variant type.__

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by variant type. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions. SNP: single nucleotide polymorphism, DNP: di-nucleotide polymorphism;  ONP: oligo-nucleotide polymorphism._

\newpage

```{r fig-protein-domain-split, fig.width = figur::convert(rep_figs$protein_domain_split, to = 'in')$w, fig.height = figur::convert(rep_figs$protein_domain_split, to = 'in')$h, fig.cap = 'Frequency of somatic mutations in FGFR1/2/3/4 genes split by protein domain.'}

rep_figs$protein_domain_split$plot

```

__Figure \@ref(fig:fig-protein-domain-split). Frequency of somatic mutations in FGFR1/2/3/4 genes split by protein domain.__

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by protein domain. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions._

\newpage

```{r tab-fgfr-class, tab.cap = 'Frequency of mutations in the FGFR1/2/3/4 genes in the GENIE BLCA and TCGA BLCA cohorts classified by mutation type, variant type, and protein domain. The table is available as a supplementary Excel file.'}

rep_tabs$fgfr_class[1, 1] %>% 
  mutate(`Classification scheme` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r fig-fgfr1-position, fig.width = figur::convert(rep_figs$fgfr1_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr1_position, to = 'in')$h, fig.cap = 'Positions of FGFR1 protein affected by somatic mutations.'}

rep_figs$fgfr1_position$plot

```

__Figure \@ref(fig:fig-fgfr1-position). Positions of FGFR1 protein affected by somatic mutations.__

_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein. Mutation localization and type are visualized in scatter plots below the FGFR1 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

\newpage

```{r fig-fgfr2-position, fig.width = figur::convert(rep_figs$fgfr2_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr2_position, to = 'in')$h, fig.cap = 'Positions of FGFR2 protein affected by somatic mutations.'}

rep_figs$fgfr2_position$plot

```

__Figure \@ref(fig:fig-fgfr2-position). Positions of FGFR2 protein affected by somatic mutations.__

_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein. Mutation localization and type are visualized in scatter plots below the FGFR2 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

\newpage

```{r fig-fgfr3-position, fig.width = figur::convert(rep_figs$fgfr3_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_position, to = 'in')$h, fig.cap = 'Positions of FGFR3 protein affected by somatic mutations.'}

rep_figs$fgfr3_position$plot

```

__Figure \@ref(fig:fig-fgfr3-position). Positions of FGFR3 protein affected by somatic mutations.__

_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein. Mutation localization and type are visualized in scatter plots below the FGFR3 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

\newpage

```{r fig-fgfr4-position, fig.width = figur::convert(rep_figs$fgfr4_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr4_position, to = 'in')$h, fig.cap = 'Positions of FGFR4 protein affected by somatic mutations.'}

rep_figs$fgfr4_position$plot

```

__Figure \@ref(fig:fig-fgfr4-position). Positions of FGFR4 protein affected by somatic mutations.__

_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein. Mutation localization and type are visualized in scatter plots below the FGFR4 protein domain schemes. Each point represents a single somatic mutation. Point color codes for protein domain. Total mutation numbers are displayed in the plot captions._

\newpage

## Co-occurrence of genetic alterations in FGF- and FGFR-coding genes with the most frequent genetic features of urothelial cancers

```{r fig-genetic-overlap, fig.width = figur::convert(rep_figs$genetic_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of the most common genetic alterations in urothelial cancer.'}

rep_figs$genetic_overlap$plot

```

__Figure \@ref(fig:fig-genetic-overlap). Co-occurrence of the most common genetic alterations in urothelial cancer.__

_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._

_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points. Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set. Note that genetic features displayed in the plots close to each other are expected to co-occur in a high fraction of cancer samples. Total numbers of samples are indicated in the plot captions._

_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features. Sample percentages and counts within the total sample sets are indicated in the heat map tiles. Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._

\newpage

```{r fig-fgfr-overlap, fig.width = figur::convert(rep_figs$fgfr_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.'}

rep_figs$fgfr_overlap$plot

```

__Figure \@ref(fig:fig-fgfr-overlap). Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.__

_Co-occurrence of somatic mutations, deletions, and amplifications of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._

_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points. Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set. Note that genetic features displayed in the plots close to each other are expected to co-occur in a high fraction of cancer samples. Total numbers of samples are indicated in the plot captions._

_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features. Sample percentages and counts within the total sample sets are indicated in the heat map tiles. Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._

\newpage

## Clinical and biological significance of genetic alterations of FGF- and FGFR-coding genes

```{r tab-fgfr3-clinic, tab.cap = 'Demographic, clinical, and pathological characteristic of cancers with WT and mutated FGFR3. Numeric features are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages and observation counts of the categories.'}

flextable::flextable(rep_tabs$fgfr3_clinic) %>% 
  set_widths(c(2.2, 3, rep(3.5, 2), 2.8, 2.8)) %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r fig-fgfr-correlation, fig.width = figur::convert(rep_figs$fgfr_correlation, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_correlation, to = 'in')$h, fig.cap = 'Correlation of mRNA levels of FGF- and FGFR-coding genes.'}

rep_figs$fgfr_correlation$plot

```

__Figure \@ref(fig:fig-fgfr-correlation). Correlation of mRNA levels of FGF- and FGFR-coding genes.__

_Pairwise correlation of log2-transformed mRNA levels of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor")] %>%  paste(collapse = ' and ')` cohorts by false discovery rate-adjusted Pearson's test. Correlation coefficients r for significant gene pairs are depicted in bubble plots. Color of the points codes for correlation sign. Point size represents absolute values of r. Numbers of analyzed cancer samples are displayed in the plot captions._

\newpage

```{r fig-fgfr-dge, fig.width = figur::convert(rep_figs$fgfr_dge, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_dge, to = 'in')$h, fig.cap = 'Expression of FGF- and FGFR-coding genes in urothelial cancers stratified by presence of FGFR3 mutations.'}

rep_figs$fgfr_dge$plot

```

__Figure \@ref(fig:fig-fgfr-dge). Expression of FGF- and FGFR-coding genes in urothelial cancers stratified by presence of FGFR3 mutations.__

_Differences in expression of FGF- and FGFR-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric. P values were corrected with the false discovery rate (FDR) method. Differentially regulated genes were identified by pFDR < 0.05 and at least 1.5 fold-regulation of gene expression in mutated samples as compared with WT specimens._

_(A) pFDR values and log2 fold-regulation estimates of gene expression in mutated vs WT samples are presented in Volcano plots. Each point represents a single gene. Point color codes for significance and regulation sign, differentially regulated genes are labeled with their symbols. The significance and fold-regulation cutoffs of differential gene expression are depicted as dashed lines. Numbers of analyzed cancer samples are indicated in the plot caption._

_(B) log2-transformed mRNA levels for genes found to be differentially regulated in both cohorts are presented in box plots. Median expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single observations are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

```{r tab-fgfr3-dge, tab.cap = "Differential expression of FGF- and FGFR-coding genes in urothelial cancers with and without FGFR3 mutations. log2-transformed expression levels are presented as medians with interquartile ranges and ranges. Statistical significance was determined by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$fgfr3_dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r fig-fgfr-tmb, fig.width = figur::convert(rep_figs$fgfr_tmb, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_tmb, to = 'in')$h, fig.cap = 'Total mutation burden and counts of copy number alterations in FGFR3 WT and FGFR3-mutated cancers.'}

rep_figs$fgfr_tmb$plot

```

__Figure \@ref(fig:fig-fgfr-tmb). Total mutation burden and counts of copy number alterations in FGFR3 WT and FGFR3-mutated cancers.__

_Total mutation burdens (TMB) were provided for the cohorts by the study authors. Counts of mutations, deletions and amplifications were computed for each cancer sample based on the published information. Note, that information on deletions and amplification was not available for the `r globals$cohort_labs["imvigor"]` cohort. TMB, and counts of mutations, deletions and amplifications were compared between cancers with and without FGFR3 mutations by Mann-Whitney test with r effect size statistic. P values were adjusted for multiple testing with the false discovery rate method. Median values with interquartile ranges are shown as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

```{r fig-fgfr-os, fig.width = figur::convert(rep_figs$fgfr_os, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_os, to = 'in')$h, fig.cap = 'Overall survival of patients with FGFR3 WT and FGFR3-mutated cancers.'}

rep_figs$fgfr_os$plot

```

__Figure \@ref(fig:fig-fgfr-os). Overall survival of patients with FGFR3 WT and FGFR3-mutated cancers.__

_Overall survival information was provided for the `r globals$cohort_labs["tcga"]` study participants. Approximate overall survival was calculated for the `r globals$cohort_labs["genie"]` cohort based on age at enrollment in years and age at the last follow-up or death in days. Overall survival was subsequently compared between patients with cancers with and without FGFR3 mutations. Statistical significance was determined by false discovery rate-adjusted Peto-Peto test. Fractions of surviving patients are presented in Kaplan-Meier plots. P values are indicated in the plots, numbers of observations and deaths are displayed in the plot captions. Numbers of observations and deaths in the FGFR3 mutation strata are presented in the plot legends._

\newpage

```{r fig-fgfr-rfs, fig.width = figur::convert(rep_figs$fgfr_rfs, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_rfs, to = 'in')$h, fig.cap = 'Disease-specific and relapse-free survival of patients with FGFR3 WT and FGFR3 mutated cancers.'}

rep_figs$fgfr_rfs$plot

```

__Figure \@ref(fig:fig-fgfr-rfs). Disease-specific and relapse-free survival of patients with FGFR3 WT and FGFR3 mutated cancers.__

_Disease-specific and relapse-free survival information was provided for the `r globals$cohort_labs["tcga"]` cohort. Differences in survival between patients with cancers with and without FGFR3 mutations were investigated by false discovery rate-adjusted Peto-Peto test. Fractions of surviving patients are presented in Kaplan-Meier plots. P values are indicated in the plots, numbers of observations and deaths are displayed in the plot captions. Numbers of observations and deaths in the FGFR3 mutation strata are presented in the plot legends._

\newpage

```{r fig-amp11q13, fig.width = figur::convert(rep_figs$amp11q13, to = 'in')$w, fig.height = figur::convert(rep_figs$amp11q13, to = 'in')$h, fig.cap = 'Expression of FGF- and FGFR-coding genes in cancers with and without amplification of the 11q13 chromosome region.'}

rep_figs$amp11q13$plot

```

__Figure \@ref(fig:fig-amp11q13). Expression of FGF- and FGFR-coding genes in cancers with and without amplification of the 11q13 chromosome region.__

_Amplification of the 11q13 chromosome region with FGF3, FGF4, FGF19 and CCND genes was identified as frequent genetic alteration in urothelial carcinoma. Differences in expression of FGF- and FGFR-coding genes between cancers with and without 11q13 amplification were assessed in the `r globals$cohort_labs["tcga"]` cohort by Mann-Whitney test with r effect size metric. P values were corrected for multiple testing with the false discovery rate method. Median log2-transformed mRNA levels with interquartile ranges are represented by boxes with whiskers spanning over 150% of the interquartile ranges for the significantly regulated genes. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

## Consensus molecular subtypes of urothelial cancers

```{r tab-cons-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes in the consensus molecular classes of urothelial cancers. The frequencies are presented as percentages of the consensus class. Statistical significance of differences between the consensus classes was determined by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$cons_genetics[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r fig-fgfr-subtypes, fig.width = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$h, fig.cap = 'Alterations of FGF- and FGFR-coding genes in consensus molecular classes of urothelial cancers.'}

rep_figs$fgfr_subtypes$plot

```

__Figure \@ref(fig:fig-fgfr-subtypes). Alterations of FGF- and FGFR-coding genes in consensus molecular classes of urothelial cancers.__

_Cancer samples from the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` data sets were assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC)._ 
_Differences in frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Note that copy number information was not available for the `r globals$cohort_labs["imvigor"]` cohort._

_(A) Percentages of cancer samples in the consensus molecular classes were visualized in a stack plot. Statistical significance of difference in class distribution between the cohorts was assessed by $\chi^2$ test with Cramer's V effect size statistic. The effect size and p value are shown in the plot caption. Total numbers of cancer samples are indicated in the Y axis._

_(B) Presence/absence of selected genetic alterations in the consensus molecular classes visualized in heat maps. Alteration names with effect sizes and p values of differences between the consensus classes are indicated in the Y axis. Significant effects are highlighted with bold font. Numbers of cancer samples in the consensus classes are displayed in the plot captions._

_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._

\newpage

```{r fig-fgfr-subdetails, fig.width = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$h, fig.cap = 'Frequency of somatic mutations of FGFFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.'}

rep_figs$fgfr_subdetails$plot

```

__Figure \@ref(fig:fig-fgfr-subdetails). Frequency of somatic mutations of FGFFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.__

_Assignment of cancer samples to consensus molecular classes and testing for differences in genetic alterations between the consensus classes were presented in Figure \@ref(fig:fig-fgfr-subtypes). Percentages of FGFR3 mutations, and amplifications of the FGF3, FGF4 and FGF19 genes in the consensus molecular classes are presented in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the consensus classes are indicated in the X axes._
_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._

\newpage

```{r tab-cons-dge, tab.cap = "Expression of FGF- and FGFR-coding genes in the consensus molecular classes of urothelial cancers. log2-transformed expression levels are presented as medians with interquartile ranges and ranges. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$cons_dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r fig-sub-fgfr-expression, fig.width = figur::convert(rep_figs$sub_fgfr_expression, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgfr_expression, to = 'in')$h, fig.cap = 'Differential expression of FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}

rep_figs$sub_fgfr_expression$plot

```

__Figure \@ref(fig:fig-sub-fgfr-expression). Differential expression of FGFR-coding genes in the consensus molecular classes of urothelial cancers.__

_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% paste(collapse = ' and ')` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes). Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method._
_Expression of FGFR1 and FGFR3 was found to differ significantly between the consensus classes in both cohorts. Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of cancer samples in the consensus classes are indicated in the X axes._ 
_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._

\newpage

```{r fig-sub-fgf-expression1, fig.width = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}

rep_figs$sub_fgf_expression1$plot

```

__Figure \@ref(fig:fig-sub-fgf-expression1). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__

_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% paste(collapse = ' and ')` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes). Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method._
_Expression of FGF2, FGF5, and FGF7 was found to differ significantly between the consensus classes in both cohorts. Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of cancer samples in the consensus classes are indicated in the X axes._ 
_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._

\newpage

```{r fig-sub-fgf-expression2, fig.width = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}

rep_figs$sub_fgf_expression2$plot

```

__Figure \@ref(fig:fig-sub-fgf-expression2). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__

_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% paste(collapse = ' and ')` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes). Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic. P values were corrected for multiple testing with the false discovery rate method._
_Expression of FGF10 and FGF18 was found to differ significantly between the consensus classes in both cohorts. Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of cancer samples in the consensus classes are indicated in the X axes._ 
_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._

\newpage

## Genetic subsets of urothelial carcinoma

```{r fig-lca-devel, fig.width = figur::convert(rep_figs$lca_devel, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_devel, to = 'in')$h, fig.cap = 'Development of genetic subsets of urothelial carcinoma in the GENIE BLCA training cohort. Prediction of the genetic subset assignment for the TCGA BLCA samples.'}

rep_figs$lca_devel$plot

```

__Figure \@ref(fig:fig-lca-devel). Development of genetic subsets of urothelial carcinoma in the GENIE BLCA training cohort. Prediction of the genetic subset assignment for the TCGA BLCA samples.__

_The genetic subsets of urothelial cancers were developed in the `r globals$cohort_labs["genie"]` training cohort by latent class modeling with binary indexes (presence/absence) of genetic alterations present in at least 5% of samples as modeling factors. Presence of FGF3, FGF4, FGF19, or CCND1 amplification was subsumed under 11q13 amplification prior to the modeling. Presence of CDKN2A, CDKN2B, or MTAP deletions was subsumed under 9p21 deletion prior to the modeling._

_(A) Selection of the genetic subset number in the `r globals$cohort_labs["genie"]` training cohort based on log-likelihood and Bayesian Information Criterion (BIC). Note the minimal BIC for the 6-subset model, which is subsequently used for definition of the genetic subsets._

_(B) The 6-subset model was trained in the `r globals$cohort_labs['genie']` cohort and used for assignment of cancer samples from the `r globals$cohort_labs["tcga"]` collective to the genetic subsets. Class assignment probabilities (posterior p) of single cancer samples are visualized in scatter plots. Note that high posterior p indicate high confidence at class assignment and stability of the modeling solution. Total observation numbers are displayed in the plot captions. Numbers of samples in the genetic subsets are indicated in the plot facets._

\newpage

```{r fig-lca-size, fig.width = figur::convert(rep_figs$lca_size, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_size, to = 'in')$h, fig.cap = 'Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort and the test TCGA BLCA collective.'}

rep_figs$lca_size$plot

```

__Figure \@ref(fig:fig-lca-size). Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort and the test TCGA BLCA collective.__

_Genetic subsets were developed in the `r globals$cohort_labs["genie"]` cohort and predicted for cancers samples in the `r globals$cohort_labs["tcga"]` as presented in Figure \@ref(fig:fig-lca-devel). Sizes of the genetic subsets expressed as percentages of cancer samples in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` collectives are presented in a stack plot._

\newpage

```{r tab-lca-genetics, tab.cap = "Frequency of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers. The frequencies are presented as percentages of samples with genetic alterations in the genetic subsets. Differences between the genetic subsets were assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

rep_tabs$lca_genetics[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(16)

```

\newpage

```{r fig-lca-genet, fig.width = figur::convert(rep_figs$lca_genet, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_genet, to = 'in')$h, fig.cap = 'Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.'}

rep_figs$lca_genet$plot

```

__Figure \@ref(fig:fig-lca-genet). Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.__

_Differences in frequency of the most common somatic mutations and copy number variants ($\geq$ 5% of samples) between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Presence of the genetic features in the genetic subsets was presented in a heat map. Significant alterations are highlighted in the Y axis with bold font. Numbers of analyzed cancer samples in the subsets are displayed in the plot captions._

\newpage

```{r fig-lca-details-mutations, fig.width = figur::convert(rep_figs$lca_details_mutations, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_mutations, to = 'in')$h, fig.cap = 'Frequency of selected somatic mutations in the genetic subsets of urothelial cancers.'}

rep_figs$lca_details_mutations$plot

```

__Figure \@ref(fig:fig-lca-details-mutations). Frequency of selected somatic mutations in the genetic subsets of urothelial cancers.__

_Differences in frequency of TP53, RB1, FGFR3, and ERBB2 mutations between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Percentages of the mutations in the genetic subsets are presented in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of cancer samples in the subsets are indicated in the X axes._

\newpage

```{r fig-lca-details-cna, fig.width = figur::convert(rep_figs$lca_details_cna, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_cna, to = 'in')$h, fig.cap = 'Frequency of selected copy number alterations in the genetic subsets of urothelial cancers.'}

rep_figs$lca_details_cna$plot

```

__Figure \@ref(fig:fig-lca-details-cna). Frequency of selected copy number alterations in the genetic subsets of urothelial cancers.__

_Differences in frequency of CCND1 amplifications (hallmark alteration for 11q13 chromosome region amplification), E2F3 amplifications, MDM2 amplifications, and CDKN2A deletions (characteristic for 9p21 chromosome region deletion) between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. Percentages of the copy number variants in the genetic subsets are presented in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of cancer samples in the subsets are indicated in the X axes._

\newpage

```{r fig-lca-tmb, fig.width = figur::convert(rep_figs$lca_tmb, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_tmb, to = 'in')$h, fig.cap = 'Total mutation burden, and counts of somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.'}

rep_figs$lca_tmb$plot

```

__Figure \@ref(fig:fig-lca-tmb). Total mutation burden, and counts of somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.__

_Total mutation burdens (TMB) were provided for the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by the study authors. Counts of mutations, deletions and amplifications were computed for each cancer sample based on the published information. TMB, and counts of mutations, deletions and amplifications were compared between the genetic subsets of urothelial cancers by Kruskal-Wallis test with $\eta^2$ effect size statistic. P values were adjusted for multiple testing with the false discovery rate method. Median values with interquartile ranges are shown as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

```{r tab-lca-dge, tab.cap = 'Expression of FGF- and FGFR-coding genes in the genetic subsets of urothelial cancers of the TCGA BLCA cohort. log2-transformed expression levels are presented as medians with interquartile ranges and ranges. Statistical significance of differences between the genetic subsets was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file.'}

rep_tabs$lca_dge[1, 1] %>% 
  mutate(Variable = '') %>% 
  set_names(' ') %>% 
  flextable %>%
  set_widths(16)

```

\newpage

```{r fig-lca-dge-fgfr, fig.width = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$h, fig.cap = 'Differential expression of FGFR-coding genes in the genetic subsets of urothelial cancers.'}

rep_figs$lca_dge_fgfr$plot

```

__Figure \@ref(fig:fig-lca-dge-fgfr). Differential expression of FGFR-coding genes in the genetic subsets of urothelial cancers.__

_Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic. Differences in expression of FGFR1, FGFR2, and FGFR3 were found significant. Median log2-transformed expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

```{r fig-lca-dge-fgf, fig.width = figur::convert(rep_figs$lca_dge_fgf, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgf, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the genetic subsets of urothelial cancers.'}

rep_figs$lca_dge_fgf$plot

```

__Figure \@ref(fig:fig-lca-dge-fgf). Differential expression of FGF-coding genes in the genetic subsets of urothelial cancers.__

_Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic. Differences in expression of FGF2, FGF5, FGF7, and FGF17 were found significant. Median log2-transformed expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Single cancer samples are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of analyzed cancer samples are indicated in the X axes._

\newpage

```{r tab-lca-clinic, tab.cap = 'Demographic, clinical, and pathological characteristic of the genetic subsets of urothelial cancers. Numeric features are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages and observation counts of the categories.'}

flextable::flextable(rep_tabs$lca_clinic) %>% 
  set_widths(c(1.5, 2, rep(2.1, 6), 2.3, 2.1)) %>% 
  merge_v(1) %>%
  footnote(1, 9:10, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r fig-lca-clinic, fig.width = figur::convert(rep_figs$lca_clinic, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_clinic, to = 'in')$h, fig.cap = 'Age and gender in the genetic subsets.'}

rep_figs$lca_clinic$plot

```

__Figure \@ref(fig:fig-lca-clinic). Age and gender in the genetic subsets.__

_Differences in age and gender between the genetic subsets of urothelial cancers in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were investigated by Kruskal-Wallis test and $\chi^2$ test, respectively. $\eta^2$ and Cramer's V served as effect size metrics. P values were corrected for multiple testing with the false discovery rate method. Median age with the interquartile range is presented in a box plot with whiskers spanning over 150% of the interquartile range and single cancer patients visualized as points. Percentages of women and men in the genetic subsets are shown in stack plots. Effect sizes and p values are displayed in the plot captions. Numbers of observations in the genetic subsets are indicated in the X axes._

\newpage

```{r fig-lca-os, fig.width = figur::convert(rep_figs$lca_os, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_os, to = 'in')$h, fig.cap = 'Overall survival in the genetic subsets of urothelial cancers.'}

rep_figs$lca_os$plot

```

__Figure \@ref(fig:fig-lca-os). Overall survival in the genetic subsets of urothelial cancers.__

_Overall survival information was provided for the `r globals$cohort_labs["tcga"]` study participants. Approximate overall survival was calculated for the `r globals$cohort_labs["genie"]` cohort based on age at enrollment in years and age at the last follow-up or death in days. Overall survival was subsequently compared between the genetic subsets by false discovery rate-corrected Peto-Peto test. Fractions of surviving patients are presented in Kaplan-Meier plots. P values for globals differences in survival between the subsets are indicated in the plots, numbers of observations and deaths are displayed in the plot captions. Numbers of observations and deaths in the FGFR3 mutation strata are presented in the plot legends._

\newpage

# Data and code availability

Data from publicly available sources were analyzed. 
The R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/FGFR-BLCA)

\newpage

# References

### 2024-05-16 report/markdown/report.Rmd
Analysis report
Figures for the MSK IMPACT cohort. Analysis summary and methods for the analysis report.

@@
                       message = FALSE, 
-                      dev = "CairoPNG", 
+                      dev = "cairo_pdf", 
                       dpi = 600)
@@
 ```
+
+\newpage
+
+# Analysis summary
+
+* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising therapy target in urothelial cancers with approval of the pan-FGFR inhibitor erdafitinib.
+However, effects of genetic alterations and modulation of expression of FGF- and FGFR-coding genes in urothelial carcinoma on tumor biology, therapy response, and clinical outcome are still not fully understood.
+
+* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations in FGF- and FGFR-coding genes in large publicly available cohorts of urothelial cancers.
+The secondary goal was an attempt to establish subsets of urothelial cancer specimens based on differences in their profiles of somatic mutations and gene copy number variants.
+
+* Four collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], and the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017].
+Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, and TCGA BLCA cohorts.
+Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort.
+Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives.
+Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA and IMvigor cohorts.
+Overall survival information was available for the MSK IMPACT, TCGA BLCA, and IMvigor cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective.
+Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days.
+Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.
+
+* Somatic mutations of _FGFR3_ belonged to the most common genetic features of urothelial cancers and were found in `r expl_genet$stats$mutation %>% filter(variable == 'FGFR3') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples.
+Somatic mutations of the remaining _FGFR1_, _FGFR2_, and _FGFR4_ receptor genes were less frequent (`r expl_genet$stats$mutation %>% filter(variable %in% c('FGFR1', 'FGFR2', 'FGFR4')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples).
+Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the 11q13 chromosome region, which codes also for an essential oncogene _CCDN1_.
+Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figures \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Tables \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).
+
+* A great majority of somatic mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ was classified as missense mutations caused by single nucleotide polymorphisms.
+Particularly for _FGFR3_, they affected protein residues were frequently located apart from the Ig-like domains responsible for ligand biding and besides the kinase domains vital for activation of downstream signaling (__Figures \@ref(fig:fig-mutation-type)__ - __\@ref(fig:fig-protein-domain-split)__, __Table \@ref(tab:tab-fgfr-class)__).
+A more detailed analysis revealed the Ig-like domain 1 and kinase domain as the mutation hot-spots of _FGFR1_.
+The kinase domain was the prime target of mutations of _FGFR2_.
+Hinge regions between the Ig-like domains 1 and 2, as well as residues around the membrane-spanning region between the Ig-like domain 3 and the kinase domain were identified as mutational hot-spots of _FGFR3_.
+In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-fgfr1-position)__ - __\@ref(fig:fig-fgfr4-position)__).
+It can be hence speculated, that somatic mutations of FGF receptors may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.
+
+* Co-occurrence and exclusivity of genetic features was investigated by analysis of pairwise Jacquard similarity coefficients for pairs of genetic alterations in urothelial cancer samples.
+Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).
+
+* Concerning co-regulation of expression of FGF- and FGFR-coding genes, _FGFR1_ and _FGFR3_ mRNA levels were found to be weakly but significantly negatively associated.
+Increased expression of _FGFR1_ in the cancer tissue correlated significantly with upregulation of genes coding for multiple FGF ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_, _FGF10_, and _FGF18_).
+This phenomenon was absent for _FGFR3_ (__Figure \@ref(fig:fig-fgfr-correlation)__).
+In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, we observed that presence of _FGFR3_ mutations translated in significantly and strongly increased _FGFR3_ mRNA levels.
+By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge)__, __Table \@ref(tab:tab-fgfr3-dge)__).
+This suggests two mechanisms leading to activation of pro-oncogenic FGFR signaling in urothelial carcinoma.
+In _FGFR3_ mutant tumors, FGFR signaling may be triggered by over-expression of FGFR3 protein, improved FGF-affinity, or even in an ligand-independent manner.
+Alternatively, concomitant over-expression of _FGFR1_ with its ligands may activated FGFR signaling in auto- or paracrine manner.
+
+* No consistent differences in total mutation burdens (TMB), counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations.
+Gene amplification numbers were in turn significantly lower in _FGFR3_ mutated neoplasms as compared with _FGF3_ wild-type tumors.
+Effect size of this difference was small (__Figure \@ref(fig:fig-fgfr-tmb)__).
+
+* There were no significant differences in age and gender distribution between patients with wild-type and _FGFR3_-mutated cancers.
+_FGFR3_ mutant cancers tended towards lower pathological stages, this effect was significant in the MSK IMPACT cohort but not in the TCGA BLCA collective (__Table \@ref(tab:tab-fgfr3-clinic)__).
+In the GENIE BLCA and MSK IMPACT cohorts, patients with _FGFR3_ mutated tumors demonstrated improved overall survival as compared with bearers of WT neoplasms.
+This phenomenon was also evident in the TCGA BLCA collective, it was only statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 2)`).
+There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA cohort, which may however result from insufficient number of events of interest (tumor-related deaths and relapses, __Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).
+
+* Analysis of effects of amplification of the 11q13 chromosome region on expression of FGF- and FGFR-coding genes was possible solely for the TCGA BLCA collective.
+Presence of this genetic aberration resulting in amplification of the _FGF3_, _FGF4_, _FGF19_, and _CCND1_ genes translated into lower expression of _FGFR1_ and significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA (__Figure \@ref(fig:fig-amp11q13)__).
+However, effects of those differences was weak and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF ligand-coding genes.
+Hence, biological relevance of the 11q13 amplification for FGFR signaling remains unclear.
+
+* Basal/squamous-like cancers followed by luminal papillary tumors constituted the largest consensus molecular classes [@Kamoun2020] of urothelial cancers in the TCGA BLCA and IMvigor cohorts (__Figure \@ref(fig:fig-fgfr-subtypes)A__).
+in both cohorts, _FGFR3_ mutated cancers were found to be significantly enriched in the papillary luminal class (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`% of specimens).
+In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the 11q13 chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens, __Figures \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).
+
+* mRNA levels of `r sub_dge$anova_significant %>% reduce(intersect) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were found to vary significant between the consensus molecular classes of urothelial carcinomas in both the TCGA BLCA and IMvigor data set (__Table \@ref(tab:tab-cons-dge)__).
+Expression of _FGFR1_ was the highest in stroma-rich and luminal non-specified tumors and the lowest in luminal papillary cancers.
+In turn, maximal mRNA levels of _FGFR3_ were detected in luminal papillary cancers and the lowest expression was observed in basal/squamous- and neuroendocrine-like tumors.
+Effect size of those differences was large (__Figure \@ref(fig:fig-sub-fgfr-expression)__).
+Expression of all significantly regulated _FGF_ ligands was the lowest in luminal papillary cancers.
+_FGF2_, _FGF7_, _FGF10_, and _FGF18_ transcript amounts followed the expression pattern of _FGFR1_ and were detected at the highest levels in stroma-rich and non-specified luminal cancers.
+_FGF5_ was expressed at the highest levels by basal/squamous-like, neuroendocrine-like and stroma-rich neoplasms.
+Effect sizes of differential regulation of these FGF-coding genes was large (__Figure \@ref(fig:fig-sub-fgf-expression1)__ and __\@ref(fig:fig-sub-fgf-expression2)__).
+Collectively, the differential gene expression analysis results support the model of activation of FGFR signaling propose above.
+While luminal papillary cancers may be driven by _FGFR3_ mutations and over-expression resulting in FGFR signaling with little dependence on the ligands, stroma-rich and non-specified luminal cancers may exploit auto- and paracrine signaling fueled by over-expression of _FGFR1_ and its ligands.
+
+* By latent class analysis (LCA, called also latent class modeling), we could establish a classification scheme of urothelial cancers based on presence/absence of the most frequent genetic alterations detected in at least 5% of cancer specimens.
+This classification scheme with six 'genetic subsets' was developed in the GENIE BLCA training cohort and technically validated in the MSK IMPACT and TCGA BLCA cohort (__Figures \@ref(fig:fig-lca-devel)__ and __\@ref(fig:fig-lca-size)__).
+The 'mutRB1' subset (`r lca_eval$n_numbers %>% map(filter, clust_id == 'mutRB1') %>% map_dbl(~.x$percent) %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens) was characterized by highly frequent _TP53_, _RB1_ and _ARID1A_ mutations, _E2F3_ amplifications, and nearly absent _FGFR3_ mutations, 11q13 amplification, and 9p21 deletions.
+The 'oligoMut' subset was hallmarked by frequent mutations of _TP53_ and the lowest overall mutation count.
+The key features of 'hyperMut' subset were the highest total mutation count, as well as frequent somatic mutations of _ARID1A_, _FGFR3_, _CREBBP_, _ERBB2_, _FBXW7_, and lacking copy number variants of 11q13 and 9p21.
+The 'del9p21' subset was characterized by relatively common deletions of the chromosome region 9p21, amplification of the 11q13 chromosome region, and highly frequent mutations of _TP53_ and _KDM6A_.
+In the 'mutFGFR3' subset, mutations of _FGFR3_ and _KDM6A_, and deletions of 9p21 were common, while mutations in _TP53_, _RB1_, and _ERBB2_, and gene amplifications were virtually absent.
+Finally, the 'ampMDM2' subset was characterized by the highest percentage of _MDM2_, highly common copy number variants of the 11p13 and 9p21 chromosome regions and essentially lacking mutations of _TP53_ and _RB1_ (__Figures \@ref(fig:fig-lca-genet1)__ - __\@ref(fig:fig-lca-details-cna2)__, __Table \@ref(tab:tab-lca-genetics)__).
+
+* Numbers of mutations differed significantly between the genetic subsets with large effect size and were the highest in hyperMut cancers and the lowest in oligoMut and mutFGFR3 tumors.
+Gene deletion numbers were significantly higher in del9p21 and ampMDM2 than in the remaining genetic subsets in the GENIE BLCA and MSK IMPACT collectives (large effect size).
+Amplification counts were the highest in del9p21, ampMDM2, and mutRB1 subsets and the lowest in mutFGFR3 cancers; those differences were large and significant in all collectives (__Figures \@ref(fig:fig-lca-tmb)__ and __\@ref(fig:fig-lca-cna-count)__).
+
+* We could observe significant differences in expression of _FGFR1_, _FGFR3_, and _FGFR3_, as well as _FGF2_, _FGF5_, FGF7_, and _FGF17_ genes between the genetic subsets of the TCGA BLCA cohort (__Table \@ref(tab:tab-lca-dge)__).
+Interestingly, mutFGFR3 cancers were found to over-express _FGFR3_ and, at the same time, to suppress _FGF2_, _FGF5_, and _FGF7_.
+The highest expression levels of _FGFR1_ were observed in mutRB1 and ampMDM2 genetic subsets; _FGFR2_ was found to be expressed in oligoMut, del9p21, and ampMDM2 at the highest levels (__Figures \@ref(fig:fig-lca-dge-fgfr)__ and __\@ref(fig:fig-lca-dge-fgf)__).
+
+* There were no significant differences in age and gender of cancer patients between the genetic clusters that were shared by the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives (__Figure \@ref(fig:fig-lca-clinic)__).
+Significantly lower pathological staging was observed in mutFGFR3 subset in the MSK IMPACT cohort, but this difference could not be corroborated in the TCGA BLCA collective.
+In the TCGA BLCA data set, mutFGFR3 and ampMDM2 genetic subsets comprised predominantly cancers assigned to the luminal papillary consensus class.
+Basal/squamous-like consensus class cancers predominated in mutRB1, oligoMut, hyperMut, and del9p21 genetic subsets (__Figure \@ref(fig:fig-lca-stage-class)__).
+Notably, consensus molecular classes of muscle-invasive urothelial carcinomas were established with bulk cancer transcriptome [@Kamoun2020; @Robertson2017].
+The disparity between the molecular consensus classes and the genetic subsets proposed by us suggests that phenotypically similar cancers may exploit highly dissimilar sets of driver genetic alterations and arise from unrelated evolutionary processes.
+
+* Significant differences in overall survival were observed between the genetic subsets in the GENIE BLCA and MSK IMPACT cohorts with the worst survival in mutRB1 subset and favorable prognosis inferred for mutFGFR3 and hyperMut cancers.
+No significant differences in overall survival were found in the TCGA cohort, which may, however, attributed to insufficient number of deaths in the genetic subsets (__Figure \@ref(fig:fig-lca-os)__).
+As investigated by univariable Cox regression, patients assigned to hyperMut subset were at roughly 50 - 75% significantly lower death risk, and patients in mutFGFR3 were at approximately 43 - 45% significantly lower death risk as compared with mutRB1 subset in both the GENIE BLCA and MSK IMPACT cohort.
+The general effect of genetic cluster assignment was found to be independent from age, sex, and pathological tumor stage (likelihood ratio test GENIE BLCA: `r lca_cox$lrt$plot_lab[[1]]`, MSK IMPACT: `r lca_cox$lrt$plot_lab[[2]]`).
+At the levels of single genetic subsets, hyperMut subset as favorable survival marker remained significant upon adjustment for age, sex, and pathological tumor stage.
+mutFGFR3 subset assignment was found independent from age and gender in the GENIE BLCA cohort but was confounded by age, sex, and pathological cancer stage in the MSK IMPACT collective (__Figure \@ref(fig:fig-lca-cox)__).
+We could not discern significant differences in disease-specific and relapse-free survival between the genetic subsets of the TCGA BLCA cohort (not shown).
+
+* Collectively, our analysis results suggest at least two mechanisms of FGFR signaling activation in urothelial cancers: the one exploiting FGFR3 mutations and largely independent on FGF ligands and operative predominantly in papillary cancers, and the other relying on over-expression of FGFR1 and FGF ligands predominantly by stroma-rich and non-specified luminal tumors.
+Furthermore, we put forward a relatively simple classification scheme of urothelial cancers based on the most common somatic mutations, gene and chromosome region deletions and amplifications established in the large GENIE BLCA collective.
+The key genetic features of those six genetic clusters such as subset-defining mutations and copy number variants, mutation and variant numbers could be reproduced in the MSK IMAPST and TCGA BLCA, i.e.
+two independent validation cohorts.
+Of clinical importance, the genetic subset classification bears consequences for overall survival in a age-, gender-, and stage-independent manner.
+
+* Our analysis bears limitations.
+Incompleteness of clinical information in the GENIE BLCA cohort precluded an in-depth characteristic of the FGFR3 mutation strata and genetic subsets.
+Unavailability of exact overall survival in this collective may have interfered with results of significant analyses.
+Detailed mutation data as well as copy number information were not provided for the IMvigor collective, which made validation of the genetic subsets impossible in this cohort of immunotherapy patients.
+Finally, gene expression data were not available for the largest GENIE BLCA and MSK IMPACT cohorts, which precluded detailed analyses of differential gene expression, biological phenotype, and consensus molecular classification based on transcriptome.
 
@@
 
-_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% paste(collapse = ', ')`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots.
+_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% collapse_and`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots.
 Alterations affecting FGF- and FGFR-coding genes are highlighted with bold font.
@@
 flextable::flextable(rep_tabs$cohorts) %>% 
-  set_widths(c(3, rep(3.5, 3), 2.8, 2.8)) %>% 
-  footnote(1, 5:6, 
+  set_widths(c(3, rep(3, 4), 2.6, 2.6)) %>% 
+  footnote(1, 6:7, 
            value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by their effects on the coding sequence.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and()` cohorts were classified by their effects on the coding sequence.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by variant type.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by variant type.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts were classified by protein domain.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by protein domain.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-```{r fig-fgfr1-position, fig.width = figur::convert(rep_figs$fgfr1_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr1_position, to = 'in')$h, fig.cap = 'Positions of FGFR1 protein affected by somatic mutations.'}
+```{r fig-fgfr1-position, fig.width = figur::convert(rep_figs$fgfr1_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr1_position, to = 'in')$h, fig.cap = 'Residues of FGFR1 protein affected by somatic mutations.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr1-position). Positions of FGFR1 protein affected by somatic mutations.__
+__Figure \@ref(fig:fig-fgfr1-position). Residues of FGFR1 protein affected by somatic mutations.__
 
-_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein.
+_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR1 protein domain schemes.
@@
 
-```{r fig-fgfr2-position, fig.width = figur::convert(rep_figs$fgfr2_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr2_position, to = 'in')$h, fig.cap = 'Positions of FGFR2 protein affected by somatic mutations.'}
+```{r fig-fgfr2-position, fig.width = figur::convert(rep_figs$fgfr2_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr2_position, to = 'in')$h, fig.cap = 'Residues of FGFR2 protein affected by somatic mutations.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr2-position). Positions of FGFR2 protein affected by somatic mutations.__
+__Figure \@ref(fig:fig-fgfr2-position). Residues of FGFR2 protein affected by somatic mutations.__
 
-_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein.
+_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR2 protein domain schemes.
@@
 
-```{r fig-fgfr3-position, fig.width = figur::convert(rep_figs$fgfr3_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_position, to = 'in')$h, fig.cap = 'Positions of FGFR3 protein affected by somatic mutations.'}
+```{r fig-fgfr3-position, fig.width = figur::convert(rep_figs$fgfr3_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_position, to = 'in')$h, fig.cap = 'Residues of FGFR3 protein affected by somatic mutations.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr3-position). Positions of FGFR3 protein affected by somatic mutations.__
+__Figure \@ref(fig:fig-fgfr3-position). Residues of FGFR3 protein affected by somatic mutations.__
 
-_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein.
+_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR3 protein domain schemes.
@@
 
-```{r fig-fgfr4-position, fig.width = figur::convert(rep_figs$fgfr4_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr4_position, to = 'in')$h, fig.cap = 'Positions of FGFR4 protein affected by somatic mutations.'}
+```{r fig-fgfr4-position, fig.width = figur::convert(rep_figs$fgfr4_position, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr4_position, to = 'in')$h, fig.cap = 'Residues of FGFR4 protein affected by somatic mutations.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr4-position). Positions of FGFR4 protein affected by somatic mutations.__
+__Figure \@ref(fig:fig-fgfr4-position). Residues of FGFR4 protein affected by somatic mutations.__
 
-_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein.
+_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR4 protein domain schemes.
@@
 
-```{r fig-genetic-overlap, fig.width = figur::convert(rep_figs$genetic_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of the most common genetic alterations in urothelial cancer.'}
+```{r fig-genetic-mds, fig.width = figur::convert(rep_figs$genetic_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_mds, to = 'in')$h, fig.cap = 'Co-occurrence of the most common genetic alterations in urothelial cancer.'}
 
-rep_figs$genetic_overlap$plot
+rep_figs$genetic_mds$plot
 
@@
 
-__Figure \@ref(fig:fig-genetic-overlap). Co-occurrence of the most common genetic alterations in urothelial cancer.__
+__Figure \@ref(fig:fig-genetic-mds). Co-occurrence of the most common genetic alterations in urothelial cancer.__
 
-_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
-
-_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points.
+_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
+_MDS results are visualized in scatter plots with single genetic alterations depicted as points.
 Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set.
@@
 
-_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features.
+\newpage
+
+```{r fig-genetic-cont-hm, fig.width = figur::convert(rep_figs$genetic_cont_hm, to = 'in')$w, fig.height = figur::convert(rep_figs$genetic_cont_hm, to = 'in')$h, fig.cap = 'Co-occurrence of selected genetic alterations in urothelial cancer.'}
+
+rep_figs$genetic_cont_hm$plot
+
+```
+
+__Figure \@ref(fig:fig-genetic-cont-hm). Co-occurrence of selected genetic alterations in urothelial cancer.__
+
+_Co-occurrence of the most common genetic alterations in urothelial cancers (at least 5% of cancer samples) was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by pairwise Jaccard's similarity coefficients J.
+Statistical significance of co-occurrence, i.e.
+$J > 0$, was assessed by false discovery rate-adjusted bootstrap tests._
+_Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features.
 Sample percentages and counts within the total sample sets are indicated in the heat map tiles.
-Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._
+J and p values are presented in the plot captions._
 
@@
 
-```{r fig-fgfr-overlap, fig.width = figur::convert(rep_figs$fgfr_overlap, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_overlap, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.'}
+```{r fig-fgfr-mds, fig.width = figur::convert(rep_figs$fgfr_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_mds, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.'}
 
-rep_figs$fgfr_overlap$plot
+rep_figs$fgfr_mds$plot
 
@@
 
-__Figure \@ref(fig:fig-fgfr-overlap). Co-occurrence of genetic alterations of FGF and FGFR genes in urothelial cancer.__
+__Figure \@ref(fig:fig-fgfr-mds). Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.__
 
-_Co-occurrence of somatic mutations, deletions, and amplifications of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by pairwise Jaccard's similarity coefficients and two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
-
-_(A) MDS results are visualized in scatter plots with single genetic alterations depicted as points.
+_Co-occurrence of somatic mutations, deletions, and amplifications of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
+_MDS results are visualized in scatter plots with single genetic alterations depicted as points.
 Point color codes for alteration type (somatic mutation, deletion, amplification), point size represents frequency of the alteration in the data set.
@@
 
-_(B) Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features.
+\newpage
+
+```{r fig-fgfr-cont-hm, fig.width = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations selected FGF- and FGFR-coding genes in urothelial cancer.'}
+
+rep_figs$fgfr_cont_hm$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr-cont-hm). Co-occurrence of genetic alterations selected FGF- and FGFR-coding genes in urothelial cancer.__
+
+_Co-occurrence of somatic mutations, deletions, and amplifications in FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by pairwise Jaccard's similarity coefficients J.
+Statistical significance of co-occurrence, i.e.
+$J > 0$, was assessed by false discovery rate-adjusted bootstrap tests._
+_Frequencies of samples with and without genetic alterations are visualized as heat maps of contingency tables for selected pairs of genetic features.
 Sample percentages and counts within the total sample sets are indicated in the heat map tiles.
-Jaccard's similarity coefficients and with their false discovery rate-corrected significance ($pFDR(J > 0)$) are presented in the plot captions._
+J and p values are presented in the plot captions._
 
@@
 ## Clinical and biological significance of genetic alterations of FGF- and FGFR-coding genes
-
-```{r tab-fgfr3-clinic, tab.cap = 'Demographic, clinical, and pathological characteristic of cancers with WT and mutated FGFR3.
-Numeric features are presented as medians with interquartile ranges and ranges.
-Qualitative variables are shown as percentages and observation counts of the categories.'}
-
-flextable::flextable(rep_tabs$fgfr3_clinic) %>% 
-  set_widths(c(2.2, 3, rep(3.5, 2), 2.8, 2.8)) %>% 
-  footnote(1, 5:6, 
-           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
-  theme_vanilla
-
-```
-
-\newpage
 
@@
 
-_Pairwise correlation of log2-transformed mRNA levels of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor")] %>%  paste(collapse = ' and ')` cohorts by false discovery rate-adjusted Pearson's test.
+_Pairwise correlation of log2-transformed mRNA levels of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor")] %>%  collapse_and` cohorts by false discovery rate-adjusted Pearson's test.
 Correlation coefficients r for significant gene pairs are depicted in bubble plots.
@@
 
-_Differences in expression of FGF- and FGFR-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric.
+_Differences in expression of FGF- and FGFR-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric in the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`.
 P values were corrected with the false discovery rate (FDR) method.
@@
 Note, that information on deletions and amplification was not available for the `r globals$cohort_labs["imvigor"]` cohort.
-TMB, and counts of mutations, deletions and amplifications were compared between cancers with and without FGFR3 mutations by Mann-Whitney test with r effect size statistic.
+TMB (A), and counts of mutations (B), deletions and amplifications (C, D) were compared between cancers with and without FGFR3 mutations by Mann-Whitney test with r effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method.
@@
 Numbers of analyzed cancer samples are indicated in the X axes._
+
+\newpage
+
+```{r tab-fgfr3-clinic, tab.cap = 'Demographic, clinical, and pathological characteristic of cancers with WT and mutated FGFR3.
+Numeric features are presented as medians with interquartile ranges and ranges.
+Qualitative variables are shown as percentages and observation counts of the categories.'}
+
+flextable::flextable(rep_tabs$fgfr3_clinic) %>% 
+  set_widths(c(2.2, 3, rep(3.5, 2), 2.8, 2.8)) %>% 
+  merge_v(1) %>% 
+  footnote(1, 5:6, 
+           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  theme_vanilla
+
+```
 
@@
 
-_Overall survival information was provided for the `r globals$cohort_labs["tcga"]` study participants.
+_Overall survival information was provided for the `r globals$cohort_labs[c("msk", "tcga", "imvigor")] %>% collapse_and` study participants.
 Approximate overall survival was calculated for the `r globals$cohort_labs["genie"]` cohort based on age at enrollment in years and age at the last follow-up or death in days.
@@
 
-_Cancer samples from the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` data sets were assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC)._
+_Cancer samples from the `r globals$cohort_labs[c("genie", "tcga")] %>% collapse_and` data sets were assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC)._
 _Differences in frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by $\chi^2$ test with Cramer's V effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
-Note that copy number information was not available for the `r globals$cohort_labs["imvigor"]` cohort._
+Note that copy number variant information was not available for the `r globals$cohort_labs["imvigor"]` cohort._
 
@@
 
-_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% paste(collapse = ' and ')` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
 Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
@@
 ```{r fig-lca-devel, fig.width = figur::convert(rep_figs$lca_devel, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_devel, to = 'in')$h, fig.cap = 'Development of genetic subsets of urothelial carcinoma in the GENIE BLCA training cohort.
-Prediction of the genetic subset assignment for the TCGA BLCA samples.'}
+Prediction of the genetic subset assignment for the MSK and TCGA BLCA samples.'}
 
@@
 
-__Figure \@ref(fig:fig-lca-devel). Development of genetic subsets of urothelial carcinoma in the GENIE BLCA training cohort. Prediction of the genetic subset assignment for the TCGA BLCA samples.__
+__Figure \@ref(fig:fig-lca-devel).
+Development of genetic subsets of urothelial carcinoma in the GENIE BLCA training cohort.
+Prediction of the genetic subset assignment for the MSK and TCGA BLCA cohort samples.__
 
@@
 _(A) Selection of the genetic subset number in the `r globals$cohort_labs["genie"]` training cohort based on log-likelihood and Bayesian Information Criterion (BIC).
-Note the minimal BIC for the 6-subset model, which is subsequently used for definition of the genetic subsets._
+Note the minimal BIC for the 6-class model, which is subsequently used for definition of the genetic subsets._
 
-_(B) The 6-subset model was trained in the `r globals$cohort_labs['genie']` cohort and used for assignment of cancer samples from the `r globals$cohort_labs["tcga"]` collective to the genetic subsets.
+_(B) The 6-class model was trained in the `r globals$cohort_labs['genie']` cohort and used for assignment of cancer samples from the `r globals$cohort_labs[c("msk", "tcga")] %>% collapse_and` test collectives to the genetic subsets.
 Class assignment probabilities (posterior p) of single cancer samples are visualized in scatter plots.
@@
 
-```{r fig-lca-size, fig.width = figur::convert(rep_figs$lca_size, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_size, to = 'in')$h, fig.cap = 'Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort and the test TCGA BLCA collective.'}
+```{r fig-lca-size, fig.width = figur::convert(rep_figs$lca_size, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_size, to = 'in')$h, fig.cap = 'Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort, and the MSK and TCGA BLCA test collectives.'}
 
@@
 
-__Figure \@ref(fig:fig-lca-size). Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort and the test TCGA BLCA collective.__
+__Figure \@ref(fig:fig-lca-size). Distribution of sizes of the genetic subsets in the training GENIE BLCA cohort, and the MSK and TCGA BLCA test collectives.__
 
-_Genetic subsets were developed in the `r globals$cohort_labs["genie"]` cohort and predicted for cancers samples in the `r globals$cohort_labs["tcga"]` as presented in Figure \@ref(fig:fig-lca-devel).
-Sizes of the genetic subsets expressed as percentages of cancer samples in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` collectives are presented in a stack plot._
+_Genetic subsets were developed in the `r globals$cohort_labs["genie"]` cohort and predicted for cancers samples in the `r globals$cohort_labs[c("msk", "tcga")] %>% collapse_and` collectives as presented in Figure \@ref(fig:fig-lca-devel).
+Sizes of the genetic subsets expressed as percentages of cancer samples in the cohort are presented in a stack plot._
 
@@
 
-```{r fig-lca-genet, fig.width = figur::convert(rep_figs$lca_genet, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_genet, to = 'in')$h, fig.cap = 'Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.'}
+```{r fig-lca-genet1, fig.width = figur::convert(rep_figs$lca_genet1, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_genet1, to = 'in')$h, fig.cap = 'Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers in the GENIE BLCA and MSK cohorts.'}
 
-rep_figs$lca_genet$plot
+rep_figs$lca_genet1$plot
 
@@
 
-__Figure \@ref(fig:fig-lca-genet). Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.__
+__Figure \@ref(fig:fig-lca-genet1). Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers in the GENIE BLCA and MSK cohorts.__
 
@@
 
-```{r fig-lca-details-mutations, fig.width = figur::convert(rep_figs$lca_details_mutations, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_mutations, to = 'in')$h, fig.cap = 'Frequency of selected somatic mutations in the genetic subsets of urothelial cancers.'}
+```{r fig-lca-genet2, fig.width = figur::convert(rep_figs$lca_genet2, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_genet2, to = 'in')$h, fig.cap = 'Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers in the TCGA BLCA cohort.'}
 
-rep_figs$lca_details_mutations$plot
+rep_figs$lca_genet2$plot
 
@@
 
-__Figure \@ref(fig:fig-lca-details-mutations). Frequency of selected somatic mutations in the genetic subsets of urothelial cancers.__
+__Figure \@ref(fig:fig-lca-genet2). Distribution of the most frequent somatic mutations and copy number alterations in the genetic subsets of urothelial cancers in the TCGA BLCA cohort.__
 
-_Differences in frequency of TP53, RB1, FGFR3, and ERBB2 mutations between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+_Differences in frequency of the most common somatic mutations and copy number variants ($\geq$ 5% of samples) between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Presence of the genetic features in the genetic subsets was presented in a heat map.
+Significant alterations are highlighted in the Y axis with bold font.
+Numbers of analyzed cancer samples in the subsets are displayed in the plot captions._
+
+\newpage
+
+```{r fig-lca-details-mutations1, fig.width = figur::convert(rep_figs$lca_details_mutations1, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_mutations1, to = 'in')$h, fig.cap = 'Frequency of somatic mutations of TP53 and RB1 in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_details_mutations1$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-details-mutations1). Frequency of somatic mutations of TP53 and RB1 in the genetic subsets of urothelial cancers.__
+
+_Differences in frequency of TP53 and RB1 mutations between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-```{r fig-lca-details-cna, fig.width = figur::convert(rep_figs$lca_details_cna, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_cna, to = 'in')$h, fig.cap = 'Frequency of selected copy number alterations in the genetic subsets of urothelial cancers.'}
+```{r fig-lca-details-mutations2, fig.width = figur::convert(rep_figs$lca_details_mutations2, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_mutations2, to = 'in')$h, fig.cap = 'Frequency of somatic mutations of FGFR3 and ERBB2 in the genetic subsets of urothelial cancers.'}
 
-rep_figs$lca_details_cna$plot
+rep_figs$lca_details_mutations2$plot
 
@@
 
-__Figure \@ref(fig:fig-lca-details-cna). Frequency of selected copy number alterations in the genetic subsets of urothelial cancers.__
+__Figure \@ref(fig:fig-lca-details-mutations2). Frequency of somatic mutations of FGFR3 and ERBB2 in the genetic subsets of urothelial cancers.__
 
-_Differences in frequency of CCND1 amplifications (hallmark alteration for 11q13 chromosome region amplification), E2F3 amplifications, MDM2 amplifications, and CDKN2A deletions (characteristic for 9p21 chromosome region deletion) between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+_Differences in frequency of FGFR3 and ERBB2 mutations between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of the mutations in the genetic subsets are presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of cancer samples in the subsets are indicated in the X axes._
+
+\newpage
+
+```{r fig-lca-details-cna1, fig.width = figur::convert(rep_figs$lca_details_cna1, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_cna1, to = 'in')$h, fig.cap = 'Frequency of amplifications of CCND1 and E3F3 in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_details_cna1$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-details-cna1). Frequency of amplifications of CCND1 and E3F3 in the genetic subsets of urothelial cancers.__
+
+_Differences in frequency of CCND1 (hallmark alteration for 11q13 chromosome region amplification) and E2F3 amplifications between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-```{r fig-lca-tmb, fig.width = figur::convert(rep_figs$lca_tmb, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_tmb, to = 'in')$h, fig.cap = 'Total mutation burden, and counts of somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.'}
+```{r fig-lca-details-cna2, fig.width = figur::convert(rep_figs$lca_details_cna2, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_details_cna2, to = 'in')$h, fig.cap = 'Frequency of amplifications of MDM2 and deletions of CDKN2A in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_details_cna2$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-details-cna2). Frequency of amplifications of MDM2 and deletions of CDKN2A in the genetic subsets of urothelial cancers.__
+
+_Differences in frequency of MDM2 amplifications and CDKN2A deletions (characteristic for 9p21 chromosome region deletion) between the genetic subsets of urothelial cancers were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+P values were corrected for multiple testing with the false discovery rate method.
+Percentages of the copy number variants in the genetic subsets are presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of cancer samples in the subsets are indicated in the X axes._
+
+\newpage
+
+```{r fig-lca-tmb, fig.width = figur::convert(rep_figs$lca_tmb, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_tmb, to = 'in')$h, fig.cap = 'Total mutation burden and counts of mutations in protein-coding genes in the genetic subsets of urothelial cancers.'}
 
@@
 
-__Figure \@ref(fig:fig-lca-tmb). Total mutation burden, and counts of somatic mutations and copy number alterations in the genetic subsets of urothelial cancers.__
+__Figure \@ref(fig:fig-lca-tmb). Total mutation burden and counts of mutations in protein-coding genes in the genetic subsets of urothelial cancers.__
 
-_Total mutation burdens (TMB) were provided for the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` cohorts by the study authors.
-Counts of mutations, deletions and amplifications were computed for each cancer sample based on the published information.
-TMB, and counts of mutations, deletions and amplifications were compared between the genetic subsets of urothelial cancers by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+_Total mutation burdens (TMB) were provided for the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by the study authors.
+Counts of mutations were computed for each cancer sample based on the published information.
+TMB and counts of mutations were compared between the genetic subsets of urothelial cancers by Kruskal-Wallis test with $\eta^2$ effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method.
+Median values with interquartile ranges are shown as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single cancer samples are visualized as points.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of analyzed cancer samples are indicated in the X axes._
+
+\newpage
+
+```{r fig-lca-cna-count, fig.width = figur::convert(rep_figs$lca_cna_count, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_cna_count, to = 'in')$h, fig.cap = 'Counts of gene deletions and amplifications in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_cna_count$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-cna-count). Counts of gene deletions and amplifications in the genetic subsets of urothelial cancers.__
+
+_Counts of deletions and amplifications were computed for cancer samples in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` based on the published information, and were compared between the genetic subsets of urothelial cancers by Kruskal-Wallis test with $\eta^2$ effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method.
@@
 
-_Differences in age and gender between the genetic subsets of urothelial cancers in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were investigated by Kruskal-Wallis test and $\chi^2$ test, respectively.
+_Differences in age (A) and gender (B) between the genetic subsets of urothelial cancers in the `r globals$cohort_labs[c("genie", "tcga")] %>% paste(collapse = ' and ')` were investigated by Kruskal-Wallis test and $\chi^2$ test, respectively.
 $\eta^2$ and Cramer's V served as effect size metrics.
 P values were corrected for multiple testing with the false discovery rate method.
-Median age with the interquartile range is presented in a box plot with whiskers spanning over 150% of the interquartile range and single cancer patients visualized as points.
+Median age with the interquartile range is presented in box plots with whiskers spanning over 150% of the interquartile range and single cancer patients visualized as points.
 Percentages of women and men in the genetic subsets are shown in stack plots.
@@
 Numbers of observations in the genetic subsets are indicated in the X axes._
+
+\newpage
+
+```{r fig-lca-stage-class, fig.width = figur::convert(rep_figs$lca_stage_class, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_stage_class, to = 'in')$h, fig.cap = 'Pathological tumor stage and consensus molecular class distribution in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_stage_class$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-stage-class). Pathological tumor stage and consensus molecular class distribution in the genetic subsets of urothelial cancers.__
+
+_Differences in distribution of pathological tumor stages (A) and consensus molecular classes (B) between the genetic subsets were investigated by $\chi^2$ test with Cramer's V effect size statistic.
+P values were adjusted for multiple testing with the false discovery rate method.
+Percentages of pathological stages and consensus classes within the genetic subsets are presented in stack plots.
+Effect sizes and p values are displayed in the plot captions.
+Numbers of complete observations are indicated in the X axes._
 
@@
 
-_Overall survival information was provided for the `r globals$cohort_labs["tcga"]` study participants.
+_Overall survival information was provided for the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` study participants.
 Approximate overall survival was calculated for the `r globals$cohort_labs["genie"]` cohort based on age at enrollment in years and age at the last follow-up or death in days.
@@
 Numbers of observations and deaths in the FGFR3 mutation strata are presented in the plot legends._
+
+\newpage
+
+```{r fig-lca-cox, fig.width = figur::convert(rep_figs$lca_cox, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_cox, to = 'in')$h, fig.cap = 'Cox modeling of overall survival in the genetic subsets of urothelial carcinoma.'}
+
+rep_figs$lca_cox$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-cox). Cox modeling of overall survival in the genetic subsets of urothelial carcinoma.__
+
+_Overall survival (`r globals$cohort_labs["msk"]`) and approximate overall survival (`r globals$cohort_labs["genie"]`) was modeled as a function of genetic subset assignment alone or as a function of genetic subset assignment and clinical confounders (`r globals$cohort_labs["genie"]`: age and gender, `r globals$cohort_labs["msk"]`: age, gender, and pathological tumor stage) with Cox proportional hazard regression.
+Significance of the genetic subset term was investigated by likelihood ratio test (LRT) between the model with genetic subset assignment only and the model with genetic subset assignment and clinical confounders as explanatory factors.
+_
+
+_(A) Numeric statistics of performance of Cox models with genetic subset assignment and Cox models with genetic subset assignment and clinical confounders as explanatory factors.
+Concordance between the observed and predicted survival was assessed by Harrell's concordance index (high values: good concordance), model calibration was investigated by integrated Brier score (IBS; low values: good calibration), and fraction of survival variance explained by the model was gauged by R^2 statistic.
+The statistic values are visualized in a bubble plot with point color coding for model type and point size representing R^2.
+Numbers of complete observations and deaths are indicated in the plot facets._
+
+_(B, C) Cox model coefficient estimates expressed as hazard ratios (HR).
+HR with 95% confidence intervals are visualized with Forest plots faceted by model type.
+Numbers of complete observations and deaths, and results of LRT are presented in the plot captions._
+
+\newpage
+
+# Methods
+
+## Software
+
+Data handling and analysis was accomplished with R version 4.2.3. 
+For general transformation tasks for matrix, tabular, and text data, base R, _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022], [_trafo_](https://github.com/PiotrTymoszuk/trafo), and _stringi_ [@Gagolewski2021] were utilized.
+General data visualization tasks were performed with _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019], and _flextable_ [@Gohel2022]. 
+R packages specific for particular analysis and visualization steps are referenced below.
+
+## Data sources
+
+Four collectives of urothelial carcinoma patients were evaluated in the current report: 
+the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
+the MSK IMPACT collective [@Clinton2022], 
+the bladder cancer TCGA collective referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
+and the IMvigor clinical study cohort [@Balar2017].
+
+The GENIE BLCA cohort [@Sweeney2017] includes n = `r nrow(genie$clinic)` cases of bladder and urinary tract cancers selected by oncotree classification 'BLAD', 'BLCA', 'BLSC', 'UA', 'UAD', 'UCA', 'UCU', 'UPA', 'URCA', 'USCC', or 'UTUC'.
+The available data sets for the GENIE BLCA cohort included basic clinical information (age at sequencing in years, sex, follow-up in days, mortality), total mutation burden (TMB) expressed as fraction of genome, detailed information on location, variant type, nucleotide sequence and protein residues affected by somatic mutations, as well as 0/1 coded absence/presence of gene deletions and amplifications.
+Approximate overall survival was computed as the time interval between the last follow-up in days and age at sequencing in years. 
+In case of multiple samples per participant, the earliest sample, i.e. the specimen with the lowest ID, was included in the current analysis. 
+The GENIE BLCA data sets were fetched from the [cBioportal API](https://www.cbioportal.org/) with tools provided by the _cbioportalR_ package [@Whiting2024] and formatted with in-house developed R scripts.
+
+The MSK IMPACT collective consisted of n = `r nrow(msk$clinic)` patients with predominantly muscle-invasive urothelial cancers [@Clinton2022].  
+In case of multiple samples available for a patient, the sample with the lowest ID was included in the analysis. 
+The clinical information included entries on participants age, gender, overall survival, cancer pathological stage, and cancer treatment. 
+The mutation data set included detailed information on mutation nucleotide sequence, variant type, and affected protein residues. 
+The copy number variant data were provided in 0/1 coded presence of gene deletions and amplifications. 
+The data sets for the MSK impact study were downloaded from [cBioportal](https://www.cbioportal.org/) and implemented in R with in-house developed scripts.
+
+Data sets for the TCGA BLCA cohort [@Liu2018; @Robertson2017] was fetched from the [cBioportal](https://www.cbioportal.org/) repository with in-house developed scripts for n = `r table(tcga$expression$tissue_type)['tumor']` predominantly muscle-invasive urothelial cancer patients.
+The clinical information consisted of demographic variables (e.g.
+age, sex, BMI, smoking history), pathological features (e.g.
+pathological staging, histology, grade, extracapsular extension), survival information (overall, tumor-specific and relapse-free survival), and TMB expressed as mutation count per MB.
+The somatic mutation data set included detailed information on location, variant type, nucleotide sequence and affected protein residues. 
+The copy number variant data set consisted of 0/1 coded presence of gene amplifications and deletions. 
+The expression data set for the TCGA BLCA collective consisted of $log_2(x + 1)$-transformed RSEM-normalized  gene-integrated transcript counts obtained by bulk cancer RNA sequencing.
+
+The IMvigor data set included a subset of `r nrow(imvigor$expression)` bladder and ureter muscle non-invasive and muscle-invasive cancers of the IMvigor 210 trial subjected to RNAseq and DNA sequencing [@Balar2017; @Mariathasan2018].
+The study participants were treated with an anti-PD-L1 drug (atezolizumab) following platinum chemotherapy. 
+The clinical data set included basic demographic, clinical and pathological features, best overall response (complete/partial response [CR/PR], stable/progressive disease [SD/PD]), overall survival, and TMB quantified as mutation count per MB.
+The mutation data set consisted of 0/1 coded basic information on presence of any somatic mutations in a particular gene. 
+The expression data set consisted of $log_2(x + 1)$-transformed gene-integrated transcript counts obtained by bulk cancer RNA sequencing. 
+The IMvigor data sets were obtained from the R package _IMvigor210CoreBiologies_.
+
+Assignment of the TCGA BLCA and IMvigor cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome log_2 mRNA expression levels (R package _consensusMIBC_) [@Kamoun2020].
+
+JSON files with domain structures of FGFR1 (ID: P11362), FGFR2 (ID: P21802), FGFR3 (ID: P22607), and FGFR4 (ID: P22455) proteins were downloaded from [UniProt](https://www.uniprot.org) and processed with the _jsonlite_ package [@Ooms2023].
+
+## Descriptive statistics, statistical inference, and effect size
+
+If not indicated otherwise, descriptive statistics for quantitative variables included median, interquartile range, and range. 
+Descriptive statistic for qualitative variables were percentage and count within categories. 
+Differences of means or medians of qualitative variables were investigated by Mann-Whitney, tow-tailed T test, Kruskal-Wallis test, and one-way ANOVA, depending on the number of analysis groups and normality of distribution of the variable.
+Correlation was assessed by Pearson's correlation test. 
+Effect size of differences in variable levels, distribution, and correlation was assessed with the following statistics [@Cohen2013]:
+
+* r for differences between two groups, non-parametric tests: small for r < 0.2, medium for 0.2 $\leq$ r < 0.4, large for r $\geq$ 0.4
+
+* Cohen's d for differences between two groups, parametric tests: small for d < 0.5, medium for 0.5 $\leq$ d < 0.8, large for d $\geq$ 0.8
+
+* $\eta^2$ for differences between three or more groups: small for $\eta^2$ < 0.06, moderate for 0.06 $\leq$ $\eta^2$ < 0.14, large for $\eta^2 \geq$ 0.14
+
+* Cramer's V for comparison of category frequencies of a numeric variable: small for r < 0.2, medium for 0.2 $\leq$ r < 0.4, large for r $\geq$ 0
+
+* Pearson's $\rho$ for correlation, parametric tests: small for $\rho$ < 0.3, medium for 0.3 $\leq$ $\rho$ < 0.5, large for r $\geq$ 0.5
+
+Hypothesis testing results were adjusted for multiple comparisons separately for each analysis step and investigated cohort with the false discovery rate method (FDR) [@Benjamini1995]. 
+Differences with p < 0.05 were considered statistically significant. 
+Statistical hypothesis testing was accomplished with the development packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [@Mangiafico2022; @Kassambara2021] internally employing statistical tools of base R, _stats_, _rstatix_, and _rcompanion_ [@Mangiafico2022; @Kassambara2021].
+
+## Survival analysis
+
+Differences in overall, disease-specific, and relapse-free survival between patients stratified by genetic features such as _FGFR3_ mutations of genetic subsets were investigated with Peto-Peto test. 
+Fractions of surviving patients were visualized with Kaplan-Meier plots. 
+Survival analyses and visualizations were done with packages _survival_ and _survminer_ [@Kassambara2016; @Therneau2000].
+
+## Co-occurrence of genetic features
+
+Co-occurrence analysis was performed in two settings: 
+(1) for the most common genetic features (mutations, amplifications, deletions) present in at least 5% of cancer samples in all cohorts and 
+(2) for all available mutations, amplifications, and deletions of FGF- and FGFR-coding genes. 
+Co-occurrence of genetic features was investigated by Jaccard's similarity coefficient J and Jaccard's distance computed with the $D = 1 - J$ formula. 
+Significance of co-occurrence, i.e. $J > 0$, was assessed by bootstrap test with B = 10000 iterations. 
+Mutual co-occurrence (e.g. between _FGFR3_ mutations and _RB1_ mutations) was depicted as a heat map of contingency table.
+
+For visualization of co-occurrence of multiple genetic features in two-dimensional Euclidean space, matrices of pairwise Jaccard's distances were subjected to multi-dimensional scaling (MDS). 
+Subsequently, the genetic features of interest were displayed in scatter plots with point size corresponding to the general frequency of the genetic alteration.
+
+Co-occurrence analysis was done with [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) and _jaccard_ R packages [@Chung2018].
+
+## Genetic subsets of urothelial cancers
+
+Definition and evaluation of genetic subsets of urothelial cancers was done for the GENIE BLCA training cohort, and the MSK IMPACT and TCGA BLCA test collectives, for which both mutation and copy number variant data were provided.
+
+Genetic subsets of urothelial cancers were developed for the GENIE BLCA training cohort with latent class modeling [@Weller2020; @Linzer2011]. 
+The explanatory variables were binary indexes (presence/absence) of genetic alterations present in at least 5% of samples in all data sets. 
+Presence of FGF3, FGF4, FGF19, or CCND1 amplification was subsumed under 11q13 amplification prior to the modeling.
+Presence of CDKN2A, CDKN2B, or MTAP deletions was subsumed under 9p21 deletion prior to the modeling.
+The latent class models were fit with R function `tune_lca()` (package [_polcaExtra_](https://github.com/PiotrTymoszuk/polcaExtra)) with the maximal iteration number set to 3000 and numbers of replicates, i.e.
+random starts of the modeling procedure, set to 5.
+Selection of the optimal subset number was motivated by the minimum of Bayesian Information Criterion (BIC). 
+With this selection criterion, a six-subset model whose classes were named after characteristic gene and chromosome alterations was chosen for definition of the genetic subsets.
+
+Subsequently, assignment of cancer samples from the MSK IMPACT and TCGA BLCA test collectives to the genetic subsets was predicted based on conditional subset assignment probabilities using `predict()` method from [_polcaExtra_](https://github.com/PiotrTymoszuk/polcaExtra) package.
+Quality of the subset assignment in the training and test collective was assessed by a visual inspection of the class assignment probabilities (posterior p) and by comparison of frequency of the subset-defining genetic features between the subsets ($\chi^2$ test with Cramer's V effect size statistic).
+
+## Surviival modeling
+
+For each of the GENIE BLCA and MSK IMPACT cohorts, two types of Cox proportional hazard models of overall survival were constructed: (1) a model with genetic subset assignment as a sole explanatory factor, and (2) a model with genetic subset assignment and clinically relevant confounders (GENIE BLCA: age and gender, MSK IMPACT: age, gender, and pathological tumor stage).
+The Cox models were constructed, checked for the proportional hazard assumption and evaluated with tools provided by R packages _survival_ [@Grambsch1994; @Therneau2000] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
+Overall Cox model performance was evaluated by Harrell's C index (high values: good concordance between the predicted and observed survival), integrated Brier score (IBS; low values: good model calibration), and R^2 as a measure of survival variance explained by the model.
+Significance of the genetic subset term and its independence from clinical confounders was determined by likelihood ratio test between the model with subset assignment and the model with cluster assignment and clinical confounders as explanatory factors (function `anova()`).
+Estimates of model coefficients were retrieved with `summary()` method.
 

### 2024-12-05 FGFR3 analysis scripts/amplification_gene_expression.R
R code update
Expansion of the set of genes of interrest. Differential gene expression analyses done with `fastTest`. Analyses of genetic alterations with `perich`.  Machine learning modeling of consensus molecular classes by Kamoun et al. Prognostic FGF signature established by Elastic Net Cox regression.

@@ -13,7 +13,8 @@
 
   ## expression data
 
-  fgfr_ampl$variables <- globals[c("receptors", "ligands")] %>%
+  fgfr_ampl$variables <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
     reduce(union)
 
   fgfr_ampl$data <- tcga$expression %>%
@@ -45,31 +46,30 @@
   insert_msg('Descriptive stats')
 
   fgfr_ampl$stats <- fgfr_ampl$data %>%
-    explore(variables = fgfr_ampl$variables,
-            split_factor = 'amp11q13',
-            what = 'table',
-            pub_styled = TRUE) %>%
-    format_desc
+    fast_num_stats(variables = fgfr_ampl$variables,
+                   split_fct = 'amp11q13')
 
 # Testing for differences between WT and amplified cancers -------
 
   insert_msg('Testing for differences')
 
-  fgfr_ampl$test <- fgfr_ampl$data %>%
-    test_two_groups(split_fct = 'amp11q13',
-                    variables = fgfr_ampl$variables,
-                    type = 'wilcox',
-                    adj_method = 'BH') %>%
-    re_adjust(method = 'none') %>%
-    mutate(eff_size = paste('r =', signif(effect_size, 2)),
-           plot_cap = paste(eff_size, significance, sep = ', '),
-           variable = response)
+  fgfr_ampl$test <-
+    f_wilcox_test(fgfr_ampl$data[fgfr_ampl$variables],
+                  f = fgfr_ampl$data$amp11q13,
+                  exact = FALSE,
+                  as_data_frame = TRUE,
+                  adj_method = 'BH',
+                  safely = TRUE) %>%
+    re_adjust %>%
+    mutate(eff_size = paste('r =', signif(biserial_r, 2)),
+           plot_cap = paste(eff_size, significance, sep = ', ')) %>%
+    as_tibble
 
   ## significant differences
 
   fgfr_ampl$significant <- fgfr_ampl$test %>%
     filter(p_adjusted < 0.05) %>%
-    .$response
+    .$variable
 
 # Box plots --------
 
@@ -103,10 +103,6 @@
                     fgfr_ampl$test) %>%
     format_res_tbl(dict = NULL,
                    remove_complete = TRUE) %>%
-    full_rbind(tibble(variable = 'Samples, N',
-                      WT = table(fgfr_ampl$data$amp11q13)['WT'],
[…]

### 2024-12-05 FGFR3 analysis scripts/clinic.R

@@ -15,7 +15,7 @@
   fgfr_clinic$lexicon <- globals$clinic_lexicon %>%
     filter(!variable %in% c('tmb_per_mb', 'bi_response', 'death')) %>%
     mutate(test_type = ifelse(format == 'numeric',
-                              'kruskal_etasq', 'cramer_v'),
+                              'wilcoxon_r', 'cramer_v'),
            plot_type = ifelse(format == 'numeric',
                               'box', 'stack'),
            ax_label = ifelse(is.na(unit),
@@ -31,7 +31,13 @@
   fgfr_clinic$data <-
     map2(fgfr_globals$genetic_groups[names(fgfr_clinic$data)],
          fgfr_clinic$data,
-         inner_join, by = 'sample_id')
+         inner_join, by = 'sample_id') %>%
+    map(map_dfc, function(x) if(is.factor(x)) droplevels(x) else x)
+
+  fgfr_clinic$data$bcan <- fgfr_clinic$data$bcan %>%
+    mutate(tissue = ifelse(tissue == 'bladder',
+                           'bladder', 'non-bladder'),
+           tissue = factor(tissue, c('bladder', 'non-bladder')))
 
   ## data set-specific variable vectors
 
@@ -39,6 +45,19 @@
     map(names) %>%
     map(~filter(fgfr_clinic$lexicon, variable %in% .x)) %>%
     map(~.x$variable)
+
+# N numbers -------
+
+  insert_msg('Total numbers of samples')
+
+  fgfr_clinic$n_numbers <- fgfr_clinic$data %>%
+    map(count, FGFR3_mutation) %>%
+    map(column_to_rownames, 'FGFR3_mutation') %>%
+    map(t) %>%
+    map(as.data.frame) %>%
+    map(mutate, variable = 'Patients, N') %>%
+    map(relocate, variable) %>%
+    map(as_tibble)
 
 # Descriptive stats -------
 
@@ -73,7 +92,13 @@
                             pub_styled = TRUE,
                             adj_method = 'BH')) %>%
     map(mutate,
-        plot_cap = paste(eff_size, significance, sep = ', '))
+        plot_cap = paste(eff_size, significance, sep = ', '),
+        significance = ifelse(stri_detect(eff_size, fixed = 'Inf'),
+                              NA, significance),
+        plot_cap = ifelse(stri_detect(eff_size, fixed = 'Inf'),
+                          NA, plot_cap),
[…]

### 2024-12-05 FGFR3 analysis scripts/disease_relapse_free_survival.R

@@ -11,24 +11,36 @@
 
   insert_msg('Analysis data')
 
-  fgfr_rfs$data <- list(tss = tcga, rfs = tcga) %>%
+  fgfr_rfs$data <- list(tss_tcga = tcga,
+                        rfs_tcga = tcga,
+                        tss_bcan = bcan) %>%
     map(~.x$clinic)
 
-  fgfr_rfs$data$tss <-
-    fgfr_rfs$data$tss[c('sample_id', 'tss_days', 'tumor_death', 'pt_stage')]
+  fgfr_rfs$data[c("tss_tcga", "tss_bcan")] <-
+    fgfr_rfs$data[c("tss_tcga", "tss_bcan")] %>%
+    map(~.x[c('sample_id',
+              'tss_days',
+              'tumor_death',
+              'pt_stage')]) %>%
+    map(set_names,
+        c('sample_id', 'time', 'event', 'pt_stage'))
 
-  fgfr_rfs$data$rfs <-
-    fgfr_rfs$data$rfs[c('sample_id', 'rfs_days', 'relapse', 'pt_stage')]
+  fgfr_rfs$data$rfs_tcga <-
+    fgfr_rfs$data$rfs[c('sample_id',
+                        'rfs_days',
+                        'relapse',
+                        'pt_stage')] %>%
+    set_names(c('sample_id', 'time', 'event', 'pt_stage'))
 
-  fgfr_rfs$data <- fgfr_rfs$data %>%
-    map(inner_join,
-        fgfr_globals$genetic_groups$tcga,
-        by = 'sample_id')
+  fgfr_rfs$data <-
+    map2(fgfr_rfs$data,
+         fgfr_globals$genetic_groups[c("tcga", "tcga", "bcan")],
+         inner_join, by = 'sample_id')
 
   ## variants of the data set without T4 cancers
 
-  fgfr_rfs$data[c('tss_wo_t4', 'rfs_wo_t4')] <-
-    fgfr_rfs$data[c("tss", "rfs")] %>%
+  fgfr_rfs$data[c('tss_tcga_wo_t4', 'rfs_tcga_wo_t4', 'tss_bcan_wo_t4')] <-
+    fgfr_rfs$data %>%
     map(filter, pt_stage != 'T4')
 
   fgfr_rfs$data <- fgfr_rfs$data %>%
@@ -43,10 +55,11 @@
 
   fgfr_rfs$n_numbers <- fgfr_rfs$data %>%
     map(~tibble(n_total = nrow(.x),
[…]

### 2024-12-05 FGFR3 analysis scripts/icb_response.R

@@ -1,5 +1,5 @@
 # Best overall response to ICB in the IMvigor cohort stratified by
-# prsence/absence of FGFR3 mutations
+# presence/absence of FGFR3 mutations
 
   insert_head()
 

### 2024-12-05 FGFR3 analysis scripts/overall_survival.R

@@ -19,10 +19,10 @@
                        fgfr_os$data,
                        inner_join, by = 'sample_id')
 
-  ## TCGA and MSK: variants of the data sets without pT4 cancers
+  ## TCGA, MSK, and BCAN: variants of the data sets without pT4 cancers
 
-  fgfr_os$data[c('msk_wo_t4', 'tcga_wo_t4')] <-
-    fgfr_os$data[c('msk', 'tcga')] %>%
+  fgfr_os$data[c('msk_wo_t4', 'tcga_wo_t4', 'bcan_wo_t4')] <-
+    fgfr_os$data[c('msk', 'tcga', 'bcan')] %>%
     map(filter, pt_stage != 'T4')
 
   fgfr_os$data <- fgfr_os$data %>%
@@ -105,7 +105,7 @@
   fgfr_os$km_plots <-
     list(x = fgfr_os$km_plots,
          y = c(globals$cohort_labs,
-               paste(globals$cohort_labs[c("msk", "tcga")],
+               paste(globals$cohort_labs[c("msk", "tcga", "bcan")],
                      'without pT4 cancers')),
          w = fgfr_os$n_captions,
          z = fgfr_os$legend_labs) %>%
@@ -123,23 +123,56 @@
 
   insert_msg('Result table')
 
-  fgfr_os$result_tbl <- fgfr_os$stats %>%
+  ## numbers of patients and events
+
+  fgfr_os$result_tbl$n_numbers <- fgfr_os$n_groups %>%
+    map(column_to_rownames, 'FGFR3_mutation') %>%
+    map(select, n_total, n_events) %>%
+    map(t) %>%
+    map(as_tibble) %>%
+    map(mutate, variable = c('Patients, N', 'Events, N')) %>%
+    map(relocate, variable)
+
+  ## median survival times and test results
+
+  fgfr_os$result_tbl$stats <- fgfr_os$stats %>%
     map(mutate,
-        median = paste0(signif(median, 3),
-                           ' [95% CI: ',
-                           signif(lower, 3),
-                           ' to ', signif(upper, 3), ']')) %>%
+        median = ifelse(is.na(median),
+                        '',
+                        paste0(signif(median, 3),
+                               ' [95% CI: ',
+                               signif(lower, 3),
+                               ' to ', signif(upper, 3), ']'))) %>%
     map(select, FGFR3_mutation, median) %>%
     map(column_to_rownames, 'FGFR3_mutation') %>%
     map(t) %>%
     map(as_tibble)
 
[…]

### 2024-12-05 FGFR3 analysis scripts/tmb.R

@@ -13,14 +13,17 @@
 
   ## TMB provided by the study authors
 
-  fgfr_tmb$data$tmb <- globals$cohort_expr %>%
+  fgfr_tmb$raw_data$tmb <- globals$cohort_expr %>%
     eval %>%
     map(~.x$clinic) %>%
-    map(select, sample_id, tmb_per_mb, any_of('msi_score'))
+    map(safely(select),
+        sample_id, tmb_per_mb, any_of('msi_score')) %>%
+    map(~.x$result) %>%
+    compact
 
-  fgfr_tmb$data$tmb <-
-    map2(fgfr_globals$genetic_groups[names(fgfr_tmb$data$tmb)],
-         fgfr_tmb$data$tmb,
+  fgfr_tmb$raw_data$tmb <-
+    map2(fgfr_globals$genetic_groups[names(fgfr_tmb$raw_data$tmb)],
+         fgfr_tmb$raw_data$tmb,
          inner_join, by = 'sample_id') %>%
     map(~filter(.x, complete.cases(.x)))
 
@@ -28,7 +31,7 @@
   ## Amplifications and deletions in the IMvigor cohort are copies
   ## of the mutation counts
 
-  fgfr_tmb$data$counts <-  globals$cohort_expr %>%
+  fgfr_tmb$raw_data$counts <-  globals$cohort_expr %>%
     eval %>%
     map(~.x[c('mutation', 'deletion', 'amplification')]) %>%
     map(compact) %>%
@@ -41,10 +44,28 @@
                         values_to = .y))) %>%
     map(reduce, left_join, by = 'sample_id')
 
-  fgfr_tmb$data <-
-    map2(fgfr_tmb$data$tmb,
-         fgfr_tmb$data$counts,
-         left_join, by = 'sample_id')
+  for(i in names(fgfr_tmb$raw_data$counts)) {
+
+    if(!is.null(fgfr_tmb$raw_data$tmb[[i]])) {
+
+      fgfr_tmb$data[[i]] <-
+        full_join(fgfr_tmb$raw_data$tmb[[i]],
+                  fgfr_tmb$raw_data$counts[[i]],
+                  by = 'sample_id')
+
+    } else {
+
+      fgfr_tmb$data[[i]] <-
+        inner_join(fgfr_globals$genetic_groups[[i]],
+                   fgfr_tmb$raw_data$counts[[i]],
+                   by = 'sample_id')
+
+    }
+
+  }
[…]

### 2024-12-05 FGFR3_analysis.R

@@ -18,6 +18,7 @@
 
   library(exda)
   library(microViz)
+  library(fastTest)
 
   library(survival)
   library(survminer)

### 2024-12-05 exec.R

@@ -6,6 +6,8 @@
                      'exploration.R',
                      'FGFR3_analysis.R',
                      'subtypes.R',
+                     'ml.R',
+                     'survival.R',
                      'latent_class.R',
                      'report.R'),
                    message = TRUE, crash = TRUE))

### 2024-12-05 exploration scripts/alteration_networks.R

@@ -1,0 +1,147 @@
+# Networks of mutations, deletions, and amplifications.
+#
+# The affinity matrices are constructed with Jaccard similarity coefficient J,
+# and the cutoff of J >= 0.2.
+
+
+  insert_head()
+
+# container -------
+
+  expl_alnet <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ## attribute table:
+
+  expl_alnet$attr_table <-
+    globals[c("ligands", "receptors", "binding_proteins")] %>%
+    map(~tibble(variable = .x)) %>%
+    compress(names_to = 'protein_type') %>%
+    mutate(protein_type = stri_replace(protein_type,
+                                       regex = 's$',
+                                       replacement = ''),
+           protein_type = stri_replace(protein_type,
+                                       fixed = '_',
+                                       replacement = ' '),
+           protein_type = factor(protein_type,
+                                 c('receptor', 'ligand', 'binding protein')))
+
+  expl_alnet$variables <- expl_alnet$attr_table$variable
+
+  expl_alnet$attr_table <-
+    list(x = c('_mut', '_amp', '_del'),
+         y = c('mutation', 'amplification', 'deletion')) %>%
+    pmap_dfr(function(x, y) expl_alnet$attr_table %>%
+               mutate(gene_symbol = variable,
+                      variable = paste0(variable, x),
+                      alteration = y))
+
+  ## analysis data: binary-coded presence of alterations
+
+  for(i in list(c('mutation', '_mut'),
+                c('amplification', '_amp'),
+                c('deletion', '_del'))) {
+
+    expl_alnet$data[[i[[1]]]] <- globals$cohort_expr %>%
+      eval %>%
+      map(~.x[[i[[1]]]]) %>%
+      compact %>%
+      map(column_to_rownames, 'sample_id') %>%
+      map(select, any_of(expl_alnet$variables)) %>%
+      map(~set_names(.x,
+                     paste0(names(.x), i[[2]])))
+
+  }
+
+  expl_alnet$data <- expl_alnet$data %>%
+    transpose %>%
+    map(compact) %>%
+    map(reduce, cbind)
+
+  ## cohort-specific variable vectors
+
+  expl_alnet$variables <- expl_alnet$data %>%
+    map(names)
+
+# Jaccard similarity and construction of graphs --------
+
+  insert_msg('Jaccard simlarity and graphs')
+
+  expl_alnet$simil_mtx <- expl_alnet$data %>%
+    map(t) %>%
[…]

### 2024-12-05 exploration scripts/co_expression_networks.R

@@ -1,0 +1,123 @@
+# Networks of co-expression of FGF ligands and FGFR receptors.
+#
+# Spearman's correlation-weighted graph objects are created for gene - gene
+# correlations of rho >= 0.3.
+# Identification of the hubs: top hub score as a metric of overall correlation
+# strength.
+
+  insert_head()
+
+# container -------
+
+  expl_exnet <- list()
+
+# analysis variables and data -------
+
+  insert_msg('Analysis variables and data')
+
+  ## attribute table: indicates if a gene codes for a ligand
+  ## or a receptor
+
+  expl_exnet$attr_table <-
+    globals[c("ligands", "receptors", "binding_proteins")] %>%
+    map(~tibble(gene_symbol = .x)) %>%
+    compress(names_to = 'protein_type') %>%
+    mutate(protein_type = stri_replace(protein_type,
+                                       regex = 's$',
+                                       replacement = ''),
+           protein_type = stri_replace(protein_type,
+                                       fixed = '_',
+                                       replacement = ' '),
+           protein_type = factor(protein_type,
+                                 c('receptor', 'ligand', 'binding protein')))
+
+  expl_exnet$variables <- expl_exnet$attr_table$gene_symbol
+
+  ## analysis data and a cohort-specific list of variables
+
+  expl_exnet$data <- globals$cohort_expr %>%
+    eval %>%
+    map(~.x$expression) %>%
+    compact %>%
+    map(column_to_rownames, 'sample_id') %>%
+    map(select, any_of(expl_exnet$variables))
+
+  expl_exnet$variables <- expl_exnet$data %>%
+    map(names) %>%
+    map(intersect, expl_exnet$variables)
+
+# Graph objects -------
+
+  insert_msg('Graph objects')
+
+  expl_exnet$graph_obj <- expl_exnet$data %>%
+    map(as_iGraph,
+        fun = cor,
+        cutoff = 0.3,
+        method = 'spearman',
+        weighted = TRUE,
+        diag = FALSE) %>%
+    map(prune_degree, cutoff = 0)
+
+# Node importance stats --------
+
+  insert_msg('Node importance stats')
+
+  expl_exnet$stats <- expl_exnet$graph_obj %>%
+    map(summary) %>%
+    map(select, -index)
+
+# appending the graphs with attributes and node importance stats -------
+
+  insert_msg('Setting the node importance')
+
+  expl_exnet$graph_obj <-
+    map2(expl_exnet$graph_obj,
+         expl_exnet$stats,
[…]

### 2024-12-05 exploration scripts/co_occurrence.R

@@ -2,7 +2,7 @@
 # FGF/FGFR genes.
 #
 # Co-occurrence is investigated by two-dimensional MDS of the Jaccard distance
-# matrices. The analysis is done for the GENIE and TCGA cohort.
+# matrices. The analysis is done for the GENIE, MSK, and TCGA cohort.
 
   insert_head()
 
@@ -27,7 +27,8 @@
     map(map,
         select,
         sample_id,
-        any_of(reduce(globals[c("receptors", "ligands")], union))) %>%
+        any_of(reduce(globals[c("receptors", "ligands", "binding_proteins")],
+                      union))) %>%
     map(map, column_to_rownames, 'sample_id') %>%
     map(~map2(., names(.),
               ~set_names(.x, paste(names(.x), .y, sep = '_')))) %>%

### 2024-12-05 exploration scripts/coexpression.R

@@ -21,7 +21,8 @@
 
   ## analysis variables
 
-  expl_corr$variables <- globals[c("receptors", "ligands")]
+  expl_corr$variables <-
+    globals[c("receptors", "ligands", "binding_proteins")]
 
   ## analysis data
 
@@ -47,12 +48,13 @@
   for(i in names(expl_corr$data)) {
 
     expl_corr$test[[i]] <- expl_corr$variable_pairs %>%
-      future_map_dfr(~correlate_variables(expl_corr$data[[i]],
-                                          variables = .x,
-                                          what = 'correlation',
-                                          type = 'pearson',
-                                          ci = TRUE),
-                     .options = furrr_options(seed = TRUE)) %>%
+      future_map(~safely(correlate_variables)(expl_corr$data[[i]],
+                                              variables = .x,
+                                              what = 'correlation',
+                                              type = 'pearson',
+                                              ci = TRUE),
+                 .options = furrr_options(seed = TRUE)) %>%
+      map_dfr(~.x$result) %>%
       re_adjust(method = 'BH') %>%
       mutate(corr_id = paste(variable1, variable2, sep = '_'),
              variable1 = factor(variable1, reduce(expl_corr$variables, c)),
@@ -70,10 +72,11 @@
         abs(estimate) >= 0.3) %>%
     map(~.x$corr_id)
 
-  ## effects shared by both cohorts
+  ## effects shared by at least two cohorts
 
   expl_corr$common_significant <- expl_corr$significant %>%
-    reduce(intersect)
+    shared_features(m = 2) %>%
+    as.character
 
 # Bubble plots for the shared significant effects -------
 
@@ -103,12 +106,12 @@
                       fill = factor(sign(estimate)),
                       size = abs(estimate))) +
            geom_point(shape = 21) +
-           geom_text(aes(label = signif(estimate, 2),
-                         color = factor(sign(estimate))),
-                     size = 2.5,
-                     vjust = -1.2,
-                     hjust = 0.5,
-                     show.legend = FALSE) +
+           #geom_text(aes(label = signif(estimate, 2),
+            #             color = factor(sign(estimate))),
+             #        size = 2.2,
+              #       vjust = -1.2,
+               #      hjust = 0.5,
+                #     show.legend = FALSE) +
            scale_fill_manual(values = c('-1' = 'steelblue',
                                         '0' = 'gray60',
                                         '1' = 'firebrick'),
@@ -123,7 +126,7 @@
                                          '0' = 'ns',
[…]

### 2024-12-05 exploration scripts/cohorts.R

@@ -10,9 +10,16 @@
 
   insert_msg('Analysis variables and data')
 
+  ## for BCAN: collapsing the non-bladder and metastasis organs
+
   expl_cohorts$data <- globals$cohort_expr %>%
     eval %>%
     map(~.x$clinic)
+
+  expl_cohorts$data$bcan <- expl_cohorts$data$bcan %>%
+    mutate(tissue = ifelse(tissue == 'bladder',
+                           'bladder', 'non-bladder'),
+           tissue = factor(tissue, c('bladder', 'non-bladder')))
 
   expl_cohorts$lexicon <- globals$clinic_lexicon %>%
     mutate(unit = ifelse(variable == 'tmb_per_mb',
@@ -59,7 +66,8 @@
            cohort = factor(cohort, unname(globals$cohort_labs)),
            death = as.character(death),
            death = car::recode(death, "'0' = 'no'; '1' = 'yes'"),
-           death = factor(death, c('no', 'yes')))
+           death = factor(death, c('no', 'yes'))) %>%
+    map_dfc(function(x) if(is.factor(x)) droplevels(x) else x)
 
 # Descriptive stats -------
 
@@ -82,15 +90,19 @@
   expl_cohorts$test_lexicon <- expl_cohorts$lexicon %>%
     filter(test_type != 'none')
 
-  expl_cohorts$test <- expl_cohorts$data %>%
-    compare_variables(variables = expl_cohorts$test_lexicon$variable,
-                      split_factor = 'cohort',
-                      what = 'eff_size',
-                      types = expl_cohorts$test_lexicon$test_type,
-                      ci = FALSE,
-                      exact = FALSE,
-                      pub_styled = TRUE,
-                      adj_method = 'BH') %>%
+  expl_cohorts$test <-
+    list(x = expl_cohorts$test_lexicon$variable,
+         y = expl_cohorts$test_lexicon$test_type) %>%
+    pmap_dfr(function(x, y) expl_cohorts$data[, c('cohort', x)] %>%
+               filter(complete.cases(.)) %>%
+               compare_variables(variables = x,
+                                 split_factor = 'cohort',
+                                 what = 'eff_size',
+                                 types = y,
+                                 ci = FALSE,
+                                 exact = FALSE,
+                                 pub_styled = TRUE)) %>%
+    re_adjust %>%
     mutate(plot_cap = paste(eff_size, significance, sep = ', '))
 
[…]

### 2024-12-05 exploration scripts/details.R

@@ -1,5 +1,5 @@
 # Details of mutation type and mutation location for FGFR1 - 4.
-# The analysis is done for the GENIE and TCGA cohorts.
+# The analysis is done for the GENIE, MSK, TCGA, and BCAN cohorts.
 
   insert_head()
 
@@ -15,12 +15,14 @@
 
   expl_fgfr$data <- list(genie = genie,
                          msk = msk,
-                         tcga = tcga) %>%
+                         tcga = tcga,
+                         bcan = bcan) %>%
     map(~.x$mutation_detail)
 
   expl_fgfr$n_totals <- list(genie = genie,
                              msk = msk,
-                             tcga = tcga) %>%
+                             tcga = tcga,
+                             bcan = bcan) %>%
     map(~.x$mutation) %>%
     map_dbl(nrow)
 
@@ -51,7 +53,8 @@
                                          "'DEL' = 'deletion';
                                          'INS' = 'insertion'"))
 
-  expl_fgfr$data[c('msk', 'tcga')] <- expl_fgfr$data[c('msk', 'tcga')] %>%
+  expl_fgfr$data[c('msk', 'tcga', 'bcan')] <-
+    expl_fgfr$data[c('msk', 'tcga', 'bcan')] %>%
     map(transmute,
         sample_id = Tumor_Sample_Barcode,
         gene_symbol = Hugo_Symbol,
@@ -73,6 +76,8 @@
                                     'Silent' = 'silent';
                                     'Splice_Region' = 'splice region';
                                     'Splice_Site' = 'splice site'"),
+        mutation_type = ifelse(mutation_type == "3'Flank",
+                               "3'UTR", mutation_type),
         variant_type = car::recode(Variant_Type,
                                    "'DEL' = 'deletion'; 'INS' = 'insertion'"))
 
@@ -91,7 +96,8 @@
                                  'splice region',
                                  'splice site',
                                  'intron',
-                                 "3'UTR")),
+                                 "3'UTR",
+                                 'silent')),
         variant_type = factor(variant_type,
                               c('SNP', 'DNP', 'ONP',
                                 'deletion', 'insertion')))
@@ -154,22 +160,22 @@
 
   expl_fgfr$stack_plots <-
     list(data = unlist(expl_fgfr$stats, recursive = FALSE),
-         fill_var = c(rep('mutation_type', 3),
-                      rep('variant_type', 3),
-                      rep('protein_domain', 3)),
[…]

### 2024-12-05 exploration scripts/dge.R

@@ -11,24 +11,20 @@
 
   expl_dge <- list()
 
-# parallel backend ------
-
-  insert_msg('Parallel backend')
-
-  plan('multisession')
-
 # Analysis data -------
 
   insert_msg('Analysis data')
 
   ## gene expression variables
 
-  expl_dge$variables <- globals[c("receptors", "ligands")]
+  expl_dge$variables <- globals[c("receptors", "ligands", "binding_proteins")]
 
-  expl_dge$split_factors <- expl_dge$variables$receptors %>%
-    paste0('_mut')
+  expl_dge$split_factors <- paste0('FGFR', 1:4, '_mut')
 
-  expl_dge$split_labels <- expl_dge$variables$receptors %>%
+  expl_dge$split_factors <- set_names(expl_dge$split_factors,
+                                      expl_dge$split_factors)
+
+  expl_dge$split_labels <- paste0('FGFR', 1:4) %>%
     html_italic %>%
     set_names(expl_dge$split_factors)
 
@@ -48,7 +44,7 @@
   expl_dge$mutation <- globals$cohort_expr %>%
     eval %>%
     map(~.x$mutation) %>%
-    map(select, sample_id, all_of(expl_dge$variables$receptors)) %>%
+    map(select, sample_id, all_of(paste0('FGFR', 1:4))) %>%
     map(set_names, c('sample_id', expl_dge$split_factors))
 
   for(i in names(expl_dge$mutation)) {
@@ -68,8 +64,14 @@
          inner_join, by = 'sample_id') %>%
     map(~filter(.x, complete.cases(.x)))
 
+  ## data set-specific expression lists
+
   expl_dge$variables <- expl_dge$variables %>%
     reduce(union)
+
+  expl_dge$variables <- expl_dge$data %>%
+    map(names) %>%
+    map(intersect, expl_dge$variables)
 
 # N numbers ------
 
@@ -92,20 +94,13 @@
   for(i in names(expl_dge$data)) {
 
     expl_dge$stats[[i]] <- expl_dge$split_factors %>%
-      future_map(~explore(expl_dge$data[[i]],
-                          variables = expl_dge$variables,
-                          split_factor = .x,
-                          what = 'table',
-                          pub_styled = TRUE),
-                 .options = furrr_options(seed = TRUE)) %>%
-      map(format_desc) %>%
-      set_names(expl_dge$split_factors)
+      map(fast_num_stats,
[…]

### 2024-12-05 exploration scripts/genetics.R

@@ -40,14 +40,45 @@
   ## components of the FGF/FGFR system are highlighted with bold font
 
   expl_genet$fgf_genes <-
-    list(receptors = globals$receptors,
-         ligands = globals$ligands)
+    globals[c("receptors", "ligands", "binding_proteins")]
 
   expl_genet$lexicon <-
     tibble(variable = reduce(expl_genet$variables, union)) %>%
     mutate(label = html_italic(variable),
            label = ifelse(variable %in% reduce(expl_genet$fgf_genes, union),
                           html_bold(label), label))
+
+  ## data for the FGF, FGFR, and FGFBP-coding genes
+  ## detected in at least two cohorts:
+  ## if a gene is not present in a cohort it is assumed all WT
+
+  expl_genet$fgf_data <- globals$cohort_expr %>%
+    eval %>%
+    map(~.x[c('mutation', 'deletion', 'amplification')]) %>%
+    transpose %>%
+    map(compact) %>%
+    map(map,
+        select,
+        any_of(reduce(expl_genet$fgf_genes, union)))
+
+  expl_genet$fgf_genes <- expl_genet$fgf_data %>%
+    map(map, colSums) %>%
+    map(map, ~.x[.x > 0]) %>%
+    map(map, names) %>%
+    map(shared_features, m = 2) %>%
+    map(as.character)
+
+  expl_genet$fgf_data <-
+    map2(expl_genet$fgf_data,
+         expl_genet$fgf_genes,
+         function(x, y) x %>%
+           map(select, any_of(y))) %>%
+    map(~map2_dfr(.x, names(.x),
+                  function(dat, coh) mutate(dat, cohort = coh))) %>%
+    map(relocate, cohort)
+
+  expl_genet$fgf_data <- expl_genet$fgf_data %>%
+    map(map_dfc, ~ifelse(is.na(.x), 0, .x))
 
 # N numbers -------
 
@@ -72,12 +103,27 @@
 
   insert_msg('Frequency of alterations')
 
-  expl_genet$stats <- expl_genet$data %>%
-    map(count_binary,
+  ## frequency of all alterations,
+  ## and alterations in the FGF, FGFR, and FGFBP genes
+
+  expl_genet[c("stats", "fgf_stats")] <-
+    expl_genet[c("data", "fgf_data")] %>%
+    map(map,
+        count_binary,
         split_fct = 'cohort') %>%
-    map(mutate,
+    map(map,
+        mutate,
[…]

### 2024-12-05 exploration scripts/subtypes.R

@@ -1,5 +1,5 @@
-# differences in distribution of the consensus molecular subtypes between
-# the cohorts. Done for the TCGA BLCA and Imvigor cohorts
+# Differences in distribution of the consensus molecular subtypes between
+# the cohorts. Done for the TCGA BLCA, Imvigor, and BCAN cohorts
 
   insert_head()
 
@@ -13,10 +13,12 @@
 
   expl_sub$data <- subtypes$assignment %>%
     map(~.x[c('sample_id', 'consensusClass')]) %>%
+    map(filter, !is.na(consensusClass), consensusClass != 'not assigned') %>%
     compress(names_to = 'cohort') %>%
     mutate(cohort = unname(globals$cohort_labs[cohort]),
            cohort = factor(cohort, unname(globals$cohort_labs)),
-           cohort = droplevels(cohort)) %>%
+           cohort = droplevels(cohort),
+           consensusClass = droplevels(consensusClass)) %>%
     filter(complete.cases(.))
 
 # Descriptive stats ------

### 2024-12-05 exploration.R

@@ -11,12 +11,14 @@
 #
 # 5) Characteristic of the mutation types and location for the FGFR genes.
 #
-# 6) Co-expression of FGF and FGFR gene mRNA in the cancer tissue.
+# 6) Co-expression of FGF and FGFR gene mRNA in the cancer tissue: correlation
+# and network analysis.
 #
 # 7) Differential expression of FGFR and FGF genes in cancer samples with and
 # without mutations in the FGFR1 - 4.
 #
-# 8) Co-occurrence of alterations in FGF and FGFR genes.
+# 8) Co-occurrence of alterations in FGF and FGFR genes: multi-dimensional
+# scaling and a network analysis.
 
 # tools -------
 
@@ -28,6 +30,10 @@
   library(exda)
   library(survival)
   library(survminer)
+  library(fastTest)
+
+  library(igraph)
+  library(graphExtra)
 
   library(jsonlite)
 
@@ -80,6 +86,12 @@
   access_cache(cache_path = './cache/expl_occur.RData',
                script_path = './exploration scripts/co_occurrence.R',
                message = 'Loading cached co-occurrence analysis results')
+
+  ## network analyses
+
+  c('./exploration scripts/co_expression_networks.R',
+    './exploration scripts/alteration_networks.R') %>%
+    source_all(message = TRUE, crash = TRUE)
 
 # END -----
 

### 2024-12-05 import scripts/bcan.R

@@ -1,0 +1,257 @@
+# Import of the BCAN study data: clinical information, 0/1-coded gene-level
+# mutation information, and Z-scores of log2 expression counts.
+
+  insert_head()
+
+# container -------
+
+  bcan <- list()
+
+# clinical information -------
+
+  insert_msg('Clinical information')
+
+  bcan$clinic$patient <-
+    read_tsv('./data/BCAN/data_clinical_patient.txt', skip = 4)
+
+  ## patient's data
+
+  bcan$clinic$patient <- bcan$clinic$patient %>%
+    transmute(
+              ## demography
+
+              patient_id = PATIENT_ID,
+              sex = factor(tolower(SEX), c('female', 'male')),
+              race = factor(RACE),
+              race = fct_recode(race, Other = 'Unknown'),
+              age = as.numeric(AGE_AT_DIAGNOSIS),
+              smoking = factor(SMOKING_STATUS),
+              smoking = fct_recode(smoking,
+                                   `current or previous` = 'Current Smoker',
+                                   `current or previous` = 'Former Smoker',
+                                   never = 'Never Smoker'),
+              smoking = factor(smoking, c('never', 'current or previous')),
+              baseline_ecog = factor(BASELINE_ECOG),
+
+              ## anatomy
+
+              primary_location = stri_extract_first(PRIMARY_TUMOR_LOCATION,
+                                                    regex = 'Bladder|Ureter|Urethra|(Renal\\s{1}pelvis)'),
+              primary_location = factor(tolower(primary_location)),
+              primary_location = fct_relevel(primary_location, 'bladder'),
+              tissue = factor(tolower(SAMPLE_SITE)),
+              tissue = fct_relevel(tissue, 'bladder'),
+              tissue_type = factor(tolower(PRIMARY_MET)),
+              tissue_type = fct_recode(tissue_type,
+                                       tumor = 'primary',
+                                       metastasis = 'met'),
+              tissue_type = fct_relevel(tissue_type, 'tumor', 'metastasis'),
+
+              ## surgery
+
+              cystectomy = ifelse(stri_detect(PRIMARY_SURGERY,
+                                              regex = '(c|C)yst(o|e)'),
+                                  'yes', 'no'),
+              cystectomy = factor(cystectomy, c('no', 'yes')),
[…]

### 2024-12-05 import scripts/combat.R

@@ -12,7 +12,9 @@
 
   ## log2-transformed, restricted to the common genes
 
-  combat$raw_expression <- list(tcga = tcga, imvigor = imvigor) %>%
+  combat$raw_expression <- list(tcga = tcga,
+                                imvigor = imvigor,
+                                bcan = bcan) %>%
     map(~.x$expression)
 
   combat$genes <- combat$raw_expression %>%
@@ -29,15 +31,18 @@
 
   insert_msg('Batch effect adjustment')
 
-  combat$expression <- pre_process(train = combat$raw_expression$tcga,
-                                   test = combat$raw_expression$imvigor)
+  combat$expression <-
+    multi_process(train = combat$raw_expression$tcga,
+                  test = combat$raw_expression[c("imvigor", "bcan")])
 
-  combat$expression <- combat$expression[c("train", "test")] %>%
+  combat$expression <-
+    list(tcga = combat$expression$train,
+         imvigor = combat$expression$test$imvigor,
+         bcan = combat$expression$test$bcan) %>%
     map(t) %>%
     map(as.data.frame) %>%
     map(rownames_to_column, 'sample_id') %>%
-    map(as_tibble) %>%
-    set_names(names(combat$raw_expression))
+    map(as_tibble)
 
 # Cachinig the results -------
 

### 2024-12-05 import scripts/imvigor.R

@@ -86,7 +86,8 @@
 
   ## expression annotation
   ## manual substitution of
-  ## the gene '102800317' according to the recent NCBI Gene
+  ## the gene '102800317' and the gene '3730'
+  ## according to the recent NCBI Gene
 
   imvigor$exp_annotation <-
     cds@featureData@data[c('entrez_id',
@@ -97,6 +98,9 @@
 
   imvigor$exp_annotation[imvigor$exp_annotation$entrez_id == '102800317', 'gene_symbol'] <-
     'TPTEP2-CSNK1E'
+
+  imvigor$exp_annotation[imvigor$exp_annotation$entrez_id == '3730', 'gene_symbol'] <-
+    'ANOS1'
 
   ## re-naming the expression data set,
   ## removing features without gene symbol assignment

### 2024-12-05 import scripts/subtypes.R

@@ -21,13 +21,18 @@
 
   subtypes$assignment <- subtypes$data %>%
     map(getConsensusClass,
+        minCor = 0.5,
         gene_id = 'hgnc_symbol') %>%
     map(rownames_to_column, 'sample_id') %>%
     map(mutate,
+        consensusClass = ifelse(is.na(consensusClass),
+                                'not assigned', consensusClass),
+        consensusClass = ifelse(separationLevel < 0.2,
+                                'not assigned', consensusClass),
         consensusClass = factor(consensusClass ,
                                 c('LumP', 'LumU', 'LumNS',
                                   'Stroma-rich', 'Ba/Sq',
-                                  'NE-like'))) %>%
+                                  'NE-like', 'not assigned'))) %>%
     map(as_tibble)
 
 # END -------

### 2024-12-05 import scripts/tcga.R

@@ -12,13 +12,17 @@
 
   tcga$expression <- read_tsv('./data/TCGA PanCancer/data_mrna_seq_v2_rsem.txt')
 
-  ## annotation
+  ## annotation, manual correction for the 3730 gene (Entrez ID)
 
   tcga$annotation <- tcga$expression %>%
     transmute(probe_id = paste0('probe_', 1:nrow(.)),
               gene_symbol = Hugo_Symbol,
               entrez_id = Entrez_Gene_Id) %>%
     filter(complete.cases(.))
+
+  tcga$annotation <- tcga$annotation %>%
+    mutate(gene_symbol = ifelse(entrez_id == '3730',
+                                'ANOS1', gene_symbol))
 
   ## integration of expression by arithmetic mean
 

### 2024-12-05 import.R

@@ -27,6 +27,7 @@
 
   library(htGLMNET)
   library(consensusMIBC)
+  library(impute)
 
   library(soucer)
 
@@ -47,16 +48,19 @@
   list(cache_path = c('./data/genie.RData',
                       './data/msk.RData',
                       './data/tcga.RData',
-                      './data/imvigor.RData'),
+                      './data/imvigor.RData',
+                      './data/bcan.RData'),
        script_path = c('./import scripts/genie.R',
                        './import scripts/msk.R',
                        './import scripts/tcga.R',
-                       './import scripts/imvigor.R'),
+                       './import scripts/imvigor.R',
+                       './import scripts/bcan.R'),
        message = paste('Loading cached',
                        c('GENIE BLCA study',
                          'MSK 2022 study',
                          'TCGA BLCA study',
-                         'IMvigor study'))) %>%
+                         'IMvigor study',
+                         'BCAN study'))) %>%
     pwalk(access_cache)
 
 # ComBat expression adjustment and assignment to the molecular subtypes -------
@@ -65,7 +69,7 @@
 
   access_cache(cache_path = './data/combat.RData',
                script_path = './import scripts/combat.R',
-               message = 'Loading cached ComBat adjutment results')
+               message = 'Loading cached ComBat adjustment results')
 
   c('./import scripts/subtypes.R') %>%
     source_all(message = TRUE, crash = TRUE)

### 2024-12-05 latent class scripts/amplifications.R

@@ -1,0 +1,142 @@
+# Enrichment and differential alteration analysis for gene amplifications in
+# Genetic Clusters.
+#
+# The analysis is done in the GENIE, MSK, and TCGA cohort by weighted
+# permutation testing with perTest/perich. The background, i.e. sample weights
+# for the permutations are obtained by counting all amplified genes per sample.
+#
+# Significant enrichment in a cluster is assumed for enrichment score >= 1.2
+# and FDR-corrected p value < 0.05.
+
+  insert_head()
+
+# container -------
+
+  lca_amp <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ## analysis data and background matrix with counts of all amplified genes
+
+  lca_amp$data <- list(genie = genie,
+                       msk = msk,
+                       tcga = tcga) %>%
+    map(~.x$amplification) %>%
+    map(column_to_rownames, 'sample_id')
+
+  lca_amp$backgrounds <- lca_amp$data
+
+  ## cluster assignment vectors
+
+  lca_amp$f_vectors <- lca_pred$assignment %>%
+    map(~set_names(.x$clust_id, .x$sample_id))
+
+  lca_amp$f_vectors <-
+    map2(lca_amp$f_vectors,
+         lca_amp$data,
+         ~.x[rownames(.y)])
+
+# Frequencies and amplifications of interest present in at least 1% samples -------
+
+  insert_msg('Frequencies of amplifications, amplifications of interest')
+
+  ## overall amplification frequencies,
+  ## mutations of interest shared by all data sets
+
+  lca_amp$frequencies <- lca_amp$data %>%
+    map(count_binary)
+
+  lca_amp$variables <- lca_amp$frequencies %>%
+    map(filter, percent >= 1) %>%
+    map(~.x$variable) %>%
+    reduce(intersect)
+
+  lca_amp$data <- lca_amp$data %>%
+    map(~.x[lca_amp$variables])
+
+  ## frequencies of mutations in genetic clusters
+
+  lca_amp$stats <-
+    map2(lca_amp$data,
+         lca_amp$f_vectors,
+         ~mutate(.x, clust_id = .y)) %>%
+    map(count_binary,
+        variables = lca_amp$variables,
+        split_fct = 'clust_id')
+
+# Permutation tests -------
+
+  insert_msg('Permutation tests')
+
+  plan('multisession')
+
+  lca_amp$test <-
[…]

### 2024-12-05 latent class scripts/clinic.R

@@ -45,6 +45,19 @@
   lca_clinic$variables$tcga <-
     lca_clinic$variables$tcga[lca_clinic$variables$tcga != 'tissue']
 
+# N numbers --------
+
+  insert_msg('N numbers of samples')
+
+  lca_clinic$n_numbers <- lca_clinic$data %>%
+    map(count, clust_id) %>%
+    map(column_to_rownames, 'clust_id') %>%
+    map(t) %>%
+    map(as_tibble) %>%
+    map(mutate,
+        variable = 'Samples, N') %>%
+    map(relocate, variable)
+
 # Descriptive stats -------
 
   insert_msg('Analysis stats')
@@ -125,6 +138,8 @@
          merge_stat_test) %>%
     map(format_res_tbl,
         dict = lca_clinic$lexicon) %>%
+    map2(., lca_clinic$n_numbers,
+         ~full_rbind(.y, .x)) %>%
     compress(names_to = 'cohort') %>%
     mutate(cohort = globals$cohort_labs[cohort]) %>%
     relocate(cohort) %>%

### 2024-12-05 latent class scripts/co_occurence_networks.R

@@ -1,0 +1,100 @@
+# Co-occurrence of the top most frequent somatic mutations, amplifications
+# and deletions investigated by a network analysis.
+#
+# The graphs are created with Jaccard similarities as affinity matrices;
+# cutoff of J >= 0.2.
+
+  insert_head()
+
+# container -------
+
+  lca_ocnet <- list()
+
+# analysis data -----
+
+  insert_msg('Analysis globals')
+
+  ## variables and lexicon
+
+  lca_ocnet$variables <- lca_globals$variables
+
+  lca_ocnet$lexicon <- lca_globals$lexicon
+
+  lca_ocnet$data <- lca_globals$data %>%
+    map(column_to_rownames, 'sample_id')
+
+# Jaccard similarity and graph objects --------
+
+  insert_msg('Similarity matrices and graph objects')
+
+  lca_ocnet$simil_mtx <- lca_ocnet$data %>%
+    map(t) %>%
+    map(calculate_dist, method = 'jaccard') %>%
+    map(~1 - .x)
+
+  lca_ocnet$graph_obj <- lca_ocnet$simil_mtx %>%
+    map(as_iGraph,
+        input_type = 'similarity',
+        weighted = TRUE,
+        diag = FALSE,
+        cutoff = 0.2) %>%
+    map(prune_degree, cutoff = 0)
+
+# Vertex importance stats -----
+
+  insert_msg('Vertex importance stats')
+
+  lca_ocnet$stats <- lca_ocnet$graph_obj %>%
+    map(summary) %>%
+    map(select, -index)
+
+# Setting the graph attributes: vertex stats and protein type -------
+
+  insert_msg('Graph attributes')
+
+  lca_ocnet$graph_obj <-
+    map2(lca_ocnet$graph_obj,
+         lca_ocnet$stats,
+         set_vertex_attributes) %>%
+    map(set_vertex_attributes,
+        lca_ocnet$lexicon)
+
+# Network plots --------
+
+  insert_msg('Network plots')
+
+  lca_ocnet$plots <-
+    list(x = lca_ocnet$graph_obj,
+         plot_title = globals$cohort_labs[names(lca_ocnet$graph_obj)]) %>%
+    pmap(plot,
+         layout = layout.fruchterman.reingold,
+         weighting_order = 1,
+         label_vertices = TRUE,
+         vertex_label_variable = 'gene_symbol',
+         vertex_fill_variable = 'alteration',
+         vertex_color_variable = NULL,
[…]

### 2024-12-05 latent class scripts/co_occurrence.R

@@ -2,7 +2,7 @@
 # and deletions.
 #
 # Co-occurrence is investigated by two-dimensional MDS of the Jaccard distance
-# matrices. The analysis is done for the GENIE and TCGA cohort.
+# matrices. The analysis is done for the GENIE, MSK and TCGA cohort.
 
   insert_head()
 
@@ -112,6 +112,33 @@
                                title_suffix = globals$cohort_labs[i]))
 
   }
+
+# Oncoplots --------
+
+  insert_msg('Oncoplots')
+
+  lca_occur$oncoplots <-
+    list(data = lca_occur$data,
+         plot_title = globals$cohort_labs[names(lca_occur$data)]) %>%
+    pmap(plot_bionco,
+         split_fct = NULL,
+         one_plot = FALSE,
+         hide_x_axis_text = TRUE,
+         cust_theme = globals$common_theme +
+           theme(axis.title.y = element_blank(),
+                 axis.text.x = element_blank(),
+                 axis.ticks.x = element_blank(),
+                 axis.line = element_blank(),
+                 panel.grid.major.x = element_blank(),
+                 strip.text.x = element_text(hjust = 0, angle = 90)),
+         color_scale = c('gray85', 'orangered3'),
+         x_lab = 'cancer sample') %>%
+    map(~.x$main +
+          theme(axis.text.y = element_markdown()) +
+          scale_y_discrete(labels = function(x) exchange(x,
+                                                         lca_occur$lexicon,
+                                                         value = 'label_html')))
+
 
 # caching the analysis results -------
 

### 2024-12-05 latent class scripts/deletions.R

@@ -1,0 +1,142 @@
+# Enrichment and differential alteration analysis for deletions in
+# Genetic Clusters.
+#
+# The analysis is done in the GENIE, MSK, and TCGA cohort by weighted
+# permutation testing with perTest/perich. The background, i.e. sample weights
+# for the permutations are obtained by counting all deleted genes per sample.
+#
+# Significant enrichment in a cluster is assumed for enrichment score >= 1.2
+# and FDR-corrected p value < 0.05.
+
+  insert_head()
+
+# container -------
+
+  lca_del <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ## analysis data and background vectors with counts of all mutated genes
+
+  lca_del$data <- list(genie = genie,
+                       msk = msk,
+                       tcga = tcga) %>%
+    map(~.x$deletion) %>%
+    map(column_to_rownames, 'sample_id')
+
+  lca_del$backgrounds <- lca_del$data
+
+  ## cluster assignment vectors
+
+  lca_del$f_vectors <- lca_pred$assignment %>%
+    map(~set_names(.x$clust_id, .x$sample_id))
+
+  lca_del$f_vectors <-
+    map2(lca_del$f_vectors,
+         lca_del$data,
+         ~.x[rownames(.y)])
+
+# Frequencies and deletions of interest present in at least 1% samples -------
+
+  insert_msg('Frequencies of mutations, mutations of interest')
+
+  ## overall deletion frequencies,
+  ## deletions of interest shared by all data sets
+
+  lca_del$frequencies <- lca_del$data %>%
+    map(count_binary)
+
+  lca_del$variables <- lca_del$frequencies %>%
+    map(filter, percent >= 1) %>%
+    map(~.x$variable) %>%
+    reduce(intersect)
+
+  lca_del$data <- lca_del$data %>%
+    map(~.x[lca_del$variables])
+
+  ## frequencies of deletions in genetic clusters
+
+  lca_del$stats <-
+    map2(lca_del$data,
+         lca_del$f_vectors,
+         ~mutate(.x, clust_id = .y)) %>%
+    map(count_binary,
+        variables = lca_del$variables,
+        split_fct = 'clust_id')
+
+# Permutation tests -------
+
+  insert_msg('Permutation tests')
+
+  plan('multisession')
+
+  lca_del$test <-
+    list(x = lca_del$data,
+         f = lca_del$f_vectors,
[…]

### 2024-12-05 latent class scripts/development.R

@@ -24,7 +24,7 @@
 
 # tuning of the poLCA models ---------
 
-  insert_msg('Tuniniig of the poLCA models')
+  insert_msg('Tuninig of the poLCA models')
 
   lca_dev$tune_object <- tune_lca(formula = lca_globals$formula,
                                   data = lca_dev$data,

### 2024-12-05 latent class scripts/dge.R

@@ -1,5 +1,8 @@
 # Differential expression of the FGF and FGFR genes in the the genetic clusters.
-# The analysis is done only for the TCGA cohort
+# The analysis is done only for the TCGA cohort.
+#
+# Statistical significance is determined by one-way ANOVA with eta-square effect
+# size metrics. FDR correction.
 
   insert_head()
 
@@ -18,7 +21,8 @@
 
   ## analysis variables
 
-  lca_dge$variables <- globals[c("receptors", "ligands")] %>%
+  lca_dge$variables <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
     reduce(union)
 
   lca_dge$variables <-
@@ -32,36 +36,33 @@
   insert_msg('Descriptive stats')
 
   lca_dge$stats <- lca_dge$data %>%
-    explore(variables = lca_dge$variables,
-            split_factor = 'clust_id',
-            what = 'table',
-            pub_styled = TRUE) %>%
-    format_desc
-
+    fast_num_stats(split_fct = 'clust_id',
+                   variables = lca_dge$variables)
 # One-way ANOVA ------
 
   insert_msg('One-way ANOVA')
 
-  lca_dge$test <- lca_dge$data %>%
-    test_anova(split_fct = 'clust_id',
-               variables = lca_dge$variables,
-               adj_method = 'BH') %>%
-    .$anova %>%
-    re_adjust(method = 'none') %>%
-    mutate(eff_size = paste('\u03B7\u00B2 =', signif(effect_size, 2)),
+  lca_dge$test <-
+    f_one_anova(lca_dge$data[lca_dge$variables],
+                f = lca_dge$data$clust_id,
+                safely = TRUE,
+                as_data_frame = TRUE,
+                adj_method = 'BH') %>%
+    re_adjust %>%
+    mutate(eff_size = paste('\u03B7\u00B2 =', signif(etasq, 2)),
            plot_cap = paste(eff_size, significance, sep = ', '),
-           variable = response,
-           ax_lab = paste(html_italic(response),
+           ax_lab = paste(html_italic(variable),
                           plot_cap,
                           sep = '<br>'),
            ax_lab = ifelse(p_adjusted < 0.05,
-                           html_bold(ax_lab), ax_lab))
+                           html_bold(ax_lab), ax_lab)) %>%
+    as_tibble
 
   ## significant effects
[…]

### 2024-12-05 latent class scripts/evaluation.R

@@ -1,4 +1,9 @@
-# Quality check for the genetic clusters
+# Quality check for the genetic clusters.
+#
+# Distribution of the class-defining features in the genetic clusters are
+# compared by chi-square test with Cramer's V effect size statistic.
+# FDR correction. The proper check of genetic background is done by
+# permutation testing in another script
 
   insert_head()
 
@@ -33,7 +38,8 @@
   lca_eval$n_total_labs <- lca_eval$data %>%
     map_dbl(nrow) %>%
     map2_chr(names(.), .,
-             ~paste(globals$cohort_labs[.x], .y, sep = '\nn = '))
+             ~paste(globals$cohort_labs[.x], .y, sep = '\nn = ')) %>%
+    set_names(names(lca_eval$data))
 
   lca_eval$n_numbers <- lca_eval$clust_objects %>%
     map(ngroups) %>%
@@ -55,7 +61,7 @@
 
   lca_eval$n_plot <- lca_eval$n_numbers %>%
     compress(names_to = 'cohort') %>%
-    mutate(cohort = factor(cohort, globals$analysis_cohorts),
+    mutate(cohort = factor(cohort, rev(globals$analysis_cohorts)),
            cohort = droplevels(cohort)) %>%
     ggplot(aes(x = percent,
                y = cohort,
@@ -80,6 +86,7 @@
 
   ## silhouette widths do not make much sense
   ## in the evidently non-Euclidean setting
+  ## neighborhood preservation is the most important metric
 
   lca_eval$clust_stats <- lca_eval$clust_objects %>%
     map(summary) %>%
@@ -105,21 +112,20 @@
 
   insert_msg('Chi-square tests')
 
-  plan('multisession')
-
   lca_eval$test <- lca_eval$data %>%
-    future_map(compare_variables,
-               variables = lca_eval$variables,
-               split_factor = 'clust_id',
-               what = 'eff_size',
-               types = 'cramer_v',
-               exact = FALSE,
-               ci = FALSE,
-               adj_method = 'BH',
-               pub_styled = TRUE,
-               .parallel = TRUE,
-               .options = furrr_options(seed = TRUE)) %>%
+    map(~f_chisq_test(.x[lca_eval$variables],
+                      f = .x[['clust_id']],
+                      as_data_frame = TRUE,
+                      adj_method = 'BH',
+                      safely = TRUE)) %>%
+    map(as_tibble)
+
[…]

### 2024-12-05 latent class scripts/genetic_plots.R

@@ -1,0 +1,320 @@
+# Plots for results of the genetic alteration enrichment analyses
+# in Genetic Clusters. We're concentrating on genes found to be significantly
+# enriched in the corresponding clusters of all three cohorts, i.e. GENIE, MSK,
+# and TCGA BLCA.
+
+  insert_head()
+
+# container -------
+
+  lca_genplots <- list()
+
+# Analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## cluster and cohort levels
+
+  lca_genplots$clust_levels <- levels(lca_pred$assignment[[1]]$clust_id)
+
+  lca_genplots$cohort_levels <- names(lca_pred$assignment)
+
+  lca_genplots$alteration_levels <- c('mutation', 'deletion', 'amplification')
+
+  ## colors for alteration types
+
+  lca_genplots$alteration_colors <-
+    c(mutation = globals$mut_status_color[["mutated"]],
+      deletion = globals$del_status_color[["deleted"]],
+      amplification = globals$amp_status_color[["amplified"]])
+
+  ## numbers of significant gene mutations, deletions and amplification
+
+  lca_genplots$n_genes <- list(mutation = lca_mut,
+                               deletion = lca_del,
+                               amplification = lca_amp) %>%
+    map(~.x$enriched) %>%
+    map(map, map_dbl, length) %>%
+    map(map,
+        compress,
+        names_to = 'cohort',
+        values_to = 'n') %>%
+    map(compress,
+        names_to = 'clust_id') %>%
+    compress(names_to = 'alteration') %>%
+    mutate(clust_id = factor(clust_id, lca_genplots$clust_levels),
+           cohort = factor(cohort, lca_genplots$cohort_levels),
+           alteraton = factor(alteration, lca_genplots$alteration_levels))
+
+  ## significant genetic alterations shared by all three cohorts
+
+  lca_genplots$common_enriched <- list(mutation = lca_mut,
+                                       deletion = lca_del,
+                                       amplification = lca_amp) %>%
+    map(~.x$enriched) %>%
+    map(map, reduce, intersect) %>%
+    map(reduce, union)
+
+  ## binary data for the common significant alterations
+
+  lca_genplots$data <- list(genie = genie,
+                            msk = msk,
+                            tcga = tcga) %>%
+    map(~.x[c('mutation', 'deletion', 'amplification')])
+
+  for(i in names(lca_genplots$data)) {
+
[…]

### 2024-12-05 latent class scripts/mutations.R

@@ -1,0 +1,142 @@
+# Enrichment and differential alteration analysis for somatic mutations in
+# Genetic Clusters.
+#
+# The analysis is done in the GENIE, MSK, and TCGA cohort by weighted
+# permutation testing with perTest/perich. The background, i.e. sample weights
+# for the permutations are obtained by counting all mutated genes per sample.
+#
+# Significant enrichment in a cluster is assumed for enrichment score >= 1.2
+# and FDR-corrected p value < 0.05.
+
+  insert_head()
+
+# container -------
+
+  lca_mut <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ## analysis data and background vectors with counts of all mutated genes
+
+  lca_mut$data <- list(genie = genie,
+                       msk = msk,
+                       tcga = tcga) %>%
+    map(~.x$mutation) %>%
+    map(column_to_rownames, 'sample_id')
+
+  lca_mut$backgrounds <- lca_mut$data
+
+  ## cluster assignment vectors
+
+  lca_mut$f_vectors <- lca_pred$assignment %>%
+    map(~set_names(.x$clust_id, .x$sample_id))
+
+  lca_mut$f_vectors <-
+    map2(lca_mut$f_vectors,
+         lca_mut$data,
+         ~.x[rownames(.y)])
+
+# Frequencies and mutations of interest present in at least 1% samples -------
+
+  insert_msg('Frequencies of mutations, mutations of interest')
+
+  ## overall mutation frequencies,
+  ## mutations of interest shared by all data sets
+
+  lca_mut$frequencies <- lca_mut$data %>%
+    map(count_binary)
+
+  lca_mut$variables <- lca_mut$frequencies %>%
+    map(filter, percent >= 1) %>%
+    map(~.x$variable) %>%
+    reduce(intersect)
+
+  lca_mut$data <- lca_mut$data %>%
+    map(~.x[lca_mut$variables])
+
+  ## frequencies of mutations in genetic clusters
+
+  lca_mut$stats <-
+    map2(lca_mut$data,
+         lca_mut$f_vectors,
+         ~mutate(.x, clust_id = .y)) %>%
+    map(count_binary,
+        variables = lca_mut$variables,
+        split_fct = 'clust_id')
+
+# Permutation tests -------
+
+  insert_msg('Permutation tests')
+
+  plan('multisession')
+
+  lca_mut$test <-
[…]

### 2024-12-05 latent class scripts/overall_survival_lca_subsets.R

@@ -1,0 +1,95 @@
+# Checking whether a combination of genetic information (Genetic Subsets) and
+# transcriptomic classification (consensus clusters of MIBC) may improve
+# prediction of overall survival.
+#
+# We're comparing three models:
+#
+# 1) a Cox model with the consensus MIBC subset as an independent variable
+#
+# 2) a Cox model with the genetic subset as an independent variable
+#
+# 3) a Cox model with the consensus and genetic subsets as independent variables
+#
+# A demonstrated both in the training data set and in 5-repeats 5-fold
+# cross-validation, there are no differences in performance between the models
+# 1 and 3, and, hence, no benefot from inclusion of the genetic information.
+
+
+  insert_head()
+
+# container -------
+
+  lca_mos <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  lca_mos$data$transcriptome <- subtypes$assignment$tcga %>%
+    select(sample_id, consensusClass)
+
+  lca_mos$data$genome <- lca_pred$assignment$tcga
+
+  lca_mos$data$clinic <- tcga$clinic %>%
+    select(sample_id, os_days, death)
+
+  lca_mos$data <- lca_mos$data %>%
+    reduce(inner_join, by = 'sample_id') %>%
+    filter(complete.cases(.))
+
+# Cox models and their cross-validation --------
+
+  insert_msg('Cox models and CV')
+
+  lca_mos$models <-
+    list(transcriptome = Surv(os_days, death) ~ consensusClass,
+         genome = Surv(os_days, death) ~ clust_id,
+         combi = Surv(os_days, death) ~ consensusClass + clust_id) %>%
+    map(~call2('coxph',
+               formula = .x,
+               data = lca_mos$data,
+               x = TRUE,
+               y = TRUE)) %>%
+    map(eval) %>%
+    map(as_coxex, lca_mos$data)
+
+  lca_mos$cv_models <- lca_mos$models %>%
+    map(cvCox, n_folds = 5, n_repeats = 5)
+
+# Assumptions, fit stats, and inference, training data ------
+
+  insert_msg('Assumptions, fit, and inference')
+
+  ## assumptions: more or less hold by all models
+
+  lca_mos$assumptions <- lca_mos$models %>%
+    map(summary, 'assumptions')
+
+  ## fit stats
+
[…]

### 2024-12-05 latent class scripts/prediction.R

@@ -32,12 +32,12 @@
   lca_pred$model <- lca_dev$tune_object$models$class_6
 
   lca_pred$model$class_names <-
-    c('mutRB1',
-      'hyperMut',
-      'mutFGFR3',
+    c('del9p21',
       'oligoMut',
       'ampMDM2',
-      'del9p21')
+      'mutRB1',
+      'mutFGFR3',
+      'hyperMut')
 
   ## predictions
 
@@ -46,8 +46,8 @@
 
   lca_pred$posteriors[c('msk', 'tcga')] <-
     lca_pred$factor_data[c("msk", "tcga")] %>%
-    map(predict,
-        object = lca_pred$model)
+    map(~predict.polcax(object = lca_pred$model,
+                        newdata = .x))
 
 # Diagnostic plots: class-assignment posterior p -------
 
@@ -97,11 +97,17 @@
         sample_id = observation,
         clust_id = clust_id)
 
-# END -----
+# caching --------
+
+  insert_msg('Caching')
 
   #lca_pred$data <- NULL
   lca_pred$factor_data <- NULL
 
   lca_pred <- compact(lca_pred)
+
+  save(lca_pred, file = './cache/lca_pred.RData')
+
+# END -----
 
   insert_tail()

### 2024-12-05 latent class scripts/subtypes.R

@@ -8,6 +8,12 @@
 
   lca_sub <- list()
 
+# parallel backend ------
+
+  insert_msg('Parallel backend')
+
+  plan('multisession')
+
 # analysis globals: data and variables -------
 
   insert_msg('Analysis variables')
@@ -19,7 +25,22 @@
   lca_sub$data <-
     inner_join(subtypes$assignment$tcga[c('sample_id', 'consensusClass')],
                lca_globals$data$tcga,
-               by = 'sample_id')
+               by = 'sample_id') %>%
+    filter(!is.na(consensusClass))
+
+  ## background matrix with all alterations
+
+  lca_sub$bcg_mtx <-
+    tcga[c("mutation", "amplification", "deletion")] %>%
+    map(column_to_rownames, 'sample_id') %>%
+    map2(., c('_mutation', '_amplification', '_deletion'),
+         ~set_names(.x,
+                    paste0(names(.x), .y)))
+
+  lca_sub$bcg_mtx <-  lca_sub$bcg_mtx %>%
+    reduce(cbind)
+
+  lca_sub$bcg_mtx <- lca_sub$bcg_mtx[rownames(lca_sub$data), ]
 
 # N numbers -------
 
@@ -50,44 +71,66 @@
 
   insert_msg('Testing for differences between the consenus classes')
 
+  ## enrichment testing with perich
+
   lca_sub$test <-
-    compare_variables(lca_sub$data,
-                      variables = lca_sub$variables,
-                      split_factor = 'consensusClass',
-                      what = 'eff_size',
-                      types = 'cramer_v',
-                      exact = FALSE,
-                      ci = FALSE,
-                      pub_styled = TRUE,
-                      adj_method = 'BH',
-                      .parallel = TRUE,
-                      .paropts = furrr_options(seed = TRUE,
-                                               globals = 'lca_sub')) %>%
-    mutate(plot_cap = paste(eff_size, significance, sep = ', '),
-           gene_symbol = stri_split_fixed(variable,
+    perTest(lca_sub$data[lca_sub$variables],
+            f = lca_sub$data$consensusClass,
+            background = lca_sub$bcg_mtx,
+            alternative = 'greater',
+            n_iter = 10000,
+            as_data_frame = TRUE,
+            compress = TRUE,
+            adj_method = 'BH',
+            .parallel = TRUE)
+
+  ## formatting the testing results
+
+  lca_sub$test <- lca_sub$test %>%
[…]

### 2024-12-05 latent class scripts/tmb.R

@@ -1,4 +1,7 @@
-# Differences in total mutation burdens between the genetic clusters
+# Differences in total mutation burdens between the genetic clusters.
+#
+# Statistical significance is determined by FDR-corrected Kruskal-Wallis test
+# with eta-square effect size statistic.
 
   insert_head()
 
@@ -74,12 +77,8 @@
   lca_tmb$stats <-
     map2(lca_tmb$data,
          lca_tmb$variables,
-         ~explore(.x,
-                  variables = .y,
-                  split_factor = 'clust_id',
-                  what = 'table',
-                  pub_styled = TRUE)) %>%
-    map(format_desc)
+         fast_num_stats,
+         split_fct = 'clust_id')
 
 # Testing for differences between the genetic strata -------
 
@@ -88,17 +87,17 @@
   lca_tmb$test <-
     map2(lca_tmb$data,
          lca_tmb$variables,
-         ~compare_variables(.x,
-                            variables = .y,
-                            split_factor = 'clust_id',
-                            what = 'eff_size',
-                            types = 'kruskal_etasq',
-                            ci = FALSE,
-                            exact = FALSE,
-                            pub_styled = TRUE,
-                            adj_method = 'BH')) %>%
+         ~f_kruskal_test(.x[.y],
+                         f = .x[['clust_id']],
+                         exact = FALSE,
+                         safely = TRUE,
+                         as_data_frame = TRUE,
+                         adj_method = 'BH')) %>%
+    map(re_adjust) %>%
     map(mutate,
-        plot_cap = paste(eff_size, significance, sep = ', '))
+        eff_size = paste('\u03B7\u00B2 =', signif(etasq, 2)),
+        plot_cap = paste(eff_size, significance, sep = ', ')) %>%
+    map(as_tibble)
 
 # Plotting ------
 

### 2024-12-05 latent_class.R

@@ -20,6 +20,8 @@
   library(caret)
 
   library(exda)
+  library(fastTest)
+  library(perich)
 
   library(survival)
   library(survminer)
@@ -58,14 +60,19 @@
   insert_msg('Co-occurrence and consensus molecular subtypes')
 
   ## general co-occurrence of the most common genetic features
-
-  access_cache(cache_path = './cache/lca_occur.RData',
-               script_path = './latent class scripts/co_occurrence.R',
-               message = 'Loading cached co-occurrence results')
-
+  ## and network analysis
+  ##
   ## presence/absence of genetic alterations in the consensus classes
 
-  c('./latent class scripts/subtypes.R') %>%
+  list(cache_path = c('./cache/lca_occur.RData',
+                      './cache/lca_sub.RData'),
+       script_path = c('./latent class scripts/co_occurrence.R',
+                       './latent class scripts/subtypes.R'),
+       message = c('Loading cached co-occurrence results',
+                   'Cached mutation enrichment for consensus classes')) %>%
+    pwalk(access_cache)
+
+  c('./latent class scripts/co_occurence_networks.R') %>%
     source_all(message = TRUE, crash = TRUE)
 
 # Definition and evaluation of the genetic classes -----
@@ -73,19 +80,20 @@
   insert_msg('Definition and evaluation of the latant classes')
 
   ## development of the clusters by latent class modeling
+  ## predictions
+  ## evaluation by comparing the class-defining variables
+  ## between the classes
 
-  access_cache(cache_path = './cache/lca_dev.RData',
-               script_path = './latent class scripts/development.R',
-               message = 'Loading cached latent class analysis')
-
-  ## prediction of the clusters in the TCAG cohort,
-  ## characteristic of the clusters
+  list(cache_path = c('./cache/lca_dev.RData'),
+       script_path = c('./latent class scripts/development.R'),
+       message = c('Loading cached latent class analysis')) %>%
+    pwalk(access_cache)
 
   c('./latent class scripts/prediction.R',
     './latent class scripts/evaluation.R') %>%
     source_all(message = TRUE, crash = TRUE)
 
-# Clinical, pathological and prognostic characteristic -------
+# Clinical, pathological, prognostic, and transcriptomic characteristic -------
 
   insert_msg('Clinical and pathological characteristic of the clusters')
 
@@ -95,7 +103,8 @@
     './latent class scripts/tmb.R',
     './latent class scripts/overall_survival.R',
[…]

### 2024-12-05 ml scripts/development.R

@@ -1,0 +1,73 @@
+# Development of and Elastic Net and Random Forest models of consensus
+# molecular classes in the TCGA cohort.
+
+  insert_head()
+
+# container --------
+
+  ml_train <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ml_train$x <- ml_globals$train_x
+  ml_train$y <- ml_globals$train_y
+
+  ml_train$cv_folds <- ml_globals$cv_folds
+  ml_train$ranger_grid <- ml_globals$ranger_grid
+
+# tuning of the lambda --------
+
+  insert_msg('Tuning')
+
+  plan('multisession')
+
+  ml_train$tune_obj$elnet <- tune_glmnet(ml_train$x,
+                                         ml_train$y,
+                                         indexes = ml_train$cv_folds,
+                                         family = 'multinomial',
+                                         alpha = 0.5)
+
+  ml_train$tune_obj$ranger <- tune_ranger(ml_train$x,
+                                          ml_train$y,
+                                          tune_grid = ml_train$ranger_grid,
+                                          tune_stat = 'prediction.error',
+                                          num.trees = 500)
+
+  plan('sequential')
+
+# OOF predictions used by the Ensemble model ---------
+
+  insert_msg('OOF predictions')
+
+  ml_train$oof_predictions$elnet <-
+    oof_glmnet(ml_globals$ens_cv_x$train,
+               ml_globals$ens_cv_x$test,
+               ml_globals$ens_cv_y$train,
+               ml_globals$ens_cv_y$test,
+               family = 'multinomial',
+               alpha = 0.5,
+               lambda = ml_train$tune_obj$elnet$best_tune$lambda)
+
+  ml_train$oof_predictions$ranger <-
+    call2(.fn = 'oof_ranger',
+          ml_globals$ens_cv_x$train,
+          ml_globals$ens_cv_x$test,
+          ml_globals$ens_cv_y$train,
+          ml_globals$ens_cv_y$test,
+          num.trees = 500,
+          !!!ml_train$tune_obj$ranger$best_tune) %>%
+    eval
+
+# Caching -------
+
+  insert_msg('Caching')
+
+  ml_train <- ml_train[c("tune_obj", "oof_predictions")]
+
+  save(ml_train, file = './cache/ml_train.RData')
+
+# END ---------
+
+  insert_tail()

### 2024-12-05 ml scripts/ensemble.R

@@ -1,0 +1,72 @@
+# Ensemble model that operates with predictions of the consensus classes
+# made by Elastic Net and Random Forest algorithms
+
+  insert_head()
+
+# container --------
+
+  ml_ens <- list()
+
+# modeling data --------
+
+  insert_msg('Modeling data')
+
+  ## class levels
+
+  ml_ens$class_levels <- ml_globals$class_levels
+
+  ## out-of-fold predictions for the SMOTE data
+
+  ml_ens$data <-
+    list(elnet = ml_train$oof_predictions$elnet %>%
+           select(-obs, -pred),
+         ranger = ml_train$oof_predictions$ranger) %>%
+    reduce(inner_join, by = c('sample_id', 'fold_id'))
+
+  ## modeling matrix and response vector
+
+  ml_ens$x <- ml_ens$data %>%
+    column_to_rownames('sample_id') %>%
+    select(ends_with('elnet'), ends_with('ranger')) %>%
+    as.matrix
+
+  ml_ens$y <- ml_ens$data$obs
+
+  ## CV folds for tuning of the lambda
+
+  ml_ens$n_rep <- 200
+
+  set.seed(1234)
+
+  ml_ens$cv_folds <- 1:ml_ens$n_rep %>%
+    map(function(x) createFolds(ml_ens$data$obs,
+                                list = FALSE,
+                                returnTrain = TRUE)) %>%
+    set_names(paste0('rep_', 1:ml_ens$n_rep))
+
+# Elastic Net ensembles: tuning and training in the TCGA cohort ---------
+
+  insert_msg('Tuning and training')
+
+  plan('multisession')
+
+  ml_ens$tune_obj <-
+    tune_glmnet(ml_ens$x,
+                ml_ens$y,
+                indexes = ml_ens$cv_folds,
+                family = 'multinomial',
+                alpha = 0.5)
+
+  plan('sequential')
+
+# Caching -------
+
+  insert_msg('Caching')
+
+  ml_ens <- ml_ens["tune_obj"]
+
+  save(ml_ens, file = './cache/ml_ens.RData')
+
+# END ---------
+
+  insert_tail()

### 2024-12-05 ml scripts/evaluation.R

@@ -1,0 +1,185 @@
+# Evaluation of performance of the ML models of consensus classes that
+# use FGFR, FGF, and FGFBP genes as explanatory factors.
+
+  insert_head()
+
+# container --------
+
+  ml_eval <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  ## predictions
+
+  ml_eval$predictions <- ml_pred$predictions
+
+  ## algorithm labels
+
+  ml_eval$algo_labs <-
+    c(elnet = 'Elastic Net',
+      ranger = 'Random Forest',
+      ensemble = 'Ensemble')
+
+# Performance stats --------
+
+  insert_msg('Performance stats')
+
+  ## global performance stats: ROC stats, accuracy, kappa, and Brier scores
+
+  ml_eval$stats <- ml_eval$predictions %>%
+    map(map, ml_summary) %>%
+    map(compress, names_to = 'cohort') %>%
+    compress(names_to = 'algorithm') %>%
+    mutate(data_set = ifelse(cohort == 'tcga',
+                             'training', 'test'),
+           data_set = factor(data_set, c('training', 'test')))
+
+  ## dot plot
+
+  ml_eval$stat_plot <- ml_eval$stats %>%
+    ggplot(aes(x = Kappa,
+               y = 2 - brier_score,
+               size = AUC,
+               fill = data_set,
+               color = data_set)) +
+    facet_grid(. ~ algorithm,
+               labeller = as_labeller(ml_eval$algo_labs)) +
+    geom_rect(aes(xmin = -Inf,
+                  ymin = 2 - min(ml_eval$stats$brier_reference),
+                  xmax = Inf,
+                  ymax = 2 - max(ml_eval$stats$brier_reference)),
+              fill = 'gray80',
+              color = NA,
+              alpha = 0.5,
+              show.legend = FALSE) +
+    geom_vline(xintercept = 0,
+               linetype = 'dashed') +
+    geom_point(shape = 21,
+               color = 'black') +
+    geom_text_repel(aes(label = paste(globals$cohort_labs[cohort],
+                                      signif(AUC, 2),
+                                      sep = '\nAUC = ')),
+                    size = 2.3,
+                    hjust = 0,
+                    box.padding = 0.6,
+                    show.legend = FALSE) +
+    scale_fill_manual(values = c(training = 'indianred3',
+                                 test = 'steelblue'),
+                      name = 'data set') +
+    scale_color_manual(values = c(training = 'indianred3',
[…]

### 2024-12-05 ml scripts/globals.R

@@ -1,0 +1,141 @@
+# Data for ML modeling of molecular subtypes with expression of FGF, FGFR,
+# and FGFBP genes. The expression levels are ComBat-processed.
+#
+# The consensus classes to be modeled are LumP, LumU, LumNS, Stroma-rich,
+# and Ba/Sq. Not assigned samples, NE-like and LumNS subset specimens are
+# excluded; the later ones because of their low frequency (n = 2 and n = 14).
+
+  insert_head()
+
+# container -------
+
+  ml_globals <- list()
+
+# analysis data ------
+
+  insert_msg('Analysis globals')
+
+  ## variables of interest and their lexicon
+
+  ml_globals$lexicon <-
+    globals[c("ligands", "receptors", "binding_proteins")] %>%
+    map(~tibble(variable = .x)) %>%
+    compress(names_to = 'protein_type')
+
+  ml_globals$data <- combat$expression[c("tcga", "imvigor", "bcan")] %>%
+    map(select, sample_id, any_of(ml_globals$lexicon$variable))
+
+  ml_globals$variables <- ml_globals$data %>%
+    map(select, -sample_id) %>%
+    map(names) %>%
+    reduce(intersect)
+
+  ml_globals$lexicon <- ml_globals$lexicon %>%
+    filter(variable %in% ml_globals$variable)
+
+  ## consensus class information
+
+  ml_globals$assignment <- subtypes$assignment %>%
+    map(~.x[c('sample_id', 'consensusClass')]) %>%
+    map(~filter(.x,
+                complete.cases(.x),
+                !consensusClass %in% c('not assigned', 'NE-like', 'LumNS'))) %>%
+    map(mutate,
+        consensusClass = droplevels(consensusClass))
+
+  ml_globals$data <-
+    map2(ml_globals$data,
+         ml_globals$assignment,
+         inner_join, by = 'sample_id') %>%
+    map(~filter(.x, complete.cases(.x)))
+
+  ml_globals$class_levels <- levels(ml_globals$data[[1]]$consensusClass)
+
+# Response vectors and model matrices ------
+
+  insert_msg('Survival objects and x matrices')
+
+  ml_globals$y <- ml_globals$data %>%
+    map(~.x$consensusClass)
+
+  ml_globals$x <- ml_globals$data %>%
+    map(column_to_rownames, 'sample_id') %>%
+    map(select, all_of(ml_globals$variables)) %>%
[…]

### 2024-12-05 ml scripts/importance.R

@@ -1,0 +1,109 @@
+# Importance of the explanatory variables in the Elastic Net and Random forest
+# models of consensus classes.
+# The models use solely FGF/FGFR/FGFBP expression as explanatory factors.
+#
+# The importance measures are simply abs(beta) of the model coefficients
+# for the Elastic Net model.
+# For the Random Forest model, we're calculating the perumtation importance.
+
+  insert_head()
+
+# container ------
+
+  ml_imp <- list()
+
+# Ranger model with permutation importance -------
+
+  insert_msg('Random forest model with permutation importance')
+
+  ## model
+
+  ml_imp$ranger_model <- ml_globals$train_data %>%
+    select(-consensusClass) %>%
+    as.matrix %>%
+    call2(.fn = 'ranger',
+          x = .,
+          y = ml_globals$train_data$consensusClass,
+          !!!ml_train$tune_obj$ranger$best_tune,
+          probability = TRUE,
+          importance = 'permutation') %>%
+    eval
+
+  ## permutation importance
+
+  ml_imp$ranger_importance <- ml_imp$ranger_model %>%
+    importance %>%
+    compress(names_to = 'variable',
+             values_to = 'importance')
+
+# Random forest importance in a bar plot --------
+
+  insert_msg('Permutation importance plot, Random Forest')
+
+  ml_imp$ranger_plot <- ml_imp$ranger_importance %>%
+    slice_max(importance, n = 15) %>%
+    ggplot(aes(x = importance,
+               y = reorder(variable, importance))) +
+    geom_bar(stat = 'identity',
+             color = 'white',
+             fill = 'aquamarine4') +
+    globals$common_theme +
+    theme(axis.title.y = element_blank(),
+          axis.text.y = element_text(face = 'italic')) +
+    labs(title = 'Random Forest',
+         x = 'permutation importance, \u0394 error')
+
+# Elastic Net model coefficients --------
+
+  insert_msg('Elastic Net model coefficients')
+
+  ml_imp$coefs <- ml_train$tune_obj$elnet$model %>%
+    coef %>%
+    map(as.matrix) %>%
+    map(as.data.frame) %>%
+    map(set_names, 'beta') %>%
+    map(rownames_to_column, 'variable') %>%
+    map(filter,
+        !variable %in% c('', '(Intercept)')) %>%
+    map(mutate, abs_beta = abs(beta)) %>%
+    map(as_tibble)
+
+# Bar plots of the Elastic Net beta coefficients -------
+
[…]

### 2024-12-05 ml scripts/predictions.R

@@ -1,0 +1,165 @@
+# Predictions of consensus class assignment by the Elastic Net, Random
+# Forest, and Ensemble models trained with FGF/FGFR/FGFBP expression data
+# in the TCGA BLCA cohort.
+
+  insert_head()
+
+# container -------
+
+  ml_pred <- list()
+
+# analysis globals -------
+
+  insert_msg('Analysis globals')
+
+  ## modeling matrices and the actual cluster assignment
+
+  ml_pred$x <- ml_globals$x
+  ml_pred$y <- ml_globals$y
+
+  ## the optimized Elastic Net and Random Forest models
+
+  ml_pred$models <- ml_train$tune_obj %>%
+    map(~.x$model)
+
+  ml_pred$models$ensemble <- ml_ens$tune_obj$model
+
+  ## class levels
+
+  ml_pred$class_levels <- ml_globals$class_levels
+
+# Elastic Net predictions ---------
+
+  insert_msg('Predictions')
+
+  ## predicted class assignment
+
+  ml_pred$predictions$elnet$class <- ml_pred$x %>%
+    map(predict,
+        object = ml_pred$models$elnet,
+        type = 'class') %>%
+    map(~.x[, 1]) %>%
+    map(factor, ml_globals$class_levels)
+
+  ## class assignment probabilities
+
+  ml_pred$predictions$elnet$p_values <-  ml_pred$x %>%
+    map(predict,
+        object = ml_pred$models$elnet,
+        type = 'response') %>%
+    map(~.x[, , 1]) %>%
+    map(as.data.frame) %>%
+    map(rownames_to_column, 'sample_id')
+
+  ## the complete prediction tables
+
+  ml_pred$predictions$elnet <-
+    list(x = ml_pred$predictions$elnet$p_values,
+         y = ml_pred$predictions$elnet$class,
+         z = ml_pred$y) %>%
+    pmap(function(x, y, z) x %>%
+           mutate(pred = y,
+                  obs = z))
+
+# Random Forest predictions ---------
+
+  insert_msg('Random Forest predictions')
+
+  # class assignment p values
+
+  ml_pred$predictions$ranger$p_values <- ml_pred$x %>%
+    map(predict,
+        object = ml_pred$models$ranger) %>%
+    map(~.x[['predictions']]) %>%
+    map(as.data.frame)
+
+  ## class assignment by nodal voting
+
[…]

### 2024-12-05 ml.R

@@ -1,0 +1,73 @@
+# Investigating if and how the consensus molecular subtypes, at least the major
+# ones (LumP, LumU, Stroma-rich, and basal/squamous) can be emulated by
+# expression of FGF, FGFR, and FGFBP genes.
+#
+# To this end, we're developing a multinomial Elastic Net model of the consensus
+# classes in the TCGA BLCA cohort; ComBat log2-transformed expression of the
+# genes of interest serve as explanatory factors. Because of an imbalance of
+# consensus class sizes, we're amplifying the minority classes
+# (LumU and Stroma-rich) by SMOTE. The model is tuned (optimal
+# lambda) by minimizing deviance in repeated cross-validation. Performance of
+# the model is evaluated in the TCGA, IMvigor, and BCAN cohorts, where
+# the consensus cluster assignment predicted by consensusMIBC package is
+# assumed as the ground truth.
+
+# tools ------
+
+  library(tidyverse)
+  library(trafo)
+  library(stringi)
+  library(rlang)
+
+  library(glmnet)
+  library(caret)
+  library(performanceEstimation)
+
+  library(furrr)
+
+  library(soucer)
+
+  select <- dplyr::select
+  reduce <- purrr::reduce
+  map <- purrr::map
+  set_rownames <- trafo::set_rownames
+  rename <- dplyr::rename
+
+  insert_head()
+
+  c('./tools/globals.R',
+    './tools/functions.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  c('./ml scripts/globals.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# analysis scripts -------
+
+  insert_msg('Analysis scripts')
+
+  ## model tuning and training
+
+  list(cache_path = c('./cache/ml_train.RData',
+                      './cache/ml_ens.RData'),
+       script_path = c('./ml scripts/development.R',
+                       './ml scripts/ensemble.R'),
+       message = c('Cached model tuning and training results',
+                   'Cached ensemble tuning and training results')) %>%
+    pwalk(access_cache)
+
+
+  ## predictions, evaluation, and variable importance
+
+  c('./ml scripts/predictions.R',
+    './ml scripts/evaluation.R',
+    './ml scripts/importance.R') %>%
+    source_all(message = TRUE, crash = TRUE)
+
+# END --------
+
+  insert_tail()

### 2024-12-05 report scripts/figures.R

@@ -10,7 +10,7 @@
 
   insert_msg('General frequency of genetic alterations')
 
-  ## facetting by the gene (?) and common X scale
+  ## faceting by the gene (?) and common X scale
 
   rep_figs$general_freq <- expl_genet$plots %>%
     map(~.x +
@@ -36,7 +36,9 @@
               .,
               ncol = 2) %>%
     plot_grid(get_legend(expl_genet$plots$mutation +
-                           theme(legend.position = 'bottom')),
+                           guides(fill = guide_legend(nrow = 2)) +
+                           theme(legend.position = 'bottom',
+                                 legend.title = element_blank())),
               nrow = 2,
               rel_heights = c(0.9, 0.1))
 
@@ -48,32 +50,33 @@
               w = 180,
               h = 180)
 
-# Frequency of alterations of FGF and FGFR genes ------
+# Frequency of alterations of FGF, FGFR, and FGFBP genes ------
 
-  insert_msg('Frequency of alterations of FGF and FGFR genes')
+  insert_msg('Frequency of alterations of FGF, FGFR, and FGFBP genes')
 
   rep_figs$fgfr_freq <- expl_genet$fgf_plots %>%
-    map(~.x +
-          scale_x_continuous(limits = expl_genet$fgf_plots %>%
-                               map(~.x$data$percent) %>%
-                               reduce(c) %>%
-                               max %>%
-                               c(0, .)) +
-          theme(plot.subtitle = element_blank(),
-                legend.position = 'none')) %>%
-    c(list(legend = get_legend(expl_genet$fgf_plots)))
-
-  rep_figs$fgfr_freq <-
-    rep_figs$fgfr_freq[c("mutation", "amplification", "deletion", "legend")] %>%
+    map2(., c('Somatic mutations', 'Deletions', 'Amplifications'),
+         ~.x +
+           labs(title = .y) +
+           scale_x_continuous(limits = expl_genet$fgf_plots %>%
+                                map(~.x$data$percent) %>%
+                                reduce(c) %>%
+                                max %>%
+                                c(0, .)) +
+           theme(plot.subtitle = element_blank(),
+                 legend.position = 'none')) %>%
+    c(list(legend = get_legend(expl_genet$fgf_plots$mutation))) %>%
     plot_grid(plotlist = .,
               ncol = 2,
-              rel_heights = c(10, 4.5)) %>%
+              align = 'hv',
+              axis = 'tblr') %>%
     as_figure(label = 'fgfr_fgf_alteration_frequency',
[…]

### 2024-12-05 report scripts/links.R

@@ -17,7 +17,11 @@
          'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools',
          'microViz' = 'https://github.com/PiotrTymoszuk/microViz',
          'polcaExtra' = 'https://github.com/PiotrTymoszuk/polcaExtra',
-         'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions') %>%
+         'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions',
+         'fastTest' = 'https://github.com/PiotrTymoszuk/fastTest',
+         'perich' = 'https://github.com/PiotrTymoszuk/perich',
+         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra',
+         'htGLMNET' = 'https://github.com/PiotrTymoszuk/htGLMNET') %>%
     compress(names_to = 'obj_name',
              values_to = 'x') %>%
     mutate(ref_name = paste0('_', obj_name, '_'))

### 2024-12-05 report scripts/tables.R

@@ -6,13 +6,38 @@
 
   rep_tabs <- list()
 
+# Genes of interest --------
+
+  insert_msg('Genes of interest')
+
+  ## FGF15 is absent from the human genoe
+
+  rep_tabs$genes <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
+    map(~tibble(`Gene symbol` = .x)) %>%
+    map(filter, `Gene symbol` != 'FGF15') %>%
+    compress(names_to = 'Protein type') %>%
+    mutate(`Protein type` = car::recode(`Protein type`,
+                                        "'binding_proteins' = 'binding proteins'"),
+           `Entrez ID` = mapIds(org.Hs.eg.db,
+                                keys = `Gene symbol`,
+                                keytype = 'SYMBOL',
+                                column = 'ENTREZID')) %>%
+    relocate(`Protein type`) %>%
+    as_mdtable(label = 'genes',
+               ref_name = 'genes',
+               caption = paste('FGFR-, FGF-, and FGFBP-coding genes',
+                               'investigated in the current report.'))
+
 # characteristic of the cohorts -------
 
   insert_msg('Cohort characteristic')
 
   rep_tabs$cohorts <- expl_cohorts$result_tbl %>%
     set_names(c('Variable',
-                globals$cohort_labs[c("genie", "msk", "tcga", "imvigor")],
+                globals$cohort_labs[c("genie", "msk",
+                                      "tcga", "imvigor",
+                                      "bcan")],
                 'Significance',
                 'Effect size')) %>%
     mdtable(label = 'cohort_characteristic',
@@ -55,8 +80,8 @@
     mdtable(label = 'genetic_fgf_fgfr_features',
             ref_name = 'fgf_genetics',
             caption = paste('Frequency of somatic mutations and copy number',
-                            'variants of FGF- and FGFR-coding genes in the',
-                            'investigated cohorts.',
+                            'variants of FGF-, FGFR-, and FGFBP-coding genes',
+                            'in the investigated cohorts.',
                             'Statistical significance for differences between',
                             'the cohorts was assessed by chi-squared test',
                             "with Cramer's V effect size statistic.",
@@ -87,7 +112,8 @@
     mdtable(label = 'fgfr_mutation_classification',
             ref_name = 'fgfr_class',
             caption = paste('Frequency of mutations in the FGFR1/2/3/4 genes',
-                            'in the GENIE BLCA and TCGA BLCA cohorts',
[…]

### 2024-12-05 subtype scripts/coexpression_networks.R

@@ -1,0 +1,127 @@
+# Co-expression networks of FGF-, FGFR-, and FGFBP-coding genes found to be
+# differentially regulated between consensus molecular classes in at least
+# two of the TCGA, IMvigor, and BCAN cohorts.
+
+  insert_head()
+
+# container -----
+
+  sub_exnet <- list()
+
+# analysis data --------
+
+  insert_msg('Analysis data')
+
+  ## variables of interest and their regulation status
+  ## in the consensus subtypes, protein product classification
+
+  sub_exnet$attr_table <- sub_dge$common_significant %>%
+    map(map,
+        ~tibble(variable = .x)) %>%
+    map2(., names(.),
+         ~compress(.x, names_to = .y)) %>%
+    reduce(full_join, by = 'variable')
+
+  sub_exnet$attr_table[-1] <- sub_exnet$attr_table[-1] %>%
+    map_dfc(~ifelse(is.na(.x), 'ns', .x)) %>%
+    map_dfc(factor, c('upregulated', 'downregulated', 'ns'))
+
+  sub_exnet$attr_table <- sub_exnet$attr_table %>%
+    mutate(protein_type = ifelse(variable %in% globals$ligands,
+                                 'ligand',
+                                 ifelse(variable %in% globals$receptors,
+                                        'receptor', 'binding protein')),
+           protein_type = factor(protein_type,
+                                 c('receptor', 'ligand', 'binding protein')))
+
+  sub_exnet$sub_levels <-
+    names(sub_exnet$attr_table)[!names(sub_exnet$attr_table) %in% c('protein_type', 'variable')]
+
+  sub_exnet$variables <- sub_exnet$attr_table$variable
+
+  ## log2 expression data
+
+  sub_exnet$data <- list(tcga = tcga,
+                         imvigor = imvigor,
+                         bcan = bcan) %>%
+    map(~.x$expression[c('sample_id', sub_exnet$variables)]) %>%
+    map(column_to_rownames, 'sample_id')
+
+# Graph objects ---------
+
+  insert_msg('Graph objects')
+
+  sub_exnet$graph_obj <- sub_exnet$data %>%
+    map(as_iGraph,
+        fun = cor,
+        cutoff = 0.3,
+        method = 'spearman',
+        weighted = TRUE,
+        diag = FALSE)
+
+# Community stats --------
+
+  insert_msg('Community stats')
+
+  sub_exnet$stats <- sub_exnet$graph_obj %>%
[…]

### 2024-12-05 subtype scripts/dge.R

@@ -4,6 +4,11 @@
 #
 # 1) Significant difference in expression between the classes by FDR-corrected
 # one-way ANOVA (pFDR < 0.05).
+#
+# 2) Differential expression in a given class as compared with LumP is
+# investigated by two-tailed T test.
+#
+# significantly differentially regulated genes are identified by pFDR(ANOVA) < 0.05
 
 
   insert_head()
@@ -12,17 +17,12 @@
 
   sub_dge <- list()
 
-# parallel backend ------
-
-  insert_msg('Parallel backend')
-
-  plan('multisession')
-
 # analysis data -------
 
   insert_msg('Analysis data')
 
-  sub_dge$variables <- globals[c("receptors", "ligands")] %>%
+  sub_dge$variables <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
     reduce(union)
 
   ## expression data
@@ -33,8 +33,9 @@
     compact %>%
     map(select, sample_id, any_of(sub_dge$variables))
 
-  sub_dge$variables <-
-    sub_dge$variables[sub_dge$variables %in% names(sub_dge$data[[1]])]
+  sub_dge$variables <- sub_dge$data %>%
+    map(names) %>%
+    map(intersect, sub_dge$variables)
 
   ## assignment to the molecular classes, setting the modeling
   ## intercept to LumP
@@ -43,68 +44,115 @@
     map2(subtypes$assignment[names(sub_dge$data)],
          sub_dge$data,
          inner_join, by = 'sample_id') %>%
+    map(filter, consensusClass != 'not assigned') %>%
+    map(mutate, consensusClass = droplevels(consensusClass)) %>%
     map(mutate,
         consensusClass = fct_relevel(consensusClass, 'LumP'))
+
+  ## BCAN: only one observation in the NE-like subset, removal of the subset
+  ## in this cohort
+
+  sub_dge$data$bcan <- sub_dge$data$bcan %>%
+    filter(consensusClass != 'NE-like') %>%
+    mutate(consensusClass = droplevels(consensusClass))
+
+  sub_dge$data <- sub_dge$data %>%
+    map(filter, !is.na(consensusClass))
+
+  ## levels for T tests
+
+  sub_dge$t_levels <- sub_dge$data %>%
+    map(filter, consensusClass != 'LumP') %>%
+    map(mutate, consensusClass = droplevels(consensusClass)) %>%
[…]

### 2024-12-05 subtype scripts/genetic.R

@@ -1,5 +1,9 @@
 # Frequency of FGFR and FGF alterations in the molecular subtypes.
-# The subtype information is available only for the TCGA and IMvigor subsets.
+# The subtype information is available only for the TCGA, IMvigor,
+# and BCAN cohorts.
+#
+# Statistical significance is terted by weighted permutations implemented
+# by perich.
 
   insert_head()
 
@@ -7,13 +11,20 @@
 
   sub_genet <- list()
 
+# parallel backend --------
+
+  insert_msg('Parallel backend')
+
+  plan('multisession')
+
 # analysis data -------
 
   insert_msg('Analysis data')
 
   ## variables
 
-  sub_genet$variables <- globals[c("receptors", "ligands")] %>%
+  sub_genet$variables <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
     reduce(union)
 
   ## data, appending with the subtype information
@@ -22,8 +33,8 @@
     eval %>%
     map(~.x[c('mutation', 'deletion', 'amplification')]) %>%
     map(compact) %>%
-    map(map, select, sample_id, any_of(sub_genet$variables)) %>%
     map(map, column_to_rownames, 'sample_id') %>%
+    map(map, ~.x[!names(.x) %in% c('sample_id', 'patient_id', 'tissue_type')]) %>%
     map(~map2(., names(.),
               ~set_names(.x, paste(names(.x), .y, sep = '_')))) %>%
     map(map, rownames_to_column, 'sample_id') %>%
@@ -33,14 +44,15 @@
     map2(map(subtypes$assignment, ~.x[c('sample_id', 'consensusClass')]),
          sub_genet$data[names(subtypes$assignment)],
          inner_join, by = 'sample_id') %>%
-    map(select, -sample_id)
+    map(select, -sample_id) %>%
+    map(filter, !is.na(consensusClass))
 
   ## removal of features completely absent from the data sets
 
   for(i in names(sub_genet$data)) {
 
     sub_genet$data[[i]] <- sub_genet$data[[i]] %>%
-      map_dfc(function(x) if(sum(as.numeric(x)) == 0) NULL else x)
+      map_dfc(function(x) if(sum(as.numeric(x), na.rm = TRUE) == 0) NULL else x)
 
   }
 
@@ -48,7 +60,11 @@
 
   sub_genet$variables <- sub_genet$data %>%
     map(select, -consensusClass) %>%
-    map(names)
+    map(names) %>%
[…]

### 2024-12-05 subtypes.R

@@ -10,6 +10,8 @@
 
   library(exda)
   library(microViz)
+  library(fastTest)
+  library(perich)
 
   library(furrr)
 
@@ -32,8 +34,13 @@
 
   insert_msg('Analysis scripts')
 
-  c('./subtype scripts/genetic.R',
-    './subtype scripts/dge.R') %>%
+  ## genetics of the consensus subtypes
+
+  access_cache(cache_path = './cache/sub_genet.RData',
+               script_path = './subtype scripts/genetic.R',
+               message = 'Cached results for genetic alterations')
+
+  c('./subtype scripts/dge.R') %>%
     source_all(message = TRUE, crash = TRUE)
 
 # END ------

### 2024-12-05 survival scripts/importance.R

@@ -1,0 +1,65 @@
+# Variable importance for the Elastic Net Cox model of overall survival
+
+  insert_head()
+
+# container -------
+
+  surv_imp <- list()
+
+# model coefficients: they translate to log HR --------
+
+  insert_msg('Model coefficients')
+
+  surv_imp$coefs <- surv_train$tune_obj$model %>%
+    coef %>%
+    as.matrix %>%
+    as.data.frame %>%
+    set_names('beta') %>%
+    rownames_to_column('variable') %>%
+    as_tibble
+
+  surv_imp$coefs <- surv_imp$coefs %>%
+    filter(beta != 0) %>%
+    mutate(hr = exp(beta))
+
+# Bar plots of the HR values ---------
+
+  insert_msg('Bar plots of the HR values')
+
+  surv_imp$coef_plot <- surv_imp$coefs %>%
+    ggplot(aes(x = beta,
+               y = reorder(html_italic(variable), beta),
+               fill = factor(sign(beta)),
+               color = factor(sign(beta)))) +
+    geom_bar(stat = 'identity',
+             color = 'white') +
+    geom_vline(xintercept = 0,
+               linetype = 'solid',
+               linewidth = 0.25) +
+    geom_text(aes(label = signif(hr, 3),
+                  x = 1.05 * beta,
+                  hjust = ifelse(beta < 0, 1, 0)),
+              size = 2.5,
+              show.legend = FALSE) +
+    scale_y_discrete(labels = function(x) stri_replace(x,
+                                                       fixed = '_sec',
+                                                       replacement = html_sup('2'))) +
+    scale_fill_manual(values = c('-1' = 'steelblue',
+                                 '1' = 'firebrick'),
+                      labels = c('-1' = 'favorable',
+                                 '1' = 'unfavorable'),
+                      name = 'risk\nassociation') +
+    scale_color_manual(values = c('-1' = 'steelblue',
+                                 '1' = 'firebrick'),
+                      labels = c('-1' = 'favorable',
+                                 '1' = 'unfavorable'),
+                      name = 'risk\nassociation') +
+    globals$common_theme +
+    theme(axis.title.y = element_blank(),
+          axis.text.y = element_markdown()) +
+    labs(title = 'Overall survival, variable importance',
+         x = 'log HR, per SD')
+
+# END -------
+
+  insert_tail()

### 2024-12-05 survival scripts/univariable.R

@@ -1,0 +1,253 @@
+# Univariable KM analysis if survival of tertile fo gene expression.
+# Statistical significance is determined by Peto-Peto test corrected for
+# multiple testing with the FDR method. The optimal cutoffs for stratification
+# of the expression values are found with `surv_cutpoint`.
+#
+# Significant effects shared by at least two cohorts:
+# unfavorable: TNFAIP6, GPC1, FIBP, FGF11, FGFBP1.
+
+  insert_head()
+
+# container --------
+
+  surv_km <- list()
+
+# parallel backend ------
+
+  insert_msg('Parallel backend')
+
+  plan('multisession')
+
+# analysis globals ---------
+
+  insert_msg('Analysis globals')
+
+  surv_km$variables <-
+    surv_globals$variables[!stri_detect(surv_globals$variables, regex = 'sec$')]
+
+  surv_km$data <- surv_globals$data
+
+  surv_km$cutpoints <- surv_km$data %>%
+    map(surv_cutpoint,
+        time = 'os_days',
+        event = 'death',
+        variables = surv_km$variables,
+        minprop = 0.1)
+
+  surv_km$data <- surv_km$cutpoints %>%
+    map(surv_categorize)
+
+  ## setting factor levels according to the
+  ## survminer's convention
+
+  for(i in names(surv_km$data)) {
+
+    surv_km$data[[i]][surv_km$variables] <-
+      surv_km$data[[i]][surv_km$variables] %>%
+      map_dfc(factor, c('high', 'low'))
+
+
+  }
+
+# Survfit objects ---------
+
+  insert_msg('Survfit objects')
+
+  for(i in names(surv_km$data)) {
+
+    surv_km$survfit_obj[[i]] <- surv_km$variables %>%
+      map(~paste('Surv(os_days, death) ~', .x)) %>%
+      map(as.formula) %>%
+      map(surv_fit,
+          data = surv_km$data[[i]]) %>%
+      set_names(surv_km$variables)
+
+  }
+
+# Numbers of patients and events in the strata --------
+
+  insert_msg('Numbers of patients in the strata')
+
+  for(i in names(surv_km$data)) {
+
+    surv_km$n_numbers[[i]] <- surv_km$variables %>%
+      set_names(surv_km$variables) %>%
+      map(~split(surv_km$data[[i]], surv_km$data[[i]][[.x]])) %>%
+      map(map,
+          ~tibble(n_total = nrow(.x),
[…]

### 2024-12-05 tools/functions.R

@@ -13,6 +13,10 @@
   library(jaccard)
   library(expm)
 
+  library(glmnet)
+  library(DescTools)
+  library(performanceEstimation)
+  library(CalibrationCurves)
   library(ranger)
 
   library(cowplot)
@@ -65,8 +69,11 @@
     if(!is.null(dict)) {
 
       df <- df %>%
-        exchange(variable = 'variable',
-                 dict = dict, ...)
+        mutate(variable = ifelse(stri_detect(variable,
+                                             regex = '^(Samples|Patients)'), #
+                                 variable,
+                                 exchange(variable,
+                                          dict = dict, ...)))
 
     }
 
@@ -225,6 +232,7 @@
                             plot_subtitle = NULL,
                             x_lab = 'protein position',
                             y_lab = NULL,
+                            y_limits = NULL,
                             palette_title = NULL,
                             reverse_split = TRUE,
                             protein_length = NULL, ...) {
@@ -314,7 +322,15 @@
 
     if(!is.null(protein_length)) {
 
-      pos_plot <- pos_plot + scale_x_continuous(limits = c(0, protein_length))
+      pos_plot <- pos_plot +
+        scale_x_continuous(limits = c(0, protein_length))
+
+    }
+
+    if(!is.null(y_limits)) {
+
+      pos_plot <- pos_plot +
+        scale_y_discrete(limits = y_limits)
 
     }
 
@@ -674,133 +690,599 @@
 
   }
 
-# Random Forests ------
+# ML modeling ---------
 
-  tune_rf <- function(data,
-                      formula,
-                      tune_grid, ...) {
+  augment_multi <- function(data,
+                            minorities = c('LumU', 'LumNS', 'Stroma-rich'),
+                            k = 7,
+                            perc.over = rep(4, 3),
+                            perc.under = rep(1, 3)) {
 
-    ## tunes a Random Forest model by minimizing the OOB prediction error
+    ## multi-class augmentation
 
-    ## construction of the models ---------
+    data_lst <- minorities %>%
+      map(~mutate(data,
+                  consensusClass = ifelse(consensusClass == .x, .x, 'other'),
+                  consensusClass = factor(consensusClass, c('other', .x)))) %>%
+      set_names(minorities)
 
-    tune_grid <- tune_grid %>%
-      mutate(model_id = paste0('rf_', 1:nrow(.)))
+    synth_lst <-
[…]

### 2024-12-05 tools/globals.R

@@ -17,6 +17,8 @@
                                            r = 2,
                                            unit = 'mm')
 
+  ## common theme for plots
+
   globals$common_theme <- theme_classic() +
     theme(axis.text = globals$common_text,
           axis.title = globals$common_text,
@@ -37,6 +39,16 @@
           plot.margin = globals$common_margin,
           panel.grid.major = element_line(color = 'gray90'))
 
+  ## common theme for networks
+
+  globals$net_theme <- theme_void() +
+    theme(plot.title = element_text(size = 8,
+                                    face = 'bold'),
+          plot.subtitle = globals$common_text,
+          legend.text = globals$common_text,
+          legend.title = globals$common_text,
+          plot.margin = globals$common_margin)
+
 # cohort labs and colors ------
 
   insert_msg('Color labs and colors')
@@ -44,12 +56,14 @@
   globals$cohort_labs <- c(genie = 'GENIE BLCA',
                            msk = 'MSK IMPACT',
                            tcga = 'TCGA BLCA',
-                           imvigor = 'IMvigor')
+                           imvigor = 'IMvigor',
+                           bcan = 'BCAN')
 
   globals$cohort_colors <- c(genie = 'indianred3',
                              msk = 'darkolivegreen4',
                              tcga = 'steelblue',
-                             imvigor = 'gray60')
+                             imvigor = 'gray60',
+                             bcan = 'plum4')
 
   ## a lexicon data frame for convenient annotation with `exchange()`
 
@@ -77,9 +91,17 @@
 
   ## gene symbols
 
-  globals$receptors <- paste0('FGFR', 1:4)
+  globals$receptors <- c(paste0('FGFR', 1:4),
+                         'FGFRL1', 'TGFBR3')
 
-  globals$ligands <- paste0('FGF', c(1:10, 15:23))
+  globals$ligands <- paste0('FGF', c(1:14, 17:23))
+
+  globals$binding_proteins <- c(paste0('FGFBP', 1:3),
+                                'KL', 'KLB', 'PTX3',
+                                'SDC1', 'SDC2', 'SDC4',
+                                'DCN', 'CD44', 'HSPG2',
+                                'GPC1', 'GPC3', 'GPC4',
[…]

### 2024-12-05 report/markdown/report.Rmd
Update of figures, tables, analysis report

@@
 ---
-title: "FGF- and FGFR-coding gene alterations in urothelial cancers"
+title: "Genetic alterations and expression patterns of genes coding for FGF, FGFR, and FGFBP proteins in urothelial cancers"
 subtitle: "Short analysis report"
@@
 
-* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising therapy target in urothelial cancers with approval of the pan-FGFR inhibitor erdafitinib.
+* Fibroblast growth factor receptor (FGFR) signaling has emerged as a promising cancer therapy target with approval of the pan-FGFR inhibitor erdafitinib for treatment of urothelial carcinoma [@Korc2009].
 However, effects of genetic alterations and modulation of expression of FGF- and FGFR-coding genes in urothelial carcinoma on tumor biology, therapy response, and clinical outcome are still not fully understood.
-
-* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations in FGF- and FGFR-coding genes in large publicly available cohorts of urothelial cancers.
+In particular, it is still not completely clear how and to which extent development and aggressiveness is driven by FGFR signaling in various form of urothelial carcinoma reflected by e.g.
+consensus molecular subtypes [@Kamoun2020].
+
+* The primary goal of the current project was to characterize frequency, co-occurrence, and clinical implications of genetic alterations and differential expression of FGF-, FGFR-, and FGF binding protein (FGFBP) coding genes (__Table \@ref(tab:tab-genes)__) [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a] in large publicly available cohorts of urothelial cancers.
 The secondary goal was an attempt to establish subsets of urothelial cancer specimens based on differences in their profiles of somatic mutations and gene copy number variants.
 
-* Four collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], and the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017].
-Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, and TCGA BLCA cohorts.
+* Five collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers [@Damrauer2022].
+Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BLCA cohorts.
 Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort.
 Binarized data on gene amplifications and deletions were obtained for the GENIE BLCA, MSK IMPACT, and TCGA BLCA collectives.
-Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA and IMvigor cohorts.
-Overall survival information was available for the MSK IMPACT, TCGA BLCA, and IMvigor cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective.
+Gene expression quantification by RNA sequencing and assignment to consensus molecular classes [@Kamoun2020] were retrieved for the TCGA BLCA, IMvigor, and BCAN cohorts.
+Overall survival information was available for the MSK IMPACT, TCGA BLCA, IMvigor, and BCAN cohorts; data on disease-specific and relapse-free survival was provided for the TCGA BLCA collective.
 Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days.
@@
 Somatic mutations of the remaining _FGFR1_, _FGFR2_, and _FGFR4_ receptor genes were less frequent (`r expl_genet$stats$mutation %>% filter(variable %in% c('FGFR1', 'FGFR2', 'FGFR4')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of cancer samples).
-Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the 11q13 chromosome region, which codes also for an essential oncogene _CCDN1_.
-Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figures \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Tables \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).
-
-* A great majority of somatic mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ was classified as missense mutations caused by single nucleotide polymorphisms.
-Particularly for _FGFR3_, they affected protein residues were frequently located apart from the Ig-like domains responsible for ligand biding and besides the kinase domains vital for activation of downstream signaling (__Figures \@ref(fig:fig-mutation-type)__ - __\@ref(fig:fig-protein-domain-split)__, __Table \@ref(tab:tab-fgfr-class)__).
-A more detailed analysis revealed the Ig-like domain 1 and kinase domain as the mutation hot-spots of _FGFR1_.
+Amplifications of the _FGF3_, _FGF4_, and _FGF19_ genes in urothelial cancers result from multiplication of the _11q13_ chromosome region, which codes also for an essential oncogene _CCDN1_.
+Amplifications of _FGF3_, _FGF4_, and _FGF19_ were detected in `r expl_genet$stats$amplification %>% filter(variable %in% c('FGF3', 'FGF4', 'FGF19')) %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens (__Figure \@ref(fig:fig-general-freq)__ and __\@ref(fig:fig-fgfr-freq)__, __Table \@ref(tab:tab-genetics)__ and __\@ref(tab:tab-fgf-genetics)__).
+
+* Mutations of _FGFR1_, _FGFR2_, _FGFR3_, and _FGFR4_ were predominantly classified as missense mutations caused by single nucleotide polymorphisms (__Figure \@ref(fig:fig-mutation-type)__ and __\@ref(fig:fig-variant-type)__, __Table \@ref(tab:tab-fgfr-class)__).
+For _FGFR1_, the mutation hot-spots were located in the Ig-like domain 1 and kinase domain.
 The kinase domain was the prime target of mutations of _FGFR2_.
-Hinge regions between the Ig-like domains 1 and 2, as well as residues around the membrane-spanning region between the Ig-like domain 3 and the kinase domain were identified as mutational hot-spots of _FGFR3_.
-In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-fgfr1-position)__ - __\@ref(fig:fig-fgfr4-position)__).
-It can be hence speculated, that somatic mutations of FGF receptors may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.
+For _FGFR3_, they affected protein residues were located between Ig-like domains 1 and 2 responsible for ligand biding, or at the membrane-spanning hinge between the ligand-binding region and the kinase domain responsible for activation of the downstream signaling.
+In case of _FGFR4_, the Ig-like domain 1, the hinge region between the Ig-like domain 3 and kinase domain, as well as the kinase domain were prime targets of mutations of _FGFR4_ (__Figures \@ref(fig:fig-protein-domain-split)
+__ - __\@ref(fig:fig-fgfr4-position)__, __Table \@ref(tab:tab-fgfr-class)__).
+It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.
 
 * Co-occurrence and exclusivity of genetic features was investigated by analysis of pairwise Jacquard similarity coefficients for pairs of genetic alterations in urothelial cancer samples.
-Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).
-
-* Concerning co-regulation of expression of FGF- and FGFR-coding genes, _FGFR1_ and _FGFR3_ mRNA levels were found to be weakly but significantly negatively associated.
-Increased expression of _FGFR1_ in the cancer tissue correlated significantly with upregulation of genes coding for multiple FGF ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_, _FGF10_, and _FGF18_).
-This phenomenon was absent for _FGFR3_ (__Figure \@ref(fig:fig-fgfr-correlation)__).
-In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, we observed that presence of _FGFR3_ mutations translated in significantly and strongly increased _FGFR3_ mRNA levels.
-By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge)__, __Table \@ref(tab:tab-fgfr3-dge)__).
-This suggests two mechanisms leading to activation of pro-oncogenic FGFR signaling in urothelial carcinoma.
-In _FGFR3_ mutant tumors, FGFR signaling may be triggered by over-expression of FGFR3 protein, improved FGF-affinity, or even in an ligand-independent manner.
-Alternatively, concomitant over-expression of _FGFR1_ with its ligands may activated FGFR signaling in auto- or paracrine manner.
-
-* No consistent differences in total mutation burdens (TMB), counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations.
+Such analysis revealed little overlap between somatic mutations of _FGFR3_ and other common alterations such as _TP53_ and _RB1_ mutations, amplification of the 11q13 chromosome region hallmarked by copy number variants of _FGF3/4/19_ and _CCND1_, as well as deletion of the 9p21 chromosome region with the _MTAP_, _CDKN2A_, and _CDKN2A_ genes (__Figures \@ref(fig:fig-genetic-mds)
+__ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).
+
+* Co-expression of FGFR-, FGF-, and FGFBP-coding genes was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__).
+Especially the network analyses point towards strong co-regulation of expression of _FGFR1_ with genes of its ligands (_FGF2_, _FGF7_, _FGF10_), and co-receptors such as _SDC2_, _DCN_ or _HSPG2_.
+Importantly, _FGFR1_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, general correlation strength with maximum of 1, degree: `r expl_exnet$stats %>% map(filter, name == 'FGFR1') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), _FGF2_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF2') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% min`), _FGF7_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF7') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`), and _FGF10_ (hub score: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$hub_score) %>% range %>% signif(2) %>% paste(collapse = ' to ')`, degree: `r expl_exnet$stats %>% map(filter, name == 'FGF10') %>% map_dbl(~.x$degree) %>% range %>% signif(2) %>% paste(collapse = ' to ')`) were identified as candidate highly connected hubs of the co-expression networks, which suggests that FGFR1 signaling as a critical player in urothelial cancer biology.
+In turn, _FGFR2_ was found to be weakly co-regulated with the non-canonical FGF receptor gene, _TGFBR3_.
+_FGFR3_ levels correlated weakly with levels of _SDC1_ and _GPC1_ transcripts that code for proteoglycan FGF binding proteins.
+
+* In a differential gene expression analysis of cancers with and without _FGFR3_ mutations, presence of _FGFR3_ mutations translated to significantly and strongly increased _FGFR3_ mRNA levels, and, to a lesser extent, upregulation of the syndecan gene _SDC1_.
+By contrast, cancers with _FGFR3_ mutations had significantly lower expression of _FGFR1_, _FGF2_, _FGF5_, and _FGF7_ (__Figure \@ref(fig:fig-fgfr-dge-volcano)__ and __\@ref(fig:fig-fgfr-dge-box)__, __Table \@ref(tab:tab-fgfr3-dge)__).
+No consistent differences in counts of protein-affecting somatic mutations, or counts of gene deletions were found for urothelial cancers with and without _FGFR3_ mutations.
 Gene amplification numbers were in turn significantly lower in _FGFR3_ mutated neoplasms as compared with _FGF3_ wild-type tumors.
-Effect size of this difference was small (__Figure \@ref(fig:fig-fgfr-tmb)__).
+Effect size of this difference was moderate (__Figure \@ref(fig:fig-fgfr-tmb)__).
 
@@
 In the GENIE BLCA and MSK IMPACT cohorts, patients with _FGFR3_ mutated tumors demonstrated improved overall survival as compared with bearers of WT neoplasms.
-This phenomenon was also evident in the TCGA BLCA collective, it was only statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 2)`).
-There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA cohort, which may however result from insufficient number of events of interest (tumor-related deaths and relapses, __Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).
-
-* Analysis of effects of amplification of the 11q13 chromosome region on expression of FGF- and FGFR-coding genes was possible solely for the TCGA BLCA collective.
-Presence of this genetic aberration resulting in amplification of the _FGF3_, _FGF4_, _FGF19_, and _CCND1_ genes translated into lower expression of _FGFR1_ and significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA (__Figure \@ref(fig:fig-amp11q13)__).
-However, effects of those differences was weak and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF ligand-coding genes.
-Hence, biological relevance of the 11q13 amplification for FGFR signaling remains unclear.
-
-* Basal/squamous-like cancers followed by luminal papillary tumors constituted the largest consensus molecular classes [@Kamoun2020] of urothelial cancers in the TCGA BLCA and IMvigor cohorts (__Figure \@ref(fig:fig-fgfr-subtypes)A__).
-in both cohorts, _FGFR3_ mutated cancers were found to be significantly enriched in the papillary luminal class (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`% of specimens).
-In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the 11q13 chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens, __Figures \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).
-
-* mRNA levels of `r sub_dge$anova_significant %>% reduce(intersect) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were found to vary significant between the consensus molecular classes of urothelial carcinomas in both the TCGA BLCA and IMvigor data set (__Table \@ref(tab:tab-cons-dge)__).
-Expression of _FGFR1_ was the highest in stroma-rich and luminal non-specified tumors and the lowest in luminal papillary cancers.
-In turn, maximal mRNA levels of _FGFR3_ were detected in luminal papillary cancers and the lowest expression was observed in basal/squamous- and neuroendocrine-like tumors.
-Effect size of those differences was large (__Figure \@ref(fig:fig-sub-fgfr-expression)__).
-Expression of all significantly regulated _FGF_ ligands was the lowest in luminal papillary cancers.
-_FGF2_, _FGF7_, _FGF10_, and _FGF18_ transcript amounts followed the expression pattern of _FGFR1_ and were detected at the highest levels in stroma-rich and non-specified luminal cancers.
-_FGF5_ was expressed at the highest levels by basal/squamous-like, neuroendocrine-like and stroma-rich neoplasms.
-Effect sizes of differential regulation of these FGF-coding genes was large (__Figure \@ref(fig:fig-sub-fgf-expression1)__ and __\@ref(fig:fig-sub-fgf-expression2)__).
-Collectively, the differential gene expression analysis results support the model of activation of FGFR signaling propose above.
-While luminal papillary cancers may be driven by _FGFR3_ mutations and over-expression resulting in FGFR signaling with little dependence on the ligands, stroma-rich and non-specified luminal cancers may exploit auto- and paracrine signaling fueled by over-expression of _FGFR1_ and its ligands.
-
-* By latent class analysis (LCA, called also latent class modeling), we could establish a classification scheme of urothelial cancers based on presence/absence of the most frequent genetic alterations detected in at least 5% of cancer specimens.
+This phenomenon was also evident in the TCGA BLCA collective and was statistically significant upon exclusion of pT4 cancer patients (entire cohort: p = `r signif(fgfr_os$test$tcga$p_adjusted, 2)`, without pT4 cancers: p = `r signif(fgfr_os$test$tcga_wo_t4$p_adjusted, 3)`).
+There were no significant differences in disease-specific or relapse-free survival between the _FGFR3_ mutation strata in the TCGA BLCA and BCAN cohorts, which may, however, result from insufficient number of relapses and tumor-related deaths (__Figures \@ref(fig:fig-fgfr-os)__ and __Figure \@ref(fig:fig-fgfr-rfs)__).
+
+* Amplification of the _11q13_ chromosome region was one of the most frequent genetic alterations in urothelial cancers and led to multiplication of _FGF3_, _FGF4_, and _FGF19_ genes.
+Analysis of effects of this amplification expression of FGFR-, FGF- and FGFBP-coding genes was possible solely for the TCGA BLCA collective.
+Cancer samples with the _11q13_ amplification had significantly higher levels of _FGF3_, _FGF4_, and _FGF19_ mRNA in the cancer tissue.
+_FGFR1_ was downregulated without formal statistical significance (__Figure \@ref(fig:fig-amp11q13)__).
+Effect sizes of those differences were weak or moderate, and, in case of _FGF3_ and _FGF4_, the expression levels were extremely low as compared with other FGF-coding genes.
+Hence, biological relevance of the _11q13_ amplification for FGFR signaling remains unclear.
+
+* Basal/squamous-like cancers followed by luminal papillary and stroma-rich tumors constituted the largest consensus molecular classes of urothelial cancers [@Kamoun2020] in the TCGA BLCA, IMvigor, and BCAN cohorts.
+In all three collectives, _FGFR3_ mutations were found to be significantly enriched in luminal papillary cancer (TCGA: `r sub_genet$stats$tcga %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, IMvigor: `r sub_genet$stats$imvigor %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)`%, BCAN: `r sub_genet$stats$bcan %>% filter(variable == 'FGFR3_mutation', consensusClass == 'LumP') %>% .$percent %>% signif(2)` of specimens).
+In the TCGA BLCA cohort, amplifications of _FGF3_, _FGF4_, and _FGF19_ genes resulting from multiplication of the _11q13_ chromosome region were found in luminal papillary (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'LumP') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), stroma-rich (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Stroma-rich') %>% .$percent %>% range %>% signif(2) %>% paste(collapse = ' to ')`%), and basal/squamous-like cancers (`r sub_genet$stats$tcga %>% filter(variable %in% c('FGF3_amplification', 'FGF4_amplification', 'FGF19_amplification'), consensusClass == 'Ba/Sq') %>% .$percent %>% .[1] %>% signif(2)`% of specimens) (__Figure \@ref(fig:fig-fgfr-subtypes)__ and __\@ref(fig:fig-fgfr-subdetails)__, __Table \@ref(tab:tab-cons-genetics)__).
+
+* Analysis of differential expression of FGFR-, FGF-, and FGFBP-coding genes was conducted for the major consensus molecular subtypes of the TCGA BLCA, IMvigor, and BCAN cohort.
+Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes (__Table \@ref(tab:tab-cons-dge)__).
+Among them, _FGFR3_, _SDC1_, _TGFBR3_, _GPC3_, and _SDC3_ were upregulated specifically in luminal papillary cancers.
+Overexpression of _KL_, a gene coding for [Klotho](https://www.genecards.org/cgi-bin/carddisp.pl?gene=KL) FGF-binding protein, was characteristic for luminal unstable tumors.
+Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class.
+High expression of _FGF5_, _FGF11_, and genes coding for binding proteins and FGF co-receptors, _CD44_, _FGFBP1_, _GPC1_ and _TNFAIP6_, was a hallmark of basal/squamous-like cancers (__Figures \@ref(fig:fig-sub-fgfr-expression)__ - __Figure \@ref(fig:fig-sub-fgf-expression2)__).
+
+* Encouraged by the profound and widespread differences in expression of FGFR-, FGF-, and FGFBP-coding genes, we asked if and to which extent expression of those genes can explain or emulate the consensus molecular subsets proposed by Kamoun and colleagues [@Kamoun2020].
+To answer this intricate question, we resorted to machine learning with two unrelated algorithms, Elastic Net regularized linear regression [@Friedman2010; @Zou2005] and Random Forest tree ensemble [@Wright2017; @Breiman2001].
+The models were trained in a synthetic data set of ComBat-processed $log_2$-transformed gene expression derived from the TCGA BLCA cohort.
+To improve identification of the minority molecular classes, LumU and stroma-rich cancers, their observations were amplified (over-sampled) in the training data set with the SMOTE algorithm [@Torgo2014; @Chawla2002; @Torgo2016].
+Performance and calibration of the machine learning models was evaluated by comparison of the model-predicted with the actual consensus molecular classes computed by a classification tool of _consensusMIBC_ R package (ground truth) [@Kamoun2020] in the unmodified TCBA BLCA data set and the independent validation IMvigor and BCAN cohorts.
+
+* As presented in __Figure \@ref(fig:fig-sub-elnet)A__ and __Table \@ref(tab:tab-ml-eval)__, the machine learning models predicted the molecular class assignment with very high accuracy, independently of the algorithm.
+Notably, fractions of correctly predicted cancer samples were lower for the luminal unstable class as compared with the remaining molecular classes (__Figure \@ref(fig:fig-sub-elnet)B__).
+Slightly worse reliability of the models at predicting luminal unstable and stroma-rich samples in the validation cohorts was also evident in results of the calibration analyses (__Figure \@ref(fig:fig-sub-cal-elnet)__ and __\@ref(fig:fig-sub-cal-ranger)__).
+__Collectively, the machine learning data suggests that the consensus molecular classification by Kamoun et al.
+can be reproduced by differences in expression of `r length(ml_globals$variables)` genes coding for FGFR, FGF, and FGFBP proteins.__ Notably, the classification tool developed by Kamoun and colleagues and distributed via _consensusMIBC_ R package employs `r consensusMIBC::centroids %>% nrow` gene variables.
+Among them, only `r intersect(consensusMIBC::centroids$hgnc_symbol, ml_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were shared with our machine learning models.
+The most influential genes for prediction of consensus molecular subtypes in our machine learning approach are presented in __Figure \@ref(fig:fig-sub-imp)__, and __Tables \@ref(tab:tab-ml-imp-elnet)__ - __\@ref(tab:tab-ml-imp-ranger)__.
+
+* Furthermore, we asked about prognostic significance of differences in FGFR, FGF, and FGFBP gene expression.
+To answer it, we constructed a multi-parameter Elastic Net Cox model [@Zou2005; @Friedman2010] in the TCGA BLCA cohort (__Table \@ref(tab:tab-surv-eval)__ and __\@ref(tab:tab-surv-imp)__).
+ComBat-adjusted $log_2$-transformed expression levels were fed into the model as first- and second-order terms.
+Among them, `r rep_tabs$surv_imp %>% filter(.data[['Coefficient beta']] != 0, !duplicated(.data[['Gene symbol']])) %>% nrow` genes had non-zero coefficients, i.e.
+contributed substantially to prediction of overall survival (__Figure \@ref(fig:fig-surv-ml)A__).
+The model delivered reliable predictions in the TCGA BLCA training collective and in the IMvigor validation cohort, but not in the BCAN data set (__Figure \@ref(fig:fig-surv-ml)BC__).
+In an univariable, model-free screening of the genes of interest as potential prognostic markers, we found reproducibly shorter overall survival of patients with high expression of _TNFAIP6_, _GPC1_, _FIBP_, _FGF11_, and _FGFBP1_ in the tumor tissue as compared with low expressors (__Figure \@ref(fig:fig-surv-km1)__ and __\@ref(fig:fig-surv-km2)__, __Table \@ref(tab:tab-surv-km)__).
+Taken together, the FGFR, FGF, and FGFBP transcriptome bears some potential for prediction of overall survival in urothelial cancer, but the models and standalone markers need additional validation, especially with inclusion of confounders such as age, gender, and staging.
+
+* By latent class analysis (LCA, called also latent class modeling), we could establish a classification scheme of urothelial cancers based on presence/absence of `r length(lca_globals$modeling_variables)` most frequent genetic alterations detected in at least 5% of cancer specimens.
 This classification scheme with six 'genetic subsets' was developed in the GENIE BLCA training cohort and technically validated in the MSK IMPACT and TCGA BLCA cohort (__Figures \@ref(fig:fig-lca-devel)__ and __\@ref(fig:fig-lca-size)__).
-The 'mutRB1' subset (`r lca_eval$n_numbers %>% map(filter, clust_id == 'mutRB1') %>% map_dbl(~.x$percent) %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens) was characterized by highly frequent _TP53_, _RB1_ and _ARID1A_ mutations, _E2F3_ amplifications, and nearly absent _FGFR3_ mutations, 11q13 amplification, and 9p21 deletions.
+The 'mutRB1' subset (`r lca_eval$n_numbers %>% map(filter, clust_id == 'mutRB1') %>% map_dbl(~.x$percent) %>% range %>% signif(2) %>% paste(collapse = ' to ')`% of specimens) was characterized by highly frequent _TP53_, _RB1_ and _ARID1A_ mutations, _E2F3_ amplifications, and near absence of _FGFR3_ mutations, _11q13_ amplifications and _9p21_ deletions.
 The 'oligoMut' subset was hallmarked by frequent mutations of _TP53_ and the lowest overall mutation count.
-The key features of 'hyperMut' subset were the highest total mutation count, as well as frequent somatic mutations of _ARID1A_, _FGFR3_, _CREBBP_, _ERBB2_, _FBXW7_, and lacking copy number variants of 11q13 and 9p21.
-The 'del9p21' subset was characterized by relatively common deletions of the chromosome region 9p21, amplification of the 11q13 chromosome region, and highly frequent mutations of _TP53_ and _KDM6A_.
-In the 'mutFGFR3' subset, mutations of _FGFR3_ and _KDM6A_, and deletions of 9p21 were common, while mutations in _TP53_, _RB1_, and _ERBB2_, and gene amplifications were virtually absent.
-Finally, the 'ampMDM2' subset was characterized by the highest percentage of _MDM2_, highly common copy number variants of the 11p13 and 9p21 chromosome regions and essentially lacking mutations of _TP53_ and _RB1_ (__Figures \@ref(fig:fig-lca-genet1)__ - __\@ref(fig:fig-lca-details-cna2)__, __Table \@ref(tab:tab-lca-genetics)__).
-
+The key features of 'hyperMut' subset were the highest total mutation count, as well as frequent somatic mutations of _ARID1A_, _FGFR3_, _CREBBP_, _ERBB2_, _FBXW7_, and nearly absent copy number variants of _11q13_ and _9p21_.
+The 'del9p21' subset was characterized by relatively common deletions of the chromosome region _9p21_, amplification of the 11q13 chromosome region, and highly frequent mutations of _TP53_ and _KDM6A_.
+In the 'mutFGFR3' subset, mutations of _FGFR3_ and _KDM6A_, and deletions of _9p21_ were common, while mutations in _TP53_, _RB1_, and _ERBB2_, and gene amplifications were virtually absent.
+Finally, the 'ampMDM2' subset was characterized by the highest percentage of _MDM2_ amplifications, highly common copy number variants of the _11p13_ and _9p21_ chromosome regions; mutations of _TP53_ and _RB1_ were essentially absent (__Figures \@ref(fig:fig-lca-genet1)__ - __\@ref(fig:fig-lca-details-cna2)__, __Table \@ref(tab:tab-lca-genetics)__).
 * Numbers of mutations differed significantly between the genetic subsets with large effect size and were the highest in hyperMut cancers and the lowest in oligoMut and mutFGFR3 tumors.
@@
 
-* We could observe significant differences in expression of _FGFR1_, _FGFR3_, and _FGFR3_, as well as _FGF2_, _FGF5_, FGF7_, and _FGF17_ genes between the genetic subsets of the TCGA BLCA cohort (__Table \@ref(tab:tab-lca-dge)__).
-Interestingly, mutFGFR3 cancers were found to over-express _FGFR3_ and, at the same time, to suppress _FGF2_, _FGF5_, and _FGF7_.
-The highest expression levels of _FGFR1_ were observed in mutRB1 and ampMDM2 genetic subsets; _FGFR2_ was found to be expressed in oligoMut, del9p21, and ampMDM2 at the highest levels (__Figures \@ref(fig:fig-lca-dge-fgfr)__ and __\@ref(fig:fig-lca-dge-fgf)__).
+* We observed significant differences in expression of _FGFR1_, _FGFR3_, _FGFR3_, _FGF5_, as well as genes coding for the binding proteins, _PTX3_, _SDC1_, _GPC1_ and _TNFAIP6_ genes between the genetic subsets of the TCGA BLCA cohort (__Table \@ref(tab:tab-lca-dge)__).
+Interestingly, mutFGFR3 cancers were found to over-express _FGFR3_ and, at the same time, to suppress _FGF5_.
+The highest expression levels of _FGFR1_ were observed in mutRB1 and ampMDM2 genetic subsets; _FGFR2_ was expressed at the highest levels in oligoMut, del9p21, and ampMDM2 subsets (__Figures \@ref(fig:fig-lca-dge-fgfr)__ and __Figure \@ref(fig:fig-lca-dge-fgfbp)__).
+At this point it has to be noted, that the significant differences in expression of the FGFR-, FGF-, and FGFBP-coding genes between genetic clusters were mostly of moderate effect size, and hence way smaller as in the case of consensus molecular subsets proposed by Kamoun and colleagues.
 
@@
 
-* Collectively, our analysis results suggest at least two mechanisms of FGFR signaling activation in urothelial cancers: the one exploiting FGFR3 mutations and largely independent on FGF ligands and operative predominantly in papillary cancers, and the other relying on over-expression of FGFR1 and FGF ligands predominantly by stroma-rich and non-specified luminal tumors.
-Furthermore, we put forward a relatively simple classification scheme of urothelial cancers based on the most common somatic mutations, gene and chromosome region deletions and amplifications established in the large GENIE BLCA collective.
-The key genetic features of those six genetic clusters such as subset-defining mutations and copy number variants, mutation and variant numbers could be reproduced in the MSK IMAPST and TCGA BLCA, i.e.
-two independent validation cohorts.
-Of clinical importance, the genetic subset classification bears consequences for overall survival in a age-, gender-, and stage-independent manner.
+* In summary, our analysis results suggest several modes of triggering of the oncogenic FGFR signaling highly specific for consensus molecular subsets of urothelial cancers (__Figure \@ref(fig:fig-summary)__).
+In papillary neoplasms, mutations of the _FGFR3_ gene affecting the hinge regions of the protein product and overexpression of co-receptors and binding proteins TGFB3, SDC1, GPC3, and SDC4 may culminate in a largely ligand-independent signaling.
+Such ligand-independent FGFR3 signaling was indeed reported for germ line mutations in the hinge region [@Moura2024].
+In stroma-rich cancers, the tightly coordinated overexpression of FGFR1, its ligands (FGF2, FGF7, FGF9, FGF10, FGF14, FGF18) and interacting proteins (DCN, SDC1, HSPG2, ANOS1, and PTX3) may fuel ligand-dependent para- and autocrine signaling.
+In basal/squamous tumors, overexpression of FGF5 along with co-activators like FGFBP1, CD44, or GPC1, may trigger ligand-dependent signaling via FGFR1.
+Interestingly, the elevated FGFPB1 in basal/squamous cancers may overcome the low availability of other FGFR1 ligands, such as FGF2 or FGF5, by forming complexes with high affinity for the receptor [@Tassi2001; @Beer2005].
+In luminal unstable cancers, overexpression of the _KL_ gene was observed, which codes for a co-receptor of FGF19, FGF21, and FGF23 [@Urakawa2006; @Goetz2007].
+Yet, those FGF-coding genes were expressed in bulk cancer samples at nearly undetectable levels (TCGA BLCA, _FGF23_: median: `r median(2^(tcga$expression$FGF23) - 1)` counts, IQR: `r  quantile(2^(tcga$expression$FGF23) - 1, c(0.25, 0.75)) %>% paste(collapse = ' to ')` counts, 95^th^ percentile: `r quantile(2^(tcga$expression$FGF23) - 1, 0.95)` copies).
+Hence, the overexpression of _KL_ in the luminal unstable subset has likely no relevance for FGFR signaling.
+__Importantly, we could demonstrate that variability of expression of the FGFR-, FGF-, and FGFBP-coding genes emulates nearly completely the consensus molecular classification system by Kamoun et al.
+This reflects the immense importance of FGFR signaling for phenotypical variability of urothelial cancers.
+It also puts forward this signaling pathway as an Achilles' heel of all urothelial cancers independent of the molecular phenotype that can be exploited by FGFR inhibitors.__
+
+* Furthermore, we put forward a relatively simple classification scheme of urothelial cancers based on the most common somatic mutations, gene and chromosome region deletions and amplifications established in the large GENIE BLCA collective.
+The key genetic features of those six genetic clusters such as subset-defining mutations and copy number variants, mutation and variant numbers could be reproduced in the independent MSK IMPACT and TCGA BLCA validation cohorts.
+Of clinical importance, the genetic subset classification bears consequences for overall survival in an age-, gender-, and stage-independent manner.
+__In particular, the hyperMut genetic subset characterized by numerous somatic mutations had the most favorable prognosis.
+We hypothesize, that the especially high mutation burden may render those tumors vulnerable to anti-cancer immunity, which may be additionally bolstered by immune checkpoint blockade.
+The discordance of consensus molecular subtypes and genetic subsets suggests that there are likely no strong genetic drivers specific for single consensus molecular subtypes.
+Instead, the phenotypical variability of urothelial neoplasms may be shaped by non-tumor intrinsic factors such as anatomical location, inflammatory milieu, immune response, or even microbiome of the urinary tract.__
 
@@
 Incompleteness of clinical information in the GENIE BLCA cohort precluded an in-depth characteristic of the FGFR3 mutation strata and genetic subsets.
-Unavailability of exact overall survival in this collective may have interfered with results of significant analyses.
-Detailed mutation data as well as copy number information were not provided for the IMvigor collective, which made validation of the genetic subsets impossible in this cohort of immunotherapy patients.
+Unavailability of exact overall survival in this collective may have interfered with significant results of survival analyses.
+Copy number information was not provided for the IMvigor and BCAN collectives, which made validation of the genetic subsets impossible in those cohorts of late stage cancers with immuno- and chemotherapy.
 Finally, gene expression data were not available for the largest GENIE BLCA and MSK IMPACT cohorts, which precluded detailed analyses of differential gene expression, biological phenotype, and consensus molecular classification based on transcriptome.
@@
 
-## Characteristic of investigated cohorts and frequency of genetic alterations
-
-```{r fig-general-freq, fig.width = figur::convert(rep_figs$general_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$general_freq, to = 'in')$h, fig.cap = 'The most frequent somatic mutations and gene copy alterations in urothelial cancer.'}
-
-rep_figs$general_freq$plot
-
-```
-
-__Figure \@ref(fig:fig-general-freq). The most frequent somatic mutations and gene copy alterations in urothelial cancer.__
-
-_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% collapse_and`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort) are presented in bar plots.
-Alterations affecting FGF- and FGFR-coding genes are highlighted with bold font.
-Numbers of analyzed cancer samples are indicated in the plot legend._
-
-\newpage
-
-```{r fig-fgfr-freq, fig.width = figur::convert(rep_figs$fgfr_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_freq, to = 'in')$h, fig.cap = 'Frequency of somatic mutations and copy number alterations of FGF- and FGFR-coding genes.'}
-
-rep_figs$fgfr_freq$plot
-
-```
-
-__Figure \@ref(fig:fig-fgfr-freq). Frequency of somatic mutations and copy number alterations of FGF- and FGFR-coding genes.__
-
-_Percentages of somatic mutations, deletions and amplifications of FGF- and FGFR-coding genes in the investigated cohort are presented in bar plots.
-Note that amplification and deletion data were not available for the `r globals$cohort_labs["imvigor"]` cohort.
-Numbers of analyzed cancer samples are indicated in the plot legend._
-
-\newpage
-
-```{r tab-cohorts, tab.cap = 'Characteristic of the investigated cohorts.
+## Characteristic of investigated cohorts
+
+```{r tab-genes, tab.cap = 'FGFR-, FGF-, and FGFBP-coding genes investigated in the current report.'}
+
+flextable::flextable(rep_tabs$genes) %>% 
+  italic(j = 2, part = 'body') %>% 
+  merge_v(1) %>% 
+  set_widths(rep(3, 3)) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-cohorts, tab.cap = "Characteristic of the investigated cohorts.
 Numeric features are presented as medians with interquartile ranges and ranges.
-Qualitative variables are shown as percentages and observation counts of the categories.'}
+Qualitative variables are shown as percentages and observation counts of the categories."}
 
 flextable::flextable(rep_tabs$cohorts) %>% 
-  set_widths(c(3, rep(3, 4), 2.6, 2.6)) %>% 
-  footnote(1, 6:7, 
+    set_widths(c(2.5, rep(2.7, 5), 2.3, 2.2)) %>% 
+  footnote(1, 7:8, 
            value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
@@
 ```
+
+\newpage
+
+## The most frequent genetic alterations and genetic alterations of genes coding for FGFR, FGF, and FGFBP proteins
+
+\newpage
+
+```{r fig-general-freq, fig.width = figur::convert(rep_figs$general_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$general_freq, to = 'in')$h, fig.cap = 'The most frequent somatic mutations and gene copy alterations in urothelial cancers.'}
+
+rep_figs$general_freq$plot
+
+```
+
+__Figure \@ref(fig:fig-general-freq). The most frequent somatic mutations and gene copy alterations in urothelial cancers.__
+
+_Percentages of somatic mutations, gene deletions and amplification present in at least 5% of cancers samples in all investigated cohorts (`r globals$cohort_labs %>% collapse_and`, note: gene amplification and deletion data were not available for the `r globals$cohort_labs[c("imvigor", "bcan")] %>% collapse_and` cohorts) are presented in bar plots.
+Alterations affecting FGF-, FGFR-, and FGFBP-coding genes are highlighted with bold font.
+Numbers of analyzed cancer samples are indicated in the plot legend._
+
+\newpage
+
+```{r fig-fgfr-freq, fig.width = figur::convert(rep_figs$fgfr_freq, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_freq, to = 'in')$h, fig.cap = 'Frequency of somatic mutations and copy number alterations of FGFR-, FGF-, and FGFBP-coding genes.'}
+
+rep_figs$fgfr_freq$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr-freq). Frequency of somatic mutations and copy number alterations of FGFR-, FGF-, and FGFBP-coding genes.__
+
+_Percentages of somatic mutations, deletions and amplifications of FGFR-, FGF-coding genes found to be affected in at lest two out of the investigated cohorts are presented in bar plots.
+Note that amplification and deletion data were not available for the `r globals$cohort_labs[c("imvigor", "bcan")] %>% collapse_and` collectives.
+Numbers of analyzed cancer samples are indicated in the plot legend._
 
@@
 
-```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes in the investigated cohorts.
+```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes in the investigated cohorts.
 Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic.
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and()` cohorts were classified by their effects on the coding sequence.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts were classified by their effects on the coding sequence.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by variant type.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts were classified by variant type.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts were classified by protein domain.
+_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts were classified by protein domain.
 Their frequency was expressed as percentage of cancer samples and presented in stack plots.
@@
 
-```{r tab-fgfr-class, tab.cap = 'Frequency of mutations in the FGFR1/2/3/4 genes in the GENIE BLCA and TCGA BLCA cohorts classified by mutation type, variant type, and protein domain.
-The table is available as a supplementary Excel file.'}
+```{r tab-fgfr-class, tab.cap = "Frequency of mutations in the FGFR1/2/3/4 genes in the GENIE BLCA, MSK, TCGA BLCA, IMvigor, and BCAN cohorts classified by mutation type, variant type, and protein domain.
+The table is available as a supplementary Excel file."}
 
@@
 
-_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein.
+_Somatic mutations in the FGFR1 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR1 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR1 protein domain schemes.
@@
 
-_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein.
+_Somatic mutations in the FGFR2 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR2 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR2 protein domain schemes.
@@
 
-_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein.
+_Somatic mutations in the FGFR3 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR3 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR3 protein domain schemes.
@@
 
-_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein.
+_Somatic mutations in the FGFR4 gene in the `r globals$cohort_labs[c("genie", "msk", "tcga", "BCAN")] %>% collapse_and` were classified by their effects on nucleotide sequence and mapped to amino acids of the FGFR4 protein.
 Mutation localization and type are visualized in scatter plots below the FGFR4 protein domain schemes.
@@
 
-## Co-occurrence of genetic alterations in FGF- and FGFR-coding genes with the most frequent genetic features of urothelial cancers
+## Co-occurrence of genetic alterations in FGFR-, FGF-, and FGFBP-coding genes with the most frequent genetic features of urothelial cancers
 
@@
 
-```{r fig-fgfr-mds, fig.width = figur::convert(rep_figs$fgfr_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_mds, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.'}
+```{r fig-fgfr-mds, fig.width = figur::convert(rep_figs$fgfr_mds, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_mds, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-mds). Co-occurrence of genetic alterations of FGF- and FGFR-coding genes in urothelial cancer.__
-
-_Co-occurrence of somatic mutations, deletions, and amplifications of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
+__Figure \@ref(fig:fig-fgfr-mds). Co-occurrence of genetic alterations of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer.__
+
+_Co-occurrence of somatic mutations, deletions, and amplifications of FGFR-, FGF-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by two-dimensional MDS (multi-dimensional scaling) of pairwise Jaccard's distances._
 _MDS results are visualized in scatter plots with single genetic alterations depicted as points.
@@
 
-```{r fig-fgfr-cont-hm, fig.width = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations selected FGF- and FGFR-coding genes in urothelial cancer.'}
+```{r fig-fgfr-cont-hm, fig.width = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_cont_hm, to = 'in')$h, fig.cap = 'Co-occurrence of genetic alterations selected FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-cont-hm). Co-occurrence of genetic alterations selected FGF- and FGFR-coding genes in urothelial cancer.__
-
-_Co-occurrence of somatic mutations, deletions, and amplifications in FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by pairwise Jaccard's similarity coefficients J.
+__Figure \@ref(fig:fig-fgfr-cont-hm). Co-occurrence of genetic alterations selected FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer.__
+
+_Co-occurrence of somatic mutations, deletions, and amplifications in FGFR-, FGF-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("genie", "msk", "tcga")] %>% collapse_and` cohorts by pairwise Jaccard's similarity coefficients J.
 Statistical significance of co-occurrence, i.e.
@@
 
-## Clinical and biological significance of genetic alterations of FGF- and FGFR-coding genes
-
-```{r fig-fgfr-correlation, fig.width = figur::convert(rep_figs$fgfr_correlation, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_correlation, to = 'in')$h, fig.cap = 'Correlation of mRNA levels of FGF- and FGFR-coding genes.'}
+## Co-regulation of genes coding for FGFR, FGF, and FGFBP proteins
+
+```{r fig-fgfr-correlation, fig.width = figur::convert(rep_figs$fgfr_correlation, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_correlation, to = 'in')$h, fig.cap = 'Correlation of mRNA levels of FGFR-, FGF-, and FGFBP-coding genes.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-correlation). Correlation of mRNA levels of FGF- and FGFR-coding genes.__
-
-_Pairwise correlation of log2-transformed mRNA levels of FGF- and FGFR-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor")] %>%  collapse_and` cohorts by false discovery rate-adjusted Pearson's test.
+__Figure \@ref(fig:fig-fgfr-correlation). Correlation of mRNA levels of FGFR-, FGF-, and FGFBP-coding genes.__
+
+_Pairwise correlation of log2-transformed mRNA levels of FGF-, FGFR-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts by false discovery rate-adjusted Pearson's test.
 Correlation coefficients r for significant gene pairs are depicted in bubble plots.
@@
 Point size represents absolute values of r.
-Numbers of analyzed cancer samples are displayed in the plot captions._
-
-\newpage
-
-```{r fig-fgfr-dge, fig.width = figur::convert(rep_figs$fgfr_dge, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_dge, to = 'in')$h, fig.cap = 'Expression of FGF- and FGFR-coding genes in urothelial cancers stratified by presence of FGFR3 mutations.'}
-
-rep_figs$fgfr_dge$plot
-
-```
-
-__Figure \@ref(fig:fig-fgfr-dge). Expression of FGF- and FGFR-coding genes in urothelial cancers stratified by presence of FGFR3 mutations.__
-
-_Differences in expression of FGF- and FGFR-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric in the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`.
+Numbers of cancer samples are displayed in the plot captions._
+
+\newpage
+
+```{r fig-fgfr-networks, fig.width = figur::convert(rep_figs$fgfr_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.'}
+
+rep_figs$fgfr_networks$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.__
+
+_Co-expression networks were constructed in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman's rho $\rho \geq$ 0.3.
+Isolated vertices of the networks were removed.
+The networks were visualized using the Fruchterman-Reingold algorithm.
+Points represent node vertices/genes, point shape codes for classification of the protein product.
+Point color represent hub scores, which measure the overall correlation strength.
+Lines represent edges, i.e.
+correlations between expression levels with $\rho \geq$ 0.3.
+Line widths correspond to $\rho$ values.
+Note the central position, connectivity, and high hub scores of the genes coding for FGFR1, its ligands FGF2, FGF7 and FGF10, and the binding proteins SDC2, DCN, PTX3 and HSPG2 preserved in all cohorts.
+This suggests that signaling via FGFR1 may serve as a major driver of urothelial cancer._
+
+\newpage
+
+## Transcriptional and clinical significance of _FGFR3_ mutations
+
+```{r fig-fgfr-dge-volcano, fig.width = figur::convert(rep_figs$fgfr_dge_volcano, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_dge_volcano, to = 'in')$h, fig.cap = 'Expression of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancers stratified by presence of FGFR3 mutations: Volcano plots.'}
+
+rep_figs$fgfr_dge_volcano$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr-dge-volcano). Expression of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancers stratified by presence of FGFR3 mutations: Volcano plots.__
+
+_Differences in $log_2$-transformed expression of FGFR-, FGF-, and FGFBP-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric in the `r globals$cohort_labs[c("tcga", "imvigor", "BCAN")] %>% collapse_and`.
 P values were corrected with the false discovery rate (FDR) method.
-Differentially regulated genes were identified by pFDR < 0.05 and at least 1.5 fold-regulation of gene expression in mutated samples as compared with WT specimens._
-
-_(A) pFDR values and log2 fold-regulation estimates of gene expression in mutated vs WT samples are presented in Volcano plots.
-Each point represents a single gene.
-Point color codes for significance and regulation sign, differentially regulated genes are labeled with their symbols.
-The significance and fold-regulation cutoffs of differential gene expression are depicted as dashed lines.
-Numbers of analyzed cancer samples are indicated in the plot caption._
-
-_(B) log2-transformed mRNA levels for genes found to be differentially regulated in both cohorts are presented in box plots.
+Differentially regulated genes were identified by pFDR < 0.05 and at least 1.25 fold-regulation of gene expression in mutated samples as compared with WT specimens._
+
+_$log_2$ fold-regulation estimates of mRNA levels in FGFR-mutated as compared with WT tumors, and FDR-corrected p values for the genes of interest are presented in Volcano plots.
+Each point represents a single gene; point color codes for significance and sign of expression regulation.
+Dashed lines represent the regulation and significance cutoffs used for identification of the differentially expressed genes.
+Numbers of WT and FGFR3-mutated samples are displayed in the plot captions._
+
+\newpage
+
+```{r fig-fgfr-dge-box, fig.width = figur::convert(rep_figs$fgfr_dge_box, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_dge_box, to = 'in')$h, fig.cap = 'Expression of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancers stratified by presence of FGFR3 mutations: selected differentially expressed genes.'}
+
+rep_figs$fgfr_dge_box$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr-dge-box). Expression of FGFR-, FGF-, and FGFBP-coding genes in urothelial cancers stratified by presence of FGFR3 mutations: selected differentially expressed genes.__
+
+_Differences in $log_2$-transformed expression of FGFR-, FGF-, and FGFBP-coding genes between cancers with and without somatic mutations in the FGFR3 gene were assessed by two-tailed T test with Cohen's d effect size metric in the `r globals$cohort_labs[c("tcga", "imvigor", "BCAN")] %>% collapse_and`.
+P values were corrected with the false discovery rate (FDR) method.
+Differentially regulated genes were identified by pFDR < 0.05 and at least 1.25 fold-regulation of gene expression in mutated samples as compared with WT specimens._
+
+_log2-transformed mRNA levels of selected genes found to be differentially regulated in at least two out of the `r globals$cohort_labs[c("tcga","imvigor", "bcan")] %>% collapse_and` collectives are presented in box plots.
 Median expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
@@
 
-_Total mutation burdens (TMB) were provided for the cohorts by the study authors.
-Counts of mutations, deletions and amplifications were computed for each cancer sample based on the published information.
-Note, that information on deletions and amplification was not available for the `r globals$cohort_labs["imvigor"]` cohort.
-TMB (A), and counts of mutations (B), deletions and amplifications (C, D) were compared between cancers with and without FGFR3 mutations by Mann-Whitney test with r effect size statistic.
+_Counts of mutations, deletions and amplifications were computed for each cancer sample based on the published information.
+Note, that information on deletions and amplification was not available for the `r globals$cohort_labs[c("imvigor", "bcan")] %>% collapse_and` cohorts.
+Counts of mutations (A), deletions, and amplifications (C, D) were compared between cancers with and without FGFR3 mutations by Mann-Whitney test with r effect size statistic.
 P values were adjusted for multiple testing with the false discovery rate method.
@@
 flextable::flextable(rep_tabs$fgfr3_clinic) %>% 
-  set_widths(c(2.2, 3, rep(3.5, 2), 2.8, 2.8)) %>% 
+  set_widths(c(2.2, 3, rep(4.5, 2), 2.5, 2.5)) %>% 
   merge_v(1) %>% 
   footnote(1, 5:6, 
-           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
+           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
            part = 'header', 
@@
 
-_Overall survival information was provided for the `r globals$cohort_labs[c("msk", "tcga", "imvigor")] %>% collapse_and` study participants.
+_Overall survival information was provided for the `r globals$cohort_labs[c("msk", "tcga", "imvigor", "bcan")] %>% collapse_and` study participants.
 Approximate overall survival was calculated for the `r globals$cohort_labs["genie"]` cohort based on age at enrollment in years and age at the last follow-up or death in days.
@@
 
-_Disease-specific and relapse-free survival information was provided for the `r globals$cohort_labs["tcga"]` cohort.
+_Disease-specific survival information was provided for the `r globals$cohort_labs[c("tcga", "bcan")] %>% collapse_and` cohorts (A).
+Relapse-free survival was available for the `r globals$cohort_labs["tcga"]` collective (B).
 Differences in survival between patients with cancers with and without FGFR3 mutations were investigated by false discovery rate-adjusted Peto-Peto test.
@@
 
-## Consensus molecular subtypes of urothelial cancers
-
-```{r tab-cons-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes in the consensus molecular classes of urothelial cancers.
+## Genetic alterations and differential expression of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular subtypes of urothelial cancers
+
+```{r tab-cons-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of urothelial cancers.
 The frequencies are presented as percentages of the consensus class.
-Statistical significance of differences between the consensus classes was determined by chi-squared test with Cramer's V effect size statistic.
+Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-```{r fig-fgfr-subtypes, fig.width = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$h, fig.cap = 'Alterations of FGF- and FGFR-coding genes in consensus molecular classes of urothelial cancers.'}
+```{r fig-fgfr-subtypes, fig.width = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subtypes, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-subtypes). Alterations of FGF- and FGFR-coding genes in consensus molecular classes of urothelial cancers.__
-
-_Cancer samples from the `r globals$cohort_labs[c("genie", "tcga")] %>% collapse_and` data sets were assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC)._
-_Differences in frequency of somatic mutations and copy number variants of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by $\chi^2$ test with Cramer's V effect size statistic.
+__Figure \@ref(fig:fig-fgfr-subtypes). Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.__
+
+_Cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` data sets were assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC)._
+_Differences in frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed by weighted permutation test.
 P values were corrected for multiple testing with the false discovery rate method.
@@
 
-_(A) Percentages of cancer samples in the consensus molecular classes were visualized in a stack plot.
-Statistical significance of difference in class distribution between the cohorts was assessed by $\chi^2$ test with Cramer's V effect size statistic.
-The effect size and p value are shown in the plot caption.
-Total numbers of cancer samples are indicated in the Y axis._
-
-_(B) Presence/absence of selected genetic alterations in the consensus molecular classes visualized in heat maps.
+_Presence/absence of selected genetic alterations in the consensus molecular classes visualized in oncoplots.
 Alteration names with effect sizes and p values of differences between the consensus classes are indicated in the Y axis.
@@
 
-```{r fig-fgfr-subdetails, fig.width = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$h, fig.cap = 'Frequency of somatic mutations of FGFFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.'}
+```{r fig-fgfr-subdetails, fig.width = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_subdetails, to = 'in')$h, fig.cap = 'Frequency of somatic mutations of FGFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-subdetails). Frequency of somatic mutations of FGFFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.__
+__Figure \@ref(fig:fig-fgfr-subdetails). Frequency of somatic mutations of FGFR3 and amplification of FGF3/4/19 in consensus molecular classes of urothelial cancer.__
 
 _Assignment of cancer samples to consensus molecular classes and testing for differences in genetic alterations between the consensus classes were presented in Figure \@ref(fig:fig-fgfr-subtypes).
-Percentages of FGFR3 mutations, and amplifications of the FGF3, FGF4 and FGF19 genes in the consensus molecular classes are presented in stack plots.
-Effect sizes and p values are displayed in the plot captions.
+Percentages of FGFR3 mutations (A), and amplifications of the FGF3, FGF4 and FGF19 genes (B) in the consensus molecular classes are presented in stack plots.
+Effect sizes and false discovery rate-corrected p values are displayed in the plot captions.
 Numbers of observations in the consensus classes are indicated in the X axes._
+
 _LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
@@
 
-```{r tab-cons-dge, tab.cap = "Expression of FGF- and FGFR-coding genes in the consensus molecular classes of urothelial cancers.
+```{r tab-cons-dge, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of urothelial cancers.
 log2-transformed expression levels are presented as medians with interquartile ranges and ranges.
@@
 
+```{r fig-sub-fgfr-dge, fig.width = figur::convert(rep_figs$sub_fgfr_dge, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgfr_dge, to = 'in')$h, fig.cap = 'Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_fgfr_dge$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-fgfr-dge). Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.__
+
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
+Differences in $log_2$-transformed gene expression between a subset and luminal papillary subset (LumP) were investigated by two-tailed T test.
+P values were corrected for multiple testing with the false discovery rate method (FDR).
+Significantly differentially regulated genes were defined by p(ANOVA) < 0.05, $\eta^2 \geq$ 0.06, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation in a subset as compared with LumP cancers._
+
+_Z-scores of $log_2$ expression levels of significantly differentially regulated genes shared by at least two cohorts are displayed in heat maps.
+Significant genes are labelled with bold font in the Y axes.
+Numbers of cancer samples in the consensus subtypes are indicated in the plot caption.
+Note: samples without reliable assignment to the consensus subtypes are not presented in the heat maps._
+
+_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
+
+\newpage
+
 ```{r fig-sub-fgfr-expression, fig.width = figur::convert(rep_figs$sub_fgfr_expression, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgfr_expression, to = 'in')$h, fig.cap = 'Differential expression of FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}
@@
 
-_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
-Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed as presented in Figure \@ref(fig:fig-sub-fgfr-dge).
 P values were corrected for multiple testing with the false discovery rate method._
-_Expression of FGFR1 and FGFR3 was found to differ significantly between the consensus classes in both cohorts.
+
+_Expression of FGFR1, FGFR3, and TGFBR3 was found to differ significantly between the consensus classes in all three cohorts.
 Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
 Single cancer samples are visualized as points.
-Effect sizes and p values are displayed in the plot captions.
+Effect sizes and ANOVA p values are displayed in the plot captions.
 Numbers of cancer samples in the consensus classes are indicated in the X axes._
+
 _LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
@@
 
-```{r fig-sub-fgf-expression1, fig.width = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}
-
-rep_figs$sub_fgf_expression1$plot
-
-```
-
-__Figure \@ref(fig:fig-sub-fgf-expression1). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__
-
-_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
-Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
+```{r fig-sub-fgfbp-expression, fig.width = figur::convert(rep_figs$sub_fgfbp_expression, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgfbp_expression, to = 'in')$h, fig.cap = 'Differential expression of FGFBP-coding genes in the consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_fgfbp_expression$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-fgfbp-expression). Differential expression of FGFBP-coding genes in the consensus molecular classes of urothelial cancers.__
+
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed as presented in Figure \@ref(fig:fig-sub-fgfr-dge).
 P values were corrected for multiple testing with the false discovery rate method._
-_Expression of FGF2, FGF5, and FGF7 was found to differ significantly between the consensus classes in both cohorts.
+
+_Expression of FGFBP1, KL, and DCN was found to differ significantly between the consensus classes in all three cohorts.
 Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
 Single cancer samples are visualized as points.
-Effect sizes and p values are displayed in the plot captions.
+Effect sizes and ANOVA p values are displayed in the plot captions.
 Numbers of cancer samples in the consensus classes are indicated in the X axes._
+
 _LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
@@
 
-```{r fig-sub-fgf-expression2, fig.width = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}
-
-rep_figs$sub_fgf_expression2$plot
-
-```
-
-__Figure \@ref(fig:fig-sub-fgf-expression2). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__
-
-_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
-Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes between the consensus molecular classes were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
+```{r fig-sub-sdc-expression, fig.width = figur::convert(rep_figs$sub_sdc_expression, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_sdc_expression, to = 'in')$h, fig.cap = 'Differential expression of FGFBP-coding genes in the consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_sdc_expression$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-sdc-expression). Differential expression of FGFBP-coding genes in the consensus molecular classes of urothelial cancers.__
+
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed as presented in Figure \@ref(fig:fig-sub-fgfr-dge).
 P values were corrected for multiple testing with the false discovery rate method._
-_Expression of FGF10 and FGF18 was found to differ significantly between the consensus classes in both cohorts.
+
+_Expression of SDC1, SDC2, and GPC3 was found to differ significantly between the consensus classes in at least two cohorts.
 Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
 Single cancer samples are visualized as points.
-Effect sizes and p values are displayed in the plot captions.
+Effect sizes and ANOVA p values are displayed in the plot captions.
 Numbers of cancer samples in the consensus classes are indicated in the X axes._
+
 _LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
+
+\newpage
+
+```{r fig-sub-fgf-expression1, fig.width = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression1, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_fgf_expression1$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-fgf-expression1). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__
+
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed as presented in Figure \@ref(fig:fig-sub-fgfr-dge).
+P values were corrected for multiple testing with the false discovery rate method._
+
+_Expression of FGF2, FGF5, and FGF7 was found to differ significantly between the consensus classes in at least two cohorts.
+Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single cancer samples are visualized as points.
+Effect sizes and ANOVA p values are displayed in the plot captions.
+Numbers of cancer samples in the consensus classes are indicated in the X axes._
+
+_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
+
+\newpage
+
+```{r fig-sub-fgf-expression2, fig.width = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_fgf_expression2, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_fgf_expression2$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-fgf-expression2). Differential expression of FGF-coding genes in the consensus molecular classes of urothelial cancers.__
+
+_Assignment of cancer samples from the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` to the consensus molecular classes was presented in Figure \@ref(fig:fig-fgfr-subtypes).
+Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the consensus molecular classes were assessed as presented in Figure \@ref(fig:fig-sub-fgfr-dge).
+P values were corrected for multiple testing with the false discovery rate method._
+
+_Expression of FGF10 and FGF18 was found to differ significantly between the consensus classes in at least two cohorts.
+Median log2-transformed mRNA levels with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single cancer samples are visualized as points.
+Effect sizes and ANOVA p values are displayed in the plot captions.
+Numbers of cancer samples in the consensus classes are indicated in the X axes._
+
+_LumP: luminal papillary; LumU: luminal genetically unstable; LumNS: luminal non-specified; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like; NE-like: neuroendocrine-like._
+
+## Differences in expression levels of FGFR-, FGF-, and FGFBP-coding genes emulate consensus molecular subsets of urothelial cancers
+
+```{r fig-sub-elnet, fig.width = figur::convert(rep_figs$sub_elnet, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_elnet, to = 'in')$h, fig.cap = 'Prediction of molecular consensus classes of urothelial cancer by machine learning models which use expression of FGFR-, FGF-, and FGFBP-coding genes as the sole explanatory factors.'}
+
+rep_figs$sub_elnet$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-elnet).
+Prediction of molecular consensus classes of urothelial cancer by machine learning models which use expression of FGFR-, FGF-, and FGFBP-coding genes as the sole explanatory factors.__
+
+_Assignment of urothelial cancer samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts to the major consensus molecular subsets (LumP, LumU, Stroma-rich, and Ba/Sq) was modeled by two independent machine learning algorithms, Elastic Net linear regression and Random Forest tree ensemble, with ComBat-processed $log_2$-transformed expression levels of genes coding for FGFR, FGF, and FGFBP proteins as the explanatory variables.
+The models were optimized (tuned) and trained in a synthetic data set derived from the `r globals$cohort_labs["tcga"]` collective with expansion (oversampling) of observations in the LumU and stroma-rich classes by the SMOTE algorithm._
+_Performance of the models was evaluated in the non-modified `r globals$cohort_labs["tcga"]` data set as well as in the IMvigor and BCAN independent validation collectives; assignment to the classes by the nearest centroid classifier from consensusMIBC R package served as the ground truth._
+
+_(A) Numeric statistics of overall non-accidental model accuracy (Cohen's $\kappa$, high values expected for a highly accurate model), overall calibration (Brier score, low scores expected for a well calibrated model), as well as sensitivity and specificity (AUC, area under the receiver-operating characteristic curve) are presented in a dot plot.
+Point size corresponds to AUC.
+Data points are labeled with the cohort names and AUC values.
+The dashed lines and gray regions visualize $\kappa$ values expected for a model that assigns to consensus molecular classes by chance._
+
+_(B) Accuracy of the models in particular consensus molecular classes shown in heat map representations of confusion matrices.
+Fields at the diagonals represent correctly classified cancer samples.
+The heat map tiles are labeled with percentages of all cancer samples and cancer sample counts._
+
+_LumP: luminal papillary; LumU: luminal genetically unstable; Stroma-rich: luminal stroma-rich; Ba/Sq: basal/squamous-like._
+
+\newpage
+
+```{r fig-sub-cal-elnet, fig.width = figur::convert(rep_figs$sub_cal_elnet, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_cal_elnet, to = 'in')$h, fig.cap = 'Calibration of the Elastic Net model of consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_cal_elnet$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-cal-elnet). Calibration of the Elastic Net model of consensus molecular classes of urothelial cancers.__
+
+_The machine learning models of consensus molecular subsets with expression of FGFR-, FGF, and FGFBP genes as the sole explanatory factors were trained in a `r globals$cohort_labs["tcga"]`-derived data set and evaluated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts as presented in Figure \@ref(fig:fig-sub-elnet)._
+_In analyses of calibration of the models, assignment to the classes by the nearest centroid classifier from consensusMIBC R package served as the ground truth._
+
+_Calibration of the Elastic Net model in particular consensus molecular classes was assessed by squares of differences between the predicted class probability and the actual class assignment (Brier squares).
+Those square distances are visualized in box plots (left), where medians with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges; single cancer samples are shown as points.
+The dashed lines represent values of Brier squares expected for a random class assignment._
+
+_Additionally, calibration of the Elastic Net model was investigated by modeling of the logistic relationship between mean predicted probabilities of class assignment and fractions of samples in the class.
+The logistic calibration curves for particular consensus classes are presented (right); line color codes for the actual consensus molecular class, the dashed lines represent ideal calibration._
+
+\newpage
+
+```{r fig-sub-cal-ranger, fig.width = figur::convert(rep_figs$sub_cal_ranger, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_cal_ranger, to = 'in')$h, fig.cap = 'Calibration of the Random Forest model of consensus molecular classes of urothelial cancers.'}
+
+rep_figs$sub_cal_ranger$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-cal-ranger). Calibration of the Random Forest model of consensus molecular classes of urothelial cancers.__
+
+_The machine learning models of consensus molecular subsets with expression of FGFR-, FGF, and FGFBP genes as the sole explanatory factors were trained in a `r globals$cohort_labs["tcga"]`-derived data set and evaluated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts as presented in Figure \@ref(fig:fig-sub-elnet)._
+_In analyses of calibration of the models, assignment to the classes by the nearest centroid classifier from consensusMIBC R package served as the ground truth._
+
+_Calibration of the Random Forest model in particular consensus molecular classes was assessed by squares of differences between the predicted class probability and the actual class assignment (Brier squares).
+Those square distances are visualized in box plots (left), where medians with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges; single cancer samples are shown as points.
+The dashed lines represent values of Brier squares expected for a random class assignment._
+
+_Additionally, calibration of the Random Forest model was investigated by modeling of the logistic relationship between mean predicted probabilities of class assignment and fractions of samples in the class.
+The logistic calibration curves for particular consensus classes are presented (right); line color codes for the actual consensus molecular class, the dashed lines represent ideal calibration._
+
+\newpage
+
+```{r fig-sub-imp, fig.width = figur::convert(rep_figs$sub_imp, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_imp, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers.'}
+
+rep_figs$sub_imp$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-imp). Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers.__
+
+_The machine learning models of consensus molecular subsets with expression of FGFR-, FGF, and FGFBP genes as the sole explanatory factors were trained in a `r globals$cohort_labs["tcga"]`-derived data set as presented in Figure \@ref(fig:fig-sub-elnet)._
+_In the Elastic Net model, absolute values of the coefficient's $\beta$ translate directly to variable importance measure; high $abs(\beta)$ is expected for a highly influential variable for prediction of a subset.
+$\beta$ values for the top 15 most influential explanatory variables in the Elastic Net model are presented as bar plots in (A)._
+_For the Random Forest model, permutation importance of explanatory variables was computed as difference in out-of-bag classification errors between the genuine model and a model with the variable of interest re-shuffled at random.
+The differences in classification errors for the top 15 most influential explanatory variables of the Random Forest model are presented in a bar plot in (B)._
+
+\newpage
+
+```{r tab-ml-eval, tab.cap = "Performance of Elastic Net and Random Forest models of LumP, LUMU, stroma-rich, and BA/Sq consensus molecular classes of urothelial cancers that uses expression of FGFR-, FGF-, and FGFBP-coding genes as the sole explanatory factors.
+The models were trained in the TCGA BLCA cohort, the gene expression values were ComBat-processed to minimize inter-cohort variability.
+Minority consensus classes were amplified by the SMOTE algorithm to improve their identification by the model.
+The models were evaluated in the TCBA BLCA, IMvigor, and BCAN cohorts; the consensus class assignment predicted by a nearest centroid algorithm from consensusMIBC package served as the ground truth.
+Metrics of model accuracy, calibration, and receiver-operating characteristic are listed."}
+
+flextable::flextable(rep_tabs$ml_eval) %>% 
+  merge_v(1) %>% 
+  set_widths(c(2.6, 2.7, rep(2.3, 6))) %>% 
+  footnote(1, 3, 
+           value = as_paragraph('Fraction of observations correctly assigned to molecular consensus classes.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  footnote(1, 4, 
+            value = as_paragraph("Cohen's \u03BA. Non-accidental concordance between the actual and predicted molecular class assignment."), 
+            part = 'header', 
+            ref_symbols = 'b') %>% 
+  footnote(1, 5, 
+           value = as_paragraph('Measure of overall model calibration.'),
+           part = 'header', 
+           ref_symbols = 'c') %>% 
+  footnote(1, 6:8, 
+           value = as_paragraph('Receiver-operating characteristic (ROC). AUC: area under the ROC curve.'), 
+           part = 'header', 
+           ref_symbols = 'd') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-ml-imp-elnet, tab.cap = "Coefficients of the Elastic Net model of consensus molecular classes of urothelial cancers.
+The Elastic Net model was developed in the TCGA BLCA cohort with expression values of genes coding for FGFR, FGF, and FGFBP proteins as the sole explanatory factors.
+Values of beta of the coefficients are listed."}
+
+flextable::flextable(rep_tabs$ml_imp_elnet) %>% 
+  italic(j = 1) %>%
+  set_widths(c(2.5, rep(2.3, 4))) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-ml-imp-ranger, tab.cap = "Permutation variable importance of the Random Forest model of consensus molecular classes of urothelial cancers.
+The Random Forest model was developed in the TCGA BLCA cohort with expression values of genes coding for FGFR, FGF, and FGFBP proteins as the sole explanatory factors.
+Permutation importance was computed as differences of errors of the genuine model and models, in which the explanatory variables were re-shuffled at random."}
+
+flextable::flextable(rep_tabs$ml_imp_ranger) %>% 
+  italic(j = 1) %>% 
+  set_widths(c(3, 3)) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+## Prognostic significance of differential expression of FGFR-, FGF-, and FGF-coding genes
+
+```{r fig-surv-ml, fig.width = figur::convert(rep_figs$surv_ml, to = 'in')$w, fig.height = figur::convert(rep_figs$surv_ml, to = 'in')$h, fig.cap = 'Development and evaluation of an Elastic Net Cox model of overall survival with expression of FGFR-, FGF-, and FGFBP-coding genes as explanatory factors.'}
+
+rep_figs$surv_ml$plot
+
+```
+
+__Figure \@ref(fig:fig-surv-ml). Development and evaluation of an Elastic Net Cox model of overall survival with expression of FGFR-, FGF-, and FGFBP-coding genes as explanatory factors.__
+
+_Overall survival was modeled as a function of expression of genes coding for FGFR, FGF, and FGFBP proteins by Elastic Net Cox regression.
+The Elastic Net model was trained in the TCGA BLCA cohort; first- and second-order terms of ComBat-processed $log_2$-transformed expression levels of the genes of interest served as explanatory factors.
+Performance of the model was evaluated in the training data set as well as in the IMvigor and BCAN independent validation collectives._
+
+_(A) Non-zero coefficients of the Elastic Net Cox model represented by hazard ratios (HR), i.e. fold-difference in risk per one standard deviation of $log_2$ gene expression value._
+
+_(B) Numeric statistics of model accuracy (Harrell's concordance index [C-index], high values expected for a highly accurate model), overall calibration (integrated Brier score [IBS], low values expected for a well calibrated models), and explanatory performance ($R^2$) shown in a dot plot.
+Points are labeled with the cohort names and $R^2$ values.
+Point size codes for $R^2$.
+The dashed lines represent values of C-index and IBS expected for a model that predicts survival at random._
+
+_(C) Overall survival in urothelial cancer patients stratified by tertiles of linear predictor scores of the Elastic Net model.
+Statistical significance of differences in survival between the tertiles is determined by Peto-Peto test corrected for multiple testing with the false discovery rate (FDR) method.
+Fractions of surviving patients are visualized with Kaplan-Meier plots.
+Total numbers of patients and deaths are displayed in the plot captions.
+Numbers of patients and deaths in the tertiles are indicated in the plot legends.
+Raw and FDR-corrected p values are shown in the plots._
+
+\newpage
+
+```{r fig-surv-km1, fig.width = figur::convert(rep_figs$surv_km1, to = 'in')$w, fig.height = figur::convert(rep_figs$surv_km1, to = 'in')$h, fig.cap = 'Candidate transcriptional markers of overall survival among the genes coding for FGFR, FGF, and FGFBP proteins.'}
+
+rep_figs$surv_km1$plot
+
+```
+
+__Figure \@ref(fig:fig-surv-km1). Candidate transcriptional markers of overall survival among the genes coding for FGFR, FGF, and FGFBP proteins.__
+
+_Urothelial cancer patients in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts were stratified by ComBat-processed $log_2$-transformed expression levels of FGFR-, FGF-, and FGFBP-coding genes.
+The stratification cutoffs for division into high and low expressors were motivated by the maxima of log-rank statistic.
+Statistical significance of differences in overall survival between the high and low expressors was determined by Peto-Peto test adjusted for multiple comparison with the false discovery rate method (FDR).
+Candidate survival markers were identified by pFDR < 0.05 in at least two cohorts._
+
+_Fractions of surviving patients in the high and low expression strata of TNFAIP6, GPC1, and FIBP genes are presented in Kaplan-Meier plots.
+Stratification cutoffs, and total numbers of patients and deaths are displayed in the plot captions.
+Numbers of patients and deaths in the expression strata are indicated in the plot legends.
+Raw and FDR-adjusted p values are shown in the plots._
+
+\newpage
+
+```{r fig-surv-km2, fig.width = figur::convert(rep_figs$surv_km2, to = 'in')$w, fig.height = figur::convert(rep_figs$surv_km2, to = 'in')$h, fig.cap = 'Candidate transcriptional markers of overall survival among the genes coding for FGFR, FGF, and FGFBP proteins.'}
+
+rep_figs$surv_km2$plot
+
+```
+
+__Figure \@ref(fig:fig-surv-km2). Candidate transcriptional markers of overall survival among the genes coding for FGFR, FGF, and FGFBP proteins.__
+
+_Urothelial cancer patients in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts were stratified by ComBat-processed $log_2$-transformed expression levels of FGFR-, FGF-, and FGFBP-coding genes.
+The stratification cutoffs for division into high and low expressors were motivated by the maxima of log-rank statistic.
+Statistical significance of differences in overall survival between the high and low expressors was determined by Peto-Peto test adjusted for multiple comparison with the false discovery rate method (FDR).
+Candidate survival markers were identified by pFDR < 0.05 in at least two cohorts._
+
+_Fractions of surviving patients in the high and low expression strata of FGF11 and FGFBP1 genes are presented in Kaplan-Meier plots.
+Stratification cutoffs, and total numbers of patients and deaths are displayed in the plot captions.
+Numbers of patients and deaths in the expression strata are indicated in the plot legends.
+Raw and FDR-adjusted p values are shown in the plots._
+
+\newpage
+
+```{r tab-surv-eval, tab.cap = "Performance of an Elastic Net Cox model using expression of FGFR-, FGF-, and FGFBP-coding genes as explanatory factors at prediction of overall survival.
+The Elastic Net model was trained in the TCGA BLCA cohort with ComBat-processed log2-transformed expression values as firs- and second-order terms.
+Metrics of concordance, explanatory performance, and overall calibration are summarized."}
+
+flextable::flextable(rep_tabs$surv_eval) %>% 
+  set_widths(c(2.5, rep(2.3, 5))) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-surv-imp, tab.cap = "Non-zero coefficients of the Elastic Net Cox model of overall survival with FGFR-, FGF-, and FGFBP-coding genes as explanatory factors."}
+
+flextable::flextable(rep_tabs$surv_imp) %>% 
+  italic(j = 1) %>% 
+  set_widths(rep(2.7, 4)) %>% 
+  footnote(1, 4, 
+           value = as_paragraph('Fold-difference in risk for one standard deviation of log2 gene expression values.'), 
+           part = 'header', 
+           ref_symbols = 'a') %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-surv-km, tab.cap = 'Univariable, model-free analysis of overall survival in urothelial cancer patients stratified by expression of FGFR-, FGF-, and FGFR-coding genes.
+Patients were stratified as low and high expressors of a gene by a cutoff value that corresponds to the maximum log rank statistics.
+Statistical significance of differences in overall survival between low and high expressors was determined by Peto-Peto test corrected for multiple comparisons with the false discovery rate (FDR) method.
+The stratification cutoffs, numbers of high and low expressors, numbers of events in the strata, median survival times with 95% confidence intervals, as well as raw and FDR-corrected p values are summarized.
+The results are shown for the genes with significant effects in at least two out of the TCGA, IMvigor, and BCAN cohorts.'}
+
+flextable::flextable(rep_tabs$surv_km) %>% 
+  italic(j = 1) %>% 
+  merge_v(1:2) %>% 
+  set_widths(c(2, rep(2, 4), 5.5, rep(2.5))) %>% 
+  theme_vanilla
+
+```
 
@@
 
-```{r tab-lca-dge, tab.cap = 'Expression of FGF- and FGFR-coding genes in the genetic subsets of urothelial cancers of the TCGA BLCA cohort.
+```{r tab-lca-dge, tab.cap = 'Expression of FGF-, FGFR-, and FGFBP-coding genes in the genetic subsets of urothelial cancers of the TCGA BLCA cohort.
 log2-transformed expression levels are presented as medians with interquartile ranges and ranges.
@@
 
-```{r fig-lca-dge-fgfr, fig.width = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$h, fig.cap = 'Differential expression of FGFR-coding genes in the genetic subsets of urothelial cancers.'}
+```{r fig-lca-dge-fgfr, fig.width = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgfr, to = 'in')$h, fig.cap = 'Differential expression of FGFR- and FGF-coding genes in the genetic subsets of urothelial cancers.'}
 
@@
 
-__Figure \@ref(fig:fig-lca-dge-fgfr). Differential expression of FGFR-coding genes in the genetic subsets of urothelial cancers.__
-
-_Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic.
-Differences in expression of FGFR1, FGFR2, and FGFR3 were found significant.
+__Figure \@ref(fig:fig-lca-dge-fgfr). Differential expression of FGFR- and FGF-coding genes in the genetic subsets of urothelial cancers.__
+
+_Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic.
+Differences in expression of FGFR1, FGFR2, FGFR3, and FGF5 were found significant.
 Median log2-transformed expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
@@
 
-```{r fig-lca-dge-fgf, fig.width = figur::convert(rep_figs$lca_dge_fgf, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgf, to = 'in')$h, fig.cap = 'Differential expression of FGF-coding genes in the genetic subsets of urothelial cancers.'}
-
-rep_figs$lca_dge_fgf$plot
-
-```
-
-__Figure \@ref(fig:fig-lca-dge-fgf). Differential expression of FGF-coding genes in the genetic subsets of urothelial cancers.__
-
-_Differences in log2-transformed mRNA levels of FGF- and FGFR-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic.
-Differences in expression of FGF2, FGF5, FGF7, and FGF17 were found significant.
+```{r fig-lca-dge-fgfbp, fig.width = figur::convert(rep_figs$lca_dge_fgfbp, to = 'in')$w, fig.height = figur::convert(rep_figs$lca_dge_fgfbp, to = 'in')$h, fig.cap = 'Differential expression of FGFBP-coding genes in the genetic subsets of urothelial cancers.'}
+
+rep_figs$lca_dge_fgfbp$plot
+
+```
+
+__Figure \@ref(fig:fig-lca-dge-fgfbp). Differential expression of FGFBP-coding genes in the genetic subsets of urothelial cancers.__
+
+_Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes were compared between the genetic subsets of urothelial cancers in the `r globals$cohort_labs["tcga"]` cohort by one-way ANOVA with $\eta^2$ effect size statistic.
+Differences in expression of PTX3, SDC1, GPC1, and TNFAIP were found significant.
 Median log2-transformed expression values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
@@
 
+## Model of FGFR signaling in urothelial cancer
+
+```{r fig-summary, fig.width = figur::convert(rep_figs$summary, to = 'in')$w, fig.height = figur::convert(rep_figs$summary, to = 'in')$h, fig.cap = 'Proposed scheme of FGFR signaling in urothelial cancers.'}
+
+rep_figs$summary$plot
+
+```
+
+__Figure \@ref(fig:fig-summary). Proposed scheme of FGFR signaling in urothelial cancers.__
+
+\newpage
+
 # Methods
@@
 
-Four collectives of urothelial carcinoma patients were evaluated in the current report: 
+Five collectives of urothelial carcinoma patients were evaluated in the current report: 
 the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
 the MSK IMPACT collective [@Clinton2022], 
-the bladder cancer TCGA collective referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
-and the IMvigor clinical study cohort [@Balar2017].
+the bladder cancer TCGA collective of muscle-invasive urothelial cancers referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
+the IMvigor cohort of chemotherapy-resistant cancers treated subsequently with immunotherapy [@Balar2017], and 
+the BCAN cohort of metastatic cancer patients administered systemic chemotherapy [@Damrauer2022].
 
@@
 
-Assignment of the TCGA BLCA and IMvigor cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome log_2 mRNA expression levels (R package _consensusMIBC_) [@Kamoun2020].
+The BCAN study information was extracted from [cBioportal](https://www.cbioportal.org/) and consisted of demographic and clinical information, staging of primary tumors, overall and disease-specific survival, RNA sequencing expression Z scores of bulk cancer samples, as well as detailed somatic mutation data of bulk cancer specimens.
+This collective consisted of metastatic cancers, treated in most patients by systemic chemotherapy. 
+The biological material was collected both in urogenital tract and at distant metastatic sites.
+
+For molecular subset assignment and machine learning modeling tasks, $log_2$-transformed gene expression levels were processed by ComBat (R package _sva_ and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)) [@Leek2012] to correct for the inter-cohort variability.
+
+Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome log_2 mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020].
+
+The following genes coding for FGF, FGFR and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a]:
+
+* FGF genes: `r globals$ligands %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`
+
+* FGFR genes: `r globals$receptors %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`
+
+* FGFBP genes: `r globals$binding_proteins %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`
+
+Those genes were returned by a systematic search of literature at [PubMed](https://pubmed.ncbi.nlm.nih.gov/) as well as a screening of physical interaction partners of FGFR and FGF proteins at [UniProt](https://www.uniprot.org/), [STRING](https://string-db.org), [BioGRID](https://thebiogrid.org), and [IntAct](https://www.ebi.ac.uk/intact).
 
@@
 Descriptive statistic for qualitative variables were percentage and count within categories. 
-Differences of means or medians of qualitative variables were investigated by Mann-Whitney, tow-tailed T test, Kruskal-Wallis test, and one-way ANOVA, depending on the number of analysis groups and normality of distribution of the variable.
+Differences of means or medians of qualitative variables were investigated by Mann-Whitney, two-tailed T test, Kruskal-Wallis test, and one-way ANOVA, depending on the number of analysis groups and normality of distribution of the variable.
 Correlation was assessed by Pearson's correlation test. 
@@
 
-* r for differences between two groups, non-parametric tests: small for r < 0.2, medium for 0.2 $\leq$ r < 0.4, large for r $\geq$ 0.4
+* biserial r for differences between two groups, non-parametric tests: small for r < 0.2, medium for 0.2 $\leq$ r < 0.4, large for r $\geq$ 0.4
 
@@
 
-* Cramer's V for comparison of category frequencies of a numeric variable: small for r < 0.2, medium for 0.2 $\leq$ r < 0.4, large for r $\geq$ 0
-
-* Pearson's $\rho$ for correlation, parametric tests: small for $\rho$ < 0.3, medium for 0.3 $\leq$ $\rho$ < 0.5, large for r $\geq$ 0.5
+* Cramer's V for comparison of category frequencies of a numeric variable: small for V < 0.2, medium for 0.2 $\leq$ V < 0.4, large for V $\geq$ 0.4
+
+* Pearson's r for correlation, parametric tests: small for r < 0.3, medium for 0.3 $\leq$ r < 0.5, large for r $\geq$ 0.5
+
+* Spearman's $\rho$ for correlation, non-parametric tests: small for $\rho$ < 0.3, medium for 0.3 $\leq$ $\rho$ < 0.5, large for $\rho$ $\geq$ 0.5
 
@@
 Differences with p < 0.05 were considered statistically significant. 
-Statistical hypothesis testing was accomplished with the development packages [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) and [@Mangiafico2022; @Kassambara2021] internally employing statistical tools of base R, _stats_, _rstatix_, and _rcompanion_ [@Mangiafico2022; @Kassambara2021].
-
-## Survival analysis
-
-Differences in overall, disease-specific, and relapse-free survival between patients stratified by genetic features such as _FGFR3_ mutations of genetic subsets were investigated with Peto-Peto test. 
-Fractions of surviving patients were visualized with Kaplan-Meier plots. 
-Survival analyses and visualizations were done with packages _survival_ and _survminer_ [@Kassambara2016; @Therneau2000].
+Statistical hypothesis testing was accomplished with the development packages [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) which partly employ statistical tools of base R, _stats_, _rstatix_, and _rcompanion_ [@Mangiafico2022; @Kassambara2021].
+
+## Differences in rates of genetic alterations
+
+Differences in rates of genetic alterations between analysis groups of interest were investigated either by $\chi^2$ test with Cramer's V effect size metric (package [_fastTest_](https://github.com/PiotrTymoszuk/fastTest)) or by weighted permutation test (package [_perich_](https://github.com/PiotrTymoszuk/perich)).
+In the later case, the actual frequency of a genetic alteration was compared with the expected frequency obtained in n = 1000 - 10000 permutations weighted by counts of all genetic alterations in the cancer samples.
+This testing procedure yielded an effect size statistic called enrichment score (ES) defined as the ratio of the actual to the expected genetic alteration number in an analysis groups. 
+By iterating over the permutations, 95% confidence intervals of ES and p values for the ES $\leq$ 0 null hypothesis were computed.
+
+## Differential gene expression
+
+Differences in $log_2$-transformed expression levels of FGF-, FGFR-, and FGFBP-coding genes between cancers with and without _FGFR3_ mutations were assessed by Welch's two-tailed T test with Cohen's d effect size statistic (function `f_t_test()`, package [_fastTest_](https://github.com/PiotrTymoszuk/fastTest)).
+The differentially expressed genes were identified by FDR-corrected p < 0.05 and at least 1.25-fold up- or downregulation as compared with WT tumors.
+
+Differences in $log_2$-transformed transcript levels of the genes of interest between consensus molecular classes or genetic subsets of urothelial cancers were explored with a two step procedure. 
+In the first step, differences of expression levels between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
+In the second step, differences between an analysis class and the baseline (luminal papillary cancers for the consensus classes, del9p21 subset for the genetic subsets) were evaluated by Welch's two-tailed T test with Cohen's d effect size.
+The significantly differentially regulated genes were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.06, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation as compared with the baseline.
+
+In the differential expression analyses, functions `f_one_anova()` and `f_t_test()` from [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) package were used.
 
@@
 
+## Gene-co-regulation networks
+
+Co-expression networks of the FGF-, FGFR-, and FGFBP-coding genes were constructed in form of weighted graphs for pairwise correlations of the expression levels with Spearman's $\rho \geq$ 0.3 (function `as_iGraph()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) with wrapper tools for _igraph_) [@Csardi2006].
+Isolated vertices were removed with `prune_degree()` function. 
+The following vertex importance statistics [@Csardi2006] were computed with `summary()` method called for the network objects: 
+degree (number of directly connected neighbor nodes), 
+hub score (eigenvector of the affinity matrix, i.e. an overall strength of correlations), and 
+betweenness (number of the shortest paths between all pairs of vertices that pass through the vertex). 
+The networks were visualized with Fruchterman - Reingold algorithm by calling `plot()` method [@Briatte2021].
+
+## Machine learning modeling of consensus molecular subsets of urothelial cancers with expression of FGFR-, FGF-, and FGFBP-coding genes as explanatory factors
+
+To check if and how differences in expression of the FGFR-, FGF-, and FGFBP-coding genes of interest emulate the consensus molecular classification proposed by Kamoun and colleagues [@Kamoun2020], we resorted to machine learning modeling with two independent algorithms, multinomial regularized Elastic Net regression [@Friedman2010; @Zou2005] and Random Forest tree ensemble [@Wright2017; @Breiman2001].
+
+The models were developed in a data set derived from the TCGA BLCA collective; the consensus molecular class assignment by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response.
+ComBat-processed $log_2$ expression levels of the genes of interest were used as the sole explanatory factors. 
+In order to improve recognition of the minority consensus classes, the LumU (luminal unstable) and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016].
+The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010].
+The optimal values of the regularization parameter $\lambda$ = `r ml_train$tune_obj$elnet$best_tune$lambda %>% signif(3)` was found by minimizing the Elastic Net model's deviance in 200 repeats 10-fold cross-validation.
+The Random Forest model of consensus molecular subsets was tuned and trained inn the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
+The optimal set of parameters that control behavior of the Random Forest model, mtry = `r ml_train$tune_obj$ranger$best_tune$mtry` (number of randomly chosen variables per tree learner), minimal node size = `r ml_train$tune_obj$ranger$best_tune$min.node.size`, and `r ml_train$tune_obj$ranger$best_tune$splitrule %>% stri_capitalize_first` splitting rule was found by minimizing the out-of-bag classification error [@Breiman2001].
+The final Random Forest model was an ensemble of 500 tree learners.
+
+Performance of the machine learning models was evaluated by comparing the predicted and actual consensus molecular class in the unmodified TCGA BLCA data set, as well as in the IMvigor and BCAN independent validation cohorts.
+The consensus classification returned by the _consensusMIBC_ R package nearest centroid tool was assumed the ground truth. 
+The following numeric statistics of model performance were computed: 
+overall accuracy (fraction of cancer samples with correct consensus molecular class prediction), 
+Cohen's $\kappa$ (non-accidental concordance between the predicted and actual class, high $\kappa$ expected for a highly accurate model) [@McHugh2012], 
+Brier score (metric of overall model calibration, low Brier scores expected for a well calibrated model) [@Brier1950; @Goldstein-Greenwood2021], 
+and receiver-operating characteristic measures such as area under the curve (AUC, measure of overall sensitivity and specificity). 
+Those evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random.
+Additionally, accuracy of prediction for single consensus molecular classes was assessed by inspection of the confusion matrices.
+
+Calibration of the machine learning models at predictions of particular consensus molecular classes was assessed with two approaches. 
+First, we compared squares of differences between the predicted class assignment probabilities and the binarized actual class assignment (1: the class of interest, 0: any other consensus molecular class) or 'Brier squares' between the consensus molecular classes.
+Second, we modeled and visualized logistic relationship between the mean predicted class assignment probability and the actual fraction of observations in the class with tools from _CalibrationCurves_ R package [@Campo2023].
+
+For the Elastic Net model, absolute $\beta$ values of the model coefficients serve as direct variable importance metrics; they were extracted from the model with `coef()` method. 
+For the Random Forest model, permutation variable importance proposed by Breiman [@Breiman2001] was computed as the difference in out-of-bag classification error of the genuine model and a model with the particular explanatory variable re-shuffled at random (function `importance()`, package _ranger_).
+
+## Model-free survival analysis
+
+Differences in overall, disease-specific, and relapse-free survival between patients stratified by genetic features such as _FGFR3_ mutations, genetic subsets, or gene expression cutoffs determined by the maxima of log-rank statistic (function `surv_cutpoint()`, package _survminer_) were investigated with Peto-Peto test.
+Fractions of surviving patients were visualized with Kaplan-Meier plots. 
+Survival analyses and visualizations were done with packages _survival_ and _survminer_ [@Kassambara2016; @Therneau2000].
+
+## Modeling of overall survival as a function of expression of FGFR-, FGF-, and FGFBP-coding genes
+
+Effects of differential expression of the FGFR-, FGF-, and FGFBP-coding genes on overall survival was modeled in the TCGA BLCA, IMvigor, and BCAN cohort by Elastic Net Cox regression [@Simon2011; @Zou2005].
+The Elastic Net Cox model was trained in the TCGA BLCA collective with ComBat-processed $log_2$ expression levels of the genes of interest. 
+The optimal value of the regularization parameter $\lambda$ = `r signif(surv_train$tune_obj$best_tune$lambda, 3)` was found by minimizing the model deviance in 200-repeats 10-fold cross-validation (a wrapper around `cv.glmnet()` and `glmnet()` function, package _glmnet_) [@Friedman2010].
+The Elastic Net Cox model was subsequently used for computation of linear predictor scores in the training cohort, and the IMvigor and BCAN validation collectives (method `predict()`, _glmnet_ package).
+For evaluation of the predictive performance, overall survival was modeled as a function of the linear predictor scores by canonical Cox regression [@Simon2011a; @Sonnweber2023]. 
+Overall performance of those Cox models was metered by Harrell's C index (high values: good concordance between the predicted and observed survival), integrated Brier score (IBS; low values: good model calibration), and R^2^ as a measure of survival variance explained by the model (method `summary()`) [@Harrell1996; @Royston2006; @Graf1999].
+Additionally, we compared overall survival of cancer patients stratified by tertiles of the linear predictor scores by Peto-Peto test and visualized with Kaplan-Meier plots. 
+The linear predictor Cox models were constructed, checked for the proportional hazard assumption, and evaluated with tools provided by R packages _survival_ [@Grambsch1994; @Therneau2000] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
+To obtain variable importance measures, non-zero coefficients of the Elastic Net Cox model were extracted and expressed as hazard ratios (method `coef()`).
+
+## Modeling of survival in genetic subsets of urothelial cancer
+
+For each of the GENIE BLCA and MSK IMPACT cohorts, two types of Cox proportional hazard models of overall survival were constructed: (1) a model with genetic subset assignment as a sole explanatory factor, and (2) a model with genetic subset assignment and clinically relevant confounders (GENIE BLCA: age and gender, MSK IMPACT: age, gender, and pathological tumor stage).
+Those Cox models were constructed, checked for the proportional hazard assumption and evaluated with tools provided by R packages _survival_ [@Grambsch1994; @Therneau2000] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions) essentially as outlined above for the Elastic Net Cox regression.
+Significance of the genetic subset term and its independence from clinical confounders was determined by likelihood ratio test between the model with subset assignment and the model with cluster assignment and clinical confounders as explanatory factors (function `anova()`).
+Estimates of model coefficients were retrieved with `summary()` method.
+
 ## Genetic subsets of urothelial cancers
@@
 
-Genetic subsets of urothelial cancers were developed for the GENIE BLCA training cohort with latent class modeling [@Weller2020; @Linzer2011]. 
-The explanatory variables were binary indexes (presence/absence) of genetic alterations present in at least 5% of samples in all data sets. 
+Genetic subsets of urothelial cancers were developed for the GENIE BLCA training cohort by latent class modeling [@Weller2020; @Linzer2011]. 
+The explanatory variables were binary indexes (presence/absence) of genetic alterations present in at least 5% of samples in all collectives. 
 Presence of FGF3, FGF4, FGF19, or CCND1 amplification was subsumed under 11q13 amplification prior to the modeling.
@@
 Selection of the optimal subset number was motivated by the minimum of Bayesian Information Criterion (BIC). 
-With this selection criterion, a six-subset model whose classes were named after characteristic gene and chromosome alterations was chosen for definition of the genetic subsets.
-
-Subsequently, assignment of cancer samples from the MSK IMPACT and TCGA BLCA test collectives to the genetic subsets was predicted based on conditional subset assignment probabilities using `predict()` method from [_polcaExtra_](https://github.com/PiotrTymoszuk/polcaExtra) package.
-Quality of the subset assignment in the training and test collective was assessed by a visual inspection of the class assignment probabilities (posterior p) and by comparison of frequency of the subset-defining genetic features between the subsets ($\chi^2$ test with Cramer's V effect size statistic).
-
-## Surviival modeling
-
-For each of the GENIE BLCA and MSK IMPACT cohorts, two types of Cox proportional hazard models of overall survival were constructed: (1) a model with genetic subset assignment as a sole explanatory factor, and (2) a model with genetic subset assignment and clinically relevant confounders (GENIE BLCA: age and gender, MSK IMPACT: age, gender, and pathological tumor stage).
-The Cox models were constructed, checked for the proportional hazard assumption and evaluated with tools provided by R packages _survival_ [@Grambsch1994; @Therneau2000] and [_coxExtensions_](https://github.com/PiotrTymoszuk/coxExtensions).
-Overall Cox model performance was evaluated by Harrell's C index (high values: good concordance between the predicted and observed survival), integrated Brier score (IBS; low values: good model calibration), and R^2 as a measure of survival variance explained by the model.
-Significance of the genetic subset term and its independence from clinical confounders was determined by likelihood ratio test between the model with subset assignment and the model with cluster assignment and clinical confounders as explanatory factors (function `anova()`).
-Estimates of model coefficients were retrieved with `summary()` method.
+With this selection criterion, a six-subset model was chosen for definition of the genetic subsets. 
+The genetic subsets were were named after their characteristic gene and chromosome alterations.
+
+Subsequently, assignment of cancer samples in the MSK IMPACT and TCGA BLCA test collectives to the genetic subsets was predicted by conditional subset assignment probabilities using `predict()` method from [_polcaExtra_](https://github.com/PiotrTymoszuk/polcaExtra) package.
+Quality of the subset assignment in the training and test collective was assessed by a visual inspection of the class assignment probabilities (posterior p) and by comparison of frequency of the subset-defining genetic genetic features between the subsets ($\chi^2$ test with Cramer's V effect size statistic).
 

### 2024-12-07 FGFR3 analysis scripts/overall_survival.R
Analysis of HPA data
Verification of protein expression by IHC data from the Human Protein Atlas (HPA)

@@ -104,7 +104,7 @@
 
   fgfr_os$km_plots <-
     list(x = fgfr_os$km_plots,
-         y = c(globals$cohort_labs,
+         y = c(globals$cohort_labs[c("genie", "msk", "tcga", "imvigor", "bcan")],
                paste(globals$cohort_labs[c("msk", "tcga", "bcan")],
                      'without pT4 cancers')),
          w = fgfr_os$n_captions,

### 2024-12-07 exploration scripts/cohorts.R

@@ -14,6 +14,7 @@
 
   expl_cohorts$data <- globals$cohort_expr %>%
     eval %>%
+    c(list(hpa = hpa)) %>%
     map(~.x$clinic)
 
   expl_cohorts$data$bcan <- expl_cohorts$data$bcan %>%

### 2024-12-07 exploration scripts/dge.R

@@ -216,10 +216,11 @@
            cust_theme = globals$common_theme +
              theme(plot.title = element_markdown(),
                    axis.title.x = element_markdown(),
+                   axis.title.y = element_markdown(),
                    legend.position = 'none'),
            x_n_labs = TRUE,
            x_lab = html_italic('FGFR3'),
-           y_lab = expression('log'[2] * ' expression')) %>%
+           y_lab = globals$cohort_axis_labs[[i]]) %>%
       map(~.x +
             scale_fill_manual(values = globals$mut_status_color)) %>%
       set_names(expl_dge$test$FGFR3_mut[[i]]$variable)

### 2024-12-07 exploration scripts/expression.R

@@ -1,0 +1,210 @@
+# Expression of FGFR, FGF, and FGFBP genes and proteins.
+# The later are investigated by IHC in the Human Protein Atlas.
+
+  insert_head()
+
+# container ------
+
+  expl_expr <- list()
+
+# analysis globals ------
+
+  insert_msg('Analysis globals')
+
+  ## variables of interest and the corresponding expression levels
+
+  expl_expr$lexicon <-
+    globals[c("receptors", "ligands", "binding_proteins")] %>%
+    map(~tibble(variable = .x)) %>%
+    compress(names_to = 'protein_type') %>%
+    mutate(protein_type = factor(protein_type,
+                                 c('receptors', 'ligands', 'binding_proteins')))
+
+  expl_expr$variables <- expl_expr$lexicon$variable
+
+  expl_expr$data <- globals$cohort_expr %>%
+    eval %>%
+    map(~.x$expression) %>%
+    c(list(hpa = hpa$ihc_score)) %>%
+    compact
+
+  expl_expr$data <- expl_expr$data %>%
+    map(select, any_of(expl_expr$variables))
+
+  expl_expr$variables <- expl_expr$data %>%
+    map(names)
+
+  ## expression units
+
+  expl_expr$txt_units <-
+    c(tcga = 'log\u2082 gene count',
+      imvigor = 'log\u2082 gene count',
+      bcan = 'Z-score',
+      hpa = 'IHC score')
+
+  ## plotting data in a long format:
+  ## for box plot panels and a heat map of Z-scores
+
+  expl_expr$plot_data <-
+    map2(expl_expr$data,
+         expl_expr$variables,
+         ~pivot_longer(.x,
+                       cols = all_of(.y),
+                       names_to = 'variable',
+                       values_to = 'exp_level'))
+
+  expl_expr$plot_data <- expl_expr$plot_data %>%
+    map(left_join,
+        expl_expr$lexicon,
+        by = 'variable') %>%
+    map(mutate,
+        z_score = zScores(exp_level))
+
+# Descriptive stats --------
+
+  insert_msg('Descriptive stats')
+
+  expl_expr$stats <- expl_expr$data %>%
+    map2_dfr(., names(.),
+             ~mutate(.x,
+                     cohort = factor(.y, names(expl_expr$data)))) %>%
+    fast_num_stats(variables = expl_expr$lexicon$variable,
+                   split_fct = 'cohort') %>%
[…]

### 2024-12-07 exploration scripts/ihc_representative.R

@@ -1,0 +1,55 @@
+# Representative IHC staining images.
+# We're showing for each of them a sample with the
+# lowest expression, median, and 75% percentile.
+
+  insert_head()
+
+# container -------
+
+  expl_ihc <- list()
+
+# analysis globals ---------
+
+  insert_msg('Genes of interest and their image paths')
+
+  expl_ihc$genes <- hpa$data$gene_symbol %>%
+    unique
+
+  expl_ihc$lexicon <- hpa$data %>%
+    filter(gene_symbol %in% expl_ihc$genes) %>%
+    mutate(gene_symbol = factor(gene_symbol, expl_ihc$genes),
+           caption = paste0(sex, ', ', age, 'y\n',
+                            grade, ' grade'),
+           caption_short = paste0(sex, ', ', age, 'y')) %>%
+    select(patient_id, gene_symbol, ihc_score,
+           image_path, caption, caption_short)
+
+  ## selection of the samples: work on that!!!
+
+  set.seed(123)
+
+  expl_ihc$lexicon <- expl_ihc$lexicon %>%
+    blast(gene_symbol)
+
+  expl_ihc$representative_idx <- expl_ihc$lexicon %>%
+    map(~.x$ihc_score) %>%
+    map(choose_representative,
+        probs = c(0, 0.6, 1))
+
+  expl_ihc$lexicon <-
+    map2(expl_ihc$lexicon,
+         expl_ihc$representative_idx,
+         ~.x[.y, ])
+
+# staining panels saved as calls to spare memory -------
+
+  insert_msg('Staining panels')
+
+  expl_ihc$panels <- expl_ihc$lexicon %>%
+    map(~call2(.fn = 'stitch_hpa_images',
+               lexicon = .x,
+               y_pos = c(0.12, 0.88)))
+
+# END -------
+
+  insert_tail()

### 2024-12-07 exploration.R

@@ -7,7 +7,8 @@
 # 3) Differences in survival between the cohorts.
 #
 # 4) General frequency of mutations and copy number alterations. Details for
-# mutations and CNA of the FGFR genes.
+# mutations and CNA of the FGFR genes. Expression levels of the genes and
+# proteins of interest
 #
 # 5) Characteristic of the mutation types and location for the FGFR genes.
 #
@@ -36,6 +37,7 @@
   library(graphExtra)
 
   library(jsonlite)
+  library(cowplot)
 
   library(microViz)
 
@@ -76,10 +78,13 @@
     source_all(message = TRUE, crash = TRUE) %>%
     print
 
+  ## overall expression levels of the features of interest: RNA and IHC
   ## co-expression and differential gene expression
   ## co-occurrence of alterations in the FGF and FGFR genes
 
-  c('./exploration scripts/coexpression.R',
+  c('./exploration scripts/expression.R',
+    './exploration scripts/ihc_representative.R',
+    './exploration scripts/coexpression.R',
     './exploration scripts/dge.R') %>%
     source_all(message = TRUE)
 

### 2024-12-07 import scripts/hpa.R

@@ -1,0 +1,149 @@
+# Processing of XML entries with patient's and sample staining information for
+# the FGFR, FGF, and FGFBP proteins. Resorting to locally saved XML files.
+
+  insert_head()
+
+# container --------
+
+  hpa <- list()
+
+# available XML files --------
+
+  insert_msg('Available XML files and gene lexicon')
+
+  ## genes and the expected file paths
+
+  hpa$annotation <-
+    tibble(gene_symbol = reduce(globals[c("receptors",
+                                          "ligands",
+                                          "binding_proteins")],
+                                union)) %>%
+    mutate(ensembl_id = mapIds(org.Hs.eg.db,
+                               keys = gene_symbol,
+                               keytype = 'SYMBOL',
+                               column = 'ENSEMBL'),
+           entrez_id =  mapIds(org.Hs.eg.db,
+                               keys = ensembl_id,
+                               keytype = 'ENSEMBL',
+                               column = 'ENTREZID')) %>%
+    filter(complete.cases(.))
+
+  hpa$annotation <- hpa$annotation %>%
+    mutate(file = paste0('./data/HPA/', ensembl_id, '.xml'),
+           exists = file.exists(file))
+
+  if(sum(hpa$annotation$exists) < 10) {
+
+    insert_msg('Too few files found, accessing HPA')
+
+    source_all('./import scripts/hpa_api_access.R',
+               message = TRUE, crash = TRUE)
+
+  }
+
+  hpa$annotation <- hpa$annotation %>%
+    filter(exists)
+
+# Extraction of the patient and staining data ------
+
+  insert_msg('Extraction')
+
+  hpa$raw_data <- hpa$annotation$file %>%
+    map(read_xml) %>%
+    set_names(hpa$annotation$gene_symbol)
+
+  hpa$raw_data <- hpa$raw_data %>%
+    future_map(safely(process_hpa_entry),
+               .options = furrr_options(seed = TRUE)) %>%
+    map(~.x$result) %>%
+    compact
+
+# Formatting ---------
+
+  insert_msg('Formatting')
+
+  hpa$data <- hpa$raw_data %>%
+    map(mutate,
+        sex = factor(tolower(sex), c('female', 'male')),
+        age = as.numeric(age),
+        staining = factor(tolower(staining),
+                          c('not detected', 'low', 'medium', 'high')),
+        intensity = factor(tolower(intensity),
[…]

### 2024-12-07 import scripts/hpa_api_access.R

@@ -1,0 +1,40 @@
+# Access to XML files with patient and staining information in the Human Protein
+# Atlas repository
+
+  insert_head()
+
+# container -------
+
+  hpa_access <- list()
+
+# genes of interest -------
+
+  insert_msg('Genes of interest')
+
+  hpa_access$lexicon <-
+    tibble(gene_symbol = reduce(globals[c("receptors",
+                                          "ligands",
+                                          "binding_proteins")],
+                                union)) %>%
+    mutate(ensembl_id = mapIds(org.Hs.eg.db,
+                               keys = gene_symbol,
+                               keytype = 'SYMBOL',
+                               column = 'ENSEMBL')) %>%
+    filter(complete.cases(.))
+
+  ## specifying the URLs and file names
+
+  hpa_access$lexicon <- hpa_access$lexicon %>%
+    mutate(url = paste0('https://www.proteinatlas.org/', ensembl_id, '.xml'),
+           file = paste0('./data/HPA/', ensembl_id, '.xml'))
+
+# Downloads ---------
+
+  insert_msg('Downloads')
+
+  hpa_access$lexicon[, c("url", "file")] %>%
+    pwalk(download_hpa)
+
+# END ------
+
+  insert_tail()

### 2024-12-07 import scripts/hpa_images.R

@@ -1,0 +1,15 @@
+# Fetches images with IHC staining from the Human Protein Atlas
+
+  insert_head()
+
+# downloading -------
+
+  insert_msg('Downloading')
+
+  hpa$data[, c("image_url", "image_path")] %>%
+    set_names(c('url', 'file')) %>%
+    pwalk(download_image)
+
+# END -------
+
+  insert_tail()

### 2024-12-07 import.R

@@ -21,6 +21,7 @@
 
   library(IMvigor210CoreBiologies)
   library(cbioportalR)
+  library(xml2)
 
   library(AnnotationDbi)
   library(org.Hs.eg.db)
@@ -28,6 +29,8 @@
   library(htGLMNET)
   library(consensusMIBC)
   library(impute)
+
+  library(furrr)
 
   library(soucer)
 
@@ -73,6 +76,14 @@
 
   c('./import scripts/subtypes.R') %>%
     source_all(message = TRUE, crash = TRUE)
+
+# Import of IHC scores for Human Protein Atlas -------
+
+  insert_msg('Import of HPA data')
+
+  access_cache(cache_path = './data/hpa.RData',
+               script_path = './import scripts/hpa.R',
+               message = paste('Cached Human Protein Atlas data'))
 
 # END ------
 

### 2024-12-07 latent class scripts/dge.R

@@ -81,9 +81,10 @@
          cust_theme = globals$common_theme,
          x_n_labs = TRUE,
          x_lab = 'Genetic cluster',
-         y_lab = expression('log'[2] * ' expression')) %>%
+         y_lab = globals$cohort_axis_labs[[i]]) %>%
     map(~.x +
-          theme(plot.title = element_markdown()) +
+          theme(plot.title = element_markdown(),
+                axis.title.y = element_markdown()) +
           scale_fill_manual(values = globals$genet_colors)) %>%
     set_names(lca_dge$test$variable)
 

### 2024-12-07 report scripts/figures.R

@@ -306,6 +306,56 @@
     pmap(as_figure,
          w = 180)
 
+# Expression of FGF, FGFR, and FGFBP genes ---------
+
+  insert_msg('Expression of genes and proteins')
+
+  rep_figs$expression <-
+    plot_grid(expl_expr$hm_plot +
+                labs(title = 'RNA expression') +
+                theme(legend.position = 'none'),
+              plot_grid(expl_expr$plots$hpa +
+                          labs(title = 'IHC, HPA') +
+                          theme(plot.subtitle = element_blank(),
+                                legend.position = 'none'),
+                        get_legend(expl_expr$hm_plot),
+                        nrow = 2,
+                        rel_heights = c(2, 1)),
+              ncol = 2,
+              rel_widths = c(1, 1.3),
+              labels = LETTERS,
+              label_size = 10) %>%
+    as_figure(label = 'expression_fgfr_fgf_fgfbp',
+              ref_name = 'expression',
+              caption = paste('Expression of genes coding FGFR, FGF, and',
+                              'FGFBP at mRNA and protein level.'),
+              w = 180,
+              h = 210)
+
+# Representative images, IHC --------
+
+  insert_msg('Representative images')
+
+  rep_figs[c('ihc_receptors', 'ihc_ligands', 'ihc_fgfbp')] <-
+    list(expl_ihc$panels[c("FGFR1", "FGFR2", "FGFR3", "FGFR4")],
+         expl_ihc$panels[c("FGF7", "FGF10")],
+         expl_ihc$panels[c("SDC2", "GPC4", "CD44", "FGFBP1")]) %>%
+    map(map, eval) %>%
+    map(~plot_grid(plotlist = .,
+                   ncol = 1))
+  rep_figs[c('ihc_receptors', 'ihc_ligands', 'ihc_fgfbp')] <-
+    rep_figs[c('ihc_receptors', 'ihc_ligands', 'ihc_fgfbp')] %>%
+    list(x = .,
+         label = names(.),
+         ref_name = names(.),
+         caption = paste('Representative immunohistochemistry staining images',
+                         'for',
+                         c('FGFR proteins', 'FGF proteins', 'FGFBP proteins'),
+                         'retrieved from the Human Protein Atlas.'),
+         h = c(160, 80, 160)) %>%
+    pmap(as_figure,
+         w = 120)
+
 # Correlation of FGF, FGFR, FGFBP gene expression --------
 
[…]

### 2024-12-07 report scripts/tables.R

@@ -37,7 +37,7 @@
     set_names(c('Variable',
                 globals$cohort_labs[c("genie", "msk",
                                       "tcga", "imvigor",
-                                      "bcan")],
+                                      "bcan", "hpa")],
                 'Significance',
                 'Effect size')) %>%
     mdtable(label = 'cohort_characteristic',
@@ -119,6 +119,32 @@
                             'The table is available as a supplementary',
                             'Excel file.'))
 
+# Expression of genes and proteins -------
+
+  insert_msg('Expression, genes and proteins')
+
+  rep_tabs$expression <- expl_expr$stats %>%
+    set_names(c('Variable', globals$cohort_labs[names(expl_expr$stats)[-1]])) %>%
+    as_mdtable(label = 'expression',
+               ref_name = 'expression',
+               caption = paste('Expression of FGFR-, FGF-, and FGFBP-coding',
+                               'genes at mRNA level was quantified in the',
+                               'TCGA BLCA, IMvigor, and BCAN cohort by',
+                               'RNA sequencing. The mRNA amounts were',
+                               'provided as log2-transformed gene counts',
+                               '(TCGA BLCA and IMvigor) or Z-scores of log2',
+                               'gene counts (BCAN).',
+                               'Protein expression of the genes of interest was',
+                               'measured in the Human Protein Atlas (HPA) by',
+                               'immunohistochemistry (IHC).',
+                               'The protein amounts are metered by the IHC score',
+                               'defined as a product of staining',
+                               'intensity and quantity of positive cells.',
+                               'Median mRNA and protein amounts with',
+                               'interquartile ranges, ranges, and numbers of',
+                               'cancer samples are listed.',
+                               'The table is available as a supplementary Excel',
+                               'file.'))
 
 # Differential gene expression, FGFR3 WT and FGFR3 mutated -------
 

### 2024-12-07 subtype scripts/dge.R

@@ -195,9 +195,10 @@
            cust_theme = globals$common_theme,
            x_n_labs = TRUE,
            x_lab = 'MIBC consensus class',
-           y_lab = expression('log'[2] * ' expression')) %>%
+           y_lab = globals$cohort_axis_labs[[i]]) %>%
       map(~.x +
-            theme(plot.title = element_markdown()) +
+            theme(plot.title = element_markdown(),
+                  axis.title.y = element_markdown()) +
             scale_fill_manual(values = globals$sub_colors)) %>%
       set_names(sub_dge$anova[[i]]$variable)
 

### 2024-12-07 tools/functions.R

@@ -23,6 +23,9 @@
   library(ggrepel)
   library(ggtext)
 
+  library(xml2)
+  library(curl)
+
   set_rownames <- trafo::set_rownames
 
 # data import ------
@@ -39,6 +42,145 @@
       mutate(gene_symbol = unlist(gene_symbol)) %>%
       filter(gene_symbol != '') %>%
       filter(complete.cases(.))
+
+  }
+
+# import of the HPA data -------
+
+  download_hpa <- function(url, file, ...) {
+
+    ## manages downloads from HPA
+
+    idx <- try(download_xml(url = url, file = file, ...),
+               silent = TRUE)
+
+    if(inherits(idx, 'try-error')) message(paste('Failed for:', url))
+
+  }
+
+  extract_patients <- function(xml_obj) {
+
+    ## extracts the nodes with patient/sample staining information
+
+    as_list(xml_find_all(xml_obj, './/patient'))
+
+  }
+
+  cancer_search <- function(x, regex = '(U|u)rothe') {
+
+    ## extracts elements of a branched list with HPA
+    ## patient's data, that are matched by a regular expression
+    ## applied to the sample/snomedParameters node
+
+    snomeds <- x %>%
+      map(~.x$sample$snomedParameters) %>%
+      map(map, attr, 'tissueDescription') %>%
+      map(unlist)
+
+    lgl_vec <- snomeds %>%
+      map(stri_detect, regex = regex) %>%
+      map_lgl(any)
+
+    x[lgl_vec]
+
+  }
+
+  cancer_format <- function(x) {
+
+    ## coerces the branched list with patient's information and IHC
+    ## staining scores into a handy table
+
+    ## removal of patients with incomplete information,
+    ## unique names for the list elements
+
+    required_names <-
+      c('sex', 'age', 'patientId', 'level', 'quantity', 'location', 'sample')
+
+    complete_patients <- x %>%
+      map(names) %>%
+      map(~required_names %in% .x) %>%
+      map_lgl(all)
+
+    x <- x[complete_patients]
+
+    ## stripping the most essential information
+
+    x <- x %>%
+      map(~.x[1:8]) %>%
+      map(set_names,
+          c('sex', 'age', 'patient_id',
+            'staining', 'intensity', 'quantity', 'location',
+            'sample'))
+
+    patient_tbl <- x %>%
+      map_dfr(as_tibble)
+
+    ## plain data frame
+    ## for each patient, we just need an entry
+    ## with the snomed pathology description.
[…]

### 2024-12-07 tools/globals.R

@@ -57,7 +57,8 @@
                            msk = 'MSK IMPACT',
                            tcga = 'TCGA BLCA',
                            imvigor = 'IMvigor',
-                           bcan = 'BCAN')
+                           bcan = 'BCAN',
+                           hpa = 'HPA')
 
   globals$cohort_colors <- c(genie = 'indianred3',
                              msk = 'darkolivegreen4',
@@ -77,13 +78,23 @@
   ## this expression includes only the studies passing the gene number
   ## criterion
 
-  globals$analysis_cohorts <- names(globals$cohort_labs)
+  globals$analysis_cohorts <-
+    globals$cohort_labs[names(globals$cohort_labs) != 'hpa'] %>%
+    names
 
   globals$cohort_expr <- globals$analysis_cohorts %>%
     map_chr(~paste(.x, .x, sep = ' = ')) %>%
     paste(collapse = ', ') %>%
     paste0('list(', ., ')') %>%
     parse_expr
+
+  ## common units
+
+  globals$cohort_axis_labs <-
+    c(tcga = 'log<sub>2</sub> gene count',
+      imvigor = 'log<sub>2</sub> gene count',
+      bcan = 'Z-score',
+      hpa = 'IHC score')
 
 # Genes of interest -------
 

### 2024-12-07 report/markdown/report.Rmd
HPA data in the report and figures

@@
 
-* Five collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (`r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (`r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers [@Damrauer2022].
-Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BLCA cohorts.
+* Five genetic and transcrional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022].
+Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts.
 Binarized indexes of presence/absence of somatic mutations in selected genes were available for the IMvigor cohort.
@@
 Approximate overall survival was calculated for the GENIE BLCA project participants based on age at enrollment in years and follow-up time in days.
+Validation of expression of the genes of interest was done with immunohistochemistry data retrieved from the [Human Protein Atlas](https://www.proteinatlas.org/).
 Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.
@@
 __ - __Figure \@ref(fig:fig-fgfr-cont-hm)__).
+
+* At mRNA level, _FGFR3_, _TGFBR3_, and _FGFR1_ were the strongest expressed receptor genes.
+_FGF7_, _FGF11_, followed by _FGF2_ were the most abundant ligand transcripts; _HSPG2_, _SDC1_, and _CD44_ were the binding protein genes with the highest mRNA counts in the urothelial cancer tissue.
+Out of `r globals[c("receptors", "ligands", "binding_proteins")] %>% reduce(union) %>% length` genes of interest, `r ncol(hpa$ihc_score) - 1` proteins were quantified by immunohistochemistry (IHC) in the Human Protein Atlas (HPA).
+The protein quantities were expressed as IHC scores defined as products of staining intensity (1: negative to 4: strong) and quantity of positive cells (1: none to 4: >75%).
+In the HPA data set, `r `r sum(colMedians(hpa$ihc_score[-1]) > 1)` FGFR, FGF, and FGFBP proteins had median IHC scores $\geq$ 2, and were hence detectable in cancer specimens.
+At the protein/IHC level, FGFR1, FGFR3, FGFRL1, and FGFR4 were the most abundant receptors.
+FGF7, FGF10, FGF3, FGF19, and FGF17 were the strongest expressed ligand proteins.
+SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).
 
@@
 flextable::flextable(rep_tabs$cohorts) %>% 
-    set_widths(c(2.5, rep(2.7, 5), 2.3, 2.2)) %>% 
-  footnote(1, 7:8, 
+  set_widths(c(2.5, rep(2.3, 6), 2, 2)) %>% 
+  fontsize(j = 1:9, size = 9, part = 'all') %>% 
+  footnote(1, 8:9, 
            value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
@@
 
-## Co-regulation of genes coding for FGFR, FGF, and FGFBP proteins
+## Expression and co-regulation of genes coding for FGFR, FGF, and FGFBP proteins
+
+```{r fig-expression, fig.width = figur::convert(rep_figs$expression, to = 'in')$w, fig.height = figur::convert(rep_figs$expression, to = 'in')$h, fig.cap = 'Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.'}
+
+rep_figs$expression$plot
+
+```
+
+__Figure \@ref(fig:fig-expression). Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.__
+
+_(A) Transcript levels of FGFR-, FGF-, and FGFBP-coding genes were determined in cancer tissue samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` by RNA sequencing and published as $log_2$ gene counts (`r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`) or Z-scores of $log_2$ gene counts (`r globals$cohort_labs["bcan"]`).
+Median Z-scores of $log_2$ mRNA expression are shown in a heat map.
+Each tile represents median expression Z-score of a gene in one cohort.
+Numbers of cancer samples are indicated in the X axis.
+White tiles represent genes not detected in a given cohort._
+
+_(B) Protein expression of FGFR-, FGF-, and FGFBP-coding genes `r ncol(hpa$intensity) - 1` FGFR-, FGF-, and FGFBP-coding genes in the Human Protein Atlas (HPA) repository was quantified by immunohistochemistry (IHC).
+The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%).
+Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single IHC specimens are visualized as points.
+Numbers of IHC samples are indicated in the Y axis._
+
+\newpage
+
+```{r fig-ihc-receptors, fig.width = figur::convert(rep_figs$ihc_receptors, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_receptors, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.'}
+
+list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
+           pattern = 'ihc_receptor') %>% 
+  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
+  include_graphics(rel_path = FALSE)
+
+```
+
+__Figure \@ref(fig:fig-ihc-receptors). Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.__
+
+_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA).
+The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
+_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
+_The images are labeled with HPA patient identifiers (ID) and IHC scores._
+
+\newpage
+
+```{r fig-ihc-ligands, fig.width = figur::convert(rep_figs$ihc_ligands, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_ligands, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.'}
+
+list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
+           pattern = 'ihc_ligand') %>% 
+  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
+  include_graphics(rel_path = FALSE)
+
+```
+
+__Figure \@ref(fig:fig-ihc-ligands). Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.__
+
+_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA).
+The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
+_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
+_The images are labeled with HPA patient identifiers (ID) and IHC scores._
+
+\newpage
+
+```{r fig-ihc-fgfbp, fig.width = figur::convert(rep_figs$ihc_fgfbp, to = 'in')$w, fig.height = figur::convert(rep_figs$ihc_fgfbp, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.'}
+
+list.files('C:/Users/piotr/Desktop/BLCA FGFR/report/figures', 
+           pattern = 'ihc_fgfbp') %>% 
+  paste0('C:/Users/piotr/Desktop/BLCA FGFR/report/figures/', .) %>% 
+  include_graphics(rel_path = FALSE)
+
+```
+
+__Figure \@ref(fig:fig-ihc-fgfbp). Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.__
+
+_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA).
+The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%)._
+_For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score._
+_The images are labeled with HPA patient identifiers (ID) and IHC scores._
+
+\newpage
+
+```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing.
+The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN).
+Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC).
+The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells.
+Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed.
+The table is available as a supplementary Excel file."}
+
+rep_tabs$expression[1, 1] %>% 
+  mutate(Variable = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(18)
+
+```
+
+\newpage
 
@@
 
-Five collectives of urothelial carcinoma patients were evaluated in the current report: 
+Five genomic and transcriptomic collectives of urothelial carcinoma patients were evaluated in the current report: 
 the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
@@
 
-The following genes coding for FGF, FGFR and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a]:
+The following genes coding for FGFR, FGFF, and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a]:
 
@@
 JSON files with domain structures of FGFR1 (ID: P11362), FGFR2 (ID: P21802), FGFR3 (ID: P22607), and FGFR4 (ID: P22455) proteins were downloaded from [UniProt](https://www.uniprot.org) and processed with the _jsonlite_ package [@Ooms2023].
+
+For verification of expression of the FGFR-, FGF-, and FGFBP-coding genes at protein level, we resorted to the [Human Protein Atlas (HPA) repository](https://www.proteinatlas.org/), which stores results of immunohistochemical (IHC) staining.
+Entries of the HPA database were retrieved as XML documents, from which patient information, staining quantification, and URLs to IHC images for the genes of interest was retrieved (R package _xml2_) [@Wickham2023].
+Protein expression in the IHC samples was expressed as the IHC score defined as a product of staining (1: negative to 4: high), staining intensity (1: negative to 4: strong), and quantity of positive cells (1: none to 4: >75%).
 

### 2024-12-10 report/markdown/report.Rmd
Figure with mutation hot spots
And an update of the scheme of FGFR signaling

@@
 It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.
+In a more detailed analysis of mutation hot spots of the _FGFR3_ gene, mutation resulting in changes of protein residues S249C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'S249C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`% of cancer samples), Y373C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'Y373C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%), and R248C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'R248C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%) were present in $\geq$ 1% of cancer samples in at least two out of the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts (__Figure \@ref(fig:fig-fgfr3-hotspots)__).
 
@@
 Total mutation numbers are displayed in the plot captions._
+
+\newpage
+
+```{r fig-fgfr3-hotspots, fig.width = figur::convert(rep_figs$fgfr3_hotspots, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr3_hotspots, to = 'in')$h, fig.cap = 'Mutational hot spots of the FGFR3 gene.'}
+
+rep_figs$fgfr3_hotspots$plot
+
+```
+
+__Figure \@ref(fig:fig-fgfr3-hotspots). Mutational hot spots of the FGFR3 gene.__
+
+_(A) Density of mutations along the FGFR3 protein sequence was calculated with the sliding window method, scaled to the percentage within all cancer samples in the cohort, and displayed in dot plots.
+Each point represents a single protein residue, point color codes for assignment of the residue to the protein domain.
+Numbers of analyzed cancer samples are indicated in the plot facets.
+The protein domain scheme is displayed above the plot._
+
+_(B) Frequency of mutations affecting particular residues of the FGFR3 protein was expressed as percentage of all cancer samples in the cohorts and displayed in dot plots.
+Each point represents a single protein residue, point color codes for assignment of the residue to the protein domain.
+Candidate mutation hot spots with mutation rates $\geq$ 1% are highlighted with enlarged symbols, and labeled with percentages and protein sequence changes.
+Numbers of analyzed cancer samples are indicated in the plot facets.
+The protein domain scheme is displayed above the plot._
 

### 2024-12-16 report/markdown/report.Rmd
Report and figure update
Figures and Report text for analyses with DepMap cell lines

@@
 
-* Five genetic and transcrional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022].
+* Five genetic and transcriptional collectives of urothelial cancer patients were evaluated: the GENIE BLCA cohort, an urothelial carcinoma subset of the large GENIE project (n = `r nrow(genie$clinic)`) [@Sweeney2017], the predominantly muscle-invasive MSK IMPACT cohort (n = `r nrow(msk$clinic)`) [@Clinton2022], the muscle-invasive TCGA BLCA collective (n = `r nrow(tcga$clinic)`) [@Liu2018; @Robertson2017], the IMvigor cohort of late stage, platinum-refractory cancers subjected to an anti-PD-L1 treatment (n = `r nrow(imvigor$clinic)`) [@Balar2017], and the BCAN cohort of metastatic urothelial cancers (n = `r nrow(bcan$clinic)`) [@Damrauer2022].
 Detailed information on somatic mutations and their effects on protein sequence was provided for the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts.
@@
 Characteristic of the investigated collectives is presented in __Table \@ref(tab:tab-cohorts)__.
+In addition, analyses of expression, knockouts and RNA inference of FGFR-, FGF- and FGFBP-coding genes, as well as analyses resistance to pan-FGFR inhibitors were performed in n = `r nrow(depmap$cell_lines)`  urothelial cancer cell lines, whose data were retrieved from the [DepMap portal](https://depmap.org/portal).
 
@@
 __ - __\@ref(fig:fig-fgfr4-position)__, __Table \@ref(tab:tab-fgfr-class)__).
+In a more detailed analysis of mutation hot spots of the _FGFR3_ gene, mutation resulting in changes of protein residues S249C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'S249C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`% of cancer samples), Y373C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'Y373C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%), and R248C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'R248C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%) were present in $\geq$ 1% of cancer samples in at least two out of the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts (__Figure \@ref(fig:fig-fgfr3-hotspots)__).
 It can be hence speculated, that somatic mutations of the _FGFR_ genes may affect both the kinase activity and the conformation of the protein resulting in a higher affinity towards ligands or ligand-independent signaling.
-In a more detailed analysis of mutation hot spots of the _FGFR3_ gene, mutation resulting in changes of protein residues S249C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'S249C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`% of cancer samples), Y373C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'Y373C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%), and R248C (`r expl_hot$frequency$protein_domain$FGFR3 %>% filter(protein_change == 'R248C') %>% .$percent %>% signif(2) %>% range %>% paste(collapse = ' to ')`%) were present in $\geq$ 1% of cancer samples in at least two out of the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts (__Figure \@ref(fig:fig-fgfr3-hotspots)__).
+In particular, constitutive, ligand-independent signaling was described for the most frequent R248C [@Naski1996], S249C [@Tomlinson2007], Y373C [@Bonaventure2007], G370C, and S371C [@Adar2002] variants of the FGFR3 protein.
 
@@
 
-* At mRNA level, _FGFR3_, _TGFBR3_, and _FGFR1_ were the strongest expressed receptor genes.
+* At mRNA level, _FGFR3_, _TGFBR3_, _FGFR1_, and _FGFR2_ were the strongest expressed receptor genes.
 _FGF7_, _FGF11_, followed by _FGF2_ were the most abundant ligand transcripts; _HSPG2_, _SDC1_, and _CD44_ were the binding protein genes with the highest mRNA counts in the urothelial cancer tissue.
+_FGFR1_, _FGFR2_ and _FGFR3_ receptor genes, as well as _SDC1_, _HSPG2_, _CD44_, _SDC4_, _GPC1_ and _FIBP_ binding protein genes were also abundantly expressed at mRNA level by urothelial cancer cell lines in the DepMap repository [@DepMap2024] (__Figure \@ref(fig:fig-expression)A__, __Table \@ref(tab:tab-expression)__).
+This later observation shows clearly, that malignant epithelium cells are equipped with the key molecules of the oncogenic FGFR signaling.
 Out of `r globals[c("receptors", "ligands", "binding_proteins")] %>% reduce(union) %>% length` genes of interest, `r ncol(hpa$ihc_score) - 1` proteins were quantified by immunohistochemistry (IHC) in the Human Protein Atlas (HPA).
@@
 FGF7, FGF10, FGF3, FGF19, and FGF17 were the strongest expressed ligand proteins.
-SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).
-
-* Co-expression of FGFR-, FGF-, and FGFBP-coding genes was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__).
+SDC2, SDC1, GPC4, CD44, and SDC4 were the most abundant FGFBP proteins (__Figures \@ref(fig:fig-expression)B__ - __\@ref(fig:fig-ihc-fgfbp)__, __Table \@ref(tab:tab-expression)__).
+
+* Co-expression of FGFR-, FGF-, and FGFBP-coding genes in the urothelial cancer tissue was explored by correlation and network analyses, which revealed a high degree co-regulation of the ligands, receptors, and binding proteins in the urothelial cancer tissue (__Figure \@ref(fig:fig-fgfr-correlation)__ and __Figure \@ref(fig:fig-fgfr-networks)__).
 Especially the network analyses point towards strong co-regulation of expression of _FGFR1_ with genes of its ligands (_FGF2_, _FGF7_, _FGF10_), and co-receptors such as _SDC2_, _DCN_ or _HSPG2_.
@@
 _FGFR3_ levels correlated weakly with levels of _SDC1_ and _GPC1_ transcripts that code for proteoglycan FGF binding proteins.
+
+* A similar co-expression network analysis was conducted for FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines in the DepMap repository.
+Its results revealed two separate communities with, respectively, _FGFR1_ and _FGFR2/3_ as candidate hubs.
+As described above for the urothelial cancer specimens, the gene community of _FGFR1_ included also genes of FGFR1 ligands (_FGF1_, _FGF2_, _FGF5_, _FGF7_, _FGF9_) and co-receptors (_CD44_, _FGFRL1_, _TNFAIP6_) (__Figure \@ref(fig:fig-cell-networks)__).
+This suggests that urothelial cancer cells employ either a ligand-dependent signaling via FGFR1 or a possibly ligand-independent signaling via FGFR2/3.
+Those signaling pathways are likely non-redundant.
+
+* Next, we sought to investigate if urothelial malignant cells depend on FGFR signaling by an analysis of CRISPR/Cas9 and RNA interference screening data for urothelial cancer cell lines in the DepMap repository [@DepMap2024; @McFarland2018; @Dempster2021].
+In this analysis, we compared effects of knockouts and RNA silencing of genes coding for FGF receptor proteins with background effects for genes without meaningful expression in the urothelial cancer cell lines.
+Growth of JMSU-1, UM-UC-13, UM-UC-11, TCCSUP, and 639-V cell lines was significantly inhibited by knockout or silencing of the _FGFR1_ gene, which points towards critical dependence on FGFR1 signaling.
+Analogically, the UM-UC-1 cell line growth was inhibited by the _FGFR2_ CRISPR/Cas9 knockout.
+UM-UC-14, RT-112, RT112/84, and SW 780 cell line proliferation was curbed by the knockout or silencing of the _FGFR3_ gene.
+Notably, the effects of _FGFR1_, _FGFR2_, and _FGFR3_ gene knockouts or interference were independent of _FGFR3_ mutation status of a cell line (__Figure \@ref(fig:fig-crispri)__).
+
+* In a similar analysis, we delved into resistance of DepMap urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (compounds: `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`) [@Yang2013] and PRISM drug screening experiments (compounds: `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`) [@Yu2016].
+The resistance was metered in the GDSC1/2 experiments by IC50, a concentration of a compound that results in 50% cell line growth inhibition; sensitive cell lines were identified by IC50 < 10 µM, i.e.
+the highest concentration of the FGFR inhibitors In the PRISM experiment, the resistance was gauged by AUC, an area under the compound's dose-response curve normalized for response of the DMSO-treated cells; sensitive cell lines were identified by AUC < 1.
+As presented in __Figure \@ref(fig:fig-resistance)__, urothelial cancer cell lines responded to pan-FGFR inhibitors in both the GDSC1/2 and PRISM experiments in a _FGFR3_ mutation-independent manner.
+_FGFR3_ wild-type cell lines RT-112, RT4, SW 780, CAL-29, and J82 showed particularly high sensitivity to multiple FGFR blockers in the GDSC1/2 experiments.
+Erdafitinib is the sole clinically approved pan-FGFR inhibitor in urothelial cancer.
+_FGFR3_ wild-type cell lines `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$name_lab %>% collapse_and` demonstrated an especially good sensitivity to Erdafitinib in the PRISM assay (AUC: `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
+By contrast, `r expl_res$sensitive$prism %>% filter(.data[['PRISM_BRD-K84868168-001-01-0_erdafitinib']] == 'resistant') %>% .$sample_id %>% exchange(depmap$cell_lines, key = 'sample_id', value = 'cell_line_name') %>% collapse_and` cell lines were found to be resistant to erdafitinib with AUC values ranging from `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), status == 'resistant') %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`.
+Collectively, the pan-FGFR inhibitor response data, and the CRISPR/Cas9 and RNA interference screening results indicate FGFR signaling addiction of urothelial cancer cells independently of _FGFR3_ mutations.
+They also argue for a broader use of pan-FGFR inhibitors in urothelial cancer patients without _FGFR3_ gene alterations.
 
@@
 * Analysis of differential expression of FGFR-, FGF-, and FGFBP-coding genes was conducted for the major consensus molecular subtypes of the TCGA BLCA, IMvigor, and BCAN cohort.
-Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes (__Table \@ref(tab:tab-cons-dge)__).
+Transcript levels of n = `r length(unique(unlist(sub_dge$common_significant)))` out of `r length(reduce(sub_dge$variables, union))` investigated genes were significantly differentially regulated between consensus molecular subtyped in at least two cohorts (__Table \@ref(tab:tab-cons-dge)__).
 Among them, _FGFR3_, _SDC1_, _TGFBR3_, _GPC3_, and _SDC3_ were upregulated specifically in luminal papillary cancers.
 Overexpression of _KL_, a gene coding for [Klotho](https://www.genecards.org/cgi-bin/carddisp.pl?gene=KL) FGF-binding protein, was characteristic for luminal unstable tumors.
-Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class.
+Genes coding for FGFR protein, its ligands (_FGF7_, _FGF14_, _FGF10_, _FGF2_, _FGF9_, _FGF18_), and interaction partners (_DCN_, _HSPG2_, _SDC2_, _PTX3_) peaked in the stroma-rich class [@Ferguson2021].
 High expression of _FGF5_, _FGF11_, and genes coding for binding proteins and FGF co-receptors, _CD44_, _FGFBP1_, _GPC1_ and _TNFAIP6_, was a hallmark of basal/squamous-like cancers (__Figures \@ref(fig:fig-sub-fgfr-expression)__ - __Figure \@ref(fig:fig-sub-fgf-expression2)__).
@@
 
-* In summary, our analysis results suggest several modes of triggering of the oncogenic FGFR signaling highly specific for consensus molecular subsets of urothelial cancers (__Figure \@ref(fig:fig-summary)__).
-In papillary neoplasms, mutations of the _FGFR3_ gene affecting the hinge regions of the protein product and overexpression of co-receptors and binding proteins TGFB3, SDC1, GPC3, and SDC4 may culminate in a largely ligand-independent signaling.
-Such ligand-independent FGFR3 signaling was indeed reported for germ line mutations in the hinge region [@Moura2024].
-In stroma-rich cancers, the tightly coordinated overexpression of FGFR1, its ligands (FGF2, FGF7, FGF9, FGF10, FGF14, FGF18) and interacting proteins (DCN, SDC1, HSPG2, ANOS1, and PTX3) may fuel ligand-dependent para- and autocrine signaling.
-In basal/squamous tumors, overexpression of FGF5 along with co-activators like FGFBP1, CD44, or GPC1, may trigger ligand-dependent signaling via FGFR1.
+* In summary, our analysis results underline the biological relevance of FGFR signaling as an oncogenic driver of urothelial carcinoma irrespectively of _FGFR3_ gene alterations.
+FGFR1 and FGFR3 emerge as the main hubs of this signaling pathway.
+Additionally, we propose several modes of triggering of the oncogenic FGFR signaling highly specific for consensus molecular subsets of urothelial cancers (__Figure \@ref(fig:fig-summary)__).
+In papillary neoplasms, mutations of the _FGFR3_ gene affecting the hinge regions of the protein product [@Naski1996; @Tomlinson2007; @Bonaventure2007; @Adar2002; @Moura2024] and, possibly, overexpression of co-receptors and binding proteins TGFB3, SDC1, GPC3, and SDC4 [@Huttlin2021] may culminate in a largely ligand-independent signaling.
+In stroma-rich cancers, the tightly coordinated overexpression of FGFR1, its ligands (FGF2, FGF7, FGF9, FGF10, FGF14, FGF18) and interacting proteins (DCN, SDC1, HSPG2, ANOS1, and PTX3) may fuel ligand-dependent para- and autocrine signaling [@Ferguson2021].
+In basal/squamous tumors, overexpression of FGF5 along with co-activators like FGFBP1, CD44, or GPC1, may trigger ligand-dependent signaling via FGFR1 [@Ferguson2021].
 Interestingly, the elevated FGFPB1 in basal/squamous cancers may overcome the low availability of other FGFR1 ligands, such as FGF2 or FGF5, by forming complexes with high affinity for the receptor [@Tassi2001; @Beer2005].
 In luminal unstable cancers, overexpression of the _KL_ gene was observed, which codes for a co-receptor of FGF19, FGF21, and FGF23 [@Urakawa2006; @Goetz2007].
-Yet, those FGF-coding genes were expressed in bulk cancer samples at nearly undetectable levels (TCGA BLCA, _FGF23_: median: `r median(2^(tcga$expression$FGF23) - 1)` counts, IQR: `r  quantile(2^(tcga$expression$FGF23) - 1, c(0.25, 0.75)) %>% paste(collapse = ' to ')` counts, 95^th^ percentile: `r quantile(2^(tcga$expression$FGF23) - 1, 0.95)` copies).
+Yet, those FGF-coding genes were expressed in bulk cancer samples at nearly undetectable levels (TCGA BLCA, _FGF23_: median: `r signif(median(2^(tcga$expression$FGF23) - 1), 2)` counts, IQR: `r  quantile(2^(tcga$expression$FGF23) - 1, c(0.25, 0.75)) %>% signif(2) %>% paste(collapse = ' to ')` counts, 95^th^ percentile: `r signif(quantile(2^(tcga$expression$FGF23) - 1, 0.95), 2)` copies).
 Hence, the overexpression of _KL_ in the luminal unstable subset has likely no relevance for FGFR signaling.
@@
 This reflects the immense importance of FGFR signaling for phenotypical variability of urothelial cancers.
-It also puts forward this signaling pathway as an Achilles' heel of all urothelial cancers independent of the molecular phenotype that can be exploited by FGFR inhibitors.__
+It also puts forward this signaling pathway as an Achilles' heel of all urothelial cancers independent of the molecular phenotype that can be exploited by pan-FGFR inhibitors.__
 
@@
 _(A) Transcript levels of FGFR-, FGF-, and FGFBP-coding genes were determined in cancer tissue samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` by RNA sequencing and published as $log_2$ gene counts (`r globals$cohort_labs[c("tcga", "imvigor")] %>% collapse_and`) or Z-scores of $log_2$ gene counts (`r globals$cohort_labs["bcan"]`).
+mRNA expression of the genes of interest in DepMap urothelial cancer cell lines was measured by RNA sequencing and provided as $log_2$ gene counts.
 Median Z-scores of $log_2$ mRNA expression are shown in a heat map.
@@
 
-```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing.
+```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohorts by RNA sequencing.
 The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN).
@@
 
-_Pairwise correlation of log2-transformed mRNA levels of FGF-, FGFR-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts by false discovery rate-adjusted Pearson's test.
-Correlation coefficients r for significant gene pairs are depicted in bubble plots.
+_Pairwise correlation of log2-transformed mRNA levels of FGF-, FGFR-, and FGFBP-coding genes was investigated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts og bulk cancer transcriptomes, and in DepMap urothelial cancer cell lines.
+Statistical significance was determined by false discovery rate (FDR) adjusted Spearman's permutation test.
+Significant effects were considered for pFDR < 0.05 and $abs(\rho)\geq$ 0.3.
+Correlation coefficients $\rho$ for gene pairs found to be significant in all four collectives are depicted in bubble plots.
 Color of the points codes for correlation sign.
-Point size represents absolute values of r.
-Numbers of cancer samples are displayed in the plot captions._
-
-\newpage
-
-```{r fig-fgfr-networks, fig.width = figur::convert(rep_figs$fgfr_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.'}
+Point size represents absolute value of $\rho$.
+Numbers of cancer samples and cell lines are displayed in the plot captions._
+
+\newpage
+
+```{r fig-fgfr-networks, fig.width = figur::convert(rep_figs$fgfr_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$fgfr_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes: cancer tissue.'}
 
@@
 
-__Figure \@ref(fig:fig-fgfr-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.__
-
-_Co-expression networks were constructed in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman's rho $\rho \geq$ 0.3.
+__Figure \@ref(fig:fig-fgfr-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes: cancer tissue.__
+
+_Co-expression networks were constructed in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>%  collapse_and` bulk cancer transcriptome cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman's rho $\rho \geq$ 0.3.
 Isolated vertices of the networks were removed.
@@
 
-## Transcriptional and clinical significance of _FGFR3_ mutations
+```{r fig-cell-networks, fig.width = figur::convert(rep_figs$cell_networks, to = 'in')$w, fig.height = figur::convert(rep_figs$cell_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGF-, FGFR-, and FGFBP-coding genes: DepMap cell lines.'}
+
+rep_figs$cell_networks$plot
+
+```
+
+__Figure \@ref(fig:fig-cell-networks). Co-expression networks of FGF-, FGFR-, and FGFBP-coding genes: DepMap cell lines.__
+
+_Co-expression network was constructed in the DepMap urothelial cancer cell line collection for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman's rho $\rho \geq$ 0.3.
+Isolated vertices of the networks were removed.
+Gene communities were identified with the Leiden algorithm and named after their FGFR-coding members.
+Candidate hub genes of the communities were selected by computing hub scores separately for each community (eigenvectors of the affinity matrix, i.e.
+metrics of the overall correlation strength)._
+
+_The networks were visualized using the Fruchterman-Reingold algorithm.
+Points represent node vertices/genes, point shape codes for classification of the protein product.
+Point color codes for the community assignment (A) or hub score (B).
+Lines represent edges, i.e.
+correlations between expression levels with $\rho \geq$ 0.3.
+Line widths correspond to $\rho$ values.
+Note the central positions and high scores of the FGFR1, FGFR2, and FGFR3 genes and their assignment to the communities.
+This suggests exclusivity, i.e that FGFR1 with its ligands (FGF2, FGF5) and accessory molecules (CD44, TNFAIP6, FGFRL1) provide oncogenic signaling in a subset of cell lines, while signaling via FGFR2/FGFR3 and their co-receptors (GPC1/3, SDC1/4, HSPG2, TGFBR3) is exploited by another, non-overlapping cell line subset._
+
+\newpage
+
+## Dependency of urothelial cancer on FGFR signaling
+
+```{r fig-crispri, fig.width = figur::convert(rep_figs$crispr_rnai, to = 'in')$w, fig.height = figur::convert(rep_figs$crispr_rnai, to = 'in')$h, fig.cap = 'CRISPR and RNA interference screening results for genes coding for FGF receptors in DepMap urothelial cancer cell lines.'}
+
+rep_figs$crispr_rnai$plot
+
+```
+
+__Figure \@ref(fig:fig-crispri). CRISPR and RNA interference screening results for genes coding for FGF receptors in DepMap urothelial cancer cell lines.__
+
+_Copy-number number-corrected effects of gene knockouts by CRISPR/Cas9 and of gene silencing by RNA interference (RNAi) on cell viability estimated by the Demeter2 and CHRONOS models were retrieved for urothelial cancer cell lines from the DepMap repository (CRISPR: n = `r nrow(depmap$crispr)`, RNAi: n = `r nrow(depmap$rnai)` cell lines).
+Effects for the FGFR-coding genes were compared with background effects of genes not expressed in the urothelial cancer cell lines (CRISPR: n = `r ncol(depmap$crispr_bcg) - 1`, RNAi: n = `r ncol(depmap$rnai_bcg) - 1` genes).
+Significant negative or positive effects were assumed for gene effect values beyond 95^th^ percentile of the non-expressed gene background effects._
+
+_Gene effects for the FGFR-coding genes and the backgrounds are shown in box plots.
+Median gene effects with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single cell lines are depicted as points; point symbol codes for the FGFR3 gene mutation status.
+Cell lines with significant positive and negative effects are highlighted by red and blue point color._
+
+\newpage
+
+```{r fig-resistance, fig.width = figur::convert(rep_figs$resistance, to = 'in')$w, fig.height = figur::convert(rep_figs$resistance, to = 'in')$h, fig.cap = 'Resistance and sensitivity to pan-FGFR inhibitors of DepMap urothelial cell lines.'}
+
+rep_figs$resistance$plot
+
+```
+
+__Figure \@ref(fig:fig-resistance). Resistance and sensitivity to pan-FGFR inhibitors of DepMap urothelial cell lines.__
+
+_Information on resistance of urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (n = `r range(nrow(depmap$gdsc) - colMissing(depmap$gdsc[-1])) %>% paste(collapse = ' to ')` cell lines) and PRISM (n = `r range(nrow(depmap$prism) - colMissing(depmap$prism[-1])) %>% paste(collapse = ' to ')`) drug screening experiments was extracted from the DepMap repository.
+The resistance values were expressed as IC50 (concentration resulting in 50% growth inhibition, GDSC) or AUC (area under the dose-response curve, relative to DMSO control, PRISM)._
+_Significant sensitivity for the GDSC data was defined as IC50 below 10µM, which was the maximum concentration of the FGFR inhibitors._ 
+_Significant sensitivity for the PRISM data was defined as AUC < 1, which indicates stronger growth inhibition by a FGFR inhibitor than the DMSO control._ 
+_IC50 and AUC for doxorubicin are presented as a positive cytotoxicity controls._
+
+_Median IC50 and AUC values for the FGFR inhibitors with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
+Points represent single cell lines.
+Point symbols codes for the FGFR3 gene mutation status, point color represent sensitivity and resistance.
+The sensitivity cutoffs are represented by dashed lines._
+
+\newpage
+
+\newpage
+
+## Biological and clinical significance of _FGFR3_ mutations
 
@@
 
-Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome log_2 mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020].
+Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome $log_2$ mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020].
 
@@
 JSON files with domain structures of FGFR1 (ID: P11362), FGFR2 (ID: P21802), FGFR3 (ID: P22607), and FGFR4 (ID: P22455) proteins were downloaded from [UniProt](https://www.uniprot.org) and processed with the _jsonlite_ package [@Ooms2023].
+
+Mutation status, baseline expression measured by RNA sequencing, and effect scores of CRISP/Cas9 and RNA interference of the FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines were downloaded from the [DepMap](https://depmap.org/portal/) repository [@DepMap2024].
+The available cell lines were `r map2_chr(depmap$cell_lines$cell_line_name, depmap$cell_lines$sample_id, ~paste0(.x, ' (ID:', .y, ')')) %>% collapse_and`. 
+The mutation status was 0/1 coded, where 1 stood for presence of at least one somatic mutation in a gene. 
+RNA sequencing information was provided as $log_2$ TPM. 
+The copy number-corrected effect scores of CRISPR/Cas9 knockouts and RNA inference were computed by Demeter2 and CHRONOS algorithms [@McFarland2018; @Dempster2021]. 
+Gene effects of the knockouts and RNA interference for the FGFR receptor genes were compared with the background effects for non-expressed genes (95^th^ percentile of expression values = 0; CRISPR/Cas9: n = `r ncol(depmap$crispr_bcg) - 1`), RNAi: n = `r ncol(depmap$rnai_bcg) - 1` non-expressed genes).
+Significant gene effects were considered for effect scores beyond the 95^th^ percentile range of the non-expressed background effects. 
+Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013].
+The resistance was measured in the GDSC1/2 experiments by IC50, the concentration resulting in 50% growth inhibition; sensitive cell lines were defined by IC50 < 10 µM, which was the highest concentration of the FGFR inhibitors in the assays.
+The resistance in the PRISM assays was gauged by AUC, i.e. area under the dose-response curve normalized for effects of DMSO control; sensitive cell lines were considered for AUC < 1.
 

### 2024-12-17 report/markdown/report.Rmd
Additional literature references
References for the Leiden algorithm and the imputation method used in the BCAN expression data set

@@
 The biological material was collected both in urogenital tract and at distant metastatic sites. 
+Missing values in the Z-score expression data set of the BCAN cohort were imputed with the 9-nearest neighbor algorithm implemented by function `impute.knn()` from _impute_ package [@Hastie2023].
 
@@
 Isolated vertices were removed with `prune_degree()` function. 
-The following vertex importance statistics [@Csardi2006] were computed with `summary()` method called for the network objects: 
+Communities of co-regulated genes in the co-expression network in the DepMap urothelial cancer cell line collective were identified with the Leiden algorithm (loss function: modularity, resolution: $\gamma$ = 0.5, iterations: n = 100, function `cluster_leiden()`, _igraph_ package) [@Csardi2006; @Traag2019].
+
+The following vertex importance statistics [@Csardi2006] were computed with `summary()` method called for entire network objects (bulk cancer cohorts) or gene community sub-networks (DepMap cell lines):
 degree (number of directly connected neighbor nodes), 

### 2024-12-19 report/markdown/report.Rmd
SHAP importance in Figures and Report
SHAP importance data included in the figures and analysis report

@@
 Among them, only `r intersect(consensusMIBC::centroids$hgnc_symbol, ml_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and` were shared with our machine learning models.
-The most influential genes for prediction of consensus molecular subtypes in our machine learning approach are presented in __Figure \@ref(fig:fig-sub-imp)__, and __Tables \@ref(tab:tab-ml-imp-elnet)__ - __\@ref(tab:tab-ml-imp-ranger)__.
+
+* The most influential genes for prediction of consensus molecular subtypes by our machine learning approach were identified with SHAP algorithm (Shapley additive explanations) [@Lundberg2017; @Covert2020] (__Table \@ref(tab:tab-shap)__).
+In such analysis for the Elastic Net model, increasing expression levels of _TGFBR3_ and _FGFR3_, and decreasing expression of _FGF7_ and _TNFAIP6_, were associated with LumP class assignment.
+Analogically, _FGFR4_ and _TNFAI_ were proposed as positive, while _CD44_ and _GPC1_ as negative predictors of the LumU class.
+_FGF7_ and _FGFR1_ were proposed as top positive predictors of stroma-rich cancers; high levels of _FGFBP1_, _CD44_, and _ANOS1_ transcripts were associated with prediction of basal/squamous class by the Elastic Net model (__Figure \@ref(fig:fig-sub-imp-elnet)__).
+For the Random Forest model, increasing levels of _TGFBR3_, _FGFR3_, and _FGFBP3_ were associated with predictions of LumP class.
+High expression of _FGFR1_, _FGFR4_, _TGFBR3_, and _KL_ predicted LumU cancers.
+High levels of _DCN_, _FGF7_, _TGFBR3_, and _FGFR1_ were identified as positive predictors of stroma-rich cancers, while _CD44_, _FGFBP1_, _TNFAIP6_, and _FGF5_ were associated with forecasts of basal/squamous tumors by the Random Forest classifier (__Figure \@ref(fig:fig-sub-imp-ranger)__).
 
@@
 Numbers of cancer samples in the consensus subtypes are indicated in the plot caption.
-Note: samples without reliable assignment to the consensus subtypes are not presented in the heat maps._
+Note: few NE-like samples (up to n = `r sub_dge$test %>% map(filter, consensusClass == 'NE-like') %>% map_dbl(~.x$n2[1]) %>% max(na.rm = TRUE)`) are removed from the plots._
 
@@
 
-```{r fig-sub-imp, fig.width = figur::convert(rep_figs$sub_imp, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_imp, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers.'}
-
-rep_figs$sub_imp$plot
-
-```
-
-__Figure \@ref(fig:fig-sub-imp). Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers.__
-
-_The machine learning models of consensus molecular subsets with expression of FGFR-, FGF, and FGFBP genes as the sole explanatory factors were trained in a `r globals$cohort_labs["tcga"]`-derived data set as presented in Figure \@ref(fig:fig-sub-elnet)._
-_In the Elastic Net model, absolute values of the coefficient's $\beta$ translate directly to variable importance measure; high $abs(\beta)$ is expected for a highly influential variable for prediction of a subset.
-$\beta$ values for the top 15 most influential explanatory variables in the Elastic Net model are presented as bar plots in (A)._
-_For the Random Forest model, permutation importance of explanatory variables was computed as difference in out-of-bag classification errors between the genuine model and a model with the variable of interest re-shuffled at random.
-The differences in classification errors for the top 15 most influential explanatory variables of the Random Forest model are presented in a bar plot in (B)._
+```{r fig-sub-imp-elnet, fig.width = figur::convert(rep_figs$sub_imp_elnet, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_imp_elnet, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers assessed by the SHAP algorithm'}
+
+rep_figs$sub_imp_elnet$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-imp-elnet). Importance of explanatory variables in the Elastic Net model of molecular consensus classes of urothelial cancers assessed by the SHAP algorithm__
+
+_Contributions of the explanatory variables, i.e.
+the FGFR-, FGF-, and FGFBP-coding genes, to prediction of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous (Ba/sq) consensus molecular classes were estimated with the SHAP algorithm (Shapley additive explanations)._
+
+_SHAP values for single highly influential genes (top 15 highest mean absolute SHAP) and observations are visualized in violin/swarm plots.
+Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene.
+The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
+Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
+
+\newpage
+
+```{r fig-sub-imp-ranger, fig.width = figur::convert(rep_figs$sub_imp_ranger, to = 'in')$w, fig.height = figur::convert(rep_figs$sub_imp_ranger, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Random Forest model of molecular consensus classes of urothelial cancers assessed by the SHAP algorithm'}
+
+rep_figs$sub_imp_ranger$plot
+
+```
+
+__Figure \@ref(fig:fig-sub-imp-ranger). Importance of explanatory variables in the Random Forest model of molecular consensus classes of urothelial cancers assessed by the SHAP algorithm__
+
+_Contributions of the explanatory variables, i.e.
+the FGFR-, FGF-, and FGFBP-coding genes, to prediction of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous (Ba/sq) consensus molecular classes were estimated with the SHAP algorithm (Shapley additive explanations)._
+
+_SHAP values for single highly influential genes (top 15 highest mean absolute SHAP) and observations are visualized in violin/swarm plots.
+Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene.
+The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
+Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
 
@@
 
-```{r tab-ml-imp-ranger, tab.cap = "Permutation variable importance of the Random Forest model of consensus molecular classes of urothelial cancers.
-The Random Forest model was developed in the TCGA BLCA cohort with expression values of genes coding for FGFR, FGF, and FGFBP proteins as the sole explanatory factors.
-Permutation importance was computed as differences of errors of the genuine model and models, in which the explanatory variables were re-shuffled at random."}
-
-flextable::flextable(rep_tabs$ml_imp_ranger) %>% 
-  italic(j = 1) %>% 
-  set_widths(c(3, 3)) %>% 
-  theme_vanilla
+```{r tab-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of consensus molecular classes of urothelial cancers estimated by the SHAP algorithm.
+Contribution of the FGFR-, FGF-, and FGFBP-coding genes to prediction of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations).
+The table is available as a supplementary Excel file."}
+
+rep_tabs$shap[1, 1] %>% 
+  mutate(Algorithm = '') %>% 
+  set_names(' ') %>% 
+  flextable %>% 
+  set_widths(18)
 
@@
 
-The models were developed in a data set derived from the TCGA BLCA collective; the consensus molecular class assignment by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response.
+The models were developed in a data set derived from the TCGA BLCA collective; assignment to luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous (Ba/Sq) consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response.
 ComBat-processed $log_2$ expression levels of the genes of interest were used as the sole explanatory factors. 
-In order to improve recognition of the minority consensus classes, the LumU (luminal unstable) and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016].
+In order to improve recognition of the minority consensus classes, the LumU and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016].
 The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010].
@@
 
+Importance of the explanatory variables, i.e.
+FGFR-, FGF-, and FGFBP-coding transcripts, and their positive and negative association with predictions of LumP, LumU, stroma-rich, and Ba/Sq classes by the machine learning models were assessed with the SHAP algorithm (Shapley additive explanations) [@Lundberg2017; @Covert2020].
+To this end, we re-fitted the Elastic Net and Random Forest models of consensus molecular classes in the unmodified TCGA BLCA data set with the optimal set of the tuning parameters. 
+Matrices of SHAP metrics for the genes and observations were computed for the models with `kernelshap()` function from _kernelshap_ package [@Mayer2023; @Mayer2023a]. 
+As the background for the SHAP procedure, we used randomly selected 30% of observations of the TCGA BLCA data set. 
+Mean absolute SHAP value served as a measure of gene importance (__Table \@ref(tab:tab-shap)__). 
+SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots (__Figure \@ref(fig:fig-sub-imp-elnet)__ and __\@ref(fig:fig-sub-imp-ranger)__) with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions.
+
 For the Elastic Net model, absolute $\beta$ values of the model coefficients serve as direct variable importance metrics; they were extracted from the model with `coef()` method. 

### 2024-12-19 ml scripts/shap.R
SHAP importance
SHAP variable importance for the Elastic Net and Random Forest models of consensus molecular clusters

@@ -1,0 +1,89 @@
+# SHAP variable importance for the Elastic Net and Random Forest models in
+# the TCGA data set.
+#
+# The background: a randomly chosen 30% of the TCGA gene expression data
+# set.
+
+
+  insert_head()
+
+# container ------
+
+  ml_shap <- list()
+
+# analysis globals --------
+
+  insert_msg('Analysis globals')
+
+  ## matrix of explanatory factors,
+  ## and responses in the TCGA training cohort
+
+  ml_shap$data <- ml_globals$data$tcga %>%
+    column_to_rownames('sample_id')
+
+  ml_shap$x <- ml_globals$x$tcga
+  ml_shap$y <- ml_globals$y$tcga
+
+  ## background matrix
+
+  ml_shap$bg_idx <- createDataPartition(ml_shap$y, p = 0.3)[[1]]
+
+  ml_shap$bg_x <- ml_shap$x[ml_shap$bg_idx, ]
+
+  ## best tunes
+
+  ml_shap$best_tunes <- ml_train$tune_obj %>%
+    map(~.x$best_tune)
+
+# Models --------
+
+  insert_msg('Models')
+
+  set.seed(12345)
+
+  ml_shap$models$elnet <-
+    glmnet(x = ml_shap$x,
+           y = ml_shap$y,
+           family = 'multinomial',
+           alpha = 0.5,
+           lambda = ml_shap$best_tunes$elnet$lambda)
+
+  ml_shap$models$ranger <-
+    ranger(x = ml_shap$x,
+           y = ml_shap$y,
+           num.trees = 500,
+           mtry = ml_shap$best_tunes$ranger$mtry,
+           min.node.size = ml_shap$best_tunes$ranger$min.node.size,
+           splitrule = ml_shap$best_tunes$ranger$splitrule,
+           probability = TRUE)
+
+# SHAP objects -------
+
+  insert_msg('SHAP objects')
+
+  ml_shap$shap_obj$elnet <-
+    kernelshap(ml_shap$models$elnet,
+               X = ml_shap$x,
+               bg_X = ml_shap$bg_x,
+               pred_fun = function(object, X, ...){
+
+                 predict(object, X, type = 'response', ...)[, , 1]
+
+               })
+
+  ml_shap$shap_obj$ranger <-
+    kernelshap(ml_shap$models$ranger,
+               X = ml_shap$x,
+               bg_X = ml_shap$bg_x)
+
[…]

### 2024-12-19 ml scripts/shap_plots.R

@@ -1,0 +1,60 @@
+# Plots of SHAP variable importance
+
+  insert_head()
+
+# container ------
+
+  ml_splots <- list()
+
+# SHAP and SHAPviz objects -------
+
+  insert_msg('SHAP and SHAPviz objects')
+
+  ## class names and algorithm labels
+
+  ml_splots$class_levels <- ml_globals$class_levels
+
+  ml_splots$algo_labs <-
+    c('elnet' = 'Elastic Net',
+      'ranger' = 'Random Forest')
+
+  ## the SHAP/SHAPviz objects
+
+  ml_splots$shap_obj <- ml_shap$shap_obj[c("elnet", "ranger")]
+
+  ml_splots$shapviz <- ml_splots$shap_obj %>%
+    map(shapviz) %>%
+    map(set_names, ml_splots$class_levels)
+
+# Mean absolute SHAP values --------
+
+  insert_msg('Mean absolute SHAP values')
+
+  ml_splots$stats <- ml_splots$shapviz %>%
+    map(get_shap_values) %>%
+    map(map, abs) %>%
+    map(map, colMeans) %>%
+    map(map,
+        compress,
+        names_to = 'variable',
+        values_to = 'importance') %>%
+    map(compress, names_to = 'class')
+
+# Beeswarms of SHAPs and bar plots of mean absolute SHAPS ------
+
+  insert_msg('Plots')
+
+  ml_splots$plots <-
+    list(shapviz_obj = ml_splots$shapviz,
+         title_suffix = ml_splots$algo_labs) %>%
+    pmap(my_shap_bee,
+         panels = TRUE,
+         add_legend = FALSE,
+         top_variables = 15,
+         point_size = 1)
+
+# END --------
+
+  ml_splots <- ml_splots[c("stats", "plots")]
+
+  insert_tail()

### 2024-12-19 ml.R

@@ -23,6 +23,10 @@
   library(caret)
   library(ranger)
   library(performanceEstimation)
+  library(exda)
+
+  library(kernelshap)
+  library(shapviz)
 
   library(furrr)
 
@@ -51,14 +55,17 @@
 
   insert_msg('Analysis scripts')
 
-  ## model tuning and training
+  ## model tuning and training, SHAP importance
 
   list(cache_path = c('./cache/ml_train.RData',
-                      './cache/ml_ens.RData'),
+                      './cache/ml_ens.RData',
+                      './cache/ml_shap.RData'),
        script_path = c('./ml scripts/development.R',
-                       './ml scripts/ensemble.R'),
+                       './ml scripts/ensemble.R',
+                       './ml scripts/shap.R'),
        message = c('Cached model tuning and training results',
-                   'Cached ensemble tuning and training results')) %>%
+                   'Cached ensemble tuning and training results',
+                   'Cached SHAP variable importance')) %>%
     pwalk(access_cache)
 
 
@@ -66,7 +73,8 @@
 
   c('./ml scripts/predictions.R',
     './ml scripts/evaluation.R',
-    './ml scripts/importance.R') %>%
+    './ml scripts/importance.R',
+    './ml scripts/shap_plots.R') %>%
     source_all(message = TRUE, crash = TRUE)
 
 # END --------

### 2024-12-19 report scripts/figures.R

@@ -769,7 +769,7 @@
   insert_msg('Molecular subtypes and expression of FGF, FGFR, FGFBP genes')
 
   ## heat map plots are presented:
-  ## not-assigned, NE-like, and LumNS samples are removed
+  ## not-assigned, NE-like samples are removed
 
   rep_figs$sub_fgfr_dge <- sub_dge$hm_plots %>%
     map(~.x +
@@ -786,7 +786,7 @@
 
     rep_figs$sub_fgfr_dge[[i]]$data <-
       rep_figs$sub_fgfr_dge[[i]]$data %>%
-      filter(!consensusClass %in% c('not assigned', 'LumNS', 'NE-like'))
+      filter(!consensusClass %in% c('not assigned', 'NE-like'))
 
   }
 
@@ -983,24 +983,30 @@
 
   insert_msg('Elastic Net model, variable importance')
 
-  rep_figs$sub_imp <- ml_imp$coef_plots %>%
-    map(~.x + theme(legend.position = 'none')) %>%
-    c(list(get_legend(ml_imp$coef_plots[[1]])),
-      list(ml_imp$ranger_plot)) %>%
-    plot_grid(plotlist = .,
-              ncol = 3,
-              align = 'hv',
-              axis = 'tblr',
-              labels = c('A', '', '',
-                         '', '', 'B'),
-              label_size = 10) %>%
-    as_figure(label = 'consensus_subset_prediction_key_genes',
-              ref_name = 'sub_imp',
-              caption = paste('Importance of explanatory variables in the',
-                              'Elastic Net model of molecular consensus',
-                              'classes of urothelial cancers.'),
-              w = 190,
-              h = 160)
+  rep_figs[c('sub_imp_elnet', 'sub_imp_ranger')] <-
+    ml_splots$plots %>%
+    map(function(x) plot_grid(plotlist = x$panels,
+                              ncol = 2,
+                              align = 'hv',
+                              axis = 'tblr') %>%
+          plot_grid(x$legend,
+                    nrow = 2,
+                    rel_heights = c(0.92, 0.08)))
+
+  rep_figs[c('sub_imp_elnet', 'sub_imp_ranger')] <-
+    rep_figs[c('sub_imp_elnet', 'sub_imp_ranger')] %>%
+    list(x = .,
+         label = paste0('consensus_subset_prediction_key_genes',
+                        c('_elastic_net', '_random_forest')),
+         ref_name = names(.),
+         caption = paste('Importance of explanatory variables in the',
[…]

### 2024-12-19 report scripts/tables.R

@@ -287,27 +287,32 @@
                                'the sole explanatory factors.',
                                'Values of beta of the coefficients are listed.'))
 
-# Permutation variable importance of the Random Forest model --------
+# SHAP variable importance -------
 
-  insert_msg('Permutation importance for the Random Forest model')
+  insert_msg('SHAP variable importance')
 
-  rep_tabs$ml_imp_ranger <- ml_imp$ranger_importance %>%
-    arrange(-importance) %>%
-    mutate(importance = as.character(signif(importance, 2))) %>%
-    set_names(c('Varaible', 'Permutation importance')) %>%
-    as_mdtable(label = 'ml_imp_ranger',
-               ref_name = 'ml_imp_ranger',
-               caption = paste('Permutation variable importance of the',
-                               'Random Forest model of consensus molecular',
-                               'classes of urothelial cancers.',
-                               'The Random Forest model was developed in the',
-                               'TCGA BLCA cohort with expression values of genes',
-                               'coding for FGFR, FGF, and FGFBP proteins as',
-                               'the sole explanatory factors.',
-                               'Permutation importance was computed as',
-                               'differences of errors of the genuine model',
-                               'and models, in which the explanatory variables',
-                               'were re-shuffled at random.'))
+  rep_tabs$shap <- ml_splots$stats %>%
+    compress(names_to = 'algorithm') %>%
+    transmute(Algorithm = car::recode(algorithm,
+                                      "'elnet' = 'Elastic Net';
+                                      'ranger' = 'Random Forest'"),
+              `Consensus class` = class,
+              `Gene symbol` = variable,
+              `Importance, mean absolute SHAP` = signif(importance, 2)) %>%
+    as_mdtable(label = 'shap',
+               ref_name = 'shap',
+               caption = paste('Importance of explanatory variables for the',
+                               'Elastic Net and Random Forest models of',
+                               'consensus molecular classes of urothelial',
+                               'cancers estimated by the SHAP algorithm.',
+                               'Contribution of the FGFR-, FGF-, and',
+                               'FGFBP-coding genes to prediction of consensus',
+                               'molecular classes was assessed by mean absolute',
+                               'values of SHAP metrics (Shapley additive',
+                               'explanations).',
+                               'The table is available as a supplementary',
+                               'Excel file.'))
+
 
 # Evaluation of the Elastic Net Cox model of overall survival --------
 

### 2024-12-19 tools/functions.R

@@ -1774,6 +1774,258 @@
 
   }
 
+  my_shap_bee <- function(shapviz_obj,
+                          top_variables = 10,
+                          title_suffix = NULL,
+                          point_size = 2,
+                          point_alpha = 0.75,
+                          point_jitter = 0.1,
+                          show_violin = TRUE,
+                          violin_alpha = 0.25,
+                          violin_fill = 'cornsilk',
+                          panels = FALSE,
+                          rel_widths = c(0.65, 0.35),
+                          add_legend = TRUE,
+                          rel_heights = c(0.85, 0.15)) {
+
+    ## a customized bee-swarm plot of SHAP values as a function of feature
+    ## value (min/max) for multinomial classification models
+
+    class_levels <- names(shapviz_obj)
+
+    plot_titles <- class_levels
+
+    if(!is.null(title_suffix)) {
+
+      plot_titles <- paste(plot_titles, title_suffix, sep = ', ')
+
+    }
+
+    ## plotting data: min/max-normalized feature values and SHAP --------
+
+    plot_data <- list()
+
+    plot_data$ft <- get_feature_values(shapviz_obj) %>%
+      map(min_max)
+
+    plot_data$shap <- get_shap_values(shapviz_obj)
+
+    plot_data$shap <-
+      map2(plot_data$shap,
+           plot_data$ft,
+           ~set_rownames(.x, rownames(.y)))
+
+    ft_names <- plot_data$ft %>%
+      map(colnames)
+
+    ## mean absolute values of SHAP per feature as a variable importance metric -------
+
+    ## the correlation of SHAP and value will be used to classify the
+    ## explanatory factors as variables that associate positively and negatively
+    ## with the outcome
+
+    shap_imp <- list()
+
+    shap_imp$absmean <- plot_data$shap %>%
+      map(abs) %>%
+      map(colMeans) %>%
+      map(compress,
+          names_to = 'variable',
+          values_to = 'meanabs_shap')
+
+    shap_imp$corr <-
+      map2(plot_data$ft,
+           plot_data$shap,
+           function(x, y) map2_dbl(as.data.frame(x),
+                                   as.data.frame(y),
+                                   cor,
+                                   method = 'spearman')) %>%
+      map(compress,
+          names_to = 'variable',
+          values_to = 'corr')
+
+    shap_imp <- shap_imp %>%
+      map(compress, names_to = 'class') %>%
[…]

### 2025-03-20 report/markdown/rebuttal_letter.Rmd
Rebuttal Letter

---
title: "FGFR1/3 Signaling as Achilles’ Heel of Phenotypic Diversity in Urothelial Carcinoma"
subtitle: "Rebuttal Letter"
author: "Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: blca.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "cairo_pdf", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

# Reviewer 1

## Point 1

### Issue

In some of the cohorts, there are patients with NMIBC (i.e. prior MSK IMPACT study with 20 - 25% NMIBC) and upper tract disease (i.e. IMvigor). 
There are prior studies suggesting genomic landscapes differ widely among NMIBC vs. MIBC and upper tract vs. bladder cancers. 
Therefore, it seems it would be beneficial to look at each of these groups separately (MIBC, NMIBC, and upper tract) rather than lump them all together. 
Can the authors provide more justification about why they have included NMIBC and upper tract disease in this manuscript instead of focusing just on MIBC?

### Response

Thanks for this constructive comment. 
We agree that urothelial tumors with different anatomical locations but also NMIBC and MIBC differ in their transcriptomic landscapes. 
Because of the brief report format of our manuscript, we decided against separate analyses of upper/lower tract malignancies and comparison of NMIBC and MIBC, and will consider them in our future works. 
Instead, we constrained the analyzed data sets to patient with MIBC of the bladder. 
If no NMIBC/MIBC information was provided, we restricted the cancer samples to T2 - T4 stage tumors. 
Please note that because of the very rudimentary clinical information for the GENIE cohort, we were not able to filter the samples by anatomical location and invasiveness. 
Yet, as evident from __Figure 1B__, there were no substantial differences in location and density of mutations of _FGFR3_ between the unfiltered GENIE cohort and the MSK IMPACT, TCGA BLCA and BCAN cohorts restricted to bladder-only MIBC.

The final numbers of investigated cancer samples were: 
n = `r length(genie$analysis_ids)` in the GENIE cohort, 
n = `r length(msk$analysis_ids)` in the MCK IMPACT cohort, 
n = `r length(tcga$analysis_ids)` in the TCGA BLCA cohort, 
n = `r length(imvigor$analysis_ids)` in the IMvigor, and 
n = `r length(bcan$analysis_ids)` in the BCAN collective.

## Point 2

### Issue

Were any of the samples pre-treated in the five cohorts used? 
What about for the drug screens, were any of those samples pre-treated?

### Response

This is certainly an important point.

Samples in the bulk cancer transcriptome cohorts were indeed obtained from patients with diverse treatment history. 
In __Supplementary Table S1__ of the revised manuscript, we present frequencies of patients with non-invasive intra-vesical treatment, surgery and distribution of surgery types, neoadjuvant treatment, adjuvant systemic treatment, systemic chemo- and immunotherapy, and radiation. 
For most of the bulk cancer data sets (GENIE, MSK IMPACT, IMvigor, BCAN), this information was published along with the molecular data or extracted from the original publications. 
For the TCGA BLCA cohort, the treatment information was derived from papers by Robertson et al. [@Robertson2017] and Moiso [@Moiso2021].

Concerning the cell lines, we extracted demographic and clinical information of the donors from the DepMap portal [@DepMap2024] and present it in __Reviewer Table \@ref(tab:tab-cell-lines)__. 
Unfortunately for none of the analyzed cell lines, treatment information was available.

## Point 3

### Issue

While the authors show they can accurately predict the consensus classifier with FGF/FGFR pathway genes - the next step is what to do with the consensus classifier information? Currently we do not make any therapeutic recommendations based on the groups, so its a bit unclear how helpful this is moving forward.

### Response

We are completely aware that, at the moment, there are no recommendations on diagnostic and treatment of the MIBC consensus classes. 
Following your suggestion, we predicted metrics of drug resistance (IC50 and AUC) for the bulk cancers in the TCGA BLCA, IMvigor, and BCAN cohorts with RIDGE linear models trained with drug response and transcriptome data of epithelial pan-cancer cell lines in the GDSC, CTRP2, and PRISM in vitro drug screening experiments [@Yang2013; @Seashore-Ludlow2015; @Corsello2020]. 
Of note, this machine learning strategy employed our development package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) and regularized linear modeling approach to drug resistance prediction proposed by Geeleher at al. [@Geeleher2014] and successfully employed in our recent publication on testicular cancer [@Torzsok2025]. 
Subsequently, we compared the predicted IC50 and AUC resistance metrics between the MIBC consensus classes (__Supplementary Figure XXX___). 
These new data shows that for most of pan-FGFR inhibitors (AZD4547, FGFR3831, FGFR0939, erdafitinib), good sensitivity was predicted for the LumP class, the predicted response was highly heterogeneous in the renaming subsets. 
In^particular, for Ba/Sq cancers resistance was forecasted, while stroma-rich were put forward as erdafitinib-responsive malignancies. 
We also compared the resistance metrics of FGFR-targeting drugs for cancers with and without mutation of _FGFR3_ (__Supplementary Figure XXX__); in this case, no differences in predicted response between WT and mutant specimens were observed.

## Point 4

### Issue

Does Kamoun et al. provide SHAP values for the 857 genes in their classifier? 
If so, would be interesting to know the value for FGF/FGFR genes. 
What about any of the other consensus classifiers?

### Response

We appreciate this point a lot. 
Unfortunately, were were not able to find SHAP variable importance measures for the nearest centroid classification tool by Kamoun and colleagues. 
As a workaround, we trained Elastic Net and Random Forest machine learning models of MIBC consensus classes in a training subset of two-thirds of observations of the TCGA BLCA cohort with the following explanatory factors:

* ComBat-adjusted $log_2$ expression levels of `r length(ml_globals$variables)` FGFR/FGF/FGFBP-related genes as outlined in the initial version of the manuscript

* Combat-adjusted $log_2$ expression levels of `r length(mibc_globals$variables)` MIBC class-defining genes proposed by Kamoun et al [@Kamoun2020] and shared by the TCGA BLCA, IMvigor, and BCAN expression data sets

In evaluation of model performance, the MIBC consensus class assignment by the genuine nearest centroid tool by Kamoun et al. served as the ground truth. 
As presented in __Supplementary Figure YYY__ predictions of consensus classes by the models with the full set of predictors by Kamoun et al. were emulated the original classification tool with very high accuracy (test collectives, accuracy: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Accuracy %>% signif(2) %>% range %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Kappa %>% signif(2) %>% range %>% paste(collapse = ' to ')`). 
Their performance was generally better that accuracy of the models employing solely the FGFR/FGF/FGFBP-related genes (test collectives, accuracy: `r mibc_comp$stats %>% filter(model == 'fgfr_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Accuracy %>% signif(2) %>% range %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r mibc_comp$stats %>% filter(model == 'fgfr_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Kappa %>% signif(2) %>% range %>% paste(collapse = ' to ')`), especially when it comes to assignment of cancer samples to LumU class. 
Please note, that this later observation corroborates findings presented in the initial manuscript, that, except of _KL_ gene, there are no good FGFR/FGF/FGFBP-related markers of LumU cancers.

For the Elastic Net and Random Forest models of the consensus MIBC classes using the full gene set by Kamoun et al. SHAP importance measures of the explanatory variables were computed. 
As shown in __Supplementary Figure ZZZZ__, except of _CD44_ whose levels were highly predictive for Ba/Sq cancers, none of the FGFR/FGF/FGFBP-related genes present in the genuine consensus class signature (`r intersect(ml_globals$variables, mibc_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`) was found to be highly influential for the MIBC consensus class forecasts.

## Point 5

### Issue

For the doses used for the cell culture experiments, do we know which concentration corresponds to a level that we would see in patients when using erdafitinib?

### Response

Thank you for this important comment. 
In in vitro experiments we used 0 to 10 µM erdafitinib, which corresponds to around 0 to 4.5 mg/L erdafitinib. 
The recommended dose for urothelial carcinoma patients with susceptible genetic alterations in the FGFR is 8 mg orally once daily [@FDA2024]. 
Treating 2D cell lines cannot be compared directly to administration in humans due to the complex differences in pharmacokinetics. 
Therefore, more research, in particular in pre-clinical animal models, is needed that can translate our findings to the clinical setting. 
However, the erdafitinib concentration range in our study is commonly used in in vitro drug screenings, and has also been used in multiple reports on erdafitinib in urothelial cancer [@Ohtsu2022; @Li2023; @Zhong2024; @Zhong2024].

## Point 6

### Issue

Isn't it plausible there is something co-expressed with the FGFR pathways that could be the true oncogenic driver? 
Or be the real cause of the different phenotypes seen? 
I'm not sure this study can prove that is not the case so this may need to be acknowledged as a limitation.

### Response

We agree with you. 
By means of statistical inference and modeling, we cannot prove causality, i.e. unequivocally demonstrate that FGFR signaling drives bladder cancer in general. 
For this reason, we re-formulated carefully the title, abstract, and conclusions of the manuscript and abstain from the 'driver' term. 
Instead, we underline diversity of activation mechanisms of FGFR signaling in urothelial cancers and which lets us advocate for a broader use of FGFR blockers, independently of _FGFR3_ mutations.

Nevertheless, inspired by you comment, we sought to carefully gauge activity of FGFR signaling by comparison of scores [@Hanzelmann2013] of `r length(sub_sig$variables)` FGFR-related gene signatures from MSig database between the consensus classes of MIBC. 
Please note, that most of those signatures included ligand and co-receptor genes, and may be hence regarded as a proxy of ligand- and co-receptor-dependent FGFR signaling. 
Interestingly, `r sub_sig$hm_plots$tcga$data %>% filter(variable_subset == 'Stroma-rich', !duplicated(variable)) %>% nrow` of those signatures were found to be significantly upregulated in stroma-rich cancers in at least two out of TCGA BLCA, IMvigor, and BCAN cohorts (__Supplementary Figure VVV__, __Supplementary Table VVV__).

## Point 7

### Issue

I understand this is a brief correspondence, but there is no mention of the limitations of this work in the manuscript.

### Response

To address this point, we have added the limitation section to the manuscript. 
As the limitations we list:

* incompleteness of the clinical and multi-omics data for the GENIE and MSK IMPACT cohorts (lacking transcriptome data), which precluded analyses of transcriptome, gene signatures, and drug response predictions

* observational character of our study based on bioinformatic analyses, which require validation in a real-world clinical setting

# Reviewer 2

## Point 8

### Issue

As a small suggestion, Authors should only briefly describe and further discuss the significant clinical impact of FGFR inhibitors in the current clinical practice.

### Response

Thanks for this constructive suggestion. 
We appended the introduction with a section on clinical relevance and important ongoing studies on FGFR inhibitors in urothelial carcinoma.
We also discuss that there is already compelling evidence that also patients without FGFR alterations may benefit from treatment with FGFR inhibitors.

# Reviewer 3

## Point 9

### Issue

Limitations: could the Authors add possible limitations of this study?

### Response

Following your and Reviewer 1 suggestion we included a listing of the following limitations in the final part of the text:

* incompleteness of the clinical and multi-omics data for the GENIE and MSK IMPACT cohorts (lacking transcriptome data), which precluded analyses of transcriptome, gene signatures, and drug response predictions

* observational character of our study based on bioinformatic analyses, which require validation in a real-world clinical setting

\newpage

# Tables for Reviewers

```{r tab-cell-lines, tab.cap = "Demographic, clinical, and pathological characteristic of donors of the DepMap urothelial cancer cell lines. Numeric variables are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages of the categories."}

flextable::flextable(rev_tabs$cell_lines) %>% 
  set_widths(c(4.5, 5.5)) %>% 
  theme_vanilla

```

\newpage

# References

### 2025-03-20 paper scripts/links.R

@@ -1,0 +1,35 @@
+# Handling of common links in the Rmarkdown project
+
+  insert_head()
+
+# container ------
+
+  proj_links <- list()
+
+# Links to development packages --------
+
+  insert_msg('Links to development packages')
+
+  proj_links <-
+    list('ExDA' = 'https://github.com/PiotrTymoszuk/ExDA',
+         'trafo' = 'https://github.com/PiotrTymoszuk/trafo',
+         'figur' = 'https://github.com/PiotrTymoszuk/figur',
+         'clustTools' = 'https://github.com/PiotrTymoszuk/clustTools',
+         'microViz' = 'https://github.com/PiotrTymoszuk/microViz',
+         'polcaExtra' = 'https://github.com/PiotrTymoszuk/polcaExtra',
+         'coxExtensions' = 'https://github.com/PiotrTymoszuk/coxExtensions',
+         'fastTest' = 'https://github.com/PiotrTymoszuk/fastTest',
+         'perich' = 'https://github.com/PiotrTymoszuk/perich',
+         'graphExtra' = 'https://github.com/PiotrTymoszuk/graphExtra',
+         'htGLMNET' = 'https://github.com/PiotrTymoszuk/htGLMNET') %>%
+    compress(names_to = 'obj_name',
+             values_to = 'x') %>%
+    mutate(ref_name = paste0('_', obj_name, '_'))
+
+  proj_links <- proj_links[c('ref_name', 'x')] %>%
+    pmap(mdlink) %>%
+    set_names(proj_links$obj_name)
+
+# END ------
+
+  insert_tail()

### 2025-03-20 paper scripts/render.R

@@ -18,6 +18,11 @@
                                         reference_docx = 'ms_template.docx'),
            output_dir = './report')
 
+  render('./report/markdown/rebuttal_letter.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './report')
+
 # END ----
 
   insert_tail()

### 2025-03-20 paper scripts/reviewer_tables.R

@@ -1,0 +1,35 @@
+# Tables for the Reviewers
+
+  insert_head()
+
+# container -------
+
+  rev_tabs <- list()
+
+# characteristic of the cell lines ------
+
+  insert_msg('Characteristic of the cell lines')
+
+  rev_tabs$cell_lines <- expl_cecli$result_tbl %>%
+    map_dfc(stri_replace_all, fixed = 'NaN%', replacement = 'not available') %>%
+    as_mdtable(label = 'cell_lines',
+               ref_name = 'cell_lines',
+               caption = paste('Demographic, clinical, and pathological',
+                               'characteristic of donors of the DepMap',
+                               'urothelial cancer cell lines.',
+                               'Numeric variables are presented as medians',
+                               'with interquartile ranges and ranges.',
+                               'Qualitative variables are shown as percentages',
+                               'of the categories.'))
+
+# Saving the tables on the disc --------
+
+  insert_msg('Saving the tables on the disc')
+
+  rev_tabs %>%
+    save_excel(path = './report/reviewer_tables.xlsx',
+               prefix = 'Reviewer Table ')
+
+# END ------
+
+  insert_tail()

### 2025-03-20 subtype scripts/drugs.R

@@ -310,7 +310,7 @@
          ~.x +
            scale_y_discrete(labels = set_names(.y$hm_label, .y$variable)))
 
-  sub_drugs$hm_plots$sub_drugs <-
+  sub_drugs$hm_plots$fgfr_drugs <-
     map2(sub_drugs$hm_plots$fgfr_drugs,
          sub_drugs$kruskal,
          ~.x +
@@ -355,7 +355,7 @@
 
   sub_drugs$result_tbl <-
     map2(sub_drugs$stats,
-         sub_drugs$test %>%
+         sub_drugs$kruskal %>%
            map(~.x[c('variable', 'title_label', 'significance', 'eff_size')]),
          left_join, by = 'variable') %>%
     compress(names_to = 'cohort') %>%

### 2025-03-23 report/markdown/rebuttal_letter.Rmd
Changes in markdown
Changes in markdown of Supplementary Material and Rebuttal Letter

@@
 In __Supplementary Table S1__ of the revised manuscript, we present frequencies of patients with non-invasive intra-vesical treatment, surgery and distribution of surgery types, neoadjuvant treatment, adjuvant systemic treatment, systemic chemo- and immunotherapy, and radiation.
+We show a clinical and pathological characteristic of the consensus classes in __Supplementary Table S2__. 
 For most of the bulk cancer data sets (GENIE, MSK IMPACT, IMvigor, BCAN), this information was published along with the molecular data or extracted from the original publications. 
@@
 [@Geeleher2014] and successfully employed in our recent publication on testicular cancer [@Torzsok2025].
-Subsequently, we compared the predicted IC50 and AUC resistance metrics between the MIBC consensus classes (__Supplementary Figure XXX___). 
+Subsequently, we compared the predicted IC50 and AUC resistance metrics between the MIBC consensus classes (__Supplementary Figure S9___, __Supplementary Table S9__). 
 These new data shows that for most of pan-FGFR inhibitors (AZD4547, FGFR3831, FGFR0939, erdafitinib), good sensitivity was predicted for the LumP class, the predicted response was highly heterogeneous in the renaming subsets.
-In^particular, for Ba/Sq cancers resistance was forecasted, while stroma-rich were put forward as erdafitinib-responsive malignancies. 
-We also compared the resistance metrics of FGFR-targeting drugs for cancers with and without mutation of _FGFR3_ (__Supplementary Figure XXX__); in this case, no differences in predicted response between WT and mutant specimens were observed.
+In particular, for Ba/Sq cancers resistance was forecasted, while stroma-rich were put forward as erdafitinib-responsive malignancies. 
+Interestingly, in a comparison of the predicted resistance to pan-FGFR inhibitors between _FGFR3_ wild-type and mutant cancers presented in __Supplementary Figure S10__, expect for AZD4547, no significant differences were observed.
+This data provides another hint that response to FGFR blockers may not be restricted to patients with genetic alterations of the _FGFR_ genes.
 
@@
 In evaluation of model performance, the MIBC consensus class assignment by the genuine nearest centroid tool by Kamoun et al. served as the ground truth. 
-As presented in __Supplementary Figure YYY__ predictions of consensus classes by the models with the full set of predictors by Kamoun et al.
-were emulated the original classification tool with very high accuracy (test collectives, accuracy: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Accuracy %>% signif(2) %>% range %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Kappa %>% signif(2) %>% range %>% paste(collapse = ' to ')`).
+As presented in __Supplementary Figures S11__ and __S12__, and __Supplementary Table S11__  predictions of consensus classes by the models with the full set of predictors by Kamoun et al.
+emulated the original classification tool with very high accuracy (test collectives, accuracy: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Accuracy %>% signif(2) %>% range %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r mibc_comp$stats %>% filter(model == 'mibc_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Kappa %>% signif(2) %>% range %>% paste(collapse = ' to ')`).
 Their performance was generally better that accuracy of the models employing solely the FGFR/FGF/FGFBP-related genes (test collectives, accuracy: `r mibc_comp$stats %>% filter(model == 'fgfr_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Accuracy %>% signif(2) %>% range %>% paste(collapse = ' to ')`, Cohen's $\kappa$: `r mibc_comp$stats %>% filter(model == 'fgfr_genes', !cohort %in% c('tcga', 'tcga_train'), algorithm != 'ensemble') %>% .$Kappa %>% signif(2) %>% range %>% paste(collapse = ' to ')`), especially when it comes to assignment of cancer samples to LumU class.
@@
 For the Elastic Net and Random Forest models of the consensus MIBC classes using the full gene set by Kamoun et al. SHAP importance measures of the explanatory variables were computed. 
-As shown in __Supplementary Figure ZZZZ__, except of _CD44_ whose levels were highly predictive for Ba/Sq cancers, none of the FGFR/FGF/FGFBP-related genes present in the genuine consensus class signature (`r intersect(ml_globals$variables, mibc_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`) was found to be highly influential for the MIBC consensus class forecasts.
+As shown in __Supplementary Figures S13__ - __S15__, and __Supplementary Table S12__, except of _CD44_ whose levels were predictive for Ba/Sq cancers, none of the FGFR/FGF/FGFBP-related genes present in the genuine consensus class signature (`r intersect(ml_globals$variables, mibc_globals$variables) %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`) was found to be highly influential for the MIBC consensus class forecasts.
 
@@
 Please note, that most of those signatures included ligand and co-receptor genes, and may be hence regarded as a proxy of ligand- and co-receptor-dependent FGFR signaling. 
-Interestingly, `r sub_sig$hm_plots$tcga$data %>% filter(variable_subset == 'Stroma-rich', !duplicated(variable)) %>% nrow` of those signatures were found to be significantly upregulated in stroma-rich cancers in at least two out of TCGA BLCA, IMvigor, and BCAN cohorts (__Supplementary Figure VVV__, __Supplementary Table VVV__).
+Interestingly, `r sub_sig$hm_plots$tcga$data %>% filter(variable_subset == 'Stroma-rich', !duplicated(variable)) %>% nrow` of those signatures were found to be significantly upregulated in stroma-rich cancers in at least two out of TCGA BLCA, IMvigor, and BCAN cohorts (__Supplementary Figure S8__, __Supplementary Table S8__).
 

### 2025-03-23 report/markdown/figure1_legend.Rmd
Changes in markdown
Changes in markdown of Supplementary Material and Rebuttal Letter

---
title: "FGFR1/3 Signaling as Achilles’ Heel of Phenotypic Diversity in Urothelial Carcinoma"
subtitle: "Figure 1 Legend"
author: "Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: blca.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "cairo_pdf", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 10)

```

\newpage

```{r fig-figure-1a, fig.width = figur::convert(fig$figure_1a, to = 'in')$w, fig.height = figur::convert(fig$figure_1a, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in MIBC consensus molecular classes of urothelial cancer.'}

fig$figure_1a$plot

```

__Figure 1A. Alterations of FGFR-, FGF-, and FGFBP-coding genes in MIBC consensus molecular classes of urothelial cancer.__

_Cancer samples from the TCGA BLCA, IMvigor, and BCAN data sets were assigned to the MIBC consensus molecular classes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC). Differences in frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes between the LumP (luminal papillary), LumU (luminal unstable), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes were assessed by weighted permutation test. P values were corrected for multiple testing with the false discovery rate method. Presence/absence of selected genetic alterations in the consensus molecular classes in the TCGA BLCA cohort was visualized in a oncoplot. Alteration names (mut: mutation, amp: amplification, del: deletion) with effect sizes and p values of differences between the consensus classes are indicated in the Y axis. Significant effects are highlighted with bold font. Numbers of cancer samples in the consensus classes are displayed in the plot caption._

\newpage

```{r fig-figure-1b, fig.width = figur::convert(fig$figure_1b, to = 'in')$w, fig.height = figur::convert(fig$figure_1b, to = 'in')$h, fig.cap = 'Mutational hot spots of the FGFR3 gene.'}

fig$figure_1b$plot

```

__Figure 1B. Mutational hot spots of the FGFR3 gene.__

_Frequency of mutations affecting residues of the FGFR3 protein was expressed as percentage of all cancer samples in the cohort and displayed in dot plots. Each point represents a single protein residue, point color codes for assignment of the residue to the protein domain. Candidate mutation hot spots with mutation rates $\geq$ 1% are highlighted with enlarged symbols and labeled with percentages and protein sequence changes. Cohort names and numbers of analyzed cancer samples are indicated in the plot facets. The protein domain scheme is displayed above the plot._

\newpage

```{r fig-figure-1c, fig.width = figur::convert(fig$figure_1c, to = 'in')$w, fig.height = figur::convert(fig$figure_1c, to = 'in')$h, fig.cap = 'Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.'}

fig$figure_1c$plot

```

__Figure 1C. Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.__

_Levels of FGFR-, FGF-, and FGFBP-coding mRNA were determined in cancer tissue samples in the TCGA BLCA, IMvigor, and BCAN by RNA sequencing and published as $log_2$ gene counts (TCGA BLCA and IMvigor) or Z-scores of $log_2$ gene counts (BCAN). mRNA expression of the genes of interest in DepMap urothelial cancer cell lines was measured by RNA sequencing and provided as $log_2$ gene counts. Median Z-scores of $log_2$ mRNA expression are shown in a heat map. Each tile represents median expression Z-score of a gene in one cohort. Numbers of cancer samples are indicated in the X axis. White tiles represent genes not detected in each cohort. Protein expression of `r expl_expr$plot_data$hpa %>% filter(complete.cases(.)) %>% .$variable %>% unique %>% length` FGFR-, FGF-, and FGFBP-coding genes in the Human Protein Atlas (HPA) repository was quantified by immunohistochemistry (IHC). The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%). Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges. Single IHC specimens are visualized as points. Numbers of IHC samples are indicated in the Y axis._

\newpage

```{r fig-figure-1d, fig.width = figur::convert(fig$figure_1d, to = 'in')$w, fig.height = figur::convert(fig$figure_1d, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.'}

fig$figure_1d$plot

```

__Figure 1D. Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.__

_Co-expression networks were constructed in the TCGA BLCA, IMvigor, and BCAN cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman’s $\rho \geq$ 0.3. Isolated vertices of the networks were removed. The networks were visualized using the Fruchterman-Reingold algorithm. Points represent node vertices/genes, point shape codes for classification of the protein product. Point color represent hub scores, which measure the overall correlation strength. Lines represent edges, i.e correlations between expression levels with $\rho \geq$ 0.3. Line widths correspond to $\rho$ values. Note the central position, connectivity, and high hub scores of the genes coding for FGFR1, its ligands FGF2, FGF7 and FGF10, and the binding proteins SDC2, DCN, PTX3 and HSPG2 preserved in all cohorts. This suggests that ligand-dependent signaling via FGFR1 may serve as a substantial signaling pathway in urothelial cancer._

\newpage

```{r fig-figure-1e, fig.width = figur::convert(fig$figure_1e, to = 'in')$w, fig.height = figur::convert(fig$figure_1e, to = 'in')$h, fig.cap = 'Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}

fig$figure_1e$plot

```

__Figure 1E. Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.__

_Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes (Figure 1A) were assessed by one-way ANOVA with $\eta^2$ effect size statistic. Differences in $log_2$-transformed gene expression between a class and LumP were investigated by two-tailed T test. P values were corrected for multiple testing with the false discovery rate method (FDR). Significantly differentially regulated genes were defined by p(ANOVA) < 0.05, $\eta^2$ 0.06, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation in a class as compared with LumP cancers. Z-scores of $log_2$ expression levels in the TCGA BLCA cohort of significantly differentially regulated genes shared by at least two cohorts are displayed in a heat map. Each tile represents a single cancer sample. Significant effects are labeled with bold font in the Y axes. Numbers of cancer samples in the consensus classes are indicated in the plot caption._

\newpage

```{r fig-figure-1f-elnet, fig.width = figur::convert(fig$figure_1f_elnet, to = 'in')$w, fig.height = figur::convert(fig$figure_1f_elnet, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.'}

fig$figure_1f_elnet$plot

```

__Figure 1F. Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.__

_Contributions of the explanatory variables, i.e. the FGFR-, FGF-, and FGFBP-coding genes, to prediction of LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by an Elastic Net machine learning model were estimated with the SHAP algorithm (Shapley additive explanations). SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots. Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene. The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots. Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._

\newpage

```{r fig-figure-1f-ranger, fig.width = figur::convert(fig$figure_1f_ranger, to = 'in')$w, fig.height = figur::convert(fig$figure_1f_ranger, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.'}

fig$figure_1f_ranger$plot

```

__Figure 1F. Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.__

_Contributions of the explanatory variables, i.e. the FGFR-, FGF-, and FGFBP-coding genes, to prediction of LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by a Random Forest machine learning model were estimated with the SHAP algorithm (Shapley additive explanations). SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots. Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene. The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots. Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._

\newpage

```{r fig-resistance, fig.width = figur::convert(fig$figure_1g, to = 'in')$w, fig.height = figur::convert(fig$figure_1g, to = 'in')$h, fig.cap = 'Resistance and sensitivity to pan-FGFR inhibitors of DepMap cancer urothelial cell lines.'}

fig$figure_1g$plot

```

__Figure 1G. Resistance and sensitivity to pan-FGFR inhibitors of DepMap cancer urothelial cell lines.__

_Information on resistance of urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (n = `r expl_res$sensitive$gdsc %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')` cell lines) and PRISM (n = `r expl_res$sensitive$prism %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')`) drug screening experiments was extracted from the DepMap repository. The resistance values were expressed as IC50 (concentration resulting in 50% growth inhibition, GDSC) or AUC (area under the dose-response curve, relative to DMSO control, PRISM). Significant sensitivity for the GDSC data was defined as IC50 below 10 µM, which was the maximum concentration of the FGFR inhibitors. Significant sensitivity for the PRISM data was defined as AUC < 1, which indicates stronger growth inhibition by a FGFR inhibitor than the DMSO control. IC50 and AUC for doxorubicin are presented as a positive cytotoxicity controls. Median IC50 and AUC values for the FGFR inhibitors with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges. Points represent single cell lines. Point symbols codes for the FGFR3 gene mutation status, point color represent sensitivity and resistance. The sensitivity cutoffs are represented by dashed lines. FGFR3 wild-type cell lines RT-112, RT4, SW 780, CAL-29, and J82 showed particularly high sensitivity to multiple FGFR blockers in the GDSC1/2 experiments. FGFR3 wild-type cell lines `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$name_lab %>% collapse_and` demonstrated a very good sensitivity to erdafitinib in the PRISM assay (AUC: `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`). By contrast, `r expl_res$sensitive$prism %>% filter(.data[['PRISM_BRD-K84868168-001-01-0_erdafitinib']] == 'resistant') %>% .$sample_id %>% exchange(depmap$cell_lines, key = 'sample_id', value = 'cell_line_name') %>% collapse_and` cell lines were found to be resistant to erdafitinib with AUC values ranging from `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), status == 'resistant') %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`._

\newpage

```{r fig-figure-1h, fig.width = figur::convert(fig$figure_1h, to = 'in')$w, fig.height = figur::convert(fig$figure_1h, to = 'in')$h, fig.cap = 'Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.'}

fig$figure_1h$plot

```

__Figure 1H. Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.__

_Confluence was measured after 96 hours in cultures treated with erdafitinib at 0, 1, 4, 7 and 10 µM. Dots represent means of three biological replicates. Mean values with standard deviations are depicted as bars with whiskers._

### 2025-03-23 report/markdown/supplementary_material.Rmd
Changes in markdown
Changes in markdown of Supplementary Material and Rebuttal Letter

---
title: "FGFR1/3 Signaling as Achilles’ Heel of Phenotypic Diversity in Urothelial Carcinoma"
subtitle: "Supplementary Material"
author: "Department of Urology, Medical University of Innsbruck"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::html_document2:
    css: "style.css"

bibliography: blca.bib

csl: frontiers_medical.csl

header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}

---

```{r, setup, include = FALSE}

library(Cairo)

opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE, 
                      dev = "png", 
                      dpi = 600)

set_flextable_defaults(font.family = "Cambria", 
                       font.size = 9)

```

\newpage

# Supplementary Methods

## Software

Data handling and analysis was accomplished with R version 4.2.3. 
For general transformation tasks for matrix, tabular, and text data, base R, _tidyverse_ package bundle [@Wickham2019], _rlang_ [@Henry2022], [_trafo_](https://github.com/PiotrTymoszuk/trafo), and _stringi_ [@Gagolewski2021] were utilized. 
General data visualization tasks were performed with _ggplot2_ [@Wickham2016], _cowplot_ [@Wilke2019], and _flextable_ [@Gohel2022].

## Data sources

Five genomic and transcriptomic collectives of urothelial carcinoma patients were evaluated in the current report: 
the urothelial cancer subset of the GENIE cohort referred to as GENIE BLCA [@Sweeney2017], 
the MSK IMPACT collective [@Clinton2022], 
the bladder cancer TCGA collective of muscle-invasive urothelial cancers referred to as TCGA BLCA [@Liu2018; @Robertson2017], 
the IMvigor cohort of chemotherapy-resistant cancers treated subsequently with immunotherapy [@Balar2017], and 
the BCAN cohort of metastatic cancer patients administered systemic chemotherapy [@Damrauer2022]. 
Except of the GENIE BLCA cohort, for which only rudimentary clinical information was available, these collectives were restricted to muscle invasive cancers of the bladder (MIBC) with the invasiveness information provided by the study authors or by selecting T2 - T4 stage cancers.

The GENIE BLCA cohort [@Sweeney2017] consisted of n = `r nrow(genie$clinic)` cases of bladder cancers selected by oncotree 'BLCA' term. 
The available data sets for the GENIE BLCA cohort included basic clinical information (age at sequencing in years, sex, follow-up in days, mortality), total mutation burden (TMB) expressed as fraction of genome, detailed information on location, variant type, nucleotide sequence and protein residues affected by somatic mutations, as well as 0/1 coded absence/presence of gene deletions and amplifications. 
Approximate overall survival was computed as the time interval between the last follow-up in days and age at sequencing in years. 
In case of multiple samples per participant, the earliest sample, i.e. the specimen with the lowest ID, was included in the current analysis. 
The GENIE BLCA data sets were fetched from the [cBioportal API](https://www.cbioportal.org/) with tools provided by the _cbioportalR_ package [@Whiting2024] and formatted with in-house developed R scripts.

The MSK IMPACT collective consisted of n = `r nrow(msk$clinic)` patients restricted to muscle-invasive cancers of the bladder [@Clinton2022]. 
In case of multiple samples available for a patient, the sample with the lowest ID was included in the analysis. 
The clinical information included entries on participants age, gender, overall survival, cancer pathological stage, and cancer treatment. 
The mutation data set included detailed information on mutation nucleotide sequence, variant type, and affected protein residues. 
The copy number variant data were provided as 0/1 coded presence of gene deletions and amplifications. 
The data sets for the MSK impact study were downloaded from [cBioportal](https://www.cbioportal.org/) and implemented in R with in-house developed scripts.

Data sets for the pan-cancer version of TCGA BLCA cohort [@Liu2018; @Robertson2017] were fetched from the [cBioportal](https://www.cbioportal.org/) repository with in-house developed scripts for n = `r table(tcga$expression$tissue_type)['tumor']` muscle-invasive bladder cancer patients. 
The clinical information consisted of demographic variables (e.g. age, sex, BMI), pathological features (e.g. pathological staging, histology, grade, extracapsular extension), survival information (overall, tumor-specific and relapse-free survival), and TMB expressed as mutation count per MB. 
Information on non-invasive treatment, surgery and smoking was extracted from the seminal publication by Robertson et al. [@Robertson2017]. 
Data on chemo-, immuno- and radiotherapy were derived from the manually curated table by Moiso [@Moiso2021]. 
The somatic mutation data set included detailed information on location, variant type, nucleotide sequence and affected protein residues. 
The copy number variant data set consisted of 0/1 coded presence of gene amplifications and deletions. 
The expression data set for the TCGA BLCA collective consisted of $log_2(x + 1)$-transformed RSEM-normalized  gene-integrated transcript counts obtained by bulk cancer RNA sequencing.

The IMvigor data set included a subset of `r nrow(imvigor$expression)` muscle-invasive bladder cancers of the IMvigor210 trial subjected to RNAseq and DNA sequencing [@Balar2017; @Mariathasan2018]. 
The study participants were treated with an anti-PD-L1 drug (atezolizumab) following chemotherapy. 
The clinical data set included basic demographic, clinical and pathological features, best overall response (complete/partial response [CR/PR], stable/progressive disease [SD/PD]), overall survival, and TMB quantified as mutation count per MB. 
The mutation data set consisted of 0/1 coded information on presence of any somatic mutations in a particular gene. 
The expression data set consisted of $log_2(x + 1)$-transformed gene-integrated transcript counts obtained by bulk cancer RNA sequencing. 
The IMvigor data sets were obtained from the R package _IMvigor210CoreBiologies_.

The BCAN study information was extracted from [cBioportal](https://www.cbioportal.org/) and consisted of demographic and clinical information, staging of primary tumors, overall and disease-specific survival, RNA sequencing expression Z scores of bulk cancer samples, as well as detailed somatic mutation data of bulk cancer specimens. 
This collective consisted of metastatic cancers, treated in most patients by systemic chemotherapy. 
For the analysis in the current report, the collective was restricted to muscle invasive primary cancers of the bladder. 
Missing values in the Z-score expression data set of the BCAN cohort were imputed with the 9-nearest neighbor algorithm implemented by function `impute.knn()` from _impute_ package [@Hastie2023].

Demographic, clinical, and pathological characteristic of the cohorts is shown in __Supplementary Table S\@ref(tab:tab-cohorts)__.

For MIBC molecular class assignment and machine learning modeling tasks, $log_2$-transformed gene expression levels were processed by ComBat (R package _sva_ and [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)) [@Leek2012] to correct for the inter-cohort variability. 
Assignment of the TCGA BLCA, IMvigor, and BCAN cancer samples to consensus molecular classes was done with the nearest centroid algorithm fed with ComBat-normalized whole-genome $log_2$ mRNA or Z-score expression levels (R package _consensusMIBC_) [@Kamoun2020]. 
Specimens without unequivocal assignment ('not-assigned') were excluded from further analyses. 
Please note that due to insufficient numbers of specimens assigned to the 
luminal non-specified (LumNS; `r subtypes$assignment %>% map(filter, consensusClass == 'LumNS') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`samples) and 
neuroendocrine-like classes (NE-like; `r subtypes$assignment %>% map(filter, consensusClass == 'NE-like') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
we restricted our analyses to the major classes, i.e. 
luminal papillary (LumP; `r subtypes$assignment %>% map(filter, consensusClass == 'LumP') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
luminal unstable (LumU; `r subtypes$assignment %>% map(filter, consensusClass == 'LumU') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), 
stroma-rich (`r subtypes$assignment %>% map(filter, consensusClass == 'Stroma-rich') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), and 
basal/squamous-like classes (Ba/Sq; `r subtypes$assignment %>% map(filter, consensusClass == 'Ba/Sq') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`). 
Demographic, clinical, and pathological characteristic of the consensus molecular classes of MIBC is presented in __Supplementary Table S\@ref(tab:tab-sub-clinic)__.

Gene signatures related to FGFR (n = `r sig$lexicon %>% nrow` features) were retrieved from a local copy of the MSig database, version 7.5.1, by a search with the `'FGF|FGFR'` regular expression. 
The search results were filtered for Reactome, PID, KEGG, and WP signatures by selection with the `'^(REACTOME|PID|KEGG|WP)'` regular expression. 
Single sample gene set enrichment analysis (ssGSEA) scores were computed for TCGA BLCA, IMvigor, and BCAN transcriptome specimens with the GSVA algorithm by Hänzelmann et al. [@Hanzelmann2013] re-implemented by open-source package [_gseaTools_](https://github.com/PiotrTymoszuk/gseaTools) (function `calculate()`).

Predictions of resistance to pan-FGFR inhibitors in the TCGA BLCA, IMvigor, and BCAN transcriptomic cohorts in form of IC50 (drug concentration resulting in 50% growth inhibition, GDSC1, GDSC2, CTRP2) and AUC (area under the dose-response curve, PRISM) were made by RIDGE regularized linear models trained with the drug response data and baseline transcriptome of epithelial cancer cell lines in GDSC1, GDSC2, CTRP2, and PRISM screening experiments [@Yang2013; @Seashore-Ludlow2015; @Corsello2020]. 
The general idea of modeling drug resistance in bulk cancer transcriptome samples by RIDGE models trained with in vitro drug response data was proposed by Geeleher and colleagues [@Geeleher2014] and used in our recent publication on testicular cancer [@Torzsok2025]. 
The models were trained with ComBat-processed $log_2$ baseline expression values of gene shared by the cell line and bulk cancer data sets. 
The IC50 values were $log$-transformed to improve normality. 
The pre-processing of expression and in vitro drug response data was accomplished by `multi_process()` function from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
Model tuning, i.e. choice of the shrinkage parameter $\lambda$ based on the minimal mean squared error in 10-fold cross-validation, and training in the cell line data sets was accomplished by function `train()` from [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET) package. 
The RIDGE models of IC50 and AUC characterized by cross-validated $R^2 \geq$ 0.13 and correlation between the predicted and observed drug response with Pearson's r $\geq$ 0.5 were subsequently used for predictions of the resistance for bulk cancer transcriptome samples. 
Those criteria were fulfilled for `r drugs$fgfr_drugs %>% length` pan-FGFR inhibitors, i.e. inhibitors targeting at least FGFR1/2/3. 
Those substances were: `r list(x = drugs$fgfr_drugs %>% exchange(drugs$lexicon), y = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'experiment'), z = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'drug_id')) %>% pmap_chr(function(x, y, z) paste0(x, ' (', y, ', ID: ', z, ')')) %>% collapse_and`. 
The predictions of IC50 and AUC were made by calling `predict()` method for the object returned by `train()` function (package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).

The following genes coding for FGFR, FGFF, and FGFBP (FGF-binding proteins) were investigated in the current report [@Korc2009; @Turati2022; @Bennett1995; @Gubbiotti2016; @Mundhenke2002; @Midorikawa2003; @Leali2012; @Xie2020; @Hagihara2000; @Pan2021; @Mohammadi2005; @Hu2009a] (__Supplementary Table S\@ref(tab:tab-genes)__):

* FGF genes: `r globals$ligands %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`

* FGFR genes: `r globals$receptors %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`

* FGFBP genes: `r globals$binding_proteins %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`

Those genes were returned by a systematic search of literature at [PubMed](https://pubmed.ncbi.nlm.nih.gov/) as well as a screening of physical interaction partners of FGFR and FGF proteins at [UniProt](https://www.uniprot.org/), [STRING](https://string-db.org), [BioGRID](https://thebiogrid.org), and [IntAct](https://www.ebi.ac.uk/intact).

JSON files with domain structures of FGFR1 (ID: P11362), FGFR2 (ID: P21802), FGFR3 (ID: P22607), and FGFR4 (ID: P22455) proteins were downloaded from [UniProt](https://www.uniprot.org) and processed with the _jsonlite_ package [@Ooms2023].

Mutation status, baseline expression measured by RNA sequencing, and effect scores of CRISP/Cas9 and RNA interference of the FGFR-, FGF-, and FGFBP-coding genes in urothelial cancer cell lines were downloaded from the [DepMap](https://depmap.org/portal/) repository [@DepMap2024]. 
No information treatment of the cell line donors or pre-treatment of the cell lines was available. 
The available cell lines were `r map2_chr(depmap$cell_lines$cell_line_name, depmap$cell_lines$sample_id, ~paste0(.x, ' (ID:', .y, ')')) %>% collapse_and`. 
The mutation status was 0/1 coded, where 1 stood for presence of at least one somatic mutation in a gene. 
RNA sequencing information was provided as $log_2$ TPM. 
Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013]. 
The resistance was measured in the GDSC1/2 experiments by IC50; sensitive cell lines were defined by IC50 < 10 µM, which was the highest concentration of the FGFR inhibitors in the assays. 
The resistance in the PRISM assays was gauged by AUC normalized for effects of DMSO control; sensitive cell lines were considered for AUC < 1.

For verification of expression of the FGFR-, FGF-, and FGFBP-coding genes at protein level, we resorted to the [Human Protein Atlas (HPA) repository](https://www.proteinatlas.org/), which stores results of immunohistochemical (IHC) staining. 
Entries of the HPA database were retrieved as XML documents, from which patient information, staining quantification, and URLs to IHC images for the genes of interest was retrieved (R package _xml2_) [@Wickham2023]. 
Protein expression in the IHC samples was expressed as the IHC score defined as a product of staining (1: negative to 4: high), staining intensity (1: negative to 4: strong), and quantity of positive cells (1: none to 4: >75%).

## Cell culture

UM-UC-3 cells were obtained from Sigma-Aldrich and cultured in Dulbecco's modified Eagle's medium, low glucose (PAN-Biotek, Aldenbach, Germany). 
5637 cells were gifted by Prof. Holm and cultured in RPMI 1640 medium (PAN-Biotek). 
RT112 cells were obtained from the German Collection of Microorganisms and Cell Cultures GmbH (DSMZ) and cultured in RPMI 1640 medium. 
UM-UC-6 and UM-UC-14 cell lines were gifted by Maxim Noeparast and cultured in Minimum Essential Medium (Gibco, Thermo Fisher Scientific) with the addition of 1% non-essential amino acids (PAN-Biotech). 
For all cell lines, medium was supplied with 10% fetal calf serum (PAN-Biotech), 1% penicillin/streptomycin (Lonza), and 1% GlutaMAX (Gibco, Thermo Fisher Scientific). 
Cell cultures were routinely tested for mycoplasma.

## In vitro erdafitinib treatment and proliferation assay

The UM-UC-3, 5637, RT112, UM-UC-6 and UM-UC-14 cell lines were seeded into 96 well plates with a density of respectively 4000, 8000, 10000, 8000 and 8000 cells per well, and treated with 0, 1, 4, 7 or 10 µM erdafitinib. 
They were cultured for 96 hours at 37°C and 5% CO2 in an IncucyteS3© live imaging system. 
During the culture period, images were made of the individual wells, and confluence was measured using the AI-generated Incucyte© Base Analysis Software. 
For the three FGFR1/2/3/4 wild-type cell lines, UM-UC-3, 5637 and RT112, images were made every 4 hours and the confluence trajectories for concentrations of erdafitinib were plotted against time by the Incucyte© Base Analysis Software.

## Descriptive statistics, statistical inference, and effect size

If not indicated otherwise, descriptive statistics for quantitative variables included median, interquartile range, and range. 
Descriptive statistic for qualitative variables were percentage and count within categories. 
Differences of means or medians of qualitative variables were investigated by Mann-Whitney, two-tailed T test, Kruskal-Wallis test, and one-way ANOVA, depending on the number of analysis groups and normality of distribution of the variable. 
Correlation was assessed by Pearson's and Spearman's correlation test. 
Effect size of differences in variable levels, distribution, and correlation was assessed with the following statistics [@Cohen2013]:

* biserial r for differences between two groups, non-parametric tests: small for r < 0.3, medium for 0.3 $\leq$ r < 0.5, large for r $\geq$ 0.5

* Cohen's d for differences between two groups, parametric tests: small for d < 0.5, medium for 0.5 $\leq$ d < 0.8, large for d $\geq$ 0.8

* $\eta^2$ for differences between three or more groups: small for $\eta^2$ < 0.06, moderate for 0.06 $\leq$ $\eta^2$ < 0.14, large for $\eta^2 \geq$ 0.14

* Cramer's V for comparison of category frequencies of a numeric variable: small for V < 0.3, medium for 0.3 $\leq$ V < 0.5, large for V $\geq$ 0.5

* Pearson's r for correlation, parametric tests: small for r < 0.3, medium for 0.3 $\leq$ r < 0.5, large for r $\geq$ 0.5

* Spearman's $\rho$ for correlation, non-parametric tests: small for $\rho$ < 0.3, medium for 0.3 $\leq$ $\rho$ < 0.5, large for $\rho$ $\geq$ 0.5

Hypothesis testing results were adjusted for multiple comparisons separately for each analysis step and investigated cohort with the false discovery rate method (FDR) [@Benjamini1995]. 
Differences with p < 0.05 were considered statistically significant. 
Statistical hypothesis testing was done with the development packages [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) and [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) which partly employ statistical tools of base R, _stats_, _rstatix_, and _rcompanion_ [@Mangiafico2022; @Kassambara2021].

## Differences in rates of genetic alterations

Differences in rates of genetic alterations between analysis groups were investigated either by $\chi^2$ test with Cramer's V effect size metric (package [_fastTest_](https://github.com/PiotrTymoszuk/fastTest)) or by weighted permutation test (package [_perich_](https://github.com/PiotrTymoszuk/perich)). 
In the later case, the actual frequency of a genetic alteration was compared with the expected frequency obtained in n = 1000 - 10000 permutations weighted by counts of all genetic alterations in the cancer samples. 
This testing procedure yielded an effect size statistic called enrichment score (ES) defined as the ratio of the actual to the expected genetic alteration number in an analysis groups. 
By iterating over the permutations, 95% confidence intervals of ES and p values for the ES = 1 null hypothesis were computed.

## Differences in gene expression, gene signature scores, and metrics of predicted drug resistance

Differences in $log_2$-transformed transcript levels of the genes of interest of ssGSEA scores of the FGFR-related gene signatures between the LumP, LumU, stroma-rich, and Ba/Sq MIBC consensus molecular classes of urothelial cancers were explored with a two step procedure. 
In the first step, differences of expression levels or ssGSEA scores between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
In the second step, differences between a given class and the LumP class were were evaluated by Welch's two-tailed T test with Cohen's d effect size. 
The significantly differentially regulated genes were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation as compared with the baseline. 
The significantly differentially regulated gene signatures were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and Cohen's d $\geq$ 0.5. 
A similar testing procedure was applied to predictions of IC50 and AUC of pan-FGFR inhibitors. 
However, due to non-normal distribution of some of the resistance metrics, Kruskal-Wallis with with $\eta^2$ statistic for testing for overall differences between the consensus subsets, and Mann-Whitney post-hoc test with biserial r effect size statistic was used for comparisons versus LumP cancers. 
The significant differences in predicted resistance were considered for pFDR(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and biserial r $\geq$ 0.3.

Differences in metrics of predicted resistance to pan-FGFR inhibitors between _FGFR3_ wild-type and _FGFR3_ mutant cancers were assessed by Mann-Whitney test with biserial r effect size statistic. 
Significant differences in the predicted drug resistance were considered for pFDR < 0.05 and biserial r $\geq$ 0.3.

In these analyses, functions `f_one_anova()`, `f_t_test()`, `f_kruskal_test()`, and `f_wilcox_test()` from [_fastTest_](https://github.com/PiotrTymoszuk/fastTest) package were used.

## Gene-co-regulation networks

Co-expression networks of the FGF-, FGFR-, and FGFBP-coding genes were constructed for bulk cancers (TCGA BLCA, IMvigor, and BCAN cohorts) and DepMap cell lines as weighted graphs whose edges were defined by pairwise correlations of the expression levels with Spearman's $\rho \geq$ 0.3 (function `as_iGraph()`, package [_graphExtra_](https://github.com/PiotrTymoszuk/graphExtra) with wrapper tools for _igraph_) [@Csardi2006]. 
Isolated vertices were removed with `prune_degree()` function. 
Communities of co-regulated genes in the co-expression network in the DepMap urothelial cancer cell line collective were identified with the Leiden algorithm (loss function: modularity, resolution: $\gamma$ = 0.5, iterations: n = 100, function `cluster_leiden()`, _igraph_ package) [@Csardi2006; @Traag2019].

The following vertex importance statistics [@Csardi2006] were computed with `summary()` method called for entire network objects (bulk cancer cohorts) or gene community sub-networks (DepMap cell lines): 
degree (number of directly connected neighbor nodes), 
hub score (eigenvector of the affinity matrix, i.e. an overall strength of correlations), and 
betweenness (number of the shortest paths between all pairs of vertices that pass through the vertex). 
The networks were visualized with Fruchterman - Reingold algorithm by calling `plot()` method [@Briatte2021].

## Machine learning modeling of consensus molecular subsets of urothelial cancers with expression of FGFR-, FGF-, and FGFBP-coding genes as explanatory factors

To check if and how differences in expression of the FGFR-, FGF- and FGFBP-coding genes of interest emulate the consensus molecular classification proposed by Kamoun and colleagues [@Kamoun2020], we resorted to machine learning modeling of LumP, LumU, stroma-rich and Ba/Sq classes in the TCGA BLCA, IMvigor and BCAN cohorts with two independent algorithms, multinomial regularized Elastic Net regression [@Friedman2010; @Zou2005] and Random Forest tree ensemble [@Wright2017; @Breiman2001].

The models were constructed for two sets of explanatory factors:

* the genuine set of MIBC consensus class-defining genes proposed by Kamoun et al. and used also be the nearest centroid classification tool of the _consensusMIBC_ package. The genes were restricted to n = `r length(mibc_globals$variables)` features detected in all investigated transcriptomic cohorts

* the set of `r length(ml_globals$variables)` FGFR-, FGF- and FGFBP-coding genes detected in all investigated transcriptomic cohorts (`r ml_globals$variables %>% map_chr(~paste0('_', .x, '_')) %>% collapse_and`)

The expression values were fed into the models in form of Z-scores of Combat-processed $log_2$ gene counts (TCGA BLCA, IMvigor) or Z-scores of $log_2$ gene counts (BCAN). 
Assignment to the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response and ground truth for evaluation of the models.

The models were developed in a training subset derived from randomly chosen two-thirds of TCGA BLCA observations (n = `r nrow(ml_globals$data$tcga_train)`). 
In order to improve recognition of the minority consensus classes, the LumU and Stroma-rich subsets, we amplified their observations by over-sampling with the SMOTE algorithm (function `smote()`, package _performanceEstimation_) [@Torgo2014; @Chawla2002; @Torgo2016]. 
The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010]. 
The optimal values of the regularization parameter $\lambda$ was found by minimizing the Elastic Net models' deviance in 200 repeats 10-fold cross-validation. 
The Random Forest models of consensus molecular subsets were tuned and trained inn the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
The optimal set of hyper-parameters that control behavior of the Random Forest models, 
mtry (number of randomly chosen variables per tree learner), 
minimal node size (minimal number of observations in the terminal nodes of the tree learner), and 
splitting rule (rule that defines partitioning of the explanatory variables at the learner's nodes) 
was found by minimizing the out-of-bag classification error [@Breiman2001]. 
The final Random Forest models consisted of 500 tree learners. 
The optimal values of the tunable hyper-parameters are listed in __Supplementary Table S\@ref(tab:tab-tuning)__.

Performance of the machine learning models was evaluated by comparing the predicted and actual consensus molecular class in the unmodified test subset of cancers in the TCGA BLCA cohort (`r ml_globals$data$tcga_test %>% nrow`), as well as in the IMvigor (n = `r ml_globals$data$imvigor %>% nrow`) and BCAN (n = `r ml_globals$data$bcan %>% nrow`) independent validation cohorts. 
The consensus classification returned by the _consensusMIBC_ R package nearest centroid tool was assumed the ground truth. 
The following numeric statistics of model performance were computed: 
overall accuracy (fraction of cancer samples with correct consensus molecular class prediction), 
Cohen's $\kappa$ (non-accidental concordance between the predicted and actual class, high $\kappa$ expected for a highly accurate model) [@McHugh2012], 
Brier score (metric of overall model calibration, low Brier scores expected for a well calibrated model) [@Brier1950; @Goldstein-Greenwood2021], 
and receiver-operating characteristic measures such as area under the curve (AUC, measure of overall sensitivity and specificity). 
Those evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random (__Supplementary Figure S\@ref(fig:fig-ml-eval)A__, __Supplementary Table S\@ref(tab:tab-ml)__) 
Additionally, accuracy of prediction for single consensus molecular classes was assessed by inspection of the confusion matrices (__Supplementary Figure S\@ref(fig:fig-confusion)__). 
Calibration of the machine learning models at predictions of particular consensus molecular classes was assessed with squares of differences between the predicted class assignment probabilities and the binarized actual class assignment (1: the class of interest, 0: any other consensus molecular class) or 'Brier squares' between the consensus molecular classes (__Supplementary Figure S\@ref(fig:fig-ml-eval)B__).

Importance of the explanatory variables, i.e. FGFR-, FGF-, and FGFBP-coding transcripts, and their positive and negative association with predictions of LumP, LumU, stroma-rich, and Ba/Sq classes by the machine learning models were assessed with the SHAP algorithm (Shapley additive explanations) [@Lundberg2017; @Covert2020]. 
To this end, we re-fitted the Elastic Net and Random Forest models of consensus molecular classes in the unmodified test subset of the TCGA BLCA cohort not used for the model training with the optimal set of the tuning parameters. 
Matrices of SHAP metrics for the genes and observations were computed for the models with `kernelshap()` function from _kernelshap_ package [@Mayer2023; @Mayer2023a]. 
As the background for the SHAP procedure, we used randomly selected 30% of observations of the test subset of the TCGA BLCA collective. 
Mean absolute SHAP value served as a measure of gene importance (__Supplementary Table S\@ref(tab:tab-shap)__). 
SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions (__Figure 1F__, __Supplementary Figure S\@ref(fig:fig-shap-elnet)__ - __S\@ref(fig:fig-shap-fgfr)__).

## Data and code availability

Data from publicly available sources were analyzed. 
The R analysis pipeline is available as a [GitHub repository](https://github.com/PiotrTymoszuk/FGFR-BLCA)

\newpage

# Supplementary Tables

```{r tab-cohorts, tab.cap = 'Characteristic of the investigated cohorts. Numeric features are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages and observation counts of the categories.'}

flextable::flextable(suppl_tabs$cohorts) %>% 
  set_widths(c(2.5, rep(2.3, 6), 2.3, 2)) %>% 
  footnote(1, 8:9, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-sub-clinic, tab.cap = "Characteristic of the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Numeric features are presented as medians with interquartile ranges and ranges. Qualitative variables are shown as percentages and observation counts of the categories."}

flextable::flextable(suppl_tabs$sub_clinic) %>% 
  merge_v(1) %>% 
  set_widths(c(2.2, 2.5, rep(2.8, 4), 2.3, 2)) %>% 
  footnote(1, 7:8, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

```

\newpage

```{r tab-genes, tab.cap = 'FGFR-, FGF-, and FGFBP-coding genes investigated in the current report.'}

flextable::flextable(suppl_tabs$genes) %>% 
  italic(j = 2, part = 'body') %>% 
  merge_v(1) %>% 
  set_widths(rep(3, 3)) %>% 
  theme_vanilla

```

\newpage

```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF-, FGFR-, and FGFBP-coding genes in the investigated cohorts. Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

suppl_tabs$genetic_alterations[1, 1] %>% 
  mutate(`Alteration type` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

\newpage

```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. The frequencies are presented as percentages of the consensus class. Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic. P values were corrected for multiple testing with the false discovery rate method. The table is available as a supplementary Excel file."}

suppl_tabs$sub_genetics[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

\newepage

```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing. The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN). Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC). The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells. Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed. The table is available as a supplementary Excel file."}

suppl_tabs$expression[1, 1] %>% 
  mutate(Variable = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

```{r tab-sub-dge, tab.cap = "Expression of FGF-, FGFR-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression levels are presented as medians with interquartile ranges and ranges. The table is available as a supplementary Excel file."}

suppl_tabs$sub_dge[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

```{r tab-sub-sig, tab.cap = "Single sample gene set enrichment analysis (ssGSEA) scores of FGFR-related gene signatures in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. ssGSEA scores are presented as medians with interquartile ranges and ranges. The table is available as a supplementary Excel file."}

suppl_tabs$sub_sig[1, 1] %>% 
  mutate(Cohort = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

\newpage

```{r tab-sub-drugs, tab.cap = "Predicted resistance metrics to pan-FGFR inhibitors in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts. The predictions of were made with RIDGE regularized linear models trained with transcriptome data of epithelial cancer cell lines treated with FGFR inhibitors in GDSC1, GDSC2, CTRP2, and PRISM drug screening experiments. Resistance metrics are presented as medians with interquartile ranges and ranges."}

flextable::flextable(suppl_tabs$sub_drugs) %>% 
  set_widths(c(1.4, 2.2, 1.6, rep(2.8, 4), 2.3, 2)) %>% 
  merge_v(1) %>% 
  footnote(1, 3, 
           as_paragraph('IC50: concentation in µM resulting in 50% inhibition of growth; AUC: area under the dose-response curve.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 8:9, 
           value = as_paragraph('Statistical significance: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple tesing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-tuning, tab.cap = "Selection of the optimal sets of hyper-parameters for Elastic Net and Random Forest models of the MIBC consensus molecular classes by cross-validation and out-of-bag prediction tuning in the training subset of the TCGA BLCA cohort."}

flextable::flextable(suppl_tabs$tuning) %>% 
  merge_v(1) %>% 
  set_widths(c(3, 2.5, 4, 4)) %>% 
  footnote(1, 1, 
           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centrouid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('lambda: shrinkage/regularization paremeter; mtry: number of randomly chosen explanatroy variables for construction of decision trees in the Random Forest ensemble; min.node.size: minimal number of observations in the terminal/leaf nodes of decision trees; splitrule: algorithm which defines splits at the nodes of decision trees.'),
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```

\newpage

```{r tab-ml, tab.cap = "Performance of Elastic Net and Random Forest models of LumP, LUMU, stroma-rich, and BA/Sq consensus molecular classes of MIBC. The models were trained in a random subset of two-thirds observations of the TCGA BLCA cohort with ComBat-processed log2 expression values of the consensus class-refining genes by Kamoun et al. and the FGFR/FGF/FGFBP genes as explanatory variables. Minority consensus classes were amplified by the SMOTE algorithm to improve their identification by the model. The models were evaluated in the training and test subsets of the TCGA BLCA, IMvigor, and BCAN cohorts; the consensus class assignment predicted by a nearest centroid classifier from consensusMIBC package served as the ground truth. Metrics of model accuracy, calibration, and receiver-operating characteristic are listed."}

flextable::flextable(suppl_tabs$ml) %>% 
  merge_v(1:2) %>% 
  set_widths(c(2.4, 2.2, 2.2, rep(2, 6))) %>% 
  footnote(1, 1, 
           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centrouid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4, 
           value = as_paragraph('Fraction of observations correctly assigned to molecular consensus classes.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5, 
            value = as_paragraph("Cohen's \u03BA. Non-accidental concordance between the actual and predicted molecular class assignment."), 
            part = 'header', 
            ref_symbols = 'c') %>% 
  footnote(1, 6, 
           value = as_paragraph('Measure of overall model calibration.'),
           part = 'header', 
           ref_symbols = 'd') %>% 
  footnote(1, 7:9, 
           value = as_paragraph('Receiver-operating characteristic (ROC). AUC: area under the ROC curve.'), 
           part = 'header', 
           ref_symbols = 'f') %>% 
  theme_vanilla

```

\newpage

```{r tab-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of MIBC consensus molecular classes estimated by the SHAP algorithm. Contribution of the genuine MIBC consensus class-defining genes and the FGFR-, FGF-, and FGFBP-coding genes to prediction of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations). The table is available as a supplementary Excel file."}

suppl_tabs$shap[1, 1] %>% 
  mutate(`Explanatory factors/genes` = '') %>% 
  set_names(' ') %>% 
  flextable %>% 
  set_widths(18)

```

\newpage

# Supplementary Figures

```{r fig-sub-genet, fig.width = figur::convert(suppl_figs$sub_genet, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_genet, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.'}

suppl_figs$sub_genet$plot

```

__Supplementary Figure S\@ref(fig:fig-sub-genet). Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.__

_Cancer samples from the IMvigor and BCAN data set was assigned to the MIBC consensus molecular subtypes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensus MIBC). Differences in frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes between the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes were assessed by weighted permutation test. P values were corrected for multiple testing with the false discovery rate method. Note that copy number variant information was not available for the IMvigor cohort. Presence/absence of selected genetic alterations in the consensus molecular classes visualized in oncoplots. Alteration names with effect sizes and p values of differences between the consensus classes are indicated in the Y axis. Significant effects are highlighted with bold font. Numbers of cancer samples in the consensus classes are displayed in the plot captions._

\newpage

```{r fig-fgfr3-domains, fig.width = figur::convert(suppl_figs$fgfr3_domains, to = 'in')$w, fig.height = figur::convert(suppl_figs$fgfr3_domains, to = 'in')$h, fig.cap = 'Frequency of somatic mutations in FGFR1/2/3/4 genes split by protein domain'}

suppl_figs$fgfr3_domains$plot

```

__Supplementary Figure S\@ref(fig:fig-fgfr3-domains). Frequency of somatic mutations in FGFR1/2/3/4 genes split by protein domain__

_Mutations in FGFR1, FGFR2, FGFR3, and FGFR4 genes in the GENIE BLCA, MSK IMPACT, TCGA BLCA, and BCAN cohorts were classified by protein domain. Their frequency was expressed as percentage of cancer samples and presented in stack plots. Numbers of analyzed cancer samples are displayed in the plot captions._

\newpage

```{r fig-ihc-receptors, fig.width = figur::convert(suppl_figs$ihc_receptors, to = 'in')$w, fig.height = figur::convert(suppl_figs$ihc_receptors, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.'}

suppl_figs$ihc_receptors$plot

```

__Supplementary Figure S\@ref(fig:fig-ihc-receptors). Representative immunohistochemistry staining images for FGFR proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%). For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score. The images are labeled with HPA patient identifiers (ID) and IHC scores._

\newpage

```{r fig-ihc-ligands, fig.width = figur::convert(suppl_figs$ihc_ligands, to = 'in')$w, fig.height = figur::convert(suppl_figs$ihc_ligands, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.'}

suppl_figs$ihc_ligands$plot

```

__Supplementary Figure S\@ref(fig:fig-ihc-ligands). Representative immunohistochemistry staining images for FGF proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%). For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score. The images are labeled with HPA patient identifiers (ID) and IHC scores._

\newpage

```{r fig-ihc-fgfbp, fig.width = figur::convert(suppl_figs$ihc_fgfbp, to = 'in')$w, fig.height = figur::convert(suppl_figs$ihc_fgfbp, to = 'in')$h, fig.cap = 'Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.'}

suppl_figs$ihc_fgfbp$plot

```

__Supplementary Figure S\@ref(fig:fig-ihc-fgfbp). Representative immunohistochemistry staining images for FGFBP proteins retrieved from the Human Protein Atlas.__

_Images of immunohistochemistry (IHC) staining for FGFR, FGF, and FGFBP in urothelial cancer samples were retrieved from the Human Protein Atlas (HPA). The staining was quantified by the IHC score which was defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%). For each protein, representative images of immunohistochemistry (IHC) staining were selected; the left image corresponds to the minimum IHC score, the middle image shows the staining close to the median IHC score, the right image corresponds to the maximal IHC score. The images are labeled with HPA patient identifiers (ID) and IHC scores._

\newpage

```{r fig-cell-networks, fig.width = figur::convert(suppl_figs$cell_networks, to = 'in')$w, fig.height = figur::convert(suppl_figs$cell_networks, to = 'in')$h, fig.cap = 'Co-expression network of FGF-, FGFR-, and FGFBP-coding genes in DepMap urothelial cancer cell lines'}

suppl_figs$cell_networks$plot

```

__Supplementary Figure S\@ref(fig:fig-cell-networks). Co-expression network of FGF-, FGFR-, and FGFBP-coding genes in DepMap urothelial cancer cell lines__

_Co-expression network was constructed in the DepMap urothelial cancer cell line collection for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman’s $rho \geq$ 0.3. Isolated vertices of the networks were removed. Gene communities were identified with the Leiden algorithm and named after their FGFR-coding members. Candidate hub genes of the communities were selected by computing hub scores separately for each community (eigenvectors of the affinity matrix, i.e. metrics of the overall correlation strength). The networks were visualized using the Fruchterman-Reingold algorithm. Points represent node vertices/genes, point shape codes for classification of the protein product. Point color codes for the community assignment (A) or hub score (B). Lines represent edges, i.e. correlations between expression levels with $\rho \geq$ 0.3. Line widths correspond to  values. Note the central positions and high scores of the FGFR1, FGFR2, and FGFR3 genes and their assignment to the communities. This suggests exclusivity, i.e that FGFR1 with its ligands (FGF2, FGF5) and accessory molecules (CD44, TNFAIP6, FGFRL1) provide oncogenic signaling in a subset of cell lines, while signaling via FGFR2/FGFR3 and their co-receptors (GPC1/3, SDC1/4, HSPG2, TGFBR3) is exploited by another, non-overlapping cell line subset._

\newpage

```{r fig-sub-fgfr-dge, fig.width = figur::convert(suppl_figs$sub_dge, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_dge, to = 'in')$h, fig.cap = 'Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}

suppl_figs$sub_dge$plot

```

__Supplementary Figure S\@ref(fig:fig-sub-fgfr-dge). Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.__

_Assignment of cancer samples from the IMvigor and BCAN collectives to the MIBC consensus molecular classes was presented in Supplementary Figure S\@ref(fig:fig-sub-genet). Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes were assessed by one-way ANOVA with  effect size statistic. Differences in $log_2$-transformed gene expression between a class and the LumP class were investigated by two-tailed T test. P values were corrected for multiple testing with the false discovery rate method (FDR). Significantly differentially regulated genes were defined by p(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation in a class as compared with LumP cancers. Z-scores of  expression levels of significantly differentially regulated genes shared by at least two cohorts are displayed in heat maps. Significant genes are labeled with bold font in the Y axes. Numbers of cancer samples in the consensus classes are indicated in the plot captions. The full analysis results are shown in Supplementary Table S\@ref(tab:tab-sub-dge)._

\newpage

```{r fig-sub-sig, fig.width = figur::convert(suppl_figs$sub_sig, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_sig, to = 'in')$h, fig.cap = 'Scores of FGFR-related gene signatures in the consensus molecular classes of urothelial cancers.'}

suppl_figs$sub_sig$plot

```

__Supplementary Figure S\@ref(fig:fig-sub-sig). Scores of FGFR-related gene signatures in the consensus molecular classes of urothelial cancers.__

_Assignment of cancer samples from the IMvigor and BCAN collectives to the MIBC consensus molecular classes was presented in Supplementary Figure S\@ref(fig:fig-sub-genet). Single sample gene set enrichment analysis (ssGSEA) scores of `r length(sub_sig$variables)` FGFR-related MSig database signatures were compared the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes by one-way ANOVA with  effect size statistic. Differences in ssGSEA scores between a class and the LumP class were investigated by two-tailed T test with Cohen's d effect size statistic. P values were corrected for multiple testing with the false discovery rate method (FDR). Significantly differentially regulated gene signatures were defined by p(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and Cohen's d $\geq$ 0.5 for regulation in a class as compared with LumP cancers. ssGSEA scores of significantly differentially regulated gene signatures shared by at least two cohorts are displayed in heat maps. Results of one-way ANOVA testing for overall differences between the consensus classes (effect sizes and p values) are indicated in the Y axes. Significant effects are labeled with bold font in the Y axes. Numbers of cancer samples in the consensus classes are indicated in the plot captions. The full analysis results with signature names and listing of the member genes are presented in Supplementary Table S\@ref(tab:tab-sub-sig)._

\newpage

```{r fig-sub-drugs, fig.width = figur::convert(suppl_figs$sub_drugs, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_drugs, to = 'in')$h, fig.cap = 'Predicted response to pan-FGFR inhibitors in the consensus molecular classes of urothelial cancers.'}

suppl_figs$sub_drugs$plot

```

__Supplementary Figure S\@ref(fig:fig-sub-drugs). Predicted response to pan-FGFR inhibitors in the consensus molecular classes of urothelial cancers.__

_Measures of resistance to pan-FGFR inhibitors (GDSC1, GDSC2, CTRP2: inhibitory concentration resulting in 50% growth inhibition [IC50], PRISM: area under the dose - response curve [AUC]) were predicted for bulk cancers in the TCGA BLCA, IMvigor, and BCAN cohorts by RIDGE linear models trained with anti-cancer drug response data and baseline transcriptomes of epithelial cancer cell lines in the GDSC1, GDSC2, CTRP2, and PRISM in vitro screening experiments._ 
_The predicted IC50 and AUC values were compared between the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes by Kruskal-Wallis test with $\eta^2$ effect size statistic._ 
_Differences in resistance metrics between a class and the LumP class were investigated by Mann-Whitney test with biserial r effect size statistic. P values were corrected for multiple testing with the false discovery rate method (FDR). Significantly differentially regulated resistance metrics were defined by p(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and r $\geq$ 0.3 for differences between a class and LumP cancers. Z-scores of IC50 and AUC values are displayed in heat maps. Results of Kruskal-Wallis tests for overall differences between the consensus classes (effect sizes and p values) are indicated in the Y axes. Significant effects are labeled with bold font in the Y axes. Numbers of cancer samples in the consensus classes are indicated in the plot captions._

\newpage

```{r fig-fgfr-drugs, fig.width = figur::convert(suppl_figs$fgfr_drugs, to = 'in')$w, fig.height = figur::convert(suppl_figs$fgfr_drugs, to = 'in')$h, fig.cap = 'Predicted response to pan-FGFR inhibitors in the urothelial cancers with and without FGFR3 mutations.'}

suppl_figs$fgfr_drugs$plot

```

__Supplementary Figure S\@ref(fig:fig-fgfr-drugs). Predicted response to pan-FGFR inhibitors in the urothelial cancers with and without FGFR3 mutations.__

_Predictions of drugs resistance (IC50: concentration resulting in 50% growth inhibition, AUC: area under the dose-response curve) were obtained for bulk cancer samples in the TCGA BLCA, IMvigor, and BCAN cohorts as presented in Supplementary Figure S\@ref(fig:fig-sub-drugs)._
_Differences in resistance metrics between cancers with and without mutations of FGFR3 were assessed by Mann-Whitney test with biserial r effect size statistic. P values were adjusted for multiple testing with the false discovery rate method (FDR). Significant differences were considered for pFDR < 0.05 and r $\geq$ 0.3. Z-scores of IC50 and AUC values are displayed in heat maps. Effect sizes and p values are indicated in the Y axes. Significant effects are labeled with bold font in the Y axes. Numbers of wild-type (WT) and mutant cancer samples are indicated in the plot captions._

\newpage

```{r fig-ml-eval, fig.width = figur::convert(suppl_figs$ml_eval, to = 'in')$w, fig.height = figur::convert(suppl_figs$ml_eval, to = 'in')$h, fig.cap = 'Evaluation of predictions of the consensus molecular classes of MIBC made by machine learning trained with the genuine consensus class-defining genes and the FGFR/FGF/FGFBP-coding genes.'}

suppl_figs$ml_eval$plot

```

__Supplementary Figure S\@ref(fig:fig-ml-eval). Evaluation of predictions of the consensus molecular classes of MIBC made by machine learning trained with the genuine consensus class-defining genes and the FGFR/FGF/FGFBP-coding genes.__

_Assignment of urothelial cancer samples in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts to the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) MIBC consensus molecular classes was modeled by two independent machine learning algorithms, Elastic Net linear regression and Random Forest tree ensemble, with ComBat-processed $log_2$-transformed expression levels of the `r mibc_globals$variables %>% length` genuine class-defining genes proposed by Kamoun et al. or the `r ml_globals$variables %>% length` genes coding for FGFR, FGF, and FGFBP proteins as the explanatory variables. The models were optimized (tuned) and trained in a synthetic data set derived from randomly chosen two-thirds of observations from the `r globals$cohort_labs["tcga"]` collective with expansion (oversampling) of observations in the LumU and stroma-rich classes by the SMOTE algorithm._ 
_Performance of the models was evaluated in the non-modified test subset of remaining observations from the `r globals$cohort_labs["tcga"]` data set as well as in the IMvigor and BCAN independent validation collectives; assignment to the classes by the nearest centroid classifier from consensusMIBC R package by Kamoun et al. served as the ground truth._

_(A) Numeric statistics of overall non-accidental model accuracy (Cohen's $\kappa$, high values expected for a highly accurate model) and overall calibration (Brier score, low scores expected for a well calibrated model) are presented in a dot plot. Data points are labeled with the cohort names; evaluations in the same cohort are connected with a line. The dashed lines and gray regions visualize $\kappa$ and Brier score values expected for a model that assigns the consensus molecular classes by chance._

_(B) Squared differences between the class assignment probability returned by the models and the actual 0/1-coded class assignment (squared distance to the outcome) were computed as measures for single cancer samples as measures of model's confidence and error at predicting the consensus molecular classes. The squared distances to the outcome are presented in box plots with medians and interquartile ranges represented by boxes and whiskers which span over 150% of the interquartile ranges. Single cancer samples are depicted as points. Note a substantial improvement of prediction (lower squared distances to the outcome) of LumU class cancers by the machine learning models that employ the full class-defining predictor set by Kamoun et al as compared with the models using solely the FGFR/FGF/FGFBP-coding genes. This suggests that there are no strong and reproducible FGFR/FGF/FGFBP-related markers of the LumU class._

\newpage

```{r fig-confusion, fig.width = figur::convert(suppl_figs$confusion, to = 'in')$w, fig.height = figur::convert(suppl_figs$confusion, to = 'in')$h, fig.cap = 'Confusion matrices of predictions of the consensus molecular classes of MIBC made by machine learning trained with the genuine consensus class-defining genes and the FGFR/FGF/FGFBP-coding genes.'}

suppl_figs$confusion$plot

```

__Supplementary Figure S\@ref(fig:fig-confusion). Confusion matrices of predictions of the consensus molecular classes of MIBC made by machine learning trained with the genuine consensus class-defining genes and the FGFR/FGF/FGFBP-coding genes.__

_Elastic Net and Random Forest (RF) models of the MIBC consensus classes were developed and evaluated in the `r globals$cohort_labs[c("tcga", "imvigor", "bcan")] %>% collapse_and` cohorts as presented in Supplementary Figure S\@ref(fig:fig-ml-eval)._ 
_Accuracy of the models in particular consensus molecular classes was visualized as heat maps of confusion matrices. Fields at the diagonals are highlighted with bold contours represent correctly classified cancer samples. The heat map tiles are labeled with percentages of all cancer samples and cancer sample counts._

_LumP: luminal papillary; LumU: luminal unstable; Ba/Sq: basal/squamous-like._

\newpage

```{r fig-shap-elnet, fig.width = figur::convert(suppl_figs$shap_elnet, to = 'in')$w, fig.height = figur::convert(suppl_figs$shap_elnet, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers with the full set of class-defining genes by Kamoun et al. assessed by the SHAP algorithm.'}

suppl_figs$shap_elnet$plot

```

__Supplementary Figure S\@ref(fig:fig-shap-elnet). Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers with the full set of class-defining genes by Kamoun et al. assessed by the SHAP algorithm.__

_The Elastic Net model of the MIBC consensus classes with the consensus class-defining genes by Kamoun at al. was developed and evaluated as presented in Supplementary Figure S\@ref(fig:fig-ml-eval)._ 
_Contributions of the explanatory variables, i.e. single genes, to prediction of the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal-squamous (Ba/Sq) consensus molecular classes were estimated with the SHAP algorithm (Shapley additive explanations). SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots. Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene. The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots. Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._

\newpage

```{r fig-shap-ranger, fig.width = figur::convert(suppl_figs$shap_ranger, to = 'in')$w, fig.height = figur::convert(suppl_figs$shap_ranger, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers with the full set of class-defining genes by Kamoun et al. assessed by the SHAP algorithm.'}

suppl_figs$shap_ranger$plot

```

__Supplementary Figure S\@ref(fig:fig-shap-ranger). Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers with the full set of class-defining genes by Kamoun et al. assessed by the SHAP algorithm.__

_The Random Forest model of the MIBC consensus classes with the consensus class-defining genes by Kamoun at al. was developed and evaluated as presented in Supplementary Figure S\@ref(fig:fig-ml-eval)._ 
_Contributions of the explanatory variables, i.e. single genes, to prediction of the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal-squamous (Ba/Sq) consensus molecular classes were estimated with the SHAP algorithm (Shapley additive explanations). SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots. Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene. The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots. Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._

\newpage

```{r fig-shap-fgfr, fig.width = figur::convert(suppl_figs$shap_fgfr, to = 'in')$w, fig.height = figur::convert(suppl_figs$shap_fgfr, to = 'in')$h, fig.cap = 'Importance of FGFR/FGF/FGFBP-coding genes for predictions of the MIBC consensus classes by machine learning models employing the full set of the class-defining genes by Kamoun et al.'}

suppl_figs$shap_fgfr$plot

```

__Supplementary Figure S\@ref(fig:fig-shap-fgfr). Importance of FGFR/FGF/FGFBP-coding genes for predictions of the MIBC consensus classes by machine learning models employing the full set of the class-defining genes by Kamoun et al.__

_Shapley additive explanations (SHAP) as measures of variable importance for predictions of the consensus molecular classes by the Elastic Net and Random Forest models employing the full class-defining predictor set by Kamoun eta al. were computed as presented in Supplementary Figures S\@ref(fig:fig-shap-elnet) and S\@ref(fig:fig-shap-elnet)._ 
_Among the class-defining genes, `r intersect(ml_globals$variables, mibc_globals$variables) %>% collapse_and` were classified as FGFR/FGF/FGFBP-coding genes._ 
_Distribution of the mean SHAP values is presented in violin plot. Single explanatory factors are depicted as points; points corresponding to the FGFR/FGF/FGFBP-coding genes among the class-defining genes are highlighted with blue (negative class assignment predictor) and red (positive predictors of class assignment)._

\newpage

```{r fig-in-vivo, fig.width = figur::convert(suppl_figs$in_vitro, to = 'in')$w, fig.height = figur::convert(suppl_figs$in_vitro, to = 'in')$h, fig.cap = 'Dose response curves of cultured FGFR1-4 wild-type cell lines (UMUC3, 5637 and RT112) after treatment with increasing concentrations of erdafitinib.'}

suppl_figs$in_vitro$plot

```

__Supplementary Figure S\@ref(fig:fig-in-vivo). Dose response curves of cultured FGFR1-4 wild-type cell lines (UMUC3, 5637 and RT112) after treatment with increasing concentrations of erdafitinib.__

_Confluence was measured after 96 hours in cell cultures treated with 0, 1, 4, 7 and 10 µM erdafitinib. Points with whiskers represent means with standard errors from three biological replicates._

\newpage

# References

### 2025-04-28 report/markdown/rebuttal_letter.Rmd
Figure/Table/Supplement reshaping
Changes in composition of figures and tables requested by a reviewer. The whole content of tables is now in Supplementary Material. Parts of Figure 1 were moved to Supplementary Material

@@
 title: "FGFR1/3 Signaling as Achilles’ Heel of Phenotypic Diversity in Urothelial Carcinoma"
-subtitle: "Rebuttal Letter"
+subtitle: "Rebuttal Letter, first revision"
 author: "Department of Urology, Medical University of Innsbruck"

### 2025-04-28 report/markdown/figure1_legend.Rmd
Figure/Table/Supplement reshaping
Changes in composition of figures and tables requested by a reviewer. The whole content of tables is now in Supplementary Material. Parts of Figure 1 were moved to Supplementary Material

@@
 
-```{r fig-figure-1a, fig.width = figur::convert(fig$figure_1a, to = 'in')$w, fig.height = figur::convert(fig$figure_1a, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in MIBC consensus molecular classes of urothelial cancer.'}
+```{r fig-figure-1a, fig.width = figur::convert(fig$figure_1a, to = 'in')$w, fig.height = figur::convert(fig$figure_1a, to = 'in')$h, fig.cap = 'Alterations of FGFR/FGF/FGFBP-coding genes in consensus molecular classes of MIBC.'}
 
@@
 
-__Figure 1A. Alterations of FGFR-, FGF-, and FGFBP-coding genes in MIBC consensus molecular classes of urothelial cancer.__
+__Figure 1A. Alterations of FGFR/FGF/FGFBP-coding genes in consensus molecular classes of MIBC.__
 
-_Cancer samples from the TCGA BLCA, IMvigor, and BCAN data sets were assigned to the MIBC consensus molecular classes based on ComBat-adjusted whole-genome mRNA levels by the nearest centroid algorithm (R package consensusMIBC).
-Differences in frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes between the LumP (luminal papillary), LumU (luminal unstable), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes were assessed by weighted permutation test.
-P values were corrected for multiple testing with the false discovery rate method.
-Presence/absence of selected genetic alterations in the consensus molecular classes in the TCGA BLCA cohort was visualized in a oncoplot.
-Alteration names (mut: mutation, amp: amplification, del: deletion) with effect sizes and p values of differences between the consensus classes are indicated in the Y axis.
+_Differences in frequency of somatic mutations and copy number variants of FGFR/FGF/FGFBP-coding genes between the LumP (luminal papillary), LumU (luminal unstable), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes of MIBC in the TCGA, BCAN and IMvigor cohorts were assessed by false discovery rate (FDR) corrected weighted permutation test.
+Presence/absence of selected genetic alterations (mut: mutation, amp: amplification, del: deletion) in the TCGA BLCA cohort was visualized in a oncoplot.
 Significant effects are highlighted with bold font.
-Numbers of cancer samples in the consensus classes are displayed in the plot caption._
+Numbers of cancer samples are displayed in the plot caption._
 
@@
 
-```{r fig-figure-1b, fig.width = figur::convert(fig$figure_1b, to = 'in')$w, fig.height = figur::convert(fig$figure_1b, to = 'in')$h, fig.cap = 'Mutational hot spots of the FGFR3 gene.'}
+```{r fig-figure-1b, fig.width = figur::convert(fig$figure_1b, to = 'in')$w, fig.height = figur::convert(fig$figure_1b, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR/FGF/FGFBP-coding genes.'}
 
@@
 
-__Figure 1B. Mutational hot spots of the FGFR3 gene.__
+__Figure 1B. Co-expression networks of FGFR/FGF/FGFBP-coding genes.__
 
-_Frequency of mutations affecting residues of the FGFR3 protein was expressed as percentage of all cancer samples in the cohort and displayed in dot plots.
-Each point represents a single protein residue, point color codes for assignment of the residue to the protein domain.
-Candidate mutation hot spots with mutation rates $\geq$ 1% are highlighted with enlarged symbols and labeled with percentages and protein sequence changes.
-Cohort names and numbers of analyzed cancer samples are indicated in the plot facets.
-The protein domain scheme is displayed above the plot._
+_Co-expression networks were constructed for the FGFR/FGF/FGFBP-coding genes, whose expression levels correlated with Spearman’s $\rho \geq$ 0.3 and visualized for the TCGA BLCA cohort with the Fruchterman-Reingold algorithm.
+Points represent genes, point shape codes for the protein product type.
+Point color represent hub scores.
+Lines represent edges, i.e correlations with $\rho \geq$ 0.3.
+Line widths correspond to $\rho$ values._
 
@@
 
-```{r fig-figure-1c, fig.width = figur::convert(fig$figure_1c, to = 'in')$w, fig.height = figur::convert(fig$figure_1c, to = 'in')$h, fig.cap = 'Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.'}
+```{r fig-figure-1c, fig.width = figur::convert(fig$figure_1c, to = 'in')$w, fig.height = figur::convert(fig$figure_1c, to = 'in')$h, fig.cap = 'Differential expression of FGFR/FGF/FGFR-coding genes in consensus molecular classes of MIBC.'}
 
@@
 
-__Figure 1C. Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.__
+__Figure 1C. Differential expression of FGFR/FGF/FGFR-coding genes in consensus molecular classes of MIBC.__
 
-_Levels of FGFR-, FGF-, and FGFBP-coding mRNA were determined in cancer tissue samples in the TCGA BLCA, IMvigor, and BCAN by RNA sequencing and published as $log_2$ gene counts (TCGA BLCA and IMvigor) or Z-scores of $log_2$ gene counts (BCAN).
-mRNA expression of the genes of interest in DepMap urothelial cancer cell lines was measured by RNA sequencing and provided as $log_2$ gene counts.
-Median Z-scores of $log_2$ mRNA expression are shown in a heat map.
-Each tile represents median expression Z-score of a gene in one cohort.
-Numbers of cancer samples are indicated in the X axis.
-White tiles represent genes not detected in each cohort.
-Protein expression of `r expl_expr$plot_data$hpa %>% filter(complete.cases(.)) %>% .$variable %>% unique %>% length` FGFR-, FGF-, and FGFBP-coding genes in the Human Protein Atlas (HPA) repository was quantified by immunohistochemistry (IHC).
-The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%).
-Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges.
-Single IHC specimens are visualized as points.
-Numbers of IHC samples are indicated in the Y axis._
+_Differences in $log_2$ mRNA levels of FGFR/FGF/FGFBP-coding genes between the LumP, LumU, stroma-rich and Ba/Sq consensus molecular classes were assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic and two-tailed post-hoc T test.
+Differentially regulated genes were defined by p(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and at least 1.25-fold regulation in a class as compared with LumP cancers.
+Z-scores of $log_2$ expression levels of differentially regulated genes shared by at least two cohorts are displayed in a heat map for TCGA BLCA cancers.
+Significant effects are labeled with bold font.
+Numbers of samples are indicated in the plot caption._
 
@@
 
-```{r fig-figure-1d, fig.width = figur::convert(fig$figure_1d, to = 'in')$w, fig.height = figur::convert(fig$figure_1d, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.'}
+```{r fig-figure-1d, fig.width = figur::convert(fig$figure_1d, to = 'in')$w, fig.height = figur::convert(fig$figure_1d, to = 'in')$h, fig.cap = 'SHAP variable importance in the Elastic Net model of consensus molecular classes of MIBC with FGFR/FGF/FGFBP-coding genes as explanatory factors.'}
 
@@
 
-__Figure 1D. Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes.__
+__Figure 1D. SHAP variable importance in the Elastic Net model of consensus molecular classes of MIBC with FGFR/FGF/FGFBP-coding genes as explanatory factors.__
 
-_Co-expression networks were constructed in the TCGA BLCA, IMvigor, and BCAN cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman’s $\rho \geq$ 0.3.
-Isolated vertices of the networks were removed.
-The networks were visualized using the Fruchterman-Reingold algorithm.
-Points represent node vertices/genes, point shape codes for classification of the protein product.
-Point color represent hub scores, which measure the overall correlation strength.
-Lines represent edges, i.e correlations between expression levels with $\rho \geq$ 0.3.
-Line widths correspond to $\rho$ values.
-Note the central position, connectivity, and high hub scores of the genes coding for FGFR1, its ligands FGF2, FGF7 and FGF10, and the binding proteins SDC2, DCN, PTX3 and HSPG2 preserved in all cohorts.
-This suggests that ligand-dependent signaling via FGFR1 may serve as a substantial signaling pathway in urothelial cancer._
+_Contributions of the FGFR-, FGF-, and FGFBP-coding genes to prediction of LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes of MIBC by the Elastic Net model were estimated with the SHAP algorithm (Shapley additive explanations).
+SHAP values for highly influential genes and observations are visualized in violin/swarm plots.
+Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed gene expression.
+Mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
+Bar colors code for association of SHAP values and gene expression (red: positive, blue: negative)._
 
@@
 
-```{r fig-figure-1e, fig.width = figur::convert(fig$figure_1e, to = 'in')$w, fig.height = figur::convert(fig$figure_1e, to = 'in')$h, fig.cap = 'Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}
+```{r fig-resistance, fig.width = figur::convert(fig$figure_1e, to = 'in')$w, fig.height = figur::convert(fig$figure_1e, to = 'in')$h, fig.cap = 'Sensitivity to pan-FGFR inhibitors of DepMap cancer urothelial cell lines.'}
 
@@
 
-__Figure 1E. Differential expression of FGFR-, FGF-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.__
+__Figure 1E. Sensitivity to pan-FGFR inhibitors of DepMap urothelial cancer cell lines.__
 
-_Differences in log2-transformed mRNA levels of FGFR-, FGF-, and FGFBP-coding genes between the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes (Figure 1A) were assessed by one-way ANOVA with $\eta^2$ effect size statistic.
-Differences in $log_2$-transformed gene expression between a class and LumP were investigated by two-tailed T test.
-P values were corrected for multiple testing with the false discovery rate method (FDR).
-Significantly differentially regulated genes were defined by p(ANOVA) < 0.05, $\eta^2$ 0.06, pFDR(T test) < 0.05, and at least 1.25-fold up- or downregulation in a class as compared with LumP cancers.
-Z-scores of $log_2$ expression levels in the TCGA BLCA cohort of significantly differentially regulated genes shared by at least two cohorts are displayed in a heat map.
-Each tile represents a single cancer sample.
-Significant effects are labeled with bold font in the Y axes.
-Numbers of cancer samples in the consensus classes are indicated in the plot caption._
+_Resistance values in the PRISM (n = `r expl_res$sensitive$prism %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')`) drug screening experiment were expressed as AUC (area under the dose-response curve).
+Significant sensitivity for was defined as AUC < 1, which indicates stronger growth inhibition than the DMSO control.
+Doxorubicin is presented as a positive cytotoxicity control.
+Median AUC values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
+Points represent single cell lines.
+Point symbols code for FGFR3 mutation, point color represents resistance/sensitivity.
+The sensitivity cutoff is represented by the dashed line._
 
@@
 
-```{r fig-figure-1f-elnet, fig.width = figur::convert(fig$figure_1f_elnet, to = 'in')$w, fig.height = figur::convert(fig$figure_1f_elnet, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.'}
+```{r fig-figure-1f, fig.width = figur::convert(fig$figure_1f, to = 'in')$w, fig.height = figur::convert(fig$figure_1f, to = 'in')$h, fig.cap = 'Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.'}
 
-fig$figure_1f_elnet$plot
+fig$figure_1f$plot
 
@@
 
-__Figure 1F. Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.__
-
-_Contributions of the explanatory variables, i.e.
-the FGFR-, FGF-, and FGFBP-coding genes, to prediction of LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by an Elastic Net machine learning model were estimated with the SHAP algorithm (Shapley additive explanations).
-SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots.
-Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene.
-The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
-Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
-
-\newpage
-
-```{r fig-figure-1f-ranger, fig.width = figur::convert(fig$figure_1f_ranger, to = 'in')$w, fig.height = figur::convert(fig$figure_1f_ranger, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.'}
-
-fig$figure_1f_ranger$plot
-
-```
-
-__Figure 1F. Importance of explanatory variables in the Random Forest model of MIBC molecular consensus classes of urothelial cancers assessed by the SHAP algorithm.__
-
-_Contributions of the explanatory variables, i.e.
-the FGFR-, FGF-, and FGFBP-coding genes, to prediction of LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by a Random Forest machine learning model were estimated with the SHAP algorithm (Shapley additive explanations).
-SHAP values for single highly influential genes (top 10 highest mean absolute SHAP) and observations are visualized in violin/swarm plots.
-Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene.
-The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
-Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
-
-\newpage
-
-```{r fig-resistance, fig.width = figur::convert(fig$figure_1g, to = 'in')$w, fig.height = figur::convert(fig$figure_1g, to = 'in')$h, fig.cap = 'Resistance and sensitivity to pan-FGFR inhibitors of DepMap cancer urothelial cell lines.'}
-
-fig$figure_1g$plot
-
-```
-
-__Figure 1G. Resistance and sensitivity to pan-FGFR inhibitors of DepMap cancer urothelial cell lines.__
-
-_Information on resistance of urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (n = `r expl_res$sensitive$gdsc %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')` cell lines) and PRISM (n = `r expl_res$sensitive$prism %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')`) drug screening experiments was extracted from the DepMap repository.
-The resistance values were expressed as IC50 (concentration resulting in 50% growth inhibition, GDSC) or AUC (area under the dose-response curve, relative to DMSO control, PRISM).
-Significant sensitivity for the GDSC data was defined as IC50 below 10 µM, which was the maximum concentration of the FGFR inhibitors.
-Significant sensitivity for the PRISM data was defined as AUC < 1, which indicates stronger growth inhibition by a FGFR inhibitor than the DMSO control.
-IC50 and AUC for doxorubicin are presented as a positive cytotoxicity controls.
-Median IC50 and AUC values for the FGFR inhibitors with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
-Points represent single cell lines.
-Point symbols codes for the FGFR3 gene mutation status, point color represent sensitivity and resistance.
-The sensitivity cutoffs are represented by dashed lines.
-FGFR3 wild-type cell lines RT-112, RT4, SW 780, CAL-29, and J82 showed particularly high sensitivity to multiple FGFR blockers in the GDSC1/2 experiments.
-FGFR3 wild-type cell lines `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$name_lab %>% collapse_and` demonstrated a very good sensitivity to erdafitinib in the PRISM assay (AUC: `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), resistance < 0.9) %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`).
-By contrast, `r expl_res$sensitive$prism %>% filter(.data[['PRISM_BRD-K84868168-001-01-0_erdafitinib']] == 'resistant') %>% .$sample_id %>% exchange(depmap$cell_lines, key = 'sample_id', value = 'cell_line_name') %>% collapse_and` cell lines were found to be resistant to erdafitinib with AUC values ranging from `r expl_res$box_plots$prism$data %>% filter(stri_detect(variable, fixed = 'erdafi'), status == 'resistant') %>%  arrange(resistance) %>% .$resistance %>% range %>% signif(2) %>% paste(collapse = ' to ')`._
-
-\newpage
-
-```{r fig-figure-1h, fig.width = figur::convert(fig$figure_1h, to = 'in')$w, fig.height = figur::convert(fig$figure_1h, to = 'in')$h, fig.cap = 'Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.'}
-
-fig$figure_1h$plot
-
-```
-
-__Figure 1H. Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.__
+__Figure \@ref(fig:fig-figure-1f).
+Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.__
 

### 2025-04-28 report/markdown/supplementary_material.Rmd
Figure/Table/Supplement reshaping
Changes in composition of figures and tables requested by a reviewer. The whole content of tables is now in Supplementary Material. Parts of Figure 1 were moved to Supplementary Material

@@
 Please note that due to insufficient numbers of specimens assigned to the 
-luminal non-specified (LumNS; `r subtypes$assignment %>% map(filter, consensusClass == 'LumNS') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`samples) and 
+luminal non-specified (LumNS; `r subtypes$assignment %>% map(filter, consensusClass == 'LumNS') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples) and 
 neuroendocrine-like classes (NE-like; `r subtypes$assignment %>% map(filter, consensusClass == 'NE-like') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
@@
 luminal papillary (LumP; `r subtypes$assignment %>% map(filter, consensusClass == 'LumP') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
-luminal unstable (LumU; `r subtypes$assignment %>% map(filter, consensusClass == 'LumU') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), 
-stroma-rich (`r subtypes$assignment %>% map(filter, consensusClass == 'Stroma-rich') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`), and 
-basal/squamous-like classes (Ba/Sq; `r subtypes$assignment %>% map(filter, consensusClass == 'Ba/Sq') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')`). 
+luminal unstable (LumU; `r subtypes$assignment %>% map(filter, consensusClass == 'LumU') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), 
+stroma-rich (`r subtypes$assignment %>% map(filter, consensusClass == 'Stroma-rich') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples), and 
+basal/squamous-like classes (Ba/Sq; `r subtypes$assignment %>% map(filter, consensusClass == 'Ba/Sq') %>% map_dbl(nrow) %>% range %>% paste(collapse = ' to ')` samples). 
 Demographic, clinical, and pathological characteristic of the consensus molecular classes of MIBC is presented in __Supplementary Table S\@ref(tab:tab-sub-clinic)__.
@@
 Those substances were: `r list(x = drugs$fgfr_drugs %>% exchange(drugs$lexicon), y = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'experiment'), z = drugs$fgfr_drugs %>% exchange(drugs$lexicon, value = 'drug_id')) %>% pmap_chr(function(x, y, z) paste0(x, ' (', y, ', ID: ', z, ')')) %>% collapse_and`.
-The predictions of IC50 and AUC were made by calling `predict()` method for the object returned by `train()` function (package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).
+The predictions of IC50 and AUC were made by calling `predict()` method for the model container object generated by `train()` function (package [_htGLMNET_](https://github.com/PiotrTymoszuk/htGLMNET)).
 
@@
 RNA sequencing information was provided as $log_2$ TPM. 
-Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013].
+Resistance of urothelial cancer cell lines to pan-FGFR inhibitors and doxorubicin considered as a positive cytotoxicity control was assessed for urothelial cancer cell lines in the GDSC1/2 (compounds: doxorubicin, `r depmap$gdsc_lexicon %>% filter(!drug_name %in% c('Doxorubicin', 'FGFR_0939'), !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source [GDSC data portal](https://www.cancerrxgene.org/)) and PRISM experiments (compounds: doxorubicin, `r depmap$prism_lexicon %>% filter(drug_name != 'doxorubicin', !duplicated(drug_name)) %>% .$drug_name %>% collapse_and`, source: [DepMap](https://depmap.org/portal/)) [@Yu2016; @Yang2013].
 The resistance was measured in the GDSC1/2 experiments by IC50; sensitive cell lines were defined by IC50 < 10 µM, which was the highest concentration of the FGFR inhibitors in the assays. 
-The resistance in the PRISM assays was gauged by AUC normalized for effects of DMSO control; sensitive cell lines were considered for AUC < 1.
+The resistance in the PRISM assays was gauged by AUC normalized to the DMSO control; sensitive cell lines were considered for AUC < 1.
 
@@
 
-Differences in $log_2$-transformed transcript levels of the genes of interest of ssGSEA scores of the FGFR-related gene signatures between the LumP, LumU, stroma-rich, and Ba/Sq MIBC consensus molecular classes of urothelial cancers were explored with a two step procedure.
-In the first step, differences of expression levels or ssGSEA scores between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
+Differences in $log_2$-transformed transcript levels of the genes of interest of ssGSEA scores of the FGFR-related gene signatures between the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes of MIBC were explored with a two step procedure.
+In the first step, differences of the expression levels or ssGSEA scores between the analysis groups were assessed by one-way ANOVA with $\eta^2$ effect size statistic. 
 In the second step, differences between a given class and the LumP class were were evaluated by Welch's two-tailed T test with Cohen's d effect size. 
@@
 The significantly differentially regulated gene signatures were defined by pFDR(ANOVA) < 0.05, $\eta^2 \geq$ 0.14, pFDR(T test) < 0.05, and Cohen's d $\geq$ 0.5. 
-A similar testing procedure was applied to predictions of IC50 and AUC of pan-FGFR inhibitors. 
+A similar testing procedure was applied to the predictions of IC50 and AUC of pan-FGFR inhibitors made for bulk cancer samples. 
 However, due to non-normal distribution of some of the resistance metrics, Kruskal-Wallis with with $\eta^2$ statistic for testing for overall differences between the consensus subsets, and Mann-Whitney post-hoc test with biserial r effect size statistic was used for comparisons versus LumP cancers.
-The significant differences in predicted resistance were considered for pFDR(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and biserial r $\geq$ 0.3.
+Significant differences in the predicted resistance were considered for pFDR(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and biserial r $\geq$ 0.3.
 
@@
 The expression values were fed into the models in form of Z-scores of Combat-processed $log_2$ gene counts (TCGA BLCA, IMvigor) or Z-scores of $log_2$ gene counts (BCAN). 
-Assignment to the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response and ground truth for evaluation of the models.
+Assignment to the LumP, LumU, stroma-rich, and Ba/Sq consensus molecular classes by the nearest centroid classifier of _consensusMIBC_ R package served as the modeling response and the ground truth for evaluation of the model predictions.
 
@@
 The Elastic Net model of consensus molecular classes was optimized (tuned) and trained in the SMOTE-amplified training data set with a wrapper around functions `cv.glmnet()` and `glmnet()` from _glmnet_ package [@Friedman2010].
-The optimal values of the regularization parameter $\lambda$ was found by minimizing the Elastic Net models' deviance in 200 repeats 10-fold cross-validation. 
-The Random Forest models of consensus molecular subsets were tuned and trained inn the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
+The optimal values of the regularization parameter $\lambda$ was found by minimizing the Elastic Net models' deviance in 200 repeats 10-fold cross-validation of the SMOTE augmented training data set. 
+The Random Forest models of consensus molecular subsets were tuned and trained in the SMOTE-amplified training data set with a wrapper around `ranger()` function from _ranger_ package [@Wright2017]. 
 The optimal set of hyper-parameters that control behavior of the Random Forest models, 
@@
 and receiver-operating characteristic measures such as area under the curve (AUC, measure of overall sensitivity and specificity). 
-Those evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random (__Supplementary Figure S\@ref(fig:fig-ml-eval)A__, __Supplementary Table S\@ref(tab:tab-ml)__)
+These evaluation metrics were computed with `multiClassSummary()` function from _caret_ package [@Kuhn2008] and an in-house developed function that calculates Brier scores for the models of interest and a nonsense model predicting the classes at random (__Supplementary Figure S\@ref(fig:fig-ml-eval)A__, __Supplementary Table S\@ref(tab:tab-ml)__)
 Additionally, accuracy of prediction for single consensus molecular classes was assessed by inspection of the confusion matrices (__Supplementary Figure S\@ref(fig:fig-confusion)__). 
@@
 As the background for the SHAP procedure, we used randomly selected 30% of observations of the test subset of the TCGA BLCA collective. 
-Mean absolute SHAP value served as a measure of gene importance (__Supplementary Table S\@ref(tab:tab-shap)__). 
-SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions (__Figure 1F__, __Supplementary Figure S\@ref(fig:fig-shap-elnet)__ - __S\@ref(fig:fig-shap-fgfr)__).
+Mean absolute SHAP value served as a measure of gene importance (__Supplementary Tables S\@ref(tab:tab-shap)__ and __S\@ref(tab:tab-kamoun-shap)__). 
+SHAP metrics for single observations and the mean absolute SHAP values were visualized as violin/swarm and bar plots with tools of _shapviz_ package [@Mayer2023] and in-house developed R functions (__Figure 1D__, __Supplementary Figure S\@ref(fig:fig-rf-shap)__ - __S\@ref(fig:fig-shap-fgfr)__).
 
@@
 flextable::flextable(suppl_tabs$cohorts) %>% 
-  set_widths(c(2.5, rep(2.3, 6), 2.3, 2)) %>% 
+  set_widths(c(2.7, rep(2.3, 6), 2.3, 2)) %>% 
   footnote(1, 8:9, 
@@
   merge_v(1) %>% 
-  set_widths(c(2.2, 2.5, rep(2.8, 4), 2.3, 2)) %>% 
+  set_widths(c(2.2, 2.7, rep(2.8, 4), 2.3, 2)) %>% 
   footnote(1, 7:8, 
@@
 P values were corrected for multiple testing with the false discovery rate method.
-The table is available as a supplementary Excel file."}
+Frequency of alterations found in at least 1% of specimens in the BCAN, GENIE BLCA, IMvigor, MSK IMPACT, and TCGA BLCA are presented."}
 
-suppl_tabs$genetic_alterations[1, 1] %>% 
-  mutate(`Alteration type` = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$genetic_alterations) %>% 
+  merge_v(1) %>% 
+  merge_v(2) %>% 
+  merge_v(3) %>% 
+  italic(j = 3) %>% 
+  set_widths(c(2.5, 3, 2.2, 3, rep(2.5, 3))) %>% 
+  theme_vanilla
 
@@
 
-```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
+```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR3 and FGF3/4/19 genes in the luminal papillary (LumP), luminal unstable (LumU), stoma-rich, and basal/squamous like (Ba/Sq) consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
 The frequencies are presented as percentages of the consensus class.
 Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-The table is available as a supplementary Excel file."}
+P values were corrected for multiple testing with the false discovery rate method."}
 
-suppl_tabs$sub_genetics[1, 1] %>% 
-  mutate(Cohort = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$sub_genetics) %>% 
+  merge_v(1:2, combine = TRUE) %>% 
+  italic(j = 2) %>% 
+  set_widths(c(2.3, 2.4, 2.3, rep(2.4, 3), rep(2.4, 2))) %>% 
+  theme_vanilla
 
@@
 
-\newepage
+\newpage
 
@@
 The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells.
-Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed.
-The table is available as a supplementary Excel file."}
+Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed."}
 
-suppl_tabs$expression[1, 1] %>% 
-  mutate(Variable = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$expression) %>% 
+  italic(i = 3:47, j = 1) %>% 
+  set_widths(c(2.5, rep(3.5, 5))) %>% 
+  theme_vanilla
 
 ```
+
+\newpage
 
@@
 P values were corrected for multiple testing with the false discovery rate method.
-log2-transformed expression levels are presented as medians with interquartile ranges and ranges.
-The table is available as a supplementary Excel file."}
+log2-transformed expression levels are presented as medians with interquartile ranges and ranges for the top genes whose expression differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}
 
-suppl_tabs$sub_dge[1, 1] %>% 
-  mutate(Cohort = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$sub_dge) %>% 
+  merge_v(1) %>% 
+  italic(j = 2, italic = suppl_tabs$sub_dge$Variable != 'Samples, N') %>% 
+  set_widths(c(1.6, 2.1, rep(3.2, 4), rep(2.3, 2))) %>% 
+  theme_vanilla
 
 ```
+
+\newpage
 
@@
 P values were corrected for multiple testing with the false discovery rate method.
-ssGSEA scores are presented as medians with interquartile ranges and ranges.
-The table is available as a supplementary Excel file."}
+ssGSEA scores are presented as medians with interquartile ranges and ranges for the top signatures whose scores differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}
 
-suppl_tabs$sub_sig[1, 1] %>% 
-  mutate(Cohort = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$sub_sig) %>% 
+  merge_v(1) %>% 
+  set_widths(c(1.6, 3.2, rep(2.9, 4), rep(2.3, 2))) %>% 
+  theme_vanilla
 
@@
 flextable::flextable(suppl_tabs$sub_drugs) %>% 
-  set_widths(c(1.4, 2.2, 1.6, rep(2.8, 4), 2.3, 2)) %>% 
+  set_widths(c(1.6, 2.2, 1.6, rep(2.9, 4), 2.3, 2)) %>% 
   merge_v(1) %>% 
   footnote(1, 3, 
-           as_paragraph('IC50: concentation in µM resulting in 50% inhibition of growth; AUC: area under the dose-response curve.'), 
+           as_paragraph('IC50: concentration in µM resulting in 50% inhibition of growth; AUC: area under the dose-response curve.'), 
            part = 'header', 
@@
   footnote(1, 8:9, 
-           value = as_paragraph('Statistical significance: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected for multiple tesing with the false discovery rate method.'), 
+           value = as_paragraph('Statistical significance: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
+P values corrected for multiple testing with the false discovery rate method.'),
            part = 'header', 
@@
   footnote(1, 1, 
-           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centrouid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
+           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centroid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
            part = 'header', 
@@
   footnote(1, 4, 
-           value = as_paragraph('lambda: shrinkage/regularization paremeter; mtry: number of randomly chosen explanatroy variables for construction of decision trees in the Random Forest ensemble; min.node.size: minimal number of observations in the terminal/leaf nodes of decision trees; splitrule: algorithm which defines splits at the nodes of decision trees.'),
+           value = as_paragraph('lambda: shrinkage/regularization parameter; mtry: number of randomly chosen explanatory variables for construction of decision trees in the Random Forest ensemble; min.node.size: minimal number of observations in the terminal/leaf nodes of decision trees; splitrule: algorithm which defines splits at the nodes of decision trees.'),
            part = 'header', 
@@
   merge_v(1:2) %>% 
-  set_widths(c(2.4, 2.2, 2.2, rep(2, 6))) %>% 
+  set_widths(c(2.6, 2.2, 3.5, rep(2, 6))) %>% 
   footnote(1, 1, 
-           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centrouid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
+           value = as_paragraph('Kamoun et al.: 818 genuine MIBC consensus class-defining genes used by the nearest centroid tool by Kamoun et al.; FGFR/FGF/FGFBP: 38 FGFR/FGF/FGFBP-coding genes.'), 
            part = 'header', 
@@
 ```{r tab-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of MIBC consensus molecular classes estimated by the SHAP algorithm.
-Contribution of the genuine MIBC consensus class-defining genes and the FGFR-, FGF-, and FGFBP-coding genes to prediction of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations).
-The table is available as a supplementary Excel file."}
+The models employed the FGFR-, FGF-, and FGFBP-coding genes as explanatory factors.
+Contribution of the explanatory factors to predictions of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations).
+Top five most influential genes for the predictions of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) classes by the Elastic Net and Random Forest algorithms are listed."}
 
-suppl_tabs$shap[1, 1] %>% 
-  mutate(`Explanatory factors/genes` = '') %>% 
-  set_names(' ') %>% 
-  flextable %>% 
-  set_widths(18)
+flextable::flextable(suppl_tabs$shap) %>% 
+  merge_v(1:2) %>% 
+  italic(j = 3) %>% 
+  set_widths(c(2.5, 2.5, 2.5, 3.5, 2.5)) %>% 
+  theme_vanilla
+
+```
+
+\newpage
+
+```{r tab-kamoun-shap, tab.cap = "Importance of explanatory variables for the Elastic Net and Random Forest models of MIBC consensus molecular classes estimated by the SHAP algorithm.
+The models employed the genuine MIBC consensus class-defining genes by Kamoun et al.
+as explanatory factors.
+Contribution of the explanatory factors to predictions of consensus molecular classes was assessed by mean absolute values of SHAP metrics (Shapley additive explanations).
+Top five most influential genes for the predictions of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) classes by the Elastic Net and Random Forest algorithms are listed."}
+
+flextable::flextable(suppl_tabs$kamoun_shap) %>% 
+   merge_v(1:2) %>% 
+  italic(j = 3) %>% 
+  set_widths(c(2.5, 2.5, 2.5, 3.5, 2.5)) %>% 
+  theme_vanilla
 
@@
 
-```{r fig-sub-genet, fig.width = figur::convert(suppl_figs$sub_genet, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_genet, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.'}
+```{r fig-sub-genet, fig.width = figur::convert(suppl_figs$sub_genet, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_genet, to = 'in')$h, fig.cap = 'Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of MIBC.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-sub-genet). Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of urothelial cancers.__
+__Supplementary Figure S\@ref(fig:fig-sub-genet). Alterations of FGFR-, FGF-, and FGFBP-coding genes in consensus molecular classes of MIBC.__
 
@@
 Numbers of analyzed cancer samples are displayed in the plot captions._
+
+\newpage
+
+```{r fig-fgfr3-hot, fig.width = figur::convert(suppl_figs$fgfr3_hot, to = 'in')$w, fig.height = figur::convert(suppl_figs$fgfr3_hot, to = 'in')$h, fig.cap = 'Mutational hot spots of the FGFR3 gene.'}
+
+suppl_figs$fgfr3_hot$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-fgfr3-hot). Mutational hot spots of the FGFR3 gene.__
+
+_Frequency of mutations affecting residues of the FGFR3 protein was expressed as percentage of all cancer samples in the `r globals$cohort_labs[c("genie", "msk", "tcga", "bcan")] %>% collapse_and` cohorts and displayed in dot plots.
+Each point represents a single protein residue, point color codes for assignment of the residue to the protein domain.
+Candidate mutation hot spots with mutation rates $\geq$ 1% are highlighted with enlarged symbols and labeled with percentages and protein sequence changes.
+Cohort names and numbers of analyzed cancer samples are indicated in the plot facets.
+The protein domain scheme is displayed above the plot._
+
+\newpage
+
+```{r fig-expression, fig.width = figur::convert(suppl_figs$expression, to = 'in')$w, fig.height = figur::convert(suppl_figs$expression, to = 'in')$h, fig.cap = 'Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.'}
+
+suppl_figs$expression$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-expression). Expression of genes coding FGFR, FGF, and FGFBP at mRNA and protein level.__ 
+_Levels of FGFR-, FGF-, and FGFBP-coding mRNA were determined in cancer tissue samples in the TCGA BLCA, IMvigor, and BCAN by RNA sequencing and published as $log_2$ gene counts (TCGA BLCA and IMvigor) or Z-scores of $log_2$ gene counts (BCAN).
+mRNA expression of the genes of interest in DepMap urothelial cancer cell lines was measured by RNA sequencing and provided as $log_2$ gene counts.
+Median Z-scores of $log_2$ mRNA expression are shown in a heat map.
+Each tile represents median expression Z-score of a gene in one cohort.
+Numbers of cancer samples are indicated in the X axis.
+White tiles represent genes not detected in each cohort.
+Protein expression of `r expl_expr$plot_data$hpa %>% filter(complete.cases(.)) %>% .$variable %>% unique %>% length` FGFR-, FGF-, and FGFBP-coding genes in the Human Protein Atlas (HPA) repository was quantified by immunohistochemistry (IHC).
+The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%).
+Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges.
+Single IHC specimens are visualized as points.
+Numbers of IHC samples are indicated in the Y axis._
 
@@
 
+```{r fig-bulk-networks, fig.width = figur::convert(suppl_figs$bulk_networks, to = 'in')$w, fig.height = figur::convert(suppl_figs$bulk_networks, to = 'in')$h, fig.cap = 'Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes in the IMvigor and BCAN cohorts.'}
+
+suppl_figs$bulk_networks$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-bulk-networks). Co-expression networks of FGFR-, FGF-, and FGFBP-coding genes in the IMvigor and BCAN cohorts.__
+
+_Co-expression networks were constructed in the IMvigor and BCAN cohorts for the FGFR-, FGF-, and FGFBP-coding genes, whose expression levels correlated with Spearman’s $\rho \geq$ 0.3.
+Isolated vertices of the networks were removed.
+The networks were visualized using the Fruchterman-Reingold algorithm.
+Points represent node vertices/genes, point shape codes for classification of the protein product.
+Point color represent hub scores, which measure the overall correlation strength.
+Lines represent edges, i.e correlations between expression levels with $\rho \geq$ 0.3.
+Line widths correspond to $\rho$ values.
+Note the central position, connectivity, and high hub scores of the genes coding for FGFR1, its ligands FGF2, FGF7 and FGF10, and the binding proteins SDC2, DCN, PTX3 and HSPG2 preserved in all cohorts.
+This suggests that ligand-dependent signaling via FGFR1 may serve as a substantial signaling pathway in urothelial cancer._
+
+\newpage
+
 ```{r fig-cell-networks, fig.width = figur::convert(suppl_figs$cell_networks, to = 'in')$w, fig.height = figur::convert(suppl_figs$cell_networks, to = 'in')$h, fig.cap = 'Co-expression network of FGF-, FGFR-, and FGFBP-coding genes in DepMap urothelial cancer cell lines'}
@@
 
-```{r fig-sub-fgfr-dge, fig.width = figur::convert(suppl_figs$sub_dge, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_dge, to = 'in')$h, fig.cap = 'Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.'}
+```{r fig-sub-fgfr-dge, fig.width = figur::convert(suppl_figs$sub_dge, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_dge, to = 'in')$h, fig.cap = 'Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of MIBC.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-sub-fgfr-dge). Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of urothelial cancers.__
+__Supplementary Figure S\@ref(fig:fig-sub-fgfr-dge). Differential expression of FGF-, FGFR-, and FGFR-coding genes in the consensus molecular classes of MIBC.__
 
@@
 Numbers of cancer samples in the consensus classes are indicated in the plot captions.
-The full analysis results are shown in Supplementary Table S\@ref(tab:tab-sub-dge)._
+The analysis results are also shown in Supplementary Table S\@ref(tab:tab-sub-dge)._
 
@@
 
-```{r fig-sub-sig, fig.width = figur::convert(suppl_figs$sub_sig, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_sig, to = 'in')$h, fig.cap = 'Scores of FGFR-related gene signatures in the consensus molecular classes of urothelial cancers.'}
+```{r fig-sub-sig, fig.width = figur::convert(suppl_figs$sub_sig, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_sig, to = 'in')$h, fig.cap = 'Scores of FGFR-related gene signatures in the consensus molecular classes of MIBC.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-sub-sig). Scores of FGFR-related gene signatures in the consensus molecular classes of urothelial cancers.__
+__Supplementary Figure S\@ref(fig:fig-sub-sig). Scores of FGFR-related gene signatures in the consensus molecular classes of MIBC__
 
@@
 Numbers of cancer samples in the consensus classes are indicated in the plot captions.
-The full analysis results with signature names and listing of the member genes are presented in Supplementary Table S\@ref(tab:tab-sub-sig)._
+The analysis results with signature names are also presented in Supplementary Table S\@ref(tab:tab-sub-sig)._
 
@@
 
-```{r fig-sub-drugs, fig.width = figur::convert(suppl_figs$sub_drugs, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_drugs, to = 'in')$h, fig.cap = 'Predicted response to pan-FGFR inhibitors in the consensus molecular classes of urothelial cancers.'}
+```{r fig-sub-drugs, fig.width = figur::convert(suppl_figs$sub_drugs, to = 'in')$w, fig.height = figur::convert(suppl_figs$sub_drugs, to = 'in')$h, fig.cap = 'Predicted response to pan-FGFR inhibitors in the consensus molecular classes of MIBC.'}
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-sub-drugs). Predicted response to pan-FGFR inhibitors in the consensus molecular classes of urothelial cancers.__
+__Supplementary Figure S\@ref(fig:fig-sub-drugs). Predicted response to pan-FGFR inhibitors in the consensus molecular classes of MIBC.__
 
@@
 
+```{r fig-rf-shap, fig.width = figur::convert(suppl_figs$rf_shap, to = 'in')$w, fig.height = figur::convert(suppl_figs$rf_shap, to = 'in')$h, fig.cap = 'SHAP importance for explanatory variables in the Random Forest model of consensus molecular classes of MIBC with FGFR/FGF/FGFBP-coding genes as explanatory factors.'}
+
+suppl_figs$rf_shap$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-rf-shap). SHAP variable importance in the Random Forest model of consensus molecular classes of MIBC with FGFR/FGF/FGFBP-coding genes as explanatory factors.__
+
+_Contributions of the FGFR-, FGF-, and FGFBP-coding genes to prediction of luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes of MIBC by the Random Forest machine learning model were estimated with the SHAP algorithm (Shapley additive explanations).
+SHAP values for highly influential genes and observations are visualized in violin/swarm plots.
+Each point represents a single observation, point color codes for the minimum/maximum-scaled $log_2$-transformed expression of the gene.
+The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
+Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
+
+\newpage
+
 ```{r fig-shap-elnet, fig.width = figur::convert(suppl_figs$shap_elnet, to = 'in')$w, fig.height = figur::convert(suppl_figs$shap_elnet, to = 'in')$h, fig.cap = 'Importance of explanatory variables in the Elastic Net model of MIBC molecular consensus classes of urothelial cancers with the full set of class-defining genes by Kamoun et al.
@@
 
-_Shapley additive explanations (SHAP) as measures of variable importance for predictions of the consensus molecular classes by the Elastic Net and Random Forest models employing the full class-defining predictor set by Kamoun eta al.
+_Shapley additive explanations (SHAP) as measures of variable importance for predictions of the consensus molecular classes by the Elastic Net and Random Forest models employing the full class-defining predictor set by Kamoun et al..
 were computed as presented in Supplementary Figures S\@ref(fig:fig-shap-elnet) and S\@ref(fig:fig-shap-elnet)._
@@
 Single explanatory factors are depicted as points; points corresponding to the FGFR/FGF/FGFBP-coding genes among the class-defining genes are highlighted with blue (negative class assignment predictor) and red (positive predictors of class assignment)._
+
+\newpage
+
+```{r fig-pani-response, fig.width = figur::convert(suppl_figs$pani_response, to = 'in')$w, fig.height = figur::convert(suppl_figs$pani_response, to = 'in')$h, fig.cap = 'Sensitivity to pan-FGFR inhibitors of DepMap urothelial cancer cell lines in the GDSC1/2 drug screening experiments.'}
+
+suppl_figs$pani_response$plot
+
+```
+
+__Supplementary Figure S\@ref(fig:fig-pani-response). Sensitivity to pan-FGFR inhibitors of DepMap urothelial cancer cell lines in the GDSC1/2 drug screening experiments.__
+
+_Information on resistance of urothelial cancer cell lines to pan-FGFR inhibitors in the GDSC1/2 (n = `r expl_res$sensitive$gdsc %>% select(-sample_id, -FGFR3_mutation) %>% map(~complete.cases(.x)) %>% map_dbl(sum) %>% range %>% paste(collapse = ' to ')` cell lines) drug screening experiments was extracted from the DepMap repository.
+The resistance values were expressed as IC50 (concentration resulting in 50% growth inhibition).
+Significant sensitivity was defined as IC50 below 10 µM.
+IC50 for doxorubicin are presented as a positive cytotoxicity controls.
+Median IC50 values for the FGFR inhibitors with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
+Points represent single cell lines.
+Point symbols code for the FGFR3 gene mutation status, point color represents sensitivity/resistance.
+The sensitivity cutoff is represented by the dashed line._
 

### 2025-04-28 paper scripts/figure.R

@@ -1,5 +1,5 @@
-# elements of figure 1 of the short correspondence.
-# the figure is subsequently stitched by hand
+# Elements of Figure 1 of the short correspondence.
+# the figure is subsequently stitched by hand.
 
   insert_head()
 
@@ -29,107 +29,38 @@
                       strip.text.x = globals$figure_text),
               label = 'figure_1a_fgfr_fgf_mutations_oncoplot_tcga',
               ref_name = 'figure_1a',
-              caption = paste('Alterations of FGFR-, FGF-, and FGFBP-coding',
-                              'genes in MIBC consensus molecular classes of',
-                              'urothelial cancer.'),
-              w = 180,
+              caption = paste('Alterations of FGFR/FGF/FGFBP-coding',
+                              'genes in consensus molecular classes of MIBC.'),
+              w = 150,
               h = 110)
 
-# Figure 1B: mutation density and mutation hotspots, FGFR3 ------
+# Figure 1B: co-expression networks ---------
 
-  insert_msg('Figure 1B: mutation density and mutation hotspots, FGFR3')
+  insert_msg('Figure 1B: co-expression networks')
 
-  ## density plot and detailed plot
+  ## showing only the TCGA cohort, the rest for the supplements
 
   fig$figure_1b <-
-    plot_grid(expl_hot$domain_density_plots_paper$FGFR3,
-              expl_hot$domain_percentage_plots_paper$protein_domain$FGFR3,
-              ncol = 2,
-              align = 'hv',
-              axis = 'tblr') %>%
-    plot_grid(get_legend(expl_fgfr$domain_plots$FGFR3 +
-                           globals$figure_theme +
-                           theme(legend.position = 'bottom',
-                                 legend.text = element_text(margin =
-                                                              ggplot2::margin(r = 10)))),
-              nrow = 2,
-              rel_heights = c(0.9, 0.1)) %>%
-    as_figure(label = 'figure_1b_fgfr3_mutation_hotspots',
-              ref_name = 'figure_1b',
-              caption = 'Mutational hot spots of the FGFR3 gene.',
-              w = 190,
-              h = 160)
-
-# Figure 1C: mRNA and protein expression of the genes of interest --------
-
-  insert_msg('Figure 1C: mRNA and protein expression')
-
-  ## left: heat map with the mRNA expression Z scores
-  ## right: box plots with IHVC scores
-
-  fig$figure_1c <-
[…]

### 2025-04-28 paper scripts/render.R

@@ -13,17 +13,26 @@
 
   insert_msg('Rendering manuscript parts')
 
+  ## figure 1 and Supplementary Material
+
   render('./report/markdown/figure1_legend.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
            output_dir = './report')
+
+  render('./report/markdown/supplementary_material.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './report')
+
+  ## Rebuttal Letters for the first and second revision
 
   render('./report/markdown/rebuttal_letter.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
          output_dir = './report')
 
-  render('./report/markdown/supplementary_material.Rmd',
+  render('./report/markdown/rebuttal_letter2.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
          output_dir = './report')

### 2025-04-28 paper scripts/supplementary_figures.R

@@ -29,8 +29,7 @@
     as_figure(label = 'consensus_subtypes_fgf_fgfr_gene_alterations',
               ref_name = 'sub_genet',
               caption = paste('Alterations of FGFR-, FGF-, and FGFBP-coding',
-                              'genes in consensus molecular classes of',
-                              'urothelial cancers.'),
+                              'genes in consensus molecular classes of MIBC.'),
               w = 180,
               h = 90)
 
@@ -64,7 +63,75 @@
               w = 180,
               h = 140)
 
-# Representative stainings for FGFR/FGF/FGFBP -------
+# Mutation density and mutation hotspots, FGFR3 ------
+
+  insert_msg('Mtation density and mutation hotspots, FGFR3')
+
+  ## density plot and detailed plot
+
+  suppl_figs$fgfr3_hot <-
+    plot_grid(expl_hot$domain_density_plots_paper$FGFR3,
+              expl_hot$domain_percentage_plots_paper$protein_domain$FGFR3,
+              ncol = 2,
+              align = 'hv',
+              axis = 'tblr') %>%
+    plot_grid(get_legend(expl_fgfr$domain_plots$FGFR3 +
+                           globals$figure_theme +
+                           theme(legend.position = 'bottom',
+                                 legend.text = element_text(margin =
+                                                              ggplot2::margin(r = 10)))),
+              nrow = 2,
+              rel_heights = c(0.9, 0.1)) %>%
+    as_figure(label = 'fgfr3_mutation_hotspots',
+              ref_name = 'fgfr3_hot',
+              caption = 'Mutational hot spots of the FGFR3 gene.',
+              w = 190,
+              h = 160)
+
+# mRNA and protein expression of the genes of interest --------
+
+  insert_msg('mRNA and protein expression')
+
+  ## left: heat map with the mRNA expression Z scores
+  ## right: box plots with IHVC scores
+
+  suppl_figs$expression <-
+    list(x = list(expl_expr$hm_plot, expl_expr$plots$hpa),
+         y = c('mRNA expression', 'Protein expression'),
+         z = c('italic', 'plain'),
+         v = c(45, 0)) %>%
+    pmap(function(x, y, z, v) x +
+           guides(x = guide_axis(angle = v)) +
+           labs(title = y) +
+           #globals$figure_theme +
+           theme(plot.subtitle = element_blank(),
+                 axis.text.y = element_text(face = z),
+                 axis.title.y = element_blank()))
+
[…]

### 2025-04-28 paper scripts/supplementary_tables.R

@@ -2,9 +2,24 @@
 
   insert_head()
 
-# container ------
+# containers ------
 
   suppl_tabs <- list()
+  suppl_globals <- list()
+
+# some globals ----
+
+  insert_msg('Table globals')
+
+  ## levels of classes and cohorts
+  ## for arr<anging the supplementary tables
+
+  suppl_globals$class_levels <- c('LumP', 'LumU', 'Stroma-rich', 'Ba/Sq')
+
+  suppl_globals$cohort_levels <- globals$analysis_cohorts
+
+  suppl_globals$cohort_lab_levels <-
+    unname(globals$cohort_labs[suppl_globals$cohort_levels])
 
 # Characteristic of the bulk cancer and IHC cohorts -------
 
@@ -61,7 +76,13 @@
 
   insert_msg('Frequency of genetic alterations in FGFR/FGF/FGFBP genes')
 
-  suppl_tabs$genetic_alterations <- expl_genet$fgf_result_tbl %>%
+  ## to meet the request by a reviewer: showing only the alterations
+  ## presenting at least 1% of the samples
+
+  suppl_tabs$genetic_alterations <-
+    map2(expl_genet$fgf_result_tbl,
+         expl_genet$fgf_top_alterations,
+         ~filter(.x, `Gene symbol` %in% .y)) %>%
     compress(names_to = 'Alteration type') %>%
     relocate(`Alteration type`) %>%
     mdtable(label = 'genetic_fgf_fgfr_features',
@@ -74,19 +95,43 @@
                             "with Cramer's V effect size statistic.",
                             'P values were corrected for multiple testing with',
                             'the false discovery rate method.',
-                            'The table is available as a supplementary Excel',
-                            'file.'))
+                            'Frequency of alterations found in at least 1% of',
+                            'specimens in the BCAN, GENIE BLCA, IMvigor,',
+                            'MSK IMPACT, and TCGA BLCA are presented.'))
 
 # Alterations of the FGFR/FGF/FGFBP genes in the consensus classes of MIBC ------
 
   insert_msg('Genetics of FGFR/FGF/FGFBP in the consensus MIBC classes')
 
+  ## showing the same genetic alterations as in the previous table
+  ## (mutations of )
+
   suppl_tabs$sub_genetics <- sub_genet$result_tbl %>%
+    mutate(Alteration = factor(Alteration,
[…]

### 2025-05-15 paper scripts/figure.R
Revision, figure and table code
Adjustment of Figure 1E and Figure 1F, and some supplementary figures. Formatting of supplementary tables according to the EAU guidelines

@@ -106,15 +106,18 @@
 
   insert_msg('Figure 1E: PRISM drug screening results')
 
+  ## violin plots, according to the rule 21 of the figure/table guidelines
   ## Results of GDSC screening go into Supplementary Material
 
-  fig$figure_1e <- expl_res$box_plots$prism +
+  fig$figure_1e <- expl_res$violin_plots$prism +
     scale_y_discrete(labels = drug_labeller) +
     globals$figure_theme +
     theme(legend.title = element_markdown(),
           strip.background.y = element_blank(),
           strip.text.y = element_blank(),
-          axis.title.y = element_blank())
+          axis.title.y = element_blank(),
+          legend.position = "bottom",
+          legend.box = "vertical")
 
   fig$figure_1e <- fig$figure_1e %>%
     as_figure(label = 'figure_1e_fgfr_pan_inhibitors_reistance_cell_lines',
@@ -122,16 +125,18 @@
               caption = paste('Sensitivity to pan-FGFR',
                               'inhibitors of DepMap urothelial cancer',
                               'cell lines.'),
-              w = 120,
-              h = 90)
+              w = 100,
+              h = 100)
 
 # Figure 1F: resistance and sensitivity, cell lines WT and FGFR3 mutant ------
 
   insert_msg('Figure 1F: cell lines, FGFR3 mutation/WT')
 
   fig$figure_1f <-
-    plot_grid(ggdraw() +
-                draw_image('./report/aux files/erdafitinib_summary.png')) %>%
+    plot_grid(iv_plots$line_plot +
+                globals$figure_theme +
+                theme(legend.position = "none",
+                      strip.text.x = element_markdown())) %>%
     as_figure(label = 'figure_1f_in_vitro_cell_lines_erdafitinib',
               ref_name = 'figure_1f',
               caption = paste('Confluence of cultured FGFR1-4 wild-type',
@@ -139,8 +144,8 @@
                               '(UM-UC-14 and UM-UC-6) UC cell lines after',
                               'treatment with different concentrations of',
                               'erdafitinib.'),
-              w = 180,
-              h = 90)
+              w = 200,
+              h = 100)
 
 # The complete Figure 1 ---------
 
@@ -172,7 +177,7 @@
     plot_grid(fig$figure_1e$plot,
               fig$figure_1f$plot,
               ncol = 2,
-              rel_widths = c(fig$figure_1e$w, fig$figure_1h$w),
[…]

### 2025-05-15 paper scripts/render.R

@@ -25,7 +25,7 @@
                                         reference_docx = 'ms_template.docx'),
          output_dir = './report')
 
-  ## Rebuttal Letters for the first and second revision
+  ## Rebuttal Letters for the first, second, and third revision
 
   render('./report/markdown/rebuttal_letter.Rmd',
          output_format = word_document2(number_sections = FALSE,
@@ -33,6 +33,11 @@
          output_dir = './report')
 
   render('./report/markdown/rebuttal_letter2.Rmd',
+         output_format = word_document2(number_sections = FALSE,
+                                        reference_docx = 'ms_template.docx'),
+         output_dir = './report')
+
+  render('./report/markdown/rebuttal_letter3.Rmd',
          output_format = word_document2(number_sections = FALSE,
                                         reference_docx = 'ms_template.docx'),
          output_dir = './report')

### 2025-05-15 paper scripts/supplementary_figures.R

@@ -277,7 +277,7 @@
               align = 'hv',
               axis = 'tblr')
 
-  ## bottom panel: box plots for erdafitinib
+  ## bottom panel: violin plots for erdafitinib
 
   suppl_figs$sub_drugs$bottom <- sub_drugs$fgfr_plots %>%
     map(~.x$erdafitinib) %>%
@@ -437,7 +437,7 @@
 
 # SHAP importance, Random Forest model, FGFR/FGF/FGFB independent variables -----------
 
-  insert_msg('Figure 1D: SHAP importances, Random Forest')
+  insert_msg('SHAP importances, Random Forest')
 
   suppl_figs$rf_shap <- ml_splots$plots$ranger$panels %>%
     map(~.x + guides(x = guide_axis(angle = 45))) %>%
@@ -529,7 +529,7 @@
 
   insert_msg('Response to pan-FGFRi, GDSC')
 
-  suppl_figs$pani_response <- expl_res$box_plots$gdsc +
+  suppl_figs$pani_response <- expl_res$violin_plots$gdsc +
     theme(legend.title = element_markdown(),
           axis.title.y = element_blank())
 

### 2025-05-15 paper scripts/supplementary_tables.R

@@ -25,14 +25,20 @@
 
   insert_msg('Charcterisric of bulk cancer and IHC cohorts')
 
+  ## we don't show hypothesis testing, according to the EAU guidelines
+  ## it is quite obvious that the cohorts are pretty different
+  ## due to the srudy designs
+
   suppl_tabs$cohorts <- expl_cohorts$result_tbl %>%
+    select(-`FDR p value`, -`Effect size`) %>%
     as_mdtable(label = 'cohorts',
                ref_name = 'cohorts',
-               caption = paste('Characteristic of the investigated cohorts.',
-                               'Numeric features are presented as medians with',
-                               'interquartile ranges and ranges.',
-                               'Qualitative variables are shown as percentages',
-                               'and observation counts of the categories.'))
+               caption = paste("Characteristic of the investigated cohorts.",
+                               "Numeric variables are presented as medians with",
+                               "interquartile ranges.",
+                               "Qualitative variables are shown as counts of",
+                               "the categories and percentages of the categories",
+                               "within the complete observation set."))
 
 # Clinical characteristic of the consensus molecular classes of MIBC ------
 
@@ -41,32 +47,47 @@
   suppl_tabs$sub_clinic <- sub_clinic$result_tbl %>%
     as_mdtable(label = 'sub_clinic',
                ref_name = 'sub_clinic',
-               caption = paste('Characteristic of the consensus molecular',
-                               'classes of MIBC in the TCGA BLCA, IMvigor,',
-                               'and BCAN cohorts.',
-                               'Numeric features are presented as medians with',
-                               'interquartile ranges and ranges.',
-                               'Qualitative variables are shown as percentages',
-                               'and observation counts of the categories.'))
+               caption = paste("Characteristic of the major consensus molecular",
+                               "classes of MIBC (luminal papillary [LumP],",
+                               "luminal unstable [LumU], stroma-rich,",
+                               "basal/squamous-like [Ba/Sq]) in the",
+                               "TCGA BLCA, IMvigor, and BCAN cohorts.",
+                               "Numeric variables are presented as medians with",
+                               "interquartile ranges.",
+                               "Qualitative variables are shown as counts of",
[…]

### 2025-05-15 paper.R

@@ -12,6 +12,7 @@
   library(cowplot)
   library(figur)
   library(flextable)
+  library(ftExtra)
   library(writexl)
 
   library(rmarkdown)

### 2025-05-15 report/markdown/figure1_legend.Rmd
Figure, table, and markdown update
Adjusted figure and tables legends, and changes to supplementary methods to meet the EAU guidelines

@@
 Doxorubicin is presented as a positive cytotoxicity control.
-Median AUC values with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
-Points represent single cell lines.
-Point symbols code for FGFR3 mutation, point color represents resistance/sensitivity.
+AUC values are depicted in violin plots; points represent single cell lines.
+Point color codes for FGFR3 mutation, point shape represents resistance/sensitivity.
 The sensitivity cutoff is represented by the dashed line._
@@
 
-__Figure \@ref(fig:fig-figure-1f).
-Confluence of cultured FGFR1-4 wild-type (UM-UC-3, 5637 and RT112) and FGFR3 mutated (UM-UC-14 and UM-UC-6) UC cell lines after treatment with different concentrations of erdafitinib.__
+__Figure \@ref(fig:fig-figure-1f). Sensitivity of *FGFR3* wild-type and *FGFR3* mutant urothelial cancer cell lines to erdafitinib.__
 
-_Confluence was measured after 96 hours in cultures treated with erdafitinib at 0, 1, 4, 7 and 10 µM.
-Dots represent means of three biological replicates.
-Mean values with standard deviations are depicted as bars with whiskers._
+_Confluence of cultures of FGFR1/2/3/4 wild-type (WT; 5637, RT-112, UM-UC-3) and FGFR3 mutant cell lines (UM-UC-6, UM-UC-14) at 96 hours following treatment with 0 (control), 1, 4, 7, and 10 µM erdafitinib.
+Points represent measurements obtained in three independent biological replicates; data points obtained in the same replicate are connected with lines._

### 2025-05-15 report/markdown/supplementary_material.Rmd
Figure, table, and markdown update
Adjusted figure and tables legends, and changes to supplementary methods to meet the EAU guidelines

@@
 
-If not indicated otherwise, descriptive statistics for quantitative variables included median, interquartile range, and range. 
-Descriptive statistic for qualitative variables were percentage and count within categories. 
+If not indicated otherwise, descriptive statistics for quantitative variables included median and interquartile range. 
+Descriptive statistic for qualitative variables were counts of observation in the categories and percentages of the complete observation set. 
 Differences of means or medians of qualitative variables were investigated by Mann-Whitney, two-tailed T test, Kruskal-Wallis test, and one-way ANOVA, depending on the number of analysis groups and normality of distribution of the variable.
@@
 
-```{r tab-cohorts, tab.cap = 'Characteristic of the investigated cohorts.
-Numeric features are presented as medians with interquartile ranges and ranges.
-Qualitative variables are shown as percentages and observation counts of the categories.'}
+```{r tab-cohorts, tab.cap = "Characteristic of the investigated cohorts.
+Numeric variables are presented as medians with interquartile ranges.
+Qualitative variables are shown as counts of the categories and percentages of the categories within the complete observation set."}
 
 flextable::flextable(suppl_tabs$cohorts) %>% 
-  set_widths(c(2.7, rep(2.3, 6), 2.3, 2)) %>% 
-  footnote(1, 8:9, 
-           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
+  set_widths(c(3, rep(2.8, 6))) %>% 
+  footnote(i = c(2, 7), 
+           j = 1, 
+           value = as_paragraph("Numeric variables: median (first quartile, third quartile) is presented."), 
+           part = "body", 
+           ref_symbols = "a") %>% 
+  footnote(i = c(3:6, 8:17), 
+           j = 1, 
+           value = as_paragraph("Qualitative variables: number of observations (percentage of complete cases) is presented."),
+           part = "body", 
+           ref_symbols = "b") %>% 
   theme_vanilla
@@
 
-```{r tab-sub-clinic, tab.cap = "Characteristic of the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
-Numeric features are presented as medians with interquartile ranges and ranges.
-Qualitative variables are shown as percentages and observation counts of the categories."}
+```{r tab-sub-clinic, tab.cap = "Characteristic of the major consensus molecular classes of MIBC (luminal papillary [LumP], luminal unstable [LumU], stroma-rich, basal/squamous-like [Ba/Sq]) in the TCGA BLCA, IMvigor, and BCAN cohorts.
+Numeric variables are presented as medians with interquartile ranges.
+Qualitative variables are shown as counts of the categories and percentages of the categories within the complete observation set.
+Statistical significance of differences between the consensus molecular classes were assessed by Kruskal-Wallis and χ² test with, respectively, η² and Cramer's V effect size statistics.
+P values were corrected for multiple testing with the false discovery rate method (FDR)."}
 
@@
   set_widths(c(2.2, 2.7, rep(2.8, 4), 2.3, 2)) %>% 
-  footnote(1, 7:8, 
-           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
+  footnote(i = c(2, 25), 
+           j = 2, 
+           value = as_paragraph("Numeric variables: median (first quartile, third quartile) is presented."),
+           part = "body", 
+           ref_symbols = "a") %>% 
+  footnote(i = c(1:nrow(suppl_tabs$sub_clinic))[!c(1:nrow(suppl_tabs$sub_clinic)) %in% c(1, 2, 15, 24, 25)], 
+           j = 2, 
+           value = as_paragraph("Qualitative variables: number of observations (percentage of complete cases) is presented."),
+           part = "body", 
+           ref_symbols = "b") %>% 
+  footnote(i = c(2, 25), 
+           j = 7, 
+           value = as_paragraph("Kruskal-Wallis test with \u03B7\u00B2 effect size statistic."),
+           part = "body", 
+           ref_symbols = "c") %>% 
+  footnote(i = c(2, 25), 
+           j = 8, 
+           value = as_paragraph("Kruskal-Wallis test with \u03B7\u00B2 effect size statistic."),
+           part = "body", 
+           ref_symbols = "c") %>% 
+  footnote(i = c(1:nrow(suppl_tabs$sub_clinic))[!c(1:nrow(suppl_tabs$sub_clinic)) %in% c(1, 2, 15, 24, 25)], 
+           j = 7, 
+           value = as_paragraph("\u03C7\u00B2 test with Cramer's V effect size statistic."),
+           part = "body", 
+           ref_symbols = "d") %>% 
+  footnote(i = c(1:nrow(suppl_tabs$sub_clinic))[!c(1:nrow(suppl_tabs$sub_clinic)) %in% c(1, 2, 15, 24, 25)], 
+           j = 8, 
+           value = as_paragraph("\u03C7\u00B2 test with Cramer's V effect size statistic."),
+           part = "body", 
+           ref_symbols = "d") %>% 
   theme_vanilla
@@
 
-```{r tab-genes, tab.cap = 'FGFR-, FGF-, and FGFBP-coding genes investigated in the current report.'}
+```{r tab-genes, tab.cap = "FGFR-, FGF-, and FGFBP-coding genes investigated in the current report."}
 
 flextable::flextable(suppl_tabs$genes) %>% 
-  italic(j = 2, part = 'body') %>% 
-  merge_v(1) %>% 
-  set_widths(rep(3, 3)) %>% 
+  set_widths(c(3, 15)) %>% 
+  colformat_md %>% 
   theme_vanilla
@@
 
-```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGF-, FGFR-, and FGFBP-coding genes in the investigated cohorts.
-Statistical significance for differences between the cohorts was assessed by chi-squared test with Cramer's V effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-Frequency of alterations found in at least 1% of specimens in the BCAN, GENIE BLCA, IMvigor, MSK IMPACT, and TCGA BLCA are presented."}
+```{r tab-fgf-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR-, FGF-, and FGFBP-coding genes.
+Counts and percentages of samples with alterations for the mutations, amplifications, and deletions found in at least 1% of specimens in the BCAN, GENIE BLCA, IMvigor, MSK IMPACT, and TCGA BLCA are presented."}
 
@@
   merge_v(2) %>% 
-  merge_v(3) %>% 
-  italic(j = 3) %>% 
-  set_widths(c(2.5, 3, 2.2, 3, rep(2.5, 3))) %>% 
+  set_widths(c(2.5, 3, 3.5, 3.5)) %>% 
+  colformat_md %>% 
   theme_vanilla
@@
 
-```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR3 and FGF3/4/19 genes in the luminal papillary (LumP), luminal unstable (LumU), stoma-rich, and basal/squamous like (Ba/Sq) consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
-The frequencies are presented as percentages of the consensus class.
+```{r tab-sub-genetics, tab.cap = "Frequency of somatic mutations and copy number variants of FGFR1/3 and FGF3/4/19 genes in the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous like (Ba/Sq) consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
+The frequencies are presented as counts and percentages of samples with alterations within the consensus class.
 Statistical significance of differences between the consensus classes was determined by weighted permutation test with Cramer's V effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method."}
+P values were corrected for multiple testing with the false discovery rate (FDR) method."}
 
 flextable::flextable(suppl_tabs$sub_genetics) %>% 
-  merge_v(1:2, combine = TRUE) %>% 
-  italic(j = 2) %>% 
-  set_widths(c(2.3, 2.4, 2.3, rep(2.4, 3), rep(2.4, 2))) %>% 
+  merge_v(c(1, 2, 5, 6), combine = TRUE) %>% 
+  merge_v(1) %>% 
+  set_widths(c(3.2, 2.5, 3, 3.5, 2.5, 2.5)) %>% 
+  colformat_md %>% 
   theme_vanilla
@@
 
-```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes at mRNA level was quantified in the TCGA BLCA, IMvigor, and BCAN cohort by RNA sequencing.
-The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA and IMvigor) or Z-scores of log2 gene counts (BCAN).
-Protein expression of the genes of interest was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC).
-The protein amounts are metered by the IHC score defined as a product of staining intensity and quantity of positive cells.
-Median mRNA and protein amounts with interquartile ranges, ranges, and numbers of cancer samples are listed."}
+```{r tab-expression, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes and proteins in the MIBC tissue and urothelial cancer cell lines.
+Expression at mRNA level was quantified in the TCGA BLCA, IMvigor and BCAN cohorts, and in DepMap cell lines by RNA sequencing.
+The mRNA amounts were provided as log2-transformed gene counts (TCGA BLCA, IMvigor, DepMap) or Z-scores of log2 gene counts (BCAN).
+Protein expression was measured in the Human Protein Atlas (HPA) by immunohistochemistry (IHC).
+The protein amounts was metered by the IHC score defined as a product of staining intensity and quantity of positive cells.
+Median mRNA and protein amounts with interquartile ranges, and numbers of cancer samples are listed.
+Numbers of all available samples in the cohorts are provided in the table caption.
+For the IHC data in the HPA cohort, numbers of specimens with evaluable staining results are indicated for each gene."}
 
 flextable::flextable(suppl_tabs$expression) %>% 
-  italic(i = 3:47, j = 1) %>% 
-  set_widths(c(2.5, rep(3.5, 5))) %>% 
+    italic(j = 1, 
+         part = "body") %>% 
+  set_widths(c(2, rep(3.3, 4), 4.5)) %>% 
   theme_vanilla
@@
 
-```{r tab-sub-dge, tab.cap = "Expression of FGF-, FGFR-, and FGFBP-coding genes in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
-Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-log2-transformed expression levels are presented as medians with interquartile ranges and ranges for the top genes whose expression differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}
+```{r tab-sub-dge, tab.cap = "Expression of FGFR-, FGF-, and FGFBP-coding genes in the consensus molecular classes of MIBC (luminal papillary [LumP], luminal unstable [LumU], stroma-rich, basal/squamous-like [Ba/Sq]) in the TCGA BLCA, IMvigor, and BCAN cohorts.
+Statistical significance of differences between the consensus molecular classes was determined by one-way ANOVA with η² effect size statistic.
+P values were corrected for multiple testing with the false discovery rate (FDR) method.
+log2-transformed expression levels are presented as medians with interquartile ranges for the top genes whose expression differentiated between the molecular consensus classes (top 10 largest η² values shared by all cohorts)."}
 
@@
   italic(j = 2, italic = suppl_tabs$sub_dge$Variable != 'Samples, N') %>% 
-  set_widths(c(1.6, 2.1, rep(3.2, 4), rep(2.3, 2))) %>% 
+  set_widths(c(1.6, 2, rep(3.3, 4), rep(1.8, 2))) %>% 
   theme_vanilla
@@
 
-```{r tab-sub-sig, tab.cap = "Single sample gene set enrichment analysis (ssGSEA) scores of FGFR-related gene signatures in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
-Statistical significance of differences between the molecular classes was determined by one-way ANOVA with eta-square effect size statistic.
-P values were corrected for multiple testing with the false discovery rate method.
-ssGSEA scores are presented as medians with interquartile ranges and ranges for the top signatures whose scores differentiated between the molecular consensus classes (top 15 largest eta-square values shared by all cohorts)."}
+```{r tab-sub-sig, tab.cap = "Single sample gene set enrichment analysis (ssGSEA) scores of FGFR-related gene signatures in the consensus molecular classes of MIBC (luminal papillary [LumP], luminal unstable [LumU], stroma-rich, basal/squamous-like [Ba/Sq]) in the TCGA BLCA, IMvigor, and BCAN cohorts.
+Statistical significance of differences between the consensus molecular classes was determined by one-way ANOVA with eta-square effect size statistic.
+P values were corrected for multiple testing with the false discovery rate (FDR) method.
+ssGSEA scores are presented as medians with interquartile ranges for the top signatures whose scores differentiating between the molecular consensus classes (top 10 largest η² values shared by all cohorts)."}
 
@@
   merge_v(1) %>% 
-  set_widths(c(1.6, 3.2, rep(2.9, 4), rep(2.3, 2))) %>% 
+  set_widths(c(1.6, 5.7, rep(2.7, 4), rep(1.6, 2))) %>% 
   theme_vanilla
@@
 
-```{r tab-sub-drugs, tab.cap = "Predicted resistance metrics to pan-FGFR inhibitors in the consensus molecular classes of MIBC in the TCGA BLCA, IMvigor, and BCAN cohorts.
-The predictions of were made with RIDGE regularized linear models trained with transcriptome data of epithelial cancer cell lines treated with FGFR inhibitors in GDSC1, GDSC2, CTRP2, and PRISM drug screening experiments.
-Resistance metrics are presented as medians with interquartile ranges and ranges."}
+```{r tab-sub-drugs, tab.cap = "Predicted metrics of resistance to pan-FGFR inhibitors in the consensus molecular classes of MIBC (luminal papillary [LumP], luminal unstable [LumU], stroma-rich, basal/squamous-like [Ba/Sq]) in the TCGA BLCA, IMvigor, and BCAN cohorts.
+The predictions of IC50 (concentration resulting in 50% growth inhibition) and AUC (area under the dose-response curve) were made by RIDGE regularized linear models trained with transcriptome data of epithelial cancer cell lines treated with FGFR inhibitors in GDSC1, GDSC2, CTRP2, and PRISM drug screening experiments.
+Statistical significance of differences between the consensus molecular classes was determined by one-way ANOVA with η² effect size statistic.
+P values were corrected for multiple testing with the false discovery rate (FDR) method.
+Resistance metrics are presented as medians with interquartile ranges."}
 
 flextable::flextable(suppl_tabs$sub_drugs) %>% 
-  set_widths(c(1.6, 2.2, 1.6, rep(2.9, 4), 2.3, 2)) %>% 
   merge_v(1) %>% 
-  footnote(1, 3, 
-           as_paragraph('IC50: concentration in µM resulting in 50% inhibition of growth; AUC: area under the dose-response curve.'), 
-           part = 'header', 
-           ref_symbols = 'a') %>% 
-  footnote(1, 8:9, 
-           value = as_paragraph('Statistical significance: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic.
-P values corrected for multiple testing with the false discovery rate method.'),
-           part = 'header', 
-           ref_symbols = 'b') %>% 
+  set_widths(c(1.6, 3.2, 2, rep(2.7, 4), rep(1.6, 2))) %>% 
+  footnote(1, 2, 
+           value = as_paragraph("Substance name, identifier in the screening experiment, and name of the screening experiment."), 
+           part = "header", 
+           ref_symbols = "a") %>% 
   theme_vanilla
@@
 
-```{r tab-tuning, tab.cap = "Selection of the optimal sets of hyper-parameters for Elastic Net and Random Forest models of the MIBC consensus molecular classes by cross-validation and out-of-bag prediction tuning in the training subset of the TCGA BLCA cohort."}
+```{r tab-tuning, tab.cap = "Selection of the optimal sets of hyper-parameters for Elastic Net and Random Forest models of the MIBC consensus molecular classes by cross-validation and out-of-bag tuning in the training subset of the TCGA BLCA cohort."}
 
@@
     merge_v(1) %>% 
-  set_widths(c(3, 2.5, 4, 4)) %>% 
+  set_widths(c(3, 2.5, 5, 4)) %>% 
   footnote(1, 1, 
@@
 
-```{r tab-ml, tab.cap = "Performance of Elastic Net and Random Forest models of LumP, LUMU, stroma-rich, and BA/Sq consensus molecular classes of MIBC.
-The models were trained in a random subset of two-thirds observations of the TCGA BLCA cohort with ComBat-processed log2 expression values of the consensus class-refining genes by Kamoun et al.
-and the FGFR/FGF/FGFBP genes as explanatory variables.
-Minority consensus classes were amplified by the SMOTE algorithm to improve their identification by the model.
+```{r tab-ml, tab.cap = "Performance of Elastic Net and Random Forest models of the luminal papillary (LumP), luminal unstable (LumU), stroma-rich, and basal/squamous-like (Ba/Sq) consensus molecular classes of MIBC.
+The models were trained in a random subset of two-thirds of observations of the TCGA BLCA cohort with ComBat-processed log2 expression values of the consensus class-defining genes by Kamoun et al.
+or the FGFR/FGF/FGFBP genes as explanatory variables.
+Minority consensus molecular classes (LumU and stroma-rich) were oversampled by the SMOTE algorithm to improve their identification by the models.
 The models were evaluated in the training and test subsets of the TCGA BLCA, IMvigor, and BCAN cohorts; the consensus class assignment predicted by a nearest centroid classifier from consensusMIBC package served as the ground truth.
@@
   footnote(1, 5, 
-            value = as_paragraph("Cohen's \u03BA. Non-accidental concordance between the actual and predicted molecular class assignment."), 
+            value = as_paragraph("Cohen's \u03BA: non-random concordance between the actual and predicted molecular class assignment; high values indicate good concordance between the predictions and the outcome."),
             part = 'header', 
@@
   footnote(1, 6, 
-           value = as_paragraph('Measure of overall model calibration.'),
+           value = as_paragraph('Measure of model calibration and confidence; low values indicate good model calibration and confidence.'),
            part = 'header', 
@@
 
-\newpage
-
 # Supplementary Figures
@@
 
-__Supplementary Figure S\@ref(fig:fig-fgfr3-domains). Frequency of somatic mutations in FGFR1/2/3/4 genes split by protein domain__
+__Supplementary Figure S\@ref(fig:fig-fgfr3-domains). Frequency of somatic mutations in *FGFR1/2/3/4* genes split by protein domain__
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-fgfr3-hot). Mutational hot spots of the FGFR3 gene.__
+__Supplementary Figure S\@ref(fig:fig-fgfr3-hot). Mutational hot spots of the *FGFR3* gene.__
 
@@
 The protein expression is shown as IHC scores defined as a product of intensity (1: negative, 4: strong) and quantity of positive cells (1: none to 4: >75%).
-Median IHC scores with interquartile ranges are visualized as boxes with whiskers spanning over 150% of the interquartile ranges.
-Single IHC specimens are visualized as points.
-Numbers of IHC samples are indicated in the Y axis._
+Distributions of IHC scores are visualized with violin plots.
+Single IHC specimens are depicted as points.
+Numbers of evaluable IHC samples are indicated in the Y axis._
 
@@
 P values were corrected for multiple testing with the false discovery rate method (FDR).
-Significantly differentially regulated resistance metrics were defined by p(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and r $\geq$ 0.3 for differences between a class and LumP cancers.
-Z-scores of IC50 and AUC values are displayed in heat maps.
-Results of Kruskal-Wallis tests for overall differences between the consensus classes (effect sizes and p values) are indicated in the Y axes.
+Significantly differentially regulated resistance metrics were defined by p(Kruskal-Wallis) < 0.05, $\eta^2 \geq$ 0.14, pFDR(Mann-Whitney) < 0.05, and r $\geq$ 0.3 for differences between a class and LumP cancers._
+
+_(A) Z-scores of IC50 and AUC values displayed in heat maps.
+Results of the Kruskal-Wallis tests for overall differences between the consensus classes (effect sizes and p values) are indicated in the Y axes.
 Significant effects are labeled with bold font in the Y axes.
 Numbers of cancer samples in the consensus classes are indicated in the plot captions._
+
+_(B) Predicted AUC values for Erdafitinib shown in violin plots.
+Red diamonds with whiskers represent medians with interquartile ranges.
+Single cancer samples are depicted as points.
+Results of the Kruskal-Wallis tests for overall differences between the consensus classes (effect sizes and p values) are displayed in the plot captions.
+Numbers of cancer samples in the consensus classes are indicated in the X axes._
 
@@
 
-__Supplementary Figure S\@ref(fig:fig-fgfr-drugs). Predicted response to pan-FGFR inhibitors in the urothelial cancers with and without FGFR3 mutations.__
+__Supplementary Figure S\@ref(fig:fig-fgfr-drugs). Predicted response to pan-FGFR inhibitors in the urothelial cancers with and without *FGFR3* mutations.__
 
@@
 _(B) Squared differences between the class assignment probability returned by the models and the actual 0/1-coded class assignment (squared distance to the outcome) were computed as measures for single cancer samples as measures of model's confidence and error at predicting the consensus molecular classes.
-The squared distances to the outcome are presented in box plots with medians and interquartile ranges represented by boxes and whiskers which span over 150% of the interquartile ranges.
-Single cancer samples are depicted as points.
+The squared distances to the outcome are presented in violin plots; single cancer samples are depicted as points.
 Note a substantial improvement of prediction (lower squared distances to the outcome) of LumU class cancers by the machine learning models that employ the full class-defining predictor set by Kamoun et al as compared with the models using solely the FGFR/FGF/FGFBP-coding genes.
@@
 The corresponding mean absolute SHAP values as metrics of overall gene importance are presented in bar plots.
-Colors of the bars code for the sign of association of SHAP values and gene expression (red: positive, blue: negative)._
+Colors of the bars code for the sign of association between SHAP values and gene expression (red: positive, blue: negative)._
 
@@
 _Among the class-defining genes, `r intersect(ml_globals$variables, mibc_globals$variables) %>% collapse_and` were classified as FGFR/FGF/FGFBP-coding genes._ 
-_Distribution of the mean SHAP values is presented in violin plot.
+_Distribution of the mean SHAP values is presented in violin plots.
 Single explanatory factors are depicted as points; points corresponding to the FGFR/FGF/FGFBP-coding genes among the class-defining genes are highlighted with blue (negative class assignment predictor) and red (positive predictors of class assignment)._
@@
 IC50 for doxorubicin are presented as a positive cytotoxicity controls.
-Median IC50 values for the FGFR inhibitors with interquartile ranges are depicted as boxes with whiskers spanning over 150% of the interquartile ranges.
-Points represent single cell lines.
-Point symbols code for the FGFR3 gene mutation status, point color represents sensitivity/resistance.
+IC50 values for the FGFR inhibitors are depicted in violin plots; points represent single cell lines.
+Point color codes for the FGFR3 gene mutation status, point shape represents sensitivity/resistance.
 The sensitivity cutoff is represented by the dashed line._
@@
 
-__Supplementary Figure S\@ref(fig:fig-in-vivo). Dose response curves of cultured FGFR1-4 wild-type cell lines (UMUC3, 5637 and RT112) after treatment with increasing concentrations of erdafitinib.__
+__Supplementary Figure S\@ref(fig:fig-in-vivo). Dose response curves of cultured *FGFR1-4* wild-type cell lines (UMUC3, 5637 and RT112) after treatment with increasing concentrations of erdafitinib.__
 
@@
 
+\newpage
+
 # References
